---
title: "Supplementary material 2"
subtitle: "Landscape-scaled technological structure in surface-collected Quina scrapers"
author:
  - Pei-Yuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Qi-Jun Ruan:
      correspondence: "yes"
      email: ruanqijun000@163.com
      orcid: 0009-0000-2143-5335
      institute:
        - YPICRA
  - Davide Delpiano:
      correspondence: "no"
      institute:
        - FERRARA
  - Marco Peresani:
      correspondence: "no"
      orcid: 0000-0001-6562-6336
      institute:
        - FERRARA
        - IGAG
  - Zhen-Xiu Jia:
      correspondence: "no"
      orcid: 0000-0002-7514-4514
      institute:
        - ALPHA
  - Li-Jing Yang:
      correspondence: "no"
      institute:
        - BCRMI
  - Ming Zhang:
      correspondence: "no"
      institute:
        - BCRMI
  - Jian Sun:
      correspondence: "no"
      institute:
        - DCRMI
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
  - Hao Li:
      correspondence: "yes"
      email: lihao@itpcas.ac.cn
      institute:
        - ALPHA
# <!-- TODO: update the author list, ORCIDs and affiliations for THIS paper -->
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - YPICRA: Yunnan Provincial Institute of Cultural Relics and Archaeology, Kunming, Yunnan, China
  - FERRARA: Department of Human Studies, Prehistoric and Anthropological Science Unit, University of Ferrara, Ferrara, Italy
  - IGAG: Consiglio Nazionale delle Ricerche—Institute of Environmental Geology and Geoengineering, Laboratory of Palynology and Palaeoecology, Research Group on Vegetation, Climate and Human Stratigraphy, Milan, Italy
  - BCRMI: Binchuan Cultural Relics Management Institute, Dali Bai Autonomous Prefecture, Yunnan, China
  - DCRMI: Dali Bai Autonomous Cultural Relics Management Institute, Dali Bai Autonomous Prefecture, Yunnan, China
date: now
format:
  html:
    number-sections: true
    code-tools: true
    embed-resources: true
    df-print: kable
    fig-align: center
    theme: cosmo
execute:
  warning: false
  message: false
  echo: false
bibliography: []
---

```{r setup}
#| include: false
knitr::opts_chunk$set(fig.align = "center", dpi = 300, out.width = "100%")
suppressPackageStartupMessages({
  library(here); library(brms); library(posterior); library(dplyr)
  library(tidyr); library(knitr); library(targets); library(ggplot2)
})

TAR_STORE <- here::here("_targets")

tar_load(c(ctx_report, barg_mf, ppc_stat_tbl, barg_figs, barg_diag_tbl),
         store = TAR_STORE)

list2env(as.list(ctx_report, all.names = TRUE), envir = globalenv())

gf  <- function(n) targets::tar_read_raw(paste0("fit_", n), store = TAR_STORE)

FIG  <- function(n) here("paper", "barg", "figures", n)
SHOW <- function(n) knitr::include_graphics(FIG(n), rel_path = FALSE)

fit_ref    <- gf("ref")
fit_prior  <- gf("prior_ref")
fit_noland <- gf("noland")
mf <- barg_mf

cf       <- coef_table(fit_ref)
icc_ref  <- icc_table(fit_ref)
icc_nol  <- icc_table(fit_noland)
cor_tb   <- cor_table(fit_ref)
ppsd_ref <- prior_post_sd(fit_ref, fit_prior)

SENS <- c(REF = "ref", S1 = "s1", S2 = "s2", S3 = "s3", S4 = "s4", S5 = "s5",
          S6 = "s6")
ppsd_s1  <- NULL
sens_long <- bind_rows(lapply(names(SENS), function(nm) {
  if (nm == "REF") {
    d <- as_draws_df(fit_ref)
  } else {
    f <- gf(SENS[[nm]])
    if (nm == "S1")
      ppsd_s1 <<- prior_post_sd(f, NULL, b_sd = setNames(rep(1, 7), resps))
    d <- as_draws_df(f); rm(f); gc()
  }
  out <- bind_rows(
    bind_rows(lapply(resps, function(r) {
      x <- icc_draws(d, r, resp_fam[[r]]); q <- eti(x)
      data.frame(Spec = nm, Quantity = paste0("ICC: ", unname(resp_lab[r])),
                 Median = median(x), lo = q[1], hi = q[2],
                 inrope = mean(x <= ICC_ROPE))
    })),
    bind_rows(lapply(resps, function(r) bind_rows(lapply(preds, function(pp) {
      x <- d[[paste0("b_", r, "_", pp)]] / link_sd[[r]]; q <- eti(x)
      data.frame(Spec = nm,
                 Quantity = paste0(unname(resp_lab[r]), " ~ ", unname(pred_lab[pp])),
                 Median = median(x), lo = q[1], hi = q[2],
                 inrope = mean(abs(x) < ROPE_SD))
    })))))
  rm(d); gc(); out
}))

dr_ref <- as_draws_df(fit_ref)
sm_ref <- summarise_draws(dr_ref, "rhat", "ess_bulk", "ess_tail")
sm_ref <- sm_ref[!is.na(sm_ref$rhat) & !startsWith(sm_ref$variable, "lp"), ]
np_ref <- nuts_params(fit_ref)
n_div  <- sum(np_ref$Value[np_ref$Parameter == "divergent__"])
n_tree <- sum(np_ref$Value[np_ref$Parameter == "treedepth__"] >= 10)

fmt  <- function(x, d = 2) formatC(x, format = "f", digits = d)
ci   <- function(m, l, h, d = 3) sprintf("%s [%s, %s]", fmt(m, d), fmt(l, d), fmt(h, d))
kt   <- function(x, caption, digits = 3, ...)
  knitr::kable(x, caption = caption, digits = digits, row.names = FALSE, ...)
rlab <- function(x) unname(resp_lab[x])
plab <- function(x) unname(pred_lab[x])
```

```{r}
#| label: q1-readout
#| echo: false
icc_dec <- icc_ref |> arrange(desc(Median))
present <- icc_dec |> filter(Verdict == "locality variation present")
undec   <- icc_dec |> filter(Verdict == "undecided")
icc_j <- icc_ref |> select(Response, ref = Median, ref_lo = CrI_lo) |>
  left_join(icc_nol |> select(Response, nol = Median), by = "Response") |>
  mutate(change = ref - nol, rel = change / nol)

icc_sens <- sens_long |>
  filter(startsWith(Quantity, "ICC: ")) |>
  mutate(Resp = sub("^ICC: ", "", Quantity),
         dec = ifelse(lo > ICC_ROPE, "present",
               ifelse(hi < ICC_ROPE, "negligible", "undecided")))
icc_stable <- icc_sens |> group_by(Resp) |>
  summarise(k = n_distinct(dec), ref = dec[Spec == "REF"], .groups = "drop")
firm  <- icc_stable |> filter(k == 1, ref == "present")
flips <- icc_stable |> filter(k > 1)
flip_detail <- icc_sens |> semi_join(flips, by = "Resp") |> filter(dec != "present")
```

```{r}
#| label: q1-readout2
#| echo: false
med_rel <- median(icc_j$rel)
n_up    <- sum(icc_j$change > 0); n_dn <- sum(icc_j$change < 0)
worst_j <- icc_j |> arrange(rel) |> slice(1)
best_j  <- icc_j |> arrange(desc(rel)) |> slice(1)
```

```{r}
#| label: q2-readout
#| echo: false
v_tab <- table(cf$Verdict)
n_und <- sum(cf$Verdict == "undecided")
n_rej <- sum(cf$Verdict == "reject the null value")
n_acc <- sum(cf$Verdict == "accept the null value")
best  <- cf |> arrange(Min_ROPE / link_sd[Response]) |> head(3)
worst2 <- cf |> arrange(desc(Min_ROPE / link_sd[Response])) |> head(3)
dir_max <- cf |> arrange(desc(Dir_certainty)) |> head(3)
ea     <- sens_long |> filter(Quantity == paste0("ICC: ", rlab("EdgeAngle")))
ea_ref <- ea |> filter(Spec == "REF")
ea_min <- ea |> arrange(Median) |> slice(1)
```

```{r}
#| label: q2-accepted
#| echo: false
acc <- cf |> filter(Verdict == "accept the null value")
```

```{r}
#| label: sens-verdict-calc
#| echo: false
sens_verdict <- sens_long |>
  mutate(is_icc = startsWith(Quantity, "ICC: "),
         Decision = ifelse(is_icc,
           ifelse(lo > ICC_ROPE, "locality variation present",
             ifelse(hi < ICC_ROPE, "negligible", "undecided")),
           rope_verdict(lo, hi, 100 * inrope, ROPE_SD))) |>
  select(Quantity, Spec, Decision) |>
  pivot_wider(names_from = Spec, values_from = Decision)
changed <- sens_verdict |>
  rowwise() |>
  mutate(n_distinct_dec = n_distinct(c_across(all_of(names(SENS))))) |>
  ungroup()
```

```{r}
#| label: sens-readout
#| echo: false
n_changed <- sum(changed$n_distinct_dec > 1)
worst_shift <- sens_long |>
  group_by(Quantity) |>
  summarise(shift = max(Median) - min(Median),
            width = mean(hi - lo), .groups = "drop") |>
  mutate(rel = shift / width) |> arrange(desc(rel))
```

```{r}
#| label: s6-identity
#| echo: false
s6_q <- local({
  f <- gf("s6")
  out <- list(cf = coef_table(f), icc = icc_table(f))
  rm(f); gc(); out
})

s6_id <- cf |>
  select(Response, Predictor, ref_med = Median, ref_mcse = MCSE_med,
         ref_sd = SD, ref_v = Verdict) |>
  left_join(s6_q$cf |> select(Response, Predictor, s6_med = Median,
                              s6_mcse = MCSE_med, s6_v = Verdict),
            by = c("Response", "Predictor")) |>
  mutate(d = s6_med - ref_med,
         mcse_pair = sqrt(ref_mcse^2 + s6_mcse^2),
         z = abs(d) / mcse_pair,
         rel_sd = abs(d) / ref_sd)

s6_icc_id <- icc_ref |>
  select(Response, ref_med = Median, ref_mcse = MCSE_med, ref_sd = SD,
         ref_v = Verdict) |>
  left_join(s6_q$icc |> select(Response, s6_med = Median, s6_mcse = MCSE_med,
                               s6_v = Verdict),
            by = "Response") |>
  mutate(d = s6_med - ref_med,
         z = abs(d) / sqrt(ref_mcse^2 + s6_mcse^2),
         rel_sd = abs(d) / ref_sd)

s6_same_slope <- all(s6_id$ref_v == s6_id$s6_v)
s6_same_icc   <- all(s6_icc_id$ref_v == s6_icc_id$s6_v)

coi_ref  <- dr_ref[["coi_GIUR"]]
zoi_ref  <- dr_ref[["zoi_GIUR"]]
p0_ref   <- zoi_ref * (1 - coi_ref)
n0_exp   <- median(p0_ref) * n_spec
```

```{=html}
<style>
.barg { background:#F3F5F7; border-left:3px solid #6BA8CE; padding:.45em .8em;
        margin:.4em 0 1em 0; font-size:.86em; color:#303238; }
caption { font-size:.86em; color:#303238; text-align:left; }
table { font-size:.84em; }
</style>
```

# Summary {#sec-summary}

## What This Document Is

This is Supplementary Material 2, the statistical supplement to the main
paper. It reports one Bayesian analysis in full: the data, the model, the
priors, how the model was fitted and checked, the posterior estimates, the
decisions drawn from them, and how far those decisions depend on choices that
could have been made differently.

The analysis follows two published guidelines, which divide the labour between
them. Kruschke (2021) specifies what a completed Bayesian analysis should
report, item by item, and @sec-barg shows where each requirement is answered.
Gelman, Vehtari, McElreath, et al. (2026) instead describe how model
development proceeds before there is a final analysis to report: fit a model,
check how well it performs, find the problems, change the model, and go round
again. They argue that a real analytical problem is better understood by
comparing several plausible models than by relying on one.

This supplement therefore follows the order in which the analysis was
conducted, from @sec-design to @sec-reproducibility, and puts the reporting
checklist at the end. A reader auditing coverage can work through the
checklist without interrupting the narrative.

The appendices give the complete tables, the model code and the diagnostics. A
reader less familiar with Bayesian terminology can begin with
@sec-reading-guide, which introduces the vocabulary used throughout.

## What Was Analysed

`r n_spec` surface-collected Quina scrapers from `r n_loc` localities in two
basins, Binchuan and Huangping. Seven attributes were taken on each
specimen: its thickness, its overall size, its GIUR, the retouched share of
its perimeter, its total scar count, its edge angle, and the number of retouch
generations recorded on it. Three further attributes describe the locality
rather than the specimen: which basin it lies in, how high it sits above the
channel, and how far it is from the channel. Each is the same for every
specimen collected from that locality.

One multilevel model takes all seven attributes at once, so that a single
locality structure is shared across them instead of seven separate analyses
being run and compared afterwards. The model gives every locality its own
baseline for every attribute, and it answers two questions. How much of the
variation in each attribute occurs between localities rather than among
specimens within a locality? And how much of that difference between
localities do the three landscape attributes explain?

## What Was Found

### Q1: Do the Localities Differ from One Another?

Yes, for `r nrow(present)` of the seven measurements:
`r paste(rlab(present$Response), collapse = ", ")`. Between
`r fmt(min(present$Median) * 100, 0)`% and
`r fmt(max(present$Median) * 100, 0)`% of the variation in those measures sits
between localities rather than between specimens within a locality, and for
`r nrow(firm)` of them the finding survives every one of the prior settings the
model was refitted under. The remaining `r nrow(undec)` are undecided
(@sec-q1).

### Q2: Do Basin, Elevation or Distance to the Channel Explain the Differences?

Not clearly. Of the `r nrow(cf)` landscape effects estimated,
`r n_rej` show an effect large enough to matter, `r n_acc`
`r if (n_acc == 1) "is" else "are"` close enough to zero to be called
practically negligible, and the remaining `r n_und` are too uncertain to
separate the two. Adding all three landscape attributes to a model that lacks
them barely moves the share of variation sitting between localities, the
median change across the seven measures being
`r sprintf("%+.1f", med_rel * 100)`% (@sec-q1, @sec-q2).

An undecided result is a finding rather than a failure to obtain one, so for
each landscape effect we report the smallest effect the data can rule out, in
the units of the measurement itself. For the three best-resolved effects those
bounds are
`r paste(sprintf("%s with %s, +/- %s %s", rlab(best$Response), tolower(plab(best$Predictor)), fmt(abs(best$Min_ROPE_resp), 0), best$Min_ROPE_unit), collapse = "; ")`.
@tbl-coef-full gives the same bound for all `r nrow(cf)` effects, in degrees,
retouch generations, or percentages of the measurement.

## What the Study Cannot Settle

Basin, height above the channel and distance to the channel are recorded once
per locality, so the three landscape attributes are estimated from
`r n_loc` localities however many specimens each of them contributes. That
number, together with the very uneven sizes of the assemblages behind it, sets
the resolution of both research questions. @sec-design-sample gives the
argument in full, and @sec-resolution-limits adds three limitations that apply
to particular measures.

# Reading Guide {#sec-reading-guide}

Eleven terms do most of the technical work in what follows. Each is glossed
again, in one clause, where it first appears in the main narrative.

| Term | What it means here |
|---|---|
| Posterior distribution | The distribution of a quantity (a slope, a share of variance) after the data have been combined with the prior. Every number in this document is a summary of one. |
| 95% credible interval | The central range containing 95% of the posterior. It says where the quantity lies, given this model and these data. A confidence interval answers a different question, about the long-run behaviour of the procedure that produced it. |
| ROPE | Region of practical equivalence: the band of values around zero that the analysis is willing to treat as no effect. Calling an effect absent means showing that the posterior sits inside that band. |
| Intraclass correlation (ICC) | The share of a measure's variation that sits between localities rather than between specimens within a locality. |
| Link scale, and dividing by the link-scale SD | Thickness is modelled on a logarithmic scale, GIUR on a log-odds scale, edge angle in degrees. A slope on one of those scales cannot be compared with a slope on another until each is divided by the spread of its own response on its own scale. Every standardised slope here has been divided in that way, which is what lets one ROPE serve all seven measures. |
| Partial pooling | A locality with one specimen is estimated partly from the other localities, so its level is pulled towards the overall level. It stops a single-specimen locality from being read as a well-measured one. |
| Prior predictive check | Simulating data from the prior alone, before the real data are used, to confirm that the prior puts no weight on physically impossible values. |
| Posterior predictive check | Simulating data from the fitted model and comparing them with the observed data, to see what the model reproduces and what it misses. |
| R-hat and effective sample size | Diagnostics of the simulation rather than of the archaeology. R-hat near 1 means the independently started chains agree; the effective sample size says how many independent draws the correlated ones are worth. |
| Rank plot | A picture of the same thing. Pool the draws from all four chains, rank them, and count how many of each chain's draws fall in each rank bin. Four flat, overlapping lines mean the chains explored the same distribution. |
| Sensitivity analysis | Refitting under different priors to find out which conclusions depend on the prior and which do not. |

---

# The Estimand and the Design {#sec-design}

Gelman, Vehtari, McElreath, et al. (2026, ch. 10) set out the chain any
empirical claim depends on. A generative theory motivates an estimand; the
theory and the estimand together justify a design; the design yields a sample;
and a statistical model, the estimator, turns that sample into an estimate.
The links exist whether or not the analyst names them, and a reader can walk
the chain backwards from a conclusion to find the one that fails. This section
names them for the present study, because three facts about the design fix the
resolution of everything that follows, and because the specimen count
overstates what the design delivers for the second research question.

## Theory and Estimand

The theory is that the technological character of a Quina assemblage varies
with where in a valley system its makers worked, so that assemblages from
different parts of a basin should differ in systematic ways.

Two quantities to be estimated follow from it, and the rest of the document
keeps them apart (@sec-goals). Q1 is the proportion of the variation in each
of the seven measures that occurs between localities rather than among
specimens within the same locality. Q2 is the difference in each measure
associated with belonging to one basin rather than the other, together with
the change associated with a one-standard-deviation increase in height above
the channel or in distance from the channel, the other two attributes held
fixed.

Q2 estimates an association. The study did not assign the localities their
elevations, it states no causal diagram, and it derives no adjustment set from
one. A slope near zero therefore means that the landscape attribute fails to
predict the measure across these `r n_loc` localities. It does not establish
that moving an assemblage's position in the landscape would leave the measure
unchanged. A causal reading would need a design that separates the landscape
variables from whatever else distinguishes one locality from another.

## Design and Sample {#sec-design-sample}

Three features of the sample constrain every result below. They follow from
how the material was collected, so no choice of model, no choice of prior and
no further specimens can overcome them.

For Q2 the unit of information is the locality, and there are `r n_loc` of
them. Basin, height above the channel and distance to the channel are recorded
once per locality and are identical for every specimen collected there. The
`r n_spec` specimens sharpen the estimate of within-locality variation and of
each locality's overall level. They add no independent information about the
three landscape slopes, so collecting more scrapers from the same localities
would describe those localities better while leaving Q2 where it is. The same
fact rules out landscape slopes that vary from locality to locality, which is
why the question of whether the landscape effects differ between the two
basins has to be put at basin level (@sec-basinx).

The basin comparison rests on a single contrast, between
`r sum(tapply(as.character(mod_dat$Basin), mod_dat$Locality, function(x) x[1]) == "Binchuan")`
localities in Binchuan and
`r sum(tapply(as.character(mod_dat$Basin), mod_dat$Locality, function(x) x[1]) == "Huangping")`
in Huangping. The width of the interval estimate is what reports the
uncertainty that follows from it.

The assemblages are very uneven. One locality contributes
`r max(table(mod_dat$Locality))` of the `r n_spec` specimens, and
`r sum(table(mod_dat$Locality) == 1)` localities contribute one specimen each.
Partial pooling estimates a poorly sampled locality partly from the others
instead of treating its level as precisely known (@sec-cor). This improves the
estimate without creating information the sample does not contain.

@sec-resolution-limits returns to these three limitations once the estimates
are in, and adds three that arise from the model rather than from the sample.

---

# The Analysis {#sec-analysis}

## Why the Analysis Is Bayesian {#sec-why-bayes}

::: {.barg}
Cf. BARG Preamble A.
:::

Four considerations led to a Bayesian treatment. The seven measures on a Quina
scraper are not on one scale -- thickness and overall size are positive and
right-skewed, the geometric index of unifacial reduction (GIUR) and the
retouched share of the perimeter are proportions with mass piled at one, the
scar count is a count, and the edge angle and the number of retouch generations
are treated as unbounded -- and a single multivariate model that gives each
measure the likelihood its scale calls for while still sharing one
locality-level random structure across all seven is straightforward to write
down and fit by MCMC, and awkward to fit any other way. Q1 asks for the share
of link-scale variance attributable to locality, an intraclass correlation
built from a variance component and a residual variance that are themselves
functions of several parameters; under MCMC that ratio is computed draw by draw
and inherits a genuine interval, with no approximation and no delta-method
step. Q2 puts a claim of absence under test, which a frequentist test can only
fail to reject, whereas a posterior distribution combined with a region of
practical equivalence can accept it, reject it, or report that the data do not
decide (@sec-decision-rule). And the variance component Q1 asks about is
estimated from `r n_loc` clusters (@sec-design-sample), a number at which
asymptotic intervals deserve little trust while posterior intervals require no
such argument.

## Goals and the Two Research Questions {#sec-goals}

::: {.barg}
Cf. BARG Preamble B.
:::

The study has two goals. The first is estimation: the size of each landscape
effect and the amount of locality-level variation, with the uncertainty on
each. The second is decision: whether each landscape effect is small enough to
treat as practically absent. Predicting new specimens is outside both.

Formal model selection is outside the scope as well. One model answers the two
research questions, and the analysis estimates its parameters instead of
comparing competing scientific hypotheses. It computes no marginal likelihood
and chooses no model on predictive performance. That matters for reading
@sec-model-series, which reports `r nrow(mf$diagnostics)` fits. The fits
measure how far the conclusions depend on reasonable alternative choices, and
none is selected as best. Gelman, Vehtari, McElreath, et al. (2026, ch. 9)
draw the same distinction, treating a series of fits as a way to understand
one model rather than as a contest among models.

Q1 asks whether technological variation exists between localities at all. The
locality-level variance component answers it through the intraclass
correlation (ICC), the proportion of a measure's total variation that occurs
between localities rather than among specimens within the same locality
(@sec-icc, @sec-q1). In plainer terms, Q1 asks how much specimens from one
locality resemble each other, relative to how much the localities differ.

Q2 asks whether basin, elevation above the channel or distance to the channel
explains that between-locality variation. The population-level slopes supply
the answer (@sec-posterior-summary, @sec-q2).

The order of the questions matters. Q2 has something to explain only if Q1
shows that the localities differ in the first place. A "yes" for Q1 alongside
an undecided result for Q2 is no contradiction: the two answer different
questions, and this document reports them separately.

## Data {#sec-data}

::: {.barg}
Cf. BARG Step 1.A.
:::

```{r}
#| label: tbl-data-vars
kt(data.frame(
  Variable = c(rlab(resps), "Basin", "Height above channel", "Distance to channel",
               "Locality"),
  Role = c(rep("response", 7), rep("predictor (locality level)", 3), "grouping factor"),
  Type = c("continuous, positive (mm)",
           "continuous, positive (mm); geometric mean of length, width and thickness",
           "proportion in [0, 1]", "proportion in [0, 1]", "count",
           "continuous (degrees)", "continuous (generations)",
           "binary: Binchuan / Huangping", "continuous (m)", "continuous (m)",
           "factor, 26 levels"),
  Observed = c(
    sprintf("%.1f - %.1f", sapply(resps, function(r) min(mod_dat[[r]])),
                            sapply(resps, function(r) max(mod_dat[[r]]))),
    sprintf("%d / %d", sum(mod_dat$Basin == "Binchuan"), sum(mod_dat$Basin == "Huangping")),
    sprintf("%.0f - %.0f", min(mod_dat$Height), max(mod_dat$Height)),
    sprintf("%.0f - %.0f", min(mod_dat$Distance), max(mod_dat$Distance)),
    sprintf("%d specimens in %d localities", n_spec, n_loc))),
  caption = "The variables entering the model, and their observed range in the analysed sample.")
```

`GMsize` is the geometric mean of the three linear dimensions,
$$\mathrm{GMsize} = (\mathrm{Length}\times\mathrm{Width}\times\mathrm{Thickness})^{1/3},$$
computed for each specimen. The definition lives in the analysis code rather
than in any measurement protocol, and it is stated here because it has a
consequence.

Thickness is one of the three factors in GMsize, so the two responses share a
component before any archaeology enters. The model is fitted with
`set_rescor(FALSE)`, which assumes the *residuals* of the seven responses are
independent given the locality intercepts, and that assumption is not credible
for this pair. The practical consequence is confined. With `rescor = FALSE`
the multivariate model factorises into seven univariate models sharing only
the correlated random intercepts, so the slopes and their intervals for each
response remain valid marginally. The invalid readings are two: any statement
about the *joint* uncertainty of the thickness slope and the size slope, and
any treatment of those two responses as independent evidence. The observed
correlation between their locality intercepts (@sec-cor) is therefore partly
mechanical.

Two filters were applied before fitting. A specimen must have all seven
measurements, and its locality must have a recorded basin, height and
distance.

```{r}
#| label: tbl-dropped
kt(drop_log |> mutate(lost = c(NA, diff(n))) |>
     rename(`specimens remaining` = n, `lost at this step` = lost),
   caption = "Complete-case filtering. One specimen was dropped, for a missing GIUR.",
   digits = 0)
```

```{r}
#| label: tbl-localities
kt(mod_dat |> count(Basin, Locality) |> count(Basin, name = "localities") |>
     left_join(mod_dat |> count(Basin, name = "specimens"), by = "Basin") |>
     mutate(`specimens per locality` = sprintf("%d - %d",
       tapply(mod_dat$Locality, mod_dat$Basin, function(x) min(table(droplevels(x)))),
       tapply(mod_dat$Locality, mod_dat$Basin, function(x) max(table(droplevels(x)))))),
   caption = "Localities and specimens by basin. Heqing, which appears under that name in the site table, is folded into Huangping: the two names refer to the same basin.",
   digits = 0)
```

The random intercept is the part of the model that enforces what
@sec-design-sample sets out, and it is why the intervals in
@sec-posterior-summary are as wide as they are.

The standardisation is specimen-weighted: `zHeight` and `zDistance` are
centred and scaled over the `r n_spec` specimens rather than over the
`r n_loc` localities, so a locality enters the centring and the scaling with
the weight of its own assemblage. A locality-weighted standardisation would
give slightly different numerical slopes on a rescaled axis. It would not
change any decision, because every slope is reported divided by the link-scale
SD of its own response and the ROPE moves with it.

## Model and Likelihoods {#sec-likelihood}

::: {.barg}
Cf. BARG Step 1.B.
:::

For specimen $i$ in locality $j$, each response $r$ has its own likelihood
with a linear predictor on its own link scale:

$$
g_r(\mu^{(r)}_{ij}) \;=\; \beta^{(r)}_{0} \;+\; \beta^{(r)}_{\text{Basin}}\,\mathrm{Basin}_j
\;+\; \beta^{(r)}_{\text{H}}\,\mathrm{zHeight}_j
\;+\; \beta^{(r)}_{\text{D}}\,\mathrm{zDistance}_j \;+\; u^{(r)}_j ,
$$

$$
\bigl(u^{(1)}_j,\dots,u^{(7)}_j\bigr)^{\!\top} \sim
\mathrm{MVNormal}\bigl(\mathbf 0,\;\mathbf D\,\boldsymbol\Omega\,\mathbf D\bigr),
\qquad \mathbf D=\mathrm{diag}(\sigma^{(1)}_u,\dots,\sigma^{(7)}_u).
$$

```{r}
#| label: tbl-likelihood
kt(data.frame(
  Response = rlab(resps),
  Likelihood = unname(fam_lab[resp_fam]),
  `Link for the mean` = unname(link_lab[resp_fam]),
  `Auxiliary parameters` = c("sigma", "sigma", "phi, zoi, coi", "phi, zoi, coi",
                             "shape", "sigma", "sigma"),
  `Link-scale SD` = round(link_sd, 3), check.names = FALSE),
  caption = "Likelihoods. The last column is the SD of the observed response on its own link scale; it is the unit in which every slope and every ROPE below is expressed.")
```

### Parameters of Theoretical Interest and Nuisance Parameters

Of the `r nrow(sm_ref)` estimated parameters, the research questions concern
the `r nrow(cf)` population-level slopes
$\beta^{(r)}_{\text{Basin}}, \beta^{(r)}_{\text{H}}, \beta^{(r)}_{\text{D}}$
(Q2) and the seven locality SDs $\sigma^{(r)}_u$ together with the ICCs
derived from them (Q1). The seven intercepts, the auxiliary parameters
(`sigma`, `phi`, `shape`, `zoi`, `coi`) and the 21 elements of
$\boldsymbol\Omega$ make the model coherent without answering any question
posed here. @sec-cor nevertheless reports the correlations in full, because
the model estimated them and BARG asks for a description of every parameter.

With `r sum(table(mod_dat$Locality) == 1)` localities of one specimen, the
locality offsets shrink hard towards the grand mean. That shrinkage is partial
pooling: a locality with little data of its own is estimated partly from the
others, and the caterpillar plot in @sec-cor shows how far each is pulled.
The behaviour is intended, and it is why the ICC comes from the fitted
variance components rather than from a between-group variance ratio of raw
locality means.

### A Trap in the Zero-One-Inflated Beta Parameterisation

For GIUR and the retouched perimeter, the three predictors enter only the beta
component's mean `mu`. The inflation parameters `zoi` (the probability that a
value is exactly 0 or exactly 1) and `coi` (the probability that such a value
is 1 rather than 0) are pure intercepts, constant across basins, elevations
and distances. Two consequences follow for every GIUR and RLI number in this
report.

1. A reported slope is the effect on the **logit of the beta-component mean,
   conditional on the value lying strictly inside (0, 1)**. The effect on the
   unconditional expected GIUR is a different quantity, because that one also
   moves with `zoi` and `coi`.
2. The ICC for these two responses likewise comes from the beta component
   alone and ignores the inflation entirely.

In this sample there are `r sum(mod_dat$GIUR == 1)` specimens with GIUR
exactly 1 and `r sum(mod_dat$GIUR == 0)` with GIUR exactly 0, and
`r sum(mod_dat$RLI == 1)` and `r sum(mod_dat$RLI == 0)` for the retouched
perimeter. There is therefore no zero inflation at all to model: the fitted
`coi` is a posterior over "all inflated values are ones", which the data
determine almost completely, and the zero-inflation half of the family is
supported by nothing. The family is effectively a one-inflated beta.

### S6, the One-Inflated Refit {#sec-oneinflated}

The paragraph above describes a defect rather than a design choice, and S6 acts
on it. Setting `coi` to `constant(1)` in `barg_priors.R` gives the family the
one-inflated beta shape the data actually show: mass at 1, none at 0, and the
zero-inflation half deleted instead of estimated. brms implements the constant
by moving `coi_GIUR` and `coi_RLI` out of the Stan `parameters` block into
`transformed parameters`, so two parameters genuinely leave the sampled space
instead of receiving a tight prior, and @tbl-diag-all shows S6 estimated over
`r mf$diagnostics$n_par[mf$diagnostics$fit == "s6"]` parameters against
`r mf$diagnostics$n_par[mf$diagnostics$fit == "ref"]` for the reference. The
reference fit shows why that is worth doing: under `beta(1, 1)` and
`r sum(mod_dat$GIUR == 1)` ones out of `r n_spec` specimens, its posterior for
GIUR's `coi` centres on `r fmt(median(coi_ref), 3)` rather than on 1, reserving
`r fmt(100 * (1 - median(coi_ref)), 1)`% of the inflation mass for an outcome
that cannot occur, which puts the probability of an exact zero at
`r fmt(median(p0_ref), 4)` and predicts **`r fmt(n0_exp, 2)`** zeros among the
`r n_spec` specimens, where the data contain none.

No conclusion can change under the refit, and the reason is stronger than a
sensitivity result. The zero-one-inflated
beta likelihood factorises: an observation contributes
$\mathrm{zoi}\cdot\mathrm{coi}$ if it equals 1,
$\mathrm{zoi}\cdot(1-\mathrm{coi})$ if it equals 0, and
$(1-\mathrm{zoi})\cdot\mathrm{Beta}(y \mid \mu, \phi)$ otherwise. The linear
predictor enters only the third term; `zoi` and `coi` enter only the first two,
take no predictors, and have their own priors. The posterior therefore
separates, and the slopes, `phi`, the locality SDs and the locality correlation
matrix come from the interior observations alone whatever value `coi` takes.
S6 therefore checks that this reasoning, and the code that implements it, are
both right, rather than whether the answers survive a different `coi`.

```{r}
#| label: tbl-s6-identity
s6_id |>
  arrange(desc(rel_sd)) |>
  head(5) |>
  transmute(Response = rlab(Response), Predictor = plab(Predictor),
            `reference median` = ref_med, `S6 median` = s6_med,
            difference = d, `difference in MCSE` = z,
            `difference / posterior SD` = rel_sd) |>
  kt(caption = sprintf("The five landscape slopes on which the reference fit and S6 disagree most, on the link scale, ordered by the difference between the two posterior medians as a fraction of that slope's own posterior SD. The other %d agree more closely. The two runs are separate samplers over different Stan programs, S6 having two fewer parameters, so exact agreement is neither expected nor desirable; agreement within a small multiple of the Monte Carlo standard error is what the factorisation predicts.", nrow(s6_id) - 5), digits = 4)
```

Across all `r nrow(s6_id)` slopes the largest difference between the two
posterior medians is `r fmt(max(abs(s6_id$d)), 4)` on the link scale, which is
**`r fmt(100 * max(s6_id$rel_sd), 1)`%** of that slope's own posterior SD and
`r fmt(max(s6_id$z), 1)` Monte Carlo standard errors; for the seven ICCs the
largest difference is `r fmt(max(abs(s6_icc_id$d)), 4)`, or
`r fmt(max(s6_icc_id$z), 1)` MCSEs. The MCSE figure is a maximum over
`r nrow(s6_id)` comparisons, so a value of three or four is ordinary behaviour
rather than a discrepancy; by the percentage of the posterior SD, the plainer
reading, the two fits agree to within a few percent of the width of the thing
being estimated. The decisions agree exactly.
`r if (s6_same_slope && s6_same_icc) sprintf("All %d slope verdicts and all %d ICC verdicts are the same under the two specifications.", nrow(s6_id), nrow(s6_icc_id)) else sprintf("%d of the %d slope verdicts and %d of the %d ICC verdicts differ, which the factorisation says should not happen and which needs explaining before either fit is used.", sum(s6_id$ref_v != s6_id$s6_v), nrow(s6_id), sum(s6_icc_id$ref_v != s6_icc_id$s6_v), nrow(s6_icc_id))`

This document therefore reports S6 as the specification the data support and
keeps the reference fit as the source of every other number. Exchanging them
would change nothing, and @tbl-s6-identity is the evidence for that claim.

## Priors {#sec-priors}

::: {.barg}
Cf. BARG Step 1.C.
:::

### Why the Slope Prior Is Scaled to Each Response

The seven responses reach their linear predictor through three different links
(log, logit, identity), and their link-scale SDs span two orders of magnitude,
from `r fmt(min(link_sd), 3)` for `r rlab(names(which.min(link_sd)))` to
`r fmt(max(link_sd), 2)` for `r rlab(names(which.max(link_sd)))`
(@tbl-likelihood). A single nominal slope prior, `normal(0, 1)` say, is
therefore a very different statement about each of them. The clearest way to
see it is the share of the prior that falls inside that response's own ROPE,
the interval of slopes the analysis is prepared to call practically zero.

```{r}
#| label: tbl-prior-rope
pir <- data.frame(Response = rlab(resps),
                  `Link-scale SD` = round(link_sd, 3),
                  `ROPE half-width` = round(ROPE_SD * link_sd, 4),
                  `normal(0, 1)` = round(prior_in_rope(rep(1, 7)), 2),
                  `normal(0, SD_link)` = round(prior_in_rope(link_sd), 2),
                  check.names = FALSE, row.names = NULL)
kt(pir, caption = "Percentage of the slope prior lying inside the response's own ROPE, under a single unscaled normal(0, 1) and under the per-response scaling used here. One nominal specification ranges, across the seven responses, from an almost dogmatic rejection of a null effect to an almost dogmatic assertion of one; scaled to each response it is 7.97% for every one of them, by construction.")
```

The diagnostic that makes this more than an aesthetic point is the ratio of
the posterior SD to the prior SD for each slope. A ratio near 1 means the
posterior is the prior and the data contributed nothing.

```{r}
#| label: tbl-ppsd
ppsd_cmp <- ppsd_s1 |>
  transmute(Response, Predictor, `normal(0, 1)` = ratio) |>
  left_join(ppsd_ref |> transmute(Response, Predictor, `normal(0, SD_link)` = ratio),
            by = c("Response", "Predictor")) |>
  mutate(Response = rlab(Response), Predictor = plab(Predictor))
kt(ppsd_cmp, caption = "Posterior SD divided by prior SD, for every slope, under an unscaled normal(0, 1) slope prior (specification S1) and under the reference prior. Values near 1 mean the data did not move the prior. The prior SD is taken from the sample_prior = 'only' draws for the reference prior and analytically for normal(0, 1).")
```

```{r}
#| label: ppsd-summary
#| echo: false
ea_old <- ppsd_cmp |> filter(Response == rlab("EdgeAngle"))
ea_new_max <- max(ea_old$`normal(0, SD_link)`)
```

Under an unscaled `normal(0, 1)` the `r rlab("EdgeAngle")` slopes have
posterior/prior SD ratios of
`r paste(fmt(sort(ea_old[["normal(0, 1)"]]), 2), collapse = ", ")`: the
posterior is very nearly the prior, so an edge-angle result read off that fit
would be mostly a restatement of the prior scale. Under the reference prior
the largest of the three is `r fmt(ea_new_max, 2)`. The unscaled specification
is kept in the sensitivity set as S1 (@sec-sensitivity), so its effect on
every quantity in the report can be read off directly.

### The Reference Prior

Every class that `get_prior()` lists receives an explicit prior, and nothing
stays at a brms default. Two reasons. BARG Step 1.C asks for the prior in
full, and an inherited default goes unreported by construction; the default
intercept prior, `student_t(3, median(y), 2.5 mad(y))`, is also read off the
data. Beyond that, `sample_prior = "only"`, which the prior predictive check
of @sec-prior-pred needs, requires every prior to be proper.

| Class | Prior | Reason |
|---|---|---|
| `b` (slopes) | `normal(0, SD_link)` per response | one prior SD equal to the whole observed spread of that response on its own link scale: an effect as large as the response's total variation is a priori about one SD out. Puts `r fmt(prior_in_rope(link_sd)[1], 2)`% of the prior inside the ROPE for all seven. |
| `Intercept` | see below | fixed from the measurement scale rather than from the sample |
| `sd` (locality) | `student_t(3, 0, 2)` | heavy-tailed half-$t$; the one class not rescaled per response, and varied in S3 and S4 |
| `sigma` | `student_t(3, 0, SD_link)` | half-$t$ scaled to the response, for the same reason as `b` |
| `phi` (zoib) | `gamma(2, 0.1)` | proper, mean 20, density zero at the origin. The brms default `gamma(0.01, 0.01)` piles mass arbitrarily close to zero, where the beta density is degenerate, and makes the prior predictive distribution unusable |
| `shape` (negbin) | `gamma(2, 0.1)` | same; the brms default `inv_gamma(0.4, 0.3)` has an extremely heavy right tail |
| `zoi`, `coi` | `beta(1, 1)` | uniform on (0, 1); this is also the brms default, stated here rather than inherited |
| `cor` (7x7) | `lkj(1)` | uniform over correlation matrices; varied in S5 |

The intercept priors are set from what a lithic analyst would have called
possible for a retouched flake tool before seeing these `r n_spec` pieces,
mapped through the link. Each is far wider than the observed spread.

```{r}
#| label: tbl-intercept-prior
kt(data.frame(
  Response = rlab(resps), Link = unname(link_lab[resp_fam]),
  Prior = unname(intercept_prior[resps]),
  `Implied 95% range on the response scale` = c(
    "6 - 100 mm", "12 - 134 mm", "GIUR 0.05 - 0.95", "0.05 - 0.95 of the perimeter",
    "3 - 143 scars", "26 - 104 degrees", "-0.9 - 6.9 generations"),
  `Observed mean` = c(sprintf("%.1f mm", mean(mod_dat$Thickness)),
                      sprintf("%.1f mm", mean(mod_dat$GMsize)),
                      sprintf("%.2f", mean(mod_dat$GIUR)),
                      sprintf("%.2f", mean(mod_dat$RLI)),
                      sprintf("%.0f", mean(mod_dat$NScar)),
                      sprintf("%.1f deg", mean(mod_dat$EdgeAngle)),
                      sprintf("%.2f", mean(mod_dat$RG))),
  check.names = FALSE),
  caption = "Intercept priors, fixed from the measurement scale rather than from the sample.")
```

One qualification attaches to the reference prior. The scale of the slope
prior and of the `sigma` prior is taken from the link-scale SD of the
response, which is a function of the data. This is a unit-setting use of the
data rather than a location-setting one: it fixes what "one unit of effect"
means without expressing any belief about where the effect lies, and it is the
standard construction behind weakly informative priors. It remains a mild
empirical-Bayes element. S1 and S2 in @sec-sensitivity vary this scale by
factors of roughly 1/8 to 2.5 and show what turns on it.

None of the priors used here is informed by previous work, because there is
none to draw on. No published posterior or meta-analytic estimate exists for
any of these seven measures in Quina assemblages; there is no prior literature
on, for instance, how edge angle varies with elevation above a channel. That
is why BARG item 5.B, sensitivity to a genuinely informed prior, has nothing
to attach to in this analysis, and the gap is itself a statement about how
little is established for these measures.

The complete prior is in Appendix A (@sec-priors-full): the rows
`prior_summary()` still labels `default` and why the sampler never reaches
them, the brms defaults that the explicit specification replaced, and the
generated Stan prior block, whose `lprior` statements are what was actually
added to the log posterior.

## Fitting and Model Checks {#sec-fitting}

brms (Bürkner 2017) fitted the model with the cmdstanr backend, running
`r mf$chains` chains of `r format(mf$iter, big.mark = ",")` iterations for
`r format(mf$ndraws, big.mark = ",")` post-warmup draws in total. Ten further
fits use identical settings; @sec-model-series says what each one is for.
@tbl-software gives versions, seeds and sampler settings, and @sec-computation
gives the convergence and resolution diagnostics for all eleven runs.

Three checks separate the fitted model from the results, and this section
states their outcomes so that a reader can decide how much of the evidence to
inspect. The prior predictive draws stay inside the physically possible range
of every measure, which is what that check is asked to establish (Appendix B,
@sec-prior-pred). The sampler converged; Appendix C (@sec-convergence) gives
R-hat, divergences and treedepth saturation for all eleven runs. On the posterior
predictive side (Appendix D, @sec-post-pred), the negative binomial reproduces
the overdispersion of the scar count, while the gaussian likelihood for
retouch generations smooths over their discreteness, a real limitation of that
family visible in @fig-post-pred. @fig-ppc-stats reports the test statistics
fixed in advance, among them the pile-up of GIUR and retouched-perimeter
values at exactly 1, against their predictive distributions.

---

# The Model Series {#sec-model-series}

Eleven fits underlie this report, and only one of them supplies the
numbers in the results tables. The other ten exist because a single fit cannot
show which of its answers depend on the analyst rather than on the scrapers.
Gelman, Vehtari, McElreath, et al. (2026, ch. 9) treat this as the step that
separates a Bayesian workflow from a Bayesian analysis: fit a series, keep the
earlier members, and use them to understand the one you report. Each member of
the series below differs from the reference in a single respect, so that any
movement in a result can be attributed to a named choice.

| Fit | How it differs from the reference | Why it was fitted | What it showed |
|---|---|---|---|
| `ref` | reference specification | supplies every number quoted outside the sensitivity appendix | @sec-q1, @sec-q2 |
| `prior_ref` | `sample_prior = "only"` | to see what the prior alone predicts, before the data are used | the prior stays inside the physically possible range of all seven measures (@sec-prior-pred) |
| `noland` | the three landscape terms removed, leaving the locality intercept alone | to test whether the landscape terms produce the between-locality variance, without depending on any single slope being well resolved | they absorb almost none of it (@tbl-icc-compare) |
| `ref_basinx` | `Basin * zHeight + Basin * zDistance` | to ask whether the answer to Q2 averages over two basins that behave differently | the contrasts are undecided, with a directional hint in Huangping (@sec-basinx) |
| `ref_rg_ln` | lognormal likelihood for retouch generations | the gaussian family smooths over the discreteness of that response, so the conclusions should be visible under an alternative | the verdicts are unchanged (@tbl-rg-alt) |
| `s1` | slope prior `normal(0, 1)`, unscaled | the specification `paper/_analysis.R` uses, and the one the per-response scaling replaced | the edge-angle slopes are prior-dominated under it (@tbl-ppsd) |
| `s2` | slope prior widened by 2.5 | the broad prior BARG 5.A asks for | @sec-5a |
| `s3` | locality SD prior `exponential(1)` | 26 clusters make this prior influential | @sec-5c |
| `s4` | locality SD prior `normal(0, 1)` truncated at zero | a lighter tail than the reference half-$t$ | @sec-5c, @sec-resolution-limits |
| `s5` | correlation prior `lkj(2)` | 21 correlations from 26 clusters are weakly identified | @sec-cor, @sec-5c |
| `s6` | `coi` fixed at 1 | a defect in the zero-one-inflated beta, found while reading the fitted parameters | the slopes and ICCs agree with the reference to within a few percent of a posterior SD (@sec-oneinflated) |

Two of these entries record a defect the analysis found in itself: `s1` the
unscaled slope prior that says something different about each response
(@sec-priors), and `s6` a `coi` parameter estimating the frequency of an
outcome that cannot occur (@sec-oneinflated). Neither defect showed up in the
research question, in the data or in the convergence diagnostics. Both
surfaced from comparing the fitted parameters with what the sample contains,
and that comparison is the part of the work a results table has no room for.

Fitting eleven models rather than one opens a familiar risk, and naming it is
part of reporting it. A reader cannot tell from the finished document how many
further specifications were tried and set aside, and a set of fits chosen after
seeing the results can be made to bracket whatever range the analyst prefers.
Two features of the design limit the room for that here. Each of `s1` to `s6`
changes exactly one line of the reference prior, so the set is determined by
the reference specification rather than by the results; and no fit was dropped,
so @tbl-diag-all lists all eleven, including the
`r sum(mf$diagnostics$divergent > 0)` that produced a handful of divergent
transitions. One thing the design cannot rule out is that a different analyst
would have chosen a different reference, and @sec-prior-dependence reports how
much of the answer that choice governs.

---

# Results {#sec-results}

::: {.barg}
Cf. BARG Step 4.E, which requires every decision to appear beside the estimate
and interval it rests on. Every decision table below gives all three.
:::

@fig-main answers both questions at a glance. The sections that follow give
the rule those answers are drawn from, then each question in turn, then how
much of the answer depends on the prior, and finally what the design cannot
settle.

```{r}
#| label: fig-main
#| fig-cap: "**A** (Q1): the locality share of link-scale variance for each response. Circles are the model with the landscape terms, triangles the intercept-only model, joined by a line. Filled circles mark responses whose 95% interval clears the 0.05 negligibility band (shaded). **B** (Q2): the 21 landscape slopes, each divided by the link-scale SD of its own response so that all seven share one axis and the ROPE is a single band at +/- 0.1. The basin slope is Huangping relative to Binchuan; height and distance are standardised, so their slopes are the change over one SD of the predictor. Rows are in the same order in both panels. Any slope whose posterior SD is still more than half its prior SD is drawn grey and hollow, because it has not been informed enough by the data to be read as an estimate."
SHOW("barg_main.png")
```

## The Decision Rule and the ROPE {#sec-decision-rule}

::: {.barg}
Cf. BARG Step 4.A.
:::

The substantive question asks how large an effect is and also whether it is
small enough to treat as absent. An interval alone answers only the first, and
so does a $p$-value, which can fail to reject and nothing more. The rule used
here is Kruschke's (2018) three-way decision against a region of practical
equivalence, or ROPE, meaning the band of values around the null value that
the analysis is willing to treat as no effect:

| Decision | Condition (either suffices) |
|---|---|
| **Reject the null value** | the 95% credible interval lies entirely outside the ROPE, **or** less than 2.5% of the posterior mass is inside it |
| **Accept the null value** | the 95% credible interval lies entirely inside the ROPE, **or** more than 95% of the posterior mass is inside it |
| **Undecided** | neither of the above |

An undecided verdict is a substantive outcome in its own right. It means the
data are compatible both with an effect large enough to matter and with none,
and @sec-min-rope turns that into a quantitative statement of what the design
was able to exclude.

### The ROPE and Its Justification {#sec-rope}

::: {.barg}
Cf. BARG Step 4.C.
:::

The ROPE half-width used in the text is 0.1 times the SD of the response on
its own link scale, so a slope counts as practically zero if moving a
predictor by one unit shifts the response by less than a tenth of that
response's total spread. Expressing the bound in link-scale SD units is what
lets one number serve seven responses on three link scales (@tbl-prior-rope).

Two qualifications attach to it. The first is that 0.1 is a convention. No
archaeological argument supports it, and none exists in the literature: nobody
has established how large a difference in edge angle or scar count a lithic
analyst would need before calling it a real difference in technological
behaviour, and this report constructs no such argument either.

The second qualification is more serious. BARG asks for ROPE limits fixed by
theoretical and practical considerations *before the results are seen*, and
this analysis fails that precondition, because the model had already been
fitted and inspected before the report was written. Choosing a half-width now,
with the posteriors in view, would allow the bound to be tuned until it
delivered a preferred verdict.

Reporting the whole curve answers the objection where defending a single
number would not. The posterior mass inside the ROPE appears below as a
continuous function of the half-width, so that a reader with a defensible
bound of their own can read the answer off at it. The equivalence $p$-value
curves of the companion analysis use the same device for the same reason.

```{r}
#| label: fig-rope
#| fig-cap: "Posterior mass inside the ROPE as a function of the ROPE half-width, in link-scale SD units, for all 21 slopes. Dashed vertical line: the 0.1 half-width quoted in the text. Dotted horizontals: the 2.5% and 95% thresholds of the decision rule. A curve that has not reached 95% by a half-width the reader considers archaeologically negligible is a slope this design cannot declare absent."
SHOW("rope_curves.png")
```

A reader who wants the mass at a particular half-width can read it off the
curve; `rope_curve()` in `barg_quantities.R` returns it numerically for any
grid of half-widths.

## Q1: Is There Technological Variation Between Localities? {#sec-q1}

Yes, for `r nrow(present)` of the seven measures under the reference prior,
and for `r nrow(firm)` of them under every prior tried.

For `r nrow(present)` responses
(`r paste(rlab(present$Response), collapse = ", ")`) the 95% interval for the
locality share of link-scale variance lies entirely above 0.05, so the rule of
@sec-decision-rule rejects the null value of a negligible locality effect.
Their posterior medians are
`r paste(fmt(present$Median, 2), collapse = ", ")`, meaning that between
`r fmt(min(present$Median) * 100, 0)`% and `r fmt(max(present$Median) * 100, 0)`%
of the link-scale variance in those measures sits between localities rather
than within them. The remaining `r nrow(undec)` are undecided: their intervals
run from near zero to values that would be substantial, and the design does
not separate the two.

```{r}
#| label: tbl-icc-decision
icc_ref |>
  arrange(desc(P_gt_rope)) |>
  transmute(Response = rlab(Response),
            `ICC median [95% ETI]` = ci(Median, CrI_lo, CrI_hi),
            `P(ICC > 0.05)` = fmt(P_gt_rope, 3),
            Decision = Verdict) |>
  kt(caption = "Q1 decisions. The ROPE for the ICC is [0, 0.05]: a locality share of link-scale variance below 5% is treated as negligible. Unlike the slope ROPE this bound is one-sided and has a natural reading -- 5% of variance is a share no archaeological argument would build on.")
```

Q1 is the only part of the analysis that returns a decisive answer, so how far
it survives the prior deserves setting out in detail (@tbl-sens-verdict).

`r if (nrow(firm) > 0) sprintf("For %s the decision is 'locality variation present' under every one of the seven priors.", paste(firm$Resp, collapse = " and ")) else ""`
`r if (nrow(flips) > 0) sprintf("For %s it is not: the 2.5%% limit sits so close to the 0.05 bound that %s pushes it just below, and the verdict becomes undecided (2.5%% limit %s under %s, against %s under the reference). The median barely moves (%s against %s), so the decision turns on a boundary rather than on a change in the estimate.", paste(unique(flip_detail$Resp), collapse = " and "), paste(unique(flip_detail$Spec), collapse = " and "), fmt(flip_detail$lo[1], 3), flip_detail$Spec[1], fmt(icc_sens$lo[icc_sens$Resp == flip_detail$Resp[1] & icc_sens$Spec == "REF"], 3), fmt(flip_detail$Median[1], 3), fmt(icc_sens$Median[icc_sens$Resp == flip_detail$Resp[1] & icc_sens$Spec == "REF"], 3)) else "No ICC decision changes under any of the seven priors."`

The safe reading is therefore the narrower one: **`r nrow(firm)` of the seven
measures show between-locality variation that no prior in the sensitivity set
can explain away**`r if (nrow(flips) > 0) sprintf(", and %s sits on the boundary", paste(unique(flip_detail$Resp), collapse = " and ")) else ""`.

The `noland` fit tests directly whether the landscape terms produce the
between-locality variation (@sec-model-series). It drops those three terms and
keeps the locality intercept. If basin, elevation and distance supplied the
between-locality signal, removing them would inflate the locality variance
component; equivalently, adding them to the intercept-only model would deflate
it. The comparison avoids the wide intervals on the individual slopes
altogether.

```{r}
#| label: tbl-icc-compare
icc_ref |> select(Response, m_ref = Median, lo_ref = CrI_lo, hi_ref = CrI_hi) |>
  left_join(icc_nol |> select(Response, m_nol = Median, lo_nol = CrI_lo, hi_nol = CrI_hi),
            by = "Response") |>
  transmute(Response = rlab(Response),
            `intercept only` = ci(m_nol, lo_nol, hi_nol, 3),
            `with landscape terms` = ci(m_ref, lo_ref, hi_ref, 3),
            `change in the median` = m_ref - m_nol,
            `as a share of the intercept-only median` = (m_ref - m_nol) / m_nol) |>
  kt(caption = "The ICC with and without the three landscape terms. A large negative change would mean the landscape terms had absorbed between-locality variance; a change near zero means they explain almost none of it.", digits = 3)
```

The three landscape terms absorb none of the between-locality variance. If
they explained it, the locality share would fall when they enter the model,
and it does not fall: the median change across the seven responses is
`r sprintf("%+.1f", med_rel * 100)`% of the intercept-only value, `r n_up` of
the seven go up and `r n_dn` go down, and the individual changes run from
`r sprintf("%+.0f", worst_j$rel * 100)`% (`r rlab(worst_j$Response)`) to
`r sprintf("%+.0f", best_j$rel * 100)`% (`r rlab(best_j$Response)`), the
largest of which starts from an intercept-only median of only
`r fmt(best_j$nol, 3)`, so that ratio is unstable rather than meaningful. The
pattern is noise scattered about zero. This is direct evidence on Q2, and it
survives however poorly any single coefficient is resolved: whatever separates
the localities, these three landscape variables do not supply it.

## Q2: Do the Landscape Variables Explain It? {#sec-q2}

The analysis identifies no landscape variable that matters, and for most of
them it also cannot establish that they do not.

Of the `r nrow(cf)` slopes, at the 0.1 SD ROPE: `r n_rej` reject the null
value, `r n_acc` accept it, and `r n_und` are undecided. The intervals are too
wide for the ROPE. That is what a design delivers when `r n_loc` localities
supply every predictor and one binary contrast represents basin
(@sec-design-sample). @tbl-coef-full gives each of the `r nrow(cf)` decisions
beside the estimate, the interval and the ROPE mass it rests on.

```{r}
#| label: prior-dom-readout
#| echo: false
dom <- ppsd_ref |> filter(ratio > 0.5)
```

`r if (nrow(dom) == 0) sprintf("No slope in Panel B is prior-dominated: the largest posterior/prior SD ratio over all %d is %s, so every point in the panel is an estimate the data have shaped.", nrow(ppsd_ref), fmt(max(ppsd_ref$ratio), 2)) else sprintf("Prior-dominated slopes, drawn grey and hollow in Panel B: %s. Their posteriors are still more than half as wide as their priors; they are shown for completeness and should not be interpreted.", paste(sprintf("%s ~ %s (ratio %s)", rlab(dom$Response), plab(dom$Predictor), fmt(dom$ratio, 2)), collapse = "; "))`

`r if (n_acc > 0) sprintf("What the analysis can state positively is confined to %s: %s.", if (n_acc == 1) "a single slope" else sprintf("%d slopes", n_acc), paste(sprintf("%s with %s (%s on the standardised axis, %s%% of the posterior inside the ROPE)", rlab(acc$Response), tolower(plab(acc$Predictor)), ci(acc$Std_median, acc$Std_lo, acc$Std_hi, 2), fmt(acc$pct_in_ROPE, 1)), collapse = "; ")) else "No slope can be positively called practically zero at this bound either."`

Direction certainty tells the same story from the other side. The largest
value over all `r nrow(cf)` slopes is `r fmt(max(cf$Dir_certainty), 3)`
(`r rlab(dir_max$Response[1])` ~ `r plab(dir_max$Predictor[1])`); a value of
0.5 would mean the posterior is split evenly about zero. Nothing here
approaches the near-certainty of sign that would be needed to claim a
directional landscape effect.

### What This Design Could Have Decided: the Smallest Acceptable ROPE {#sec-min-rope}

For each slope there is a smallest ROPE half-width at which the 95% interval
would have fitted entirely inside, namely $\max(|\text{lower}|,
|\text{upper}|)$. Anything wider than that, this design *can* declare
practically zero; anything narrower, it cannot. Translating that bound into
the response's own units turns "we could not decide" into a quantitative
statement of what the study was able to rule out. The last column of
@tbl-coef-full gives that bound for every slope, in the response's own units:
an effect smaller than it is what the study can rule out, an effect larger than
it is what the study cannot.

The slopes with a *small* minimum acceptable ROPE are where the "undecided"
label most misleads. For the three best-resolved slopes the 95% interval sits
within `r fmt(max(best$Min_ROPE / link_sd[best$Response]), 2)` SD, so the
analysis declines to call them zero at a tenth of an SD while still ruling out
anything larger than the following:
`r paste(sprintf("%s with %s, +/- %s %s", rlab(best$Response), tolower(plab(best$Predictor)), fmt(abs(best$Min_ROPE_resp), 0), best$Min_ROPE_unit), collapse = "; ")`.
That is a real, if modest, constraint. At the other extreme,
`r paste(sprintf("%s ~ %s", rlab(worst2$Response), plab(worst2$Predictor)), collapse = "; ")`
are effectively uninformative: their intervals span
`r fmt(max(worst2$Min_ROPE / link_sd[worst2$Response]), 1)` SD or more, and no
bound a reader would accept as negligible lies inside them.

## Are the Landscape Effects the Same in Both Basins? {#sec-basinx}

::: {.barg}
Cf. BARG Step 5. A structural extension rather than a prior change.
:::

@sec-q2 asks whether the landscape explains the differences between localities
and answers no. The obvious follow-up is whether that "no" is uniform, or
whether it averages over two basins that behave differently.

The localities cannot answer it. All three predictors are locality attributes,
so a slope that varies by locality is unidentified: within a locality the
predictor never moves, and $a_j + b_j x_j$ collapses to a single number
however many specimens that locality contributes (@sec-design-sample). The
basin is the finest level at which the question can be put, because height and
distance do vary across the localities inside each basin.

```{r}
#| label: basinx-readout
#| echo: false
bx      <- basin_slopes(gf("ref_basinx"))
bx_dif  <- bx |> filter(Quantity == "difference")
bx_bin  <- bx |> filter(Quantity == "Binchuan")
bx_hua  <- bx |> filter(Quantity == "Huangping")
bx_rej  <- bx_dif |> filter(Verdict == "reject the null value")
bx_wide <- median((bx_hua$CrI_hi - bx_hua$CrI_lo) /
                  (bx_bin$CrI_hi - bx_bin$CrI_lo))
bx_joint_h <- basin_joint_positive(gf("ref_basinx"), "zHeight")
bx_loc <- mod_dat |> distinct(Locality, Basin, Height) |>
  group_by(Basin) |> summarise(n = n(), lo = min(Height), hi = max(Height),
                               .groups = "drop")
gl <- function(b, f) bx_loc[[f]][bx_loc$Basin == b]
```

Binchuan's `r gl("Binchuan", "n")` localities span
`r gl("Binchuan", "lo")`--`r gl("Binchuan", "hi")` m above the channel and
Huangping's `r gl("Huangping", "n")` span
`r gl("Huangping", "lo")`--`r gl("Huangping", "hi")` m, so there is a gradient
inside each basin to fit a slope to.

`ref_basinx` therefore replaces the two shared gradient slopes with one of each
per basin, writing `Basin * zHeight + Basin * zDistance` and leaving the
reference model otherwise untouched. With two basins this is a fixed contrast
rather than a random slope, since two groups supply no variance to estimate.
Binchuan is the reference level of the factor, so the interaction coefficient
gives the Huangping slope minus the Binchuan one, and the Huangping slope
itself forms draw by draw as their sum.

### The Difference Between the Two Basins

```{r}
#| label: tbl-basinx-diff
bx_dif |>
  transmute(Response = rlab(Response),
            Gradient = ifelse(Gradient == "zHeight", "Height", "Distance"),
            `std median` = Median, `2.5%` = CrI_lo, `97.5%` = CrI_hi,
            `% in ROPE` = pct_in_ROPE, Verdict) |>
  arrange(desc(abs(`std median`))) |>
  kt(caption = "The 14 basin contrasts: the Huangping gradient slope minus the Binchuan one, standardised by the link-scale SD of the response so that the ROPE is the same band used everywhere else in this report.")
```

Of the `r nrow(bx_dif)` contrasts, `r nrow(bx_rej)` clears the ROPE and
`r sum(bx_dif$Verdict == "undecided")` are undecided.
`r if (nrow(bx_rej) > 0) sprintf("The one that clears it is %s with %s, and its interval reaches %s at the near end. One result out of %d comparisons, with a limit that close to zero, is what chance produces.", paste(sprintf("%s with %s", rlab(bx_rej$Response), tolower(ifelse(bx_rej$Gradient == "zHeight", "height", "distance"))), collapse = "; "), "the gradient", fmt(min(abs(bx_rej$CrI_lo)), 2), nrow(bx_dif)) else "None clears it."`
Taken one at a time, then, the contrasts do not establish that the basins
differ.

### The Two Basins' Slopes Side by Side

The contrasts give only part of the picture, because a difference can be
undecided while the two slopes it separates look nothing alike. Binchuan's
slopes are estimated from its `r gl("Binchuan", "n")` localities and
Huangping's from its `r gl("Huangping", "n")`, on the same standardised axis as
the contrasts above.

The pattern is one-sided.
`r sum(bx_bin$Median > 0)` of the `r nrow(bx_bin)` Binchuan slopes are positive
and `r sum(bx_bin$Median < 0)` negative, scattered about zero, and
`r sum(bx_bin$excl_zero)` have an interval excluding zero. All
`r nrow(bx_hua)` Huangping slopes are positive, and `r sum(bx_hua$excl_zero)`
have an interval excluding zero. On its face that is the shape the question was
asking about: a landscape signal in one basin and none in the other.

Three things argue for reading it as a hint rather than a finding. The
Huangping intervals are `r fmt(bx_wide, 1)` times as wide as Binchuan's, so
correspondingly less information sits behind the larger medians; a big point
estimate with a wide interval is what a small sample looks like. The intervals
that exclude zero do so by a margin of
`r fmt(min(bx_hua$CrI_lo[bx_hua$excl_zero]), 2)` at the near end. The
consistency of sign is also weaker evidence than it appears: the posterior
probability that all `r length(resps)` Huangping height slopes are
*simultaneously* positive is **`r fmt(100 * bx_joint_h, 0)`%**, well short of
the near certainty that `r nrow(bx_hua)` positive medians might suggest.

This design could ask the question and could not answer it. Huangping
contributes `r gl("Huangping", "n")` localities, and the number of localities
sets the resolution of every landscape question in this report
(@sec-design-sample). The Q2 answer therefore survives intact, with one
addition: the answer was tested for basin-dependence, the test came back
undecided, and a directional hint remains that a larger Huangping sample could
settle.

## How Far the Conclusions Depend on the Prior {#sec-prior-dependence}

Six fits differ from the reference in exactly one prior each, and Appendix F
(@sec-sensitivity) reports every ICC and every slope under all seven. The
summary is short.

Of the `r nrow(changed)` quantities, **`r n_changed`**
`r if (n_changed == 1) "receives" else "receive"` a different
decision under at least one of the seven priors`r if (n_changed > 0) paste0(": ", paste(changed$Quantity[changed$n_distinct_dec > 1], collapse = "; ")) else ""`.
`r if (n_changed > 0) "Those rows are the ones a reader should treat as prior-dependent; the rest are not." else "No conclusion in this report turns on the choice among the seven priors."`
The largest movement of a
posterior median across the seven priors, relative to the average width of that
quantity's own 95% interval, is
**`r fmt(worst_shift$rel[1] * 100, 1)`%** of the interval width, for
`r worst_shift$Quantity[1]`; for
`r sum(worst_shift$rel < 0.1)` of the `r nrow(worst_shift)` quantities the
movement is under a tenth of the interval width.

The edge-angle ICC is the quantity most exposed to the prior, for a reason
that deserves stating in its own right, and @sec-resolution-limits gives it.

## What This Study Can and Cannot Resolve {#sec-resolution-limits}

The three limits set by the design itself are stated in @sec-design-sample:
`r n_loc` localities rather than `r n_spec` specimens as the effective sample
for Q2, one basin contrast, and assemblages of very unequal size. Three
further limits arise from the model rather than from the sample.

1. The 21 correlations are weakly identified. Estimating a 7 x 7 correlation
   matrix from `r n_loc` clusters asks a great deal, so @fig-cor describes the
   fitted structure rather than reporting a set of findings. S5 shows how much
   the LKJ concentration moves them.

2. The `sd` prior is unscaled across responses, and the results show where
   that matters. The reference `student_t(3, 0, 2)` on the locality SD keeps
   one scale for all seven responses, unlike the slope prior (@sec-priors).
   For edge angle, measured in degrees with a link-scale SD of
   `r fmt(link_sd[["EdgeAngle"]], 1)`, a prior of scale 2 is restrictive; for
   log thickness, whose link-scale SD is `r fmt(link_sd[["Thickness"]], 2)`,
   the same prior is extremely loose. The sensitivity analysis shows the
   consequence in the numbers. The edge-angle ICC is
   `r ci(ea_ref$Median, ea_ref$lo, ea_ref$hi, 3)` under the reference prior
   and `r ci(ea_min$Median, ea_min$lo, ea_min$hi, 3)` under `r ea_min$Spec`,
   the largest prior-driven movement of any ICC in the set, and it happens to
   the response whose measurement scale is furthest from the prior's.
   `r ea_min$Spec` replaces the heavy-tailed half-$t$ with a prior whose tail
   dies quickly, which on a scale measured in degrees is a strong statement.
   The refinement this calls for is the one already applied to the slopes:
   scale the locality SD prior to each response's own link-scale SD. That
   specification fell outside the one-at-a-time set fitted here, and it is the
   clearest open refinement of this model. Until someone makes it, the
   edge-angle ICC should be read as prior-sensitive. The measures behind the
   Q1 answer are far less exposed, because their link-scale SDs are small
   enough that `student_t(3, 0, 2)` is uninformative for them:
   `r paste(firm$Resp, collapse = " and ")` keep the same decision
   under all seven priors`r if (nrow(flips) > 0) sprintf(" while %s changes only because its 2.5%% limit is already touching the bound", paste(unique(flip_detail$Resp), collapse = " and ")) else ""`.

3. Retouch generations are discrete and modelled as continuous
   (@sec-post-pred), and GMsize contains Thickness (@sec-data) while the model
   assumes independent residuals across responses.

---

# Reproducibility {#sec-reproducibility}

::: {.barg}
Cf. BARG Step 6.
:::

## Install the Software

```
R                4.5.2 or later
CmdStan          2.39.0     (install with cmdstanr::install_cmdstan())
R packages       brms, cmdstanr, posterior, bayesplot, here, readxl,
                 dplyr, tidyr, ggplot2, patchwork, knitr, quarto (CLI)
```

The backend must be cmdstanr, and the CmdStan version above is part of the
specification; the caption of @tbl-software gives the reason.

## Retrieve the Files

| File | What it is |
|---|---|
| `data/Quina_scraper_surface.xlsx` | specimen measurements, sheet "Quina scraper" |
| `data/Site_information.xlsx` | locality attributes: basin, height above channel, distance to channel |
| `paper/barg/barg_data.R` | builds the analysis frame; defines `link_sd`, the ROPE unit |
| `paper/barg/barg_priors.R` | all seven prior specifications, in full |
| `paper/barg/barg_models.R` | the eleven model specifications, the sampler call, the diagnostics |
| `paper/barg/barg_quantities.R` | every derived quantity: ICC, ROPE curves, HDI, MCSE, decisions |
| `paper/barg/barg_context.R` | assembles the environments the pipeline runs the above in |
| `paper/barg/barg_theme.R` | the figure style, copied from the manuscript pipeline |
| `paper/barg/barg_figures.R` | every figure |
| `paper/barg/barg_dummy_data.R` | the synthetic dataset described below |
| `paper/barg/barg_report.qmd` | this document |
| `_targets.R` | the pipeline: what depends on what, and what is rebuilt when |
| `_targets/` | the store, with the eleven fits and the pipeline metadata |
| `paper/barg/figures/*.png` | the figures written by `barg_figures.R`, at 300 dpi |

The scripts in that list produce every result in this document:
`barg_quantities.R` the derived quantities, `barg_figures.R` the figures, and
the code chunks of `barg_report.qmd` itself the tables, reading the stored
fits. No result is transcribed by hand. @fig-mcmc-rank is the one figure drawn
inside the document rather than by `barg_figures.R`, for the reason given in
@sec-convergence. The tables that list rather than compute, such as the prior
specification of @sec-priors and the model series of @sec-model-series, are
written by hand and marked as such by having no code chunk. The rendered
document omits the code; a reader who wants it goes to the archive.

`paper/_analysis.R` fits the same model for the two summary tables of the
supplementary information, leaving the slope prior at `normal(0, 1)`, which is
specification S1 of @sec-sensitivity. None of the files above uses it, and
none of them modifies it.

**Archive DOI:** `<10.17605/OSF.IO/XXXXXX>`, to be minted on OSF or Zenodo at
submission. The deposit will contain the four source files above, the
`paper/barg/` scripts, the `_targets/` store with the fits, and the rendered
HTML.

If the specimen data cannot be released, `paper/barg/barg_dummy_data.R`
generates a synthetic dataset with the same columns, types, ranges, locality
structure and missingness pattern as the real one, so that a reader can verify
that the scripts run end to end without access to the measurements. It is
simulated from the fitted model and reproduces no individual specimen.

## Run the Analysis

```bash
Rscript -e "targets::tar_make()"
```

That is the whole command. The pipeline fits the eleven models, derives the
quantities, draws the figures and renders this document and the two manuscript
documents, in dependency order.

Only the fitting is expensive, and it runs once. Every later `tar_make()`
compares the stored hash of each target against its inputs and rebuilds only
what has changed: edit a prior in `barg_priors.R` and the eleven fits are
refitted; edit a figure in `barg_figures.R` and no fit is touched.

`targets::tar_visnetwork()` shows the graph and what in it is out of date, and
`targets::tar_make(names = "fit_ref")` builds one target and its dependencies
without touching the rest.

```{r}
#| label: fit-times-readout
#| echo: false
r_by_fit <- if (!is.null(mf$env_by_fit))
  mf$env_by_fit[names(mf$minutes), "r_version"] else
  rep(mf$r_version, length(mf$minutes))
r_all <- unique(sub("^R version ", "", r_by_fit))
```

The eleven runs took between `r fmt(min(mf$minutes, na.rm = TRUE), 1)` and
`r fmt(max(mf$minutes, na.rm = TRUE), 1)` minutes each on the machine that
produced this document, a median of
`r fmt(median(mf$minutes, na.rm = TRUE), 1)`,
`r if (length(r_all) == 1) sprintf("all of them under %s", r_all) else sprintf("across %d R versions (%s), the cache having been assembled over more than one session", length(r_all), paste(r_all, collapse = ", "))`.

## Session Information

```{r}
#| label: session
#| attr-output: "style='max-height:420px;overflow:auto;'"
cat("CmdStan version:", mf$cmdstan, "\n")
cat("Fits produced at:", format(mf$fitted_at, "%Y-%m-%d %H:%M:%S %Z"), "\n\n")
sessionInfo()
```

---

# Appendix A. Priors in Full {#sec-priors-full}

@sec-priors states the prior and the argument for scaling the slope prior to
each response. This appendix gives the complete record: every row
`prior_summary()` returns, the rows it still labels `default` and why the
sampler never reaches them, the generated Stan prior block, and the brms
defaults that the explicit specification replaced.

```{r}
#| label: tbl-prior-summary
ps <- as.data.frame(prior_summary(fit_ref))[
  , c("prior", "class", "coef", "group", "resp", "dpar", "lb", "ub", "source")]
n_inherit <- sum(ps$prior == "")
n_flat    <- sum(ps$prior == "" & ps$coef == "")
ps |>
  filter(prior != "") |>
  arrange(class, resp) |>
  select(-source) |>
  kt(caption = "prior_summary(fit_ref), every non-empty row.")
```

`r n_inherit` further rows have an empty `prior` field. Every one of them
names a specific `coef` and inherits the prior of its class; they are omitted
above. Rows that specify no `coef` and are still empty, meaning genuinely flat
priors, number `r n_flat`. That is the condition `sample_prior = "only"`
imposes, and it is why the prior predictive draws of @sec-prior-pred exist at
all: with a flat prior anywhere, brms cannot sample from the prior. The prior
predictive fit ran, so the condition is met.

```{r}
#| label: prior-source
#| echo: false
n_user <- sum(ps$prior != "" & ps$source == "user")
n_def  <- sum(ps$prior != "" & ps$source != "user")
def_rows <- ps |> filter(prior != "", source != "user") |>
  transmute(class, resp, group, prior)
```

`r n_user` of the `r n_user + n_def` non-empty rows are marked `user`. The
remaining `r n_def` are marked `default`, which needs explaining:

```{r}
#| label: tbl-prior-default-rows
kt(def_rows, caption = "The rows prior_summary() still labels 'default'.")
```

They are not priors the sampler uses. The `sd` rows name no `group`: they sit
one level above the `sd` + `Locality` rows that were set explicitly, and brms
takes the most specific match, so these parents are never reached. The `L` row
is brms's own restatement of the user-supplied `lkj(1)` in the Cholesky
parameterisation it actually samples in.

The authoritative record is the generated Stan program rather than
`prior_summary()`. Its prior block, extracted verbatim, is the complete list of
what was added to the log posterior:

```{r}
#| label: stan-lprior
lines <- strsplit(brms::stancode(fit_ref), "\n")[[1]]
lp    <- grep("lprior \\+=", lines)
last  <- max(lp)
while (grepl("^\\s*-", lines[last + 1])) last <- last + 1
cat(paste(trimws(lines[min(lp):last]), collapse = "\n"))
```

## What the brms Defaults Would Have Been {#sec-brms-defaults}

Leaving a class at its default means inheriting a prior without reporting it,
and two of the defaults here are read off the data rather than set
independently of it. They are listed so that the cost of the explicit
specification in @sec-priors is visible.

```{r}
#| label: tbl-default-priors
gp <- brms::get_prior(fit_ref$formula, data = mod_dat)
gp <- as.data.frame(gp)
gp <- gp[gp$prior != "" & !(gp$class %in% c("b", "sd")), ]
gp |>
  transmute(Class = class, Response = ifelse(resp == "", "--", rlab(resp)),
            Group = ifelse(group == "", "--", group),
            `brms default` = prior,
            Comment = case_when(
              class == "Intercept" ~ "data-dependent: centred and scaled on the observed response",
              class == "cor"       ~ "the 7 x 7 LKJ prior",
              class == "phi"       ~ "near-improper; mass arbitrarily close to zero",
              class == "shape"     ~ "very heavy right tail",
              class %in% c("zoi", "coi") ~ "uniform on (0, 1); proper, but unstated",
              TRUE                 ~ "half-t scaled on the observed response")) |>
  kt(caption = "The brms defaults for every class other than b and sd. None of them is used by the reference fit; each was replaced by an explicit prior. The Intercept and sigma rows are read off the data, which would make them empirical priors rather than the vague priors a reader would assume.")
```

# Appendix B. Prior Predictive Check {#sec-prior-pred}

::: {.barg}
Cf. BARG Step 1.E.
:::

The reference prior was refitted with `sample_prior = "only"`, so the draws
below come from the prior alone with the data unseen. The question is not
whether the simulated data look like the observed data. They should not; if
they did, the prior would be doing the data's work. The question is whether
they are physically possible: no negative thicknesses, edge angles inside the
range a goniometer can return, GIUR inside its definitional bounds, scar
counts within reach of a single flake.

```{r}
#| label: fig-prior-pred
#| fig-cap: "Prior predictive distributions (grey) against the observed distribution (dark), one panel per response. Axes are clipped to the physically possible range of each measure; the table below reports what falls outside."
SHOW("ppc_prior.png")
```

```{r}
#| label: tbl-prior-pred
ppd <- brms::posterior_predict(fit_prior, ndraws = 400)
bound <- list(Thickness = c(0, 150), GMsize = c(0, 250), GIUR = c(0, 1), RLI = c(0, 1),
              NScar = c(0, 400), EdgeAngle = c(0, 180), RG = c(0, 12))
kt(bind_rows(lapply(resps, function(r) {
  y <- as.vector(if (is.list(ppd)) ppd[[r]] else ppd[, , r])
  b <- bound[[r]]
  data.frame(Response = rlab(r),
             `physically possible range` = sprintf("%g - %g", b[1], b[2]),
             `prior-predictive median` = median(y),
             `prior-predictive 2.5%` = unname(quantile(y, 0.025)),
             `prior-predictive 97.5%` = unname(quantile(y, 0.975)),
             `% outside the possible range` = 100 * mean(y < b[1] | y > b[2]),
             check.names = FALSE)
})), caption = "Prior predictive draws against the physically possible range of each measure. GIUR and the retouched perimeter cannot leave [0, 1] under the zero-one-inflated beta, so their zero is structural.", digits = 2)
```

The prior predictive distributions are wide, deliberately so, since they
encode ignorance about the effects; but they stay inside the possible range of
every measure, and the prior passes.

# Appendix C. Computation, Convergence and Resolution {#sec-computation}

## Software and Settings {#sec-software}

::: {.barg}
Cf. BARG Step 2.A.
:::

```{r}
#| label: tbl-software
kt(data.frame(
  Item = c("R", "brms", "Stan backend", "CmdStan", "posterior", "chains",
           "iterations per chain", "warmup per chain", "post-warmup draws",
           "seed", "adapt_delta", "max_treedepth"),
  Value = c(mf$r_version, mf$brms, mf$backend, mf$cmdstan,
            as.character(utils::packageVersion("posterior")),
            mf$chains, mf$iter, mf$warmup, format(mf$ndraws, big.mark = ","),
            mf$seed, "0.8 (default)", "10 (default)")),
  caption = sprintf("Computational settings. Chains, iterations, warmup, seed, sampler settings, brms and CmdStan are identical for all eleven runs, %s. The backend is locked to cmdstanr: an rstan fit of the same model is not bit-for-bit reproducible against a cmdstanr fit, and the CmdStan version pins the C++ that sessionInfo() does not capture.", if (length(r_all) == 1) "as is the R version the R row names" else sprintf("while the R row lists the %d distinct R versions in the cache, which was assembled over more than one session", length(r_all))))
```

All results in this report come from the full run recorded above:
`r mf$chains` chains of `r format(mf$iter, big.mark = ",")` iterations and
`r format(mf$ndraws, big.mark = ",")` post-warmup draws, rather than from a
reduced pipeline test.

## Chain Convergence {#sec-convergence}

::: {.barg}
Cf. BARG Step 2.B.
:::

For the reference fit: divergent transitions **`r n_div`**, iterations at
maximum treedepth **`r n_tree`**, largest $\hat R$ over all
`r nrow(sm_ref)` estimated parameters **`r fmt(max(sm_ref$rhat), 4)`**.

The conclusions rest on more than the reference fit: the intercept-only
model supplies the Q1 comparison and the prior variants supply the sensitivity
analysis. The same diagnostics therefore appear for all eleven runs.

```{r}
#| label: tbl-diag-all
mf$diagnostics |>
  transmute(Fit = fit, `parameters` = n_par, `max Rhat` = max_rhat,
            `min bulk ESS` = round(min_ess_bulk),
            `min tail ESS` = round(min_ess_tail),
            `divergent transitions` = divergent,
            `at max treedepth` = max_treedepth) |>
  kt(caption = sprintf("Diagnostics for all eleven runs, each of %s post-warmup draws.",
                       format(mf$ndraws, big.mark = ",")), digits = 4)
```

```{r}
#| label: div-readout
#| echo: false
d_any <- mf$diagnostics |> filter(divergent > 0)
```

`r if (nrow(d_any) == 0) "No run produced a divergent transition." else sprintf("%d of the eleven runs produced a small number of divergent transitions (%s) against %s draws each, which is at most %s%% of the transitions in any run. Two observations bear on how much weight to put on them: the affected runs' Rhat and ESS are indistinguishable from the clean ones, and their posteriors coincide with the reference in @fig-sens-icc. The usual explanation for divergences at this rate in a hierarchical model with 26 clusters and a 7 x 7 correlation matrix is mild curvature in the posterior near zero variance, but this report has not located them in parameter space and so offers that as an inference rather than as a demonstration. All eleven runs are reported here in full.", nrow(d_any), paste(sprintf("%s: %d", d_any$fit, d_any$divergent), collapse = ", "), format(mf$ndraws, big.mark = ","), fmt(100 * max(d_any$divergent) / mf$ndraws, 3))`

@tbl-diag-all reduces each run to its worst value on each diagnostic, which is
the right summary for deciding whether a run is usable and the wrong one for
seeing how the chains behaved. @fig-mcmc-rank shows the chains themselves for
the reference fit. It uses rank plots rather than trace plots, on the argument
of Vehtari, Gelman, Simpson, et al. (2021), who introduced the $\hat R$ this
report quotes and proposed the rank plot alongside it: with
`r format(mf$ndraws, big.mark = ",")` draws a trace plot is an unreadable band
of ink, whereas ranking the draws across chains and counting each chain's
share of each rank bin stays legible at any chain length and exposes chains
that differ in location or in scale.

```{r}
#| label: rank-prep
#| echo: false
NBIN  <- 20L
nchn  <- length(unique(dr_ref$.chain))
ndrw  <- nrow(dr_ref)

icc_mat <- vapply(resps, function(r) icc_draws(dr_ref, r, resp_fam[[r]]),
                  numeric(ndrw))
colnames(icc_mat) <- paste0("ICC: ", rlab(resps))
sel      <- unique(c(sm_ref$variable[which.min(sm_ref$ess_bulk)],
                     sm_ref$variable[which.min(sm_ref$ess_tail)], "lp__"))
par_mat  <- vapply(sel, function(v) as.numeric(dr_ref[[v]]), numeric(ndrw))
rank_mat <- cbind(icc_mat, par_mat)

rank_long <- bind_rows(lapply(colnames(rank_mat), function(v) {
  rk <- rank(rank_mat[, v], ties.method = "random")
  data.frame(Quantity = v, chain = factor(dr_ref$.chain),
             bin = ceiling(rk / (ndrw / NBIN)))
})) |>
  count(Quantity, chain, bin) |>
  mutate(Quantity = factor(Quantity, levels = colnames(rank_mat)))

rank_ncol <- ceiling(sqrt(ncol(rank_mat)))
```

The `r ncol(rank_mat)` quantities in the figure are chosen by a rule fixed in
the code rather than by inspection. The seven ICCs are there because Q1 rests
on them and because the smallest effective sample size anywhere in the
analysis is one of them (@sec-resolution). The rest are the sampled parameter
with the lowest bulk ESS, the sampled parameter with the lowest tail ESS, and
the log posterior density, which come to `r length(sel)` panels because
`r if (length(sel) < 3) "one parameter is worst on both counts" else "the three are distinct"`.
A reader expecting the figure to reveal a problem will be disappointed, which
is its purpose: with $\hat R$ at `r fmt(max(sm_ref$rhat), 4)` and `r n_div`
divergent transitions the chains should look interchangeable, and the figure
lets a reader confirm that instead of taking the summary on trust.

```{r}
#| label: fig-mcmc-rank
#| fig-width: 7.5
#| fig-height: 5.6
#| fig-cap: "Rank plots for the reference fit. The draws of each quantity are pooled across the four chains and ranked; each line counts how many of one chain's draws fall in each of 20 equal rank bins. Under convergence every chain contributes an equal share of every bin, so the four lines should be flat and overlapping around the dashed expected count. A chain running high at one end of the axis and low at the other has explored a different part of the distribution from the others. The seven ICCs are derived draw by draw and diagnosed exactly like sampled parameters; the remaining panels are the log posterior density and the sampled parameters with the lowest bulk and tail effective sample sizes."
ggplot(rank_long, aes(bin, n, colour = chain)) +
  geom_hline(yintercept = ndrw / (NBIN * nchn), linetype = "dashed",
             linewidth = 0.3, colour = GREY) +
  geom_step(direction = "mid", linewidth = 0.4) +
  facet_wrap(~ Quantity, ncol = rank_ncol,
             labeller = label_wrap_gen(width = 24)) +
  scale_colour_manual(values = PAL[seq_len(nchn)], name = "chain") +
  scale_x_continuous(breaks = c(1, NBIN / 2, NBIN)) +
  labs(x = sprintf("rank of the draw among all %s, in %d equal bins",
                   format(ndrw, big.mark = ","), NBIN),
       y = "draws per chain") +
  fig_theme +
  theme(legend.title = element_text(size = 8.5))
```

```{r}
#| label: fig-convergence
#| fig-cap: "Rhat (A) and bulk effective sample size (B) for every estimated parameter, and for the seven ICCs, which are derived quantities computed draw by draw and diagnosed like any other."
SHOW("convergence.png")
```

## Chain Resolution {#sec-resolution}

::: {.barg}
Cf. BARG Step 2.C.
:::

Kruschke (2021) suggests an ESS of about 10,000, with the explicit
qualification that the figure aims at the stability of HDI limits and that
quantile-based limits are less demanding. This report is quantile-based
throughout (@sec-posterior-summary), so it gives both: the ESS, and the Monte
Carlo standard error of each quantile it reports.

```{r}
#| label: conv-prep
#| echo: false
full_tbl <- bind_rows(
  bind_rows(lapply(sm_ref$variable, function(v) {
    x <- dr_ref[[v]]; s <- draw_summary(x)
    d <- sm_ref[sm_ref$variable == v, ]
    data.frame(Parameter = v, s, rhat = d$rhat,
               ess_bulk = d$ess_bulk, ess_tail = d$ess_tail)
  })),
  icc_ref |> transmute(Parameter = paste0("ICC_", Response), Median, CrI_lo, CrI_hi,
                       Mode, HDI_lo, HDI_hi, MCSE_lo, MCSE_med, MCSE_hi, SD,
                       rhat, ess_bulk, ess_tail))

is_offset <- startsWith(full_tbl$Parameter, "r_Locality__")
is_cor    <- startsWith(full_tbl$Parameter, "cor_Locality__")
shown_tbl <- full_tbl[!is_offset & !is_cor, ]
```

The table below prints the `r nrow(shown_tbl)` quantities this report quotes
or discusses. The `r nrow(full_tbl)` quantities of the full model also include
`r sum(is_offset)` locality offsets and `r sum(is_cor)` locality-intercept
correlations, which @fig-caterpillar and @tbl-cor already give; printing them
again here would bury the rows a reader needs in a table of
`r nrow(full_tbl)`. The summaries below the table are computed over all
`r nrow(full_tbl)`, and the archive contains the complete version.

```{r}
#| label: tbl-convergence
#| tbl-cap: "Convergence and resolution for the slopes, the intercepts, the auxiliary parameters and the locality SDs, and for the seven derived ICCs. Median and equal-tailed limits are given with their Monte Carlo standard errors; mode and HDI limits are given alongside, following the reference BARG document. The locality offsets and the locality-intercept correlations are left out of the printed table: @fig-caterpillar draws the offsets, @tbl-cor gives the correlations with their own diagnostics, and the complete table is deposited with the archive (@sec-reproducibility)."
shown_tbl |>
  transmute(Parameter,
            `median` = Median, `2.5%` = CrI_lo, `97.5%` = CrI_hi,
            `mode` = Mode, `HDI lo` = HDI_lo, `HDI hi` = HDI_hi,
            `MCSE 2.5%` = MCSE_lo, `MCSE median` = MCSE_med, `MCSE 97.5%` = MCSE_hi,
            `Rhat` = rhat, `ESS bulk` = round(ess_bulk), `ESS tail` = round(ess_tail)) |>
  kt(caption = NULL, digits = 4)
```

```{r}
#| label: ess-summary
#| echo: false
min_bulk <- min(full_tbl$ess_bulk, na.rm = TRUE)
min_tail <- min(full_tbl$ess_tail, na.rm = TRUE)
worst    <- full_tbl$Parameter[which.min(full_tbl$ess_bulk)]
rel_all  <- with(full_tbl, MCSE_med / SD)
rel_all  <- rel_all[is.finite(rel_all)]
mcse_rel <- max(rel_all)
mcse_par <- full_tbl$Parameter[which.max(replace(with(full_tbl, MCSE_med / SD),
                                                 !is.finite(with(full_tbl, MCSE_med / SD)), -Inf))]
n_below_10k <- sum(full_tbl$ess_bulk < 10000, na.rm = TRUE)
```

Smallest bulk ESS: **`r format(round(min_bulk), big.mark = ",")`**
(`r worst`); smallest tail ESS:
**`r format(round(min_tail), big.mark = ",")`**.
`r n_below_10k` of the `r nrow(full_tbl)` quantities have a bulk ESS below
Kruschke's 10,000, so the recommendation is not met everywhere even at
`r format(mf$ndraws, big.mark = ",")` draws. What matters for a
quantile-based report is the Monte Carlo error of the quantiles themselves.
The largest MCSE of a reported median, as a fraction of that parameter's own
posterior SD, is **`r fmt(mcse_rel, 4)`** (`r mcse_par`), so the simulation
noise in the numbers this report quotes is about
**`r format(round(1 / mcse_rel), big.mark = ",")` times smaller** than the
posterior uncertainty it describes. @tbl-convergence gives the per-parameter
MCSEs for anyone who wants to check a particular one.

## How Many Digits the Simulation Supports {#sec-digits}

Gelman, Vehtari, McElreath, et al. (2026, sec. 11.6) argue that the number of
digits to report should follow from the Monte Carlo error and the posterior
uncertainty rather than from a printing default, and they take two significant
digits as the usual answer. Since the previous section has already computed
the Monte Carlo standard error of every quantile this report quotes, the rule
can be applied to this document rather than merely cited.

```{r}
#| label: digits-audit
#| echo: false
dig <- full_tbl |>
  filter(is.finite(MCSE_med), MCSE_med > 0, is.finite(SD), SD > 0) |>
  mutate(
    d_mcse = floor(-log10(2 * MCSE_med)),
    d_post = floor(-log10(SD)) + 2)
d_min   <- min(dig$d_mcse)
d_med   <- median(dig$d_mcse)
n_bound <- sum(dig$d_post > dig$d_mcse)
n_free  <- sum(dig$d_post <= dig$d_mcse)
n_over3 <- sum(dig$d_mcse < 3)
```

Each of the `r nrow(dig)` quantities in the model has two ceilings. The
simulation supports decimal places up to the point where the Monte Carlo
standard error stays below half a unit in the last digit. The posterior
supports decimal places up to two significant digits of its own standard
deviation, beyond which further digits describe the width of a distribution
the data have not resolved that finely. Across the `r nrow(dig)` quantities the
first ceiling runs from `r d_min` to `r max(dig$d_mcse)` decimal places with a
median of `r d_med`; it is the binding one for `r n_bound` quantities, and the
posterior's own width binds first for the other `r n_free`.

The audit is unflattering to the tables. They print three decimal places by
default and four in @tbl-convergence, while `r n_over3` of the `r nrow(dig)`
quantities support fewer than three. Trimming a shared `digits` argument would
not fix it, because the ceiling varies from `r d_min` to
`r max(dig$d_mcse)` across rows measured on three different link scales, so no
single setting is right for all of them. The per-row answer is already in the
table: the three MCSE columns beside every quantile say how far that row can
be read, and they are printed for this reason. A reader comparing two numbers
that differ only in the third decimal place should look at that row's MCSE
before treating the difference as real.

Where a single figure has to be chosen, two decimal places is what the rule
endorses for a typical quantity here, and it is what the summaries in
@sec-summary and @tbl-sens-full use. Buying one further digit would mean
halving the Monte Carlo error, which needs four times the iterations.

## What This Report Does Not Check: Calibration {#sec-sbc}

Convergence diagnostics establish that the sampler explored the posterior of
the model that was coded. They cannot establish that the coded model is the
one intended, nor that the posterior it returns is calibrated. Simulation-based
calibration checking (Modrák, Moon, Kim, et al. 2025) is the tool for that: draw
parameters from the prior, simulate a dataset from them, refit, and check that
the rank of each drawn parameter within its own posterior is uniform across
many repetitions. This report does not perform it, and the honest account has
two parts, of which only the second is a real defence.

The first part is cost, and it is a weaker excuse than it looks. One fit of
this model takes about `r fmt(median(mf$minutes, na.rm = TRUE), 0)` minutes on
the machine that produced this document, taking the median over the eleven
runs (@sec-reproducibility), so the few hundred refits a calibration check needs
come to roughly
`r fmt(median(mf$minutes, na.rm = TRUE) * 300 / (60 * 24), 1)` days of
continuous computation, and less than that in parallel. That is affordable.
Gelman, Vehtari, McElreath, et al. (2026, sec. 14.3) add that a handful of
replications beats none, because ranks at the extremes of the range reveal
gross problems even in small numbers. Cost alone therefore does not justify
omitting the check.

The second part is specific to this model.
Calibration checking draws its parameters from the prior, so it checks the
model where the prior puts its mass. The intercept priors here are
deliberately wide, spanning 6 to 100 mm of thickness and 26 to 104 degrees of
edge angle before the data are seen (@tbl-intercept-prior), which
is what @sec-prior-pred asks of them. Datasets simulated from those priors
would include assemblages no lithic analyst would recognise, and calibration
verified over that range says little about the much narrower region the
posterior actually occupies. Gelman, Vehtari, McElreath, et al. (2026, sec.
14.3) identify exactly this tension between weakly informative priors and
calibration checking, and point to posterior SBC (Säilynoja, Schmitt, Bürkner
and Vehtari 2026), which conditions the simulation on the observed data and so
checks calibration where the answers live. They recommend it in particular for
assessing the size of any bias when convergence diagnostics show only faint
signs of trouble, which is the position of the
`r sum(mf$diagnostics$divergent > 0)` runs in @tbl-diag-all with a handful of
divergent transitions.

Posterior SBC on the reference fit is therefore the computational check this
analysis most clearly still owes, and a reader should treat the evidence in
this appendix as evidence about the sampler rather than about the calibration
of the model.

# Appendix D. Posterior Predictive Checks {#sec-post-pred}

::: {.barg}
Cf. BARG Step 3.A. Kruschke (2021) treats this check as a precondition for
reporting parameter estimates at all.
:::

```{r}
#| label: fig-post-pred
#| fig-cap: "Posterior predictive distributions (grey, 100 draws) against the observed distribution (dark), one panel per response."
SHOW("ppc_posterior.png")
```

```{r}
#| label: fig-ppc-basin
#| fig-cap: "Posterior predictive check by basin: the observed median of each response within each basin, against the distribution of that same statistic in the replicated datasets."
SHOW("ppc_group_basin.png")
```

## The Checks That Were Specified in Advance {#sec-ppc-stats}

```{r}
#| label: fig-ppc-stats
#| fig-cap: "Targeted posterior predictive statistics. Diamond: the observed value. Open circle and bar: the median and 95% interval of the same statistic over the replicated datasets. Green means the observation falls inside the interval."
SHOW("ppc_stats.png")
```

```{r}
#| label: tbl-ppc-stats
ppc_stat_tbl |>
  mutate(Response = rlab(Response),
         inside = observed >= rep_lo & observed <= rep_hi,
         verdict = ifelse(inside, "reproduced", "NOT reproduced")) |>
  transmute(Response, Statistic, observed,
            `replicated median` = rep_median, `2.5%` = rep_lo, `97.5%` = rep_hi,
            verdict) |>
  kt(caption = "Observed test statistics against their posterior predictive distributions.")
```

```{r}
#| label: ppc-readout
#| echo: false
gs <- function(r, s) ppc_stat_tbl[ppc_stat_tbl$Response == r & ppc_stat_tbl$Statistic == s, ]
rg_neg <- gs("RG", "share below zero"); rg_min <- gs("RG", "minimum")
gi_one <- gs("GIUR", "share exactly 1"); rl_one <- gs("RLI", "share exactly 1")
ns_disp <- gs("NScar", "variance / mean")
```

### Does the Gaussian Likelihood for Retouch Generations Predict Negative Values?

The observed share of retouch-generation values below zero is
`r fmt(rg_neg$observed, 3)`; over the replicated datasets the share has median
`r fmt(rg_neg$rep_median, 3)` and 95% interval
[`r fmt(rg_neg$rep_lo, 3)`, `r fmt(rg_neg$rep_hi, 3)`]. The observed minimum
is `r fmt(rg_min$observed, 2)` against a replicated interval of
[`r fmt(rg_min$rep_lo, 2)`, `r fmt(rg_min$rep_hi, 2)`]. The gaussian's
support extends below zero, while the fitted mean sits far enough above it
relative to the residual SD that the model rarely generates impossible values
in practice. The gaussian misses the *discreteness* instead: retouch
generations are recorded as a small number of counted generations averaged
over at most three edges, so the observed distribution is a spike train and
the replicated one is smooth. @fig-post-pred shows it, and it is a real
limitation of the family, though it leaves a location parameter unbiased.

The check is a judgement call rather than a pass or a failure, so the
`ref_rg_ln` fit refits the same response as a lognormal for comparison
(@sec-model-series), letting a reader see directly whether the choice of
family moves the conclusions:

```{r}
#| label: tbl-rg-alt
ls_ln  <- sd(log(mod_dat$RG))
dr_ln  <- local({ f <- gf("ref_rg_ln"); d <- as_draws_df(f); rm(f); gc(); d })
rg_alt <- bind_rows(lapply(preds, function(pp) {
  x <- dr_ln[[paste0("b_RG_", pp)]] / ls_ln
  q <- eti(x); p <- 100 * mean(abs(x) < ROPE_SD)
  data.frame(Family = "lognormal (comparison)", Predictor = plab(pp),
             `standardised median` = median(x), `2.5%` = q[1], `97.5%` = q[2],
             `% in ROPE` = p,
             Verdict = rope_verdict(q[1], q[2], p, ROPE_SD),
             check.names = FALSE, row.names = NULL)
}))
bind_rows(
  cf |> filter(Response == "RG") |>
    transmute(Family = "gaussian (reference)", Predictor = plab(Predictor),
              `standardised median` = Std_median,
              `2.5%` = Std_lo, `97.5%` = Std_hi,
              `% in ROPE` = pct_in_ROPE, Verdict),
  rg_alt) |>
  arrange(Predictor, Family) |>
  kt(caption = "Retouch generations: the reference gaussian fit against a lognormal refit of the same response. Both are shown on their own standardised axis, so the ROPE is +/- 0.1 in each case.")
```

### Do the Zero-One-Inflated Beta Fits Reproduce the Pile-Up at 1?

For GIUR the observed share of exact ones is `r fmt(gi_one$observed, 3)`
against a replicated interval of
[`r fmt(gi_one$rep_lo, 3)`, `r fmt(gi_one$rep_hi, 3)`]; for the retouched
perimeter, `r fmt(rl_one$observed, 3)` against
[`r fmt(rl_one$rep_lo, 3)`, `r fmt(rl_one$rep_hi, 3)`]. The share of exact
zeros is 0 in the data and 0 in the replications. That agreement shows only
that the zero-inflation half of the family is estimating a probability the
data have pinned at essentially zero, and says nothing about whether the model
captures zero inflation. See the note in @sec-likelihood.

### Is the Overdispersion of the Scar Count Captured?

The observed variance-to-mean ratio is `r fmt(ns_disp$observed, 2)`, well
above the value of 1 a Poisson would imply; the replicated interval is
[`r fmt(ns_disp$rep_lo, 2)`, `r fmt(ns_disp$rep_hi, 2)`]. The negative
binomial captures it.

# Appendix E. Full Posterior Summaries {#sec-posterior-summary}

::: {.barg}
Cf. BARG Step 3.B.
:::

::: {.barg}
Every interval in this report is a 95% equal-tailed (percentile) credible
interval, running from the 2.5th to the 97.5th posterior percentile, and every
point summary in the text is the posterior median. That is the quantile-based
convention. The density-based pair, the posterior mode and the 95%
highest-density interval, appears alongside it in @tbl-convergence and
@tbl-coef-full, following the reference BARG document, so that a reader who
prefers that pair need recompute nothing. @sec-digits says how many of the
printed decimal places the simulation supports.
:::

```{r}
#| label: tbl-coef-full
#| tbl-cap: "The 21 landscape slopes in full, each with the decision it supports beside the estimate, the interval and the ROPE mass that decision rests on. `median`, `2.5%`, `97.5%` are on the link scale; `std` columns divide by the link-scale SD of the response, which is the axis the 0.1 ROPE is defined on; `response scale` translates the median to a percentage change per unit of the predictor (log and logit links) or leaves it in the response's own units (identity link). 'Direction certainty' is max(P(beta > 0), P(beta < 0)): 0.5 means the posterior is split evenly about zero, 1.0 that the sign is certain. The last column is the smallest ROPE half-width at which the 95% interval would have fitted entirely inside, in the response's own units (@sec-min-rope)."
cf |>
  transmute(Response = rlab(Response), Predictor = plab(Predictor), Link,
            median = Median, `2.5%` = CrI_lo, `97.5%` = CrI_hi,
            mode = Mode, `HDI lo` = HDI_lo, `HDI hi` = HDI_hi,
            `std median` = Std_median, `std 2.5%` = Std_lo, `std 97.5%` = Std_hi,
            `% in ROPE` = fmt(pct_in_ROPE, 1),
            `direction certainty` = fmt(Dir_certainty, 3),
            Decision = Verdict,
            `response scale [95% ETI]` = sprintf("%s [%s, %s]",
              fmt(Resp_median, 1), fmt(Resp_lo, 1), fmt(Resp_hi, 1)),
            unit = Min_ROPE_unit,
            `min ROPE, response scale` = sprintf("+/- %s %s",
              fmt(abs(Min_ROPE_resp), 1), Min_ROPE_unit)) |>
  kt(caption = NULL, digits = 3)
```

```{r}
#| label: fig-prior-post
#| fig-cap: "Prior (grey) against posterior (blue) for every slope, on the standardised axis. The shaded band is the ROPE. Where the two curves nearly coincide, the data have said little."
SHOW("prior_posterior.png")
```

## Locality-Level Variance and Correlation Structure {#sec-icc}

::: {.barg}
Cf. BARG Step 3.B, for the parameters that answer Q1.
:::

The intraclass correlation is
$\mathrm{ICC}^{(r)} = \sigma_u^{(r)2} / (\sigma_u^{(r)2} + \sigma_{\text{resid}}^{(r)2})$,
computed **draw by draw** so that it acquires a full posterior. The residual
variance is taken on the link scale, and its form depends on the family:
$\sigma^2$ for the gaussian and lognormal responses;
$\log(1 + 1/\mu + 1/\text{shape})$ evaluated at the intercept for the
negative binomial; and, for the two zero-one-inflated beta responses,
$\psi'(\mu\phi) + \psi'((1-\mu)\phi)$, the logit-scale variance of the beta
component. For those two the ICC therefore describes the beta component alone
and ignores the inflation (@sec-likelihood).

```{r}
#| label: tbl-icc
icc_ref |>
  transmute(Response = rlab(Response), Family = unname(fam_lab[Family]),
            `locality SD` = sd_median, `SD 2.5%` = sd_lo, `SD 97.5%` = sd_hi,
            `ICC median` = Median, `2.5%` = CrI_lo, `97.5%` = CrI_hi,
            `ICC mode` = Mode, `HDI lo` = HDI_lo, `HDI hi` = HDI_hi,
            `P(ICC > 0.05)` = P_gt_rope, Rhat = rhat,
            `ESS bulk` = round(ess_bulk), Verdict) |>
  arrange(desc(`ICC median`)) |>
  kt(caption = "Locality SD and intraclass correlation, with diagnostics for the derived quantity.", digits = 3)
```

## The 21 Locality-Intercept Correlations {#sec-cor}

::: {.barg}
Cf. BARG Step 1.B and 3.B. These 21 parameters are part of the model and no
research question is about them, so they are reported here in full rather than
in the main narrative.
:::

```{r}
#| label: fig-cor
#| fig-cap: "Posterior median of each of the 21 locality-intercept correlations. Bold entries are those whose 95% equal-tailed interval excludes zero."
SHOW("correlation_matrix.png")
```

```{r}
#| label: tbl-cor
cor_tb |>
  transmute(`Response A` = rlab(A), `Response B` = rlab(B),
            median = Median, `2.5%` = CrI_lo, `97.5%` = CrI_hi,
            mode = Mode, `HDI lo` = HDI_lo, `HDI hi` = HDI_hi,
            `P(rho > 0)` = P_positive, Rhat = rhat, `ESS bulk` = round(ess_bulk)) |>
  arrange(desc(abs(median))) |>
  kt(caption = "All 21 correlations among the locality intercepts, ordered by absolute posterior median. With 26 clusters these are weakly identified; the LKJ(1) prior is uniform over correlation matrices and S5 varies it.", digits = 3)
```

```{r}
#| label: fig-caterpillar
#| fig-cap: "Locality offsets on the link scale, one panel per response, with 95% equal-tailed intervals. Localities are in one fixed order in every panel, largest assemblage first, with the specimen count in brackets. Reading down each panel shows partial pooling at work: the smaller the assemblage, the harder the offset is pulled towards zero and the wider its interval."
SHOW("locality_intercepts.png")
```

# Appendix F. Sensitivity Analysis in Full {#sec-sensitivity}

::: {.barg}
Cf. BARG Step 5.
:::

Seven fits, identical in every respect except the prior, each changing one
element of the reference specification. A full factorial would have made any
movement impossible to attribute, so the design here changes one thing at a
time (@sec-model-series). S1 is the unscaled `normal(0, 1)` slope prior
discussed in @sec-priors, which is also the specification `paper/_analysis.R`
uses for the main paper's summary tables (@sec-reproducibility).

S6 differs in kind from the rest. S1 to S5 vary how much the data are allowed
to say about a quantity the report decides on; S6 removes a parameter the data
cannot inform, so its curves in @fig-sens-icc and @fig-sens-slopes coincide
with the reference by construction (@sec-oneinflated).

```{r}
#| label: tbl-sens-specs
kt(data.frame(
  Fit = c("REF", names(prior_specs)[-1]),
  `Prior change` = c(prior_specs$REF$label,
                     sapply(prior_specs[-1], function(x) x$label)),
  Rationale = c(prior_specs$REF$note, sapply(prior_specs[-1], function(x) x$note)),
  check.names = FALSE, row.names = NULL),
  caption = "The seven prior specifications. REF is the reference; each of S1 to S6 changes exactly one line of it.")
```

## Broad Prior {#sec-5a}

::: {.barg}
Cf. BARG Step 5.A.
:::

S2 widens every slope prior by a factor of 2.5, to
`normal(0, 2.5 x SD_link)`. This is the broad prior the guideline asks for: if
the posterior barely moves when the prior becomes two and a half times more
permissive, the result rests on the data instead of on the prior. S3 and S4
play the same role for the locality SD, and S5 for the correlation matrix.

## Variations of the Default Priors {#sec-5c}

::: {.barg}
Cf. BARG Step 5.C.
:::

```{r}
#| label: fig-sens-icc
#| fig-cap: "Q1 under all seven priors: the cumulative posterior distribution of each ICC. The shaded band is the ICC ROPE [0, 0.05]. A curve's height where it crosses the right edge of the band is P(ICC <= 0.05)."
SHOW("sensitivity_icc.png")
```

```{r}
#| label: fig-sens-slopes
#| fig-cap: "Q2 under all seven priors, for the slopes most exposed to the prior: the three edge-angle slopes (the response that was prior-dominated under the original specification) and the height slope of every response. Cumulative distributions on the standardised axis, with the ROPE shaded."
SHOW("sensitivity_slopes.png")
```

The curves are cumulative distributions rather than densities, because this
report's summaries are quantiles: a horizontal slice at 0.025 and 0.975 reads
off the interval limits directly, and the vertical distance between two curves
at the ROPE edge is the change in the quantity the decision rule uses.

BARG asks for the numbers as well as the pictures, and for every parameter
rather than only the plotted ones:

```{r}
#| label: tbl-sens-full
#| tbl-cap: "Posterior median and 95% equal-tailed interval for every ICC and every slope, under each of the seven priors. Slopes are standardised by the link-scale SD of their response."
sens_long |>
  mutate(cell = ci(Median, lo, hi, 2)) |>
  select(Quantity, Spec, cell) |>
  pivot_wider(names_from = Spec, values_from = cell) |>
  kt(caption = NULL)
```

## Do the Conclusions Change? {#sec-5e}

::: {.barg}
Cf. BARG Step 5.E. Some movement is inevitable, so the criterion is whether
the numbers moved enough, relative to the width of the posterior interval and
relative to the ROPE, to change a conclusion.
:::

@sec-prior-dependence gives the summary and quotes the largest movement of a
posterior median across the seven priors. The table here gives the detail
behind the decisions: what each quantity is judged to be under each prior.
@tbl-sens-full above supplies the estimates those judgements come from.

```{r}
#| label: tbl-sens-verdict
kt(changed |> select(-n_distinct_dec),
   caption = "The decision each quantity receives under each prior.")
```

---

# Appendix G. BARG Conformance {#sec-barg}

Kruschke (2021) sets out a checklist for reporting a Bayesian analysis. The
table below places every item on it against the section that answers it. The
same checklist can be walked from the other direction: each section of this
document opens with a shaded box naming the item it addresses.

This table sits at the end rather than at the front because the narrative of
this document follows the order of the work rather than the order of the
checklist (@sec-summary). A reader auditing coverage should start here; a
reader following the argument will have passed every row already.

| BARG item | How it is addressed | Where |
|---|---|---|
| **Preamble A.** Why a Bayesian approach is used | Seven responses on three link scales share one locality-level structure in a single model; derived quantities acquire genuine intervals; a null value can be accepted rather than merely not rejected; no large-sample approximation is needed in a design with few clusters | @sec-why-bayes |
| **Preamble B.** The goals of the analysis | Estimation of each landscape effect and of the locality variance, and decisions about null values. Prediction and model comparison are not goals | @sec-goals |
| **1.A.** Data variables, their roles and types | Seven responses, three locality-level predictors, one grouping factor, with the observed range of each; the complete-case filter is audited step by step; the estimand and the design that produced the sample are stated separately | @sec-design, @sec-data, @tbl-data-vars, @tbl-dropped |
| **1.B.** The likelihood and its parameters, separating those of theoretical interest from nuisance parameters | One multivariate model over four distributional families, with the linear predictor written out; the slopes and the locality SDs answer the questions, the intercepts, dispersion parameters and 21 correlations do not | @sec-likelihood, @tbl-likelihood |
| **1.C.** The prior distribution and its justification | Every class given an explicit prior, the slope prior scaled to each response's own link-scale SD, with the share of prior mass inside each response's ROPE as the argument for scaling it | @sec-priors, @sec-priors-full |
| **1.D.** Formal specification of the model | model formula / Stan code available from the analysis code or archive | analysis code or archive |
| **1.E.** Prior predictive check | The reference prior refitted with `sample_prior = "only"`, and the simulated data checked against the physically possible range of each measure | @sec-prior-pred, @tbl-prior-pred |
| **2.A.** Software, versions and settings | R, brms, backend and CmdStan versions, chains, iterations, warmup, seed, `adapt_delta` and `max_treedepth`, for all eleven runs | @sec-software, @tbl-software |
| **2.B.** MCMC chain convergence | R-hat, divergent transitions and treedepth saturation, for the reference fit and for all eleven runs, with rank plots of the chains behind the reference fit | @sec-convergence, @tbl-diag-all, @fig-mcmc-rank |
| **2.C.** MCMC chain resolution | Bulk and tail effective sample size for every parameter and every derived ICC, plotted in @fig-convergence and tabulated with the Monte Carlo standard error of each quantile the report quotes; the locality offsets and correlations omitted from the printed table are in the archive; an audit of how many decimal places that error supports | @sec-resolution, @sec-digits, @tbl-convergence, @fig-convergence |
| **2.D.** Accuracy of any approximation used in place of exact sampling | Not applicable: every posterior in this document is a full MCMC sample, with no variational, Laplace or INLA approximation anywhere | Not applicable |
| **3.A.** Posterior predictive check | Full predictive distributions, a check of the within-basin medians, and test statistics fixed in advance | @sec-post-pred, @sec-ppc-stats, @tbl-ppc-stats |
| **3.B.** Summary of the posterior of every parameter | Posterior median and 95% equal-tailed interval, with mode and 95% HDI alongside, for the 21 slopes, the seven locality SDs and ICCs, and the 21 correlations | @sec-posterior-summary, @sec-icc, @sec-cor |
| **3.C.** Bayes factors and posterior model probabilities | Not applicable: no model comparison is performed, so no marginal likelihood, no bridge sampling and no Bayes factor appears anywhere in this document | Not applicable (@sec-goals) |
| **4.A.** The decision procedure and its rule | Kruschke's (2018) three-way decision (reject, accept, undecided) against a region of practical equivalence | @sec-decision-rule |
| **4.B.** The loss or utility function, where decisions are made by formal decision theory | Not applicable: no cost attaches to being wrong about a landscape slope, so no loss function is specified; the ROPE curve reports the whole family of decisions instead of committing to one | Not applicable (@sec-rope) |
| **4.C.** ROPE limits and their justification | Half-width 0.1 of each response's link-scale SD, with the convention acknowledged and the posterior mass reported as a continuous function of the half-width | @sec-rope, @fig-rope |
| **4.D.** Bayes-factor decision thresholds | Not applicable: no Bayes factor is used, so no threshold on one is needed | Not applicable (@sec-goals) |
| **4.E.** Each decision reported with the estimate and interval it rests on | Every decision table gives the median, the 95% interval, the posterior mass inside the ROPE and the verdict in the same row | @sec-q1, @sec-q2 |
| **5.A.** Sensitivity to a broad prior | S2 widens every slope prior by a factor of 2.5; S3, S4 and S5 do the same job for the locality SD and for the correlation matrix | @sec-5a, @tbl-sens-specs |
| **5.B.** Sensitivity to a genuinely informed prior drawn from previous research | Not applicable: no published posterior or meta-analytic estimate exists for any of these seven measures in Quina assemblages, so no informed prior can be built; the gap is discussed where the priors are set out | Not applicable (@sec-priors) |
| **5.C.** Variations of the default priors | Six fits, each differing from the reference in exactly one prior, with every ICC and every slope reported under all seven specifications; the eleven fits and the purpose of each are set out together | @sec-model-series, @sec-5c, @tbl-sens-full |
| **5.D.** Sensitivity of the Bayes factor to the prior | Not applicable: no Bayes factor is computed. The item is singled out in the guideline because Bayes factors are prior-dependent in a way posterior estimates are not, which is part of why this analysis is framed as estimation | Not applicable (@sec-goals) |
| **5.E.** Whether the conclusions change | The decision each quantity receives under each of the seven priors, and the largest movement of a posterior median relative to that quantity's own interval width; and, for the one structural question the design can pose, whether the landscape slopes differ between the two basins | @sec-prior-dependence, @sec-basinx, @sec-5e |
| **6.** Reproducibility of data, code and computational environment | Software versions, the file list, the archive DOI, a synthetic dataset for the case where the specimens cannot be released, run instructions and `sessionInfo()` | @sec-reproducibility |


## What the Checklist Does Not Ask For

Three parts of this document answer to the second standard rather than to this
one, and a reader auditing BARG coverage will not find rows for them.
@sec-design states the estimand and the design before any model appears, so
that the resolution of both research questions is fixed in the open rather
than inferred from wide intervals later. @sec-model-series sets out all eleven
fits together with the reason each exists, including the two that record a
defect the analysis found in itself. @sec-sbc says which computational check
this report has not performed, why, and what a follow-up should do instead.
None of the three is required by Kruschke (2021), and each is standard in
Gelman, Vehtari, McElreath, et al. (2026).

# References

Bürkner, P.-C. (2017). brms: An R package for Bayesian multilevel models
using Stan. *Journal of Statistical Software*, 80(1), 1--28.

Gelman, A., Vehtari, A., & McElreath, R., with Simpson, D., Margossian, C. C.,
Yao, Y., Kennedy, L., Gabry, J., Bürkner, P.-C., Modrák, M., & Leos Barajas,
V. (2026). *Bayesian Workflow*. CRC Press.

Kruschke, J. K. (2018). Rejecting or accepting parameter values in Bayesian
estimation. *Advances in Methods and Practices in Psychological Science*,
1(2), 270--280.

Kruschke, J. K. (2021). Bayesian analysis reporting guidelines. *Nature Human
Behaviour*, 5, 1282--1291.

Modrák, M., Moon, A. H., Kim, S., Bürkner, P.-C., Huurre, N., Faltejsková, K.,
Gelman, A., & Vehtari, A. (2025). Simulation-based calibration checking for
Bayesian computation: the choice of test quantities shapes sensitivity.
*Bayesian Analysis*, 20, 461--488.

Säilynoja, T., Schmitt, M., Bürkner, P.-C., & Vehtari, A. (2026). Posterior
SBC: simulation-based calibration checking conditional on data. *Statistics
and Computing*, 36, 78.

Vehtari, A., Gelman, A., Simpson, D., Carpenter, B., & Bürkner, P.-C. (2021).
Rank-normalization, folding, and localization: an improved $\hat R$ for
assessing convergence of MCMC. *Bayesian Analysis*, 16(2), 667--718.
