---
title: "The Middle Paleolithic Quina valleys in southeastern margin of the Tibetan Plateau"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Qijun Ruan:
      correspondence: "yes"
      orcid: 0009-0000-2143-5335
      institute:
        - YPICRA
   - Zhenxiu Jia:
      correspondence: "no"
      orcid: 0000-0002-7514-4514
      institute:
        - ALPHA
   - Marco Peresani:
      correspondence: "no"
      orcid: 0000-0001-6562-6336
      institute:
        - FERRARA
        - IGAG
   - Davide Delpiano:
      correspondence: "yes"
      institute:
        - FERRARA
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
# <!-- TODO: update the author list, ORCIDs and affiliations for THIS paper -->
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - YPICRA: Yunnan Provincial Institute of Cultural Relics and Archaeology, Kunming, China
  - FERRARA: Department of Human Studies, Prehistoric and Anthropological Science Unit, University of Ferrara, Ferrara, Italy
  - IGAG: Consiglio Nazionale delle Ricerche-Institute of Environmental Geology and Geoengineering, Laboratory of Palynology and Palaeoecology, Research Group on Vegetation, Climate and Human Stratigraphy, Milan, Italy
title-block-published: "Last updated"
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
  tbl-title: "Table"
  tbl-prefix: "Table"
number-sections: true
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/elsevier-harvard.csl"
keywords:
  - Middle Paleolithic
  - Southeastern Tibetan Plateau
  - Quina technology
  - Quina valleys
abstract: |
  The Middle Paleolithic (~300–40 ka) was a pivotal period in human evolution, characterized by diverse lithic technologies. Conventional views have held that East Asia experienced slow technological development and lacked remarkable Middle Paleolithic innovations. Recent discoveries, however, demonstrate a more complex scenario, with distinctive Middle Paleolithic technologies identified, such as discoidal technology, Levallois technology and Quina technology. Here we present stone artifacts surface-collected from the Caifeng river and Sangyuan river in northwestern Yunnan province, geographically located in southeastern margin of the Tibetan Plateau. These artifacts mainly collected on the second and third river terraces. Diagnostic Quina technological elements have been identified, including plan-secant core reduction pattern, scalariform tool retouching, and branched reduction strategy. Our results demonstrate the existence of “Quina valleys” and expand the spatial range of Quina technological behavior in southwestern China. In addition, our study reinforces the technological diversity and complexity of the Chinese Middle Paleolithic.
editor_options:
  chunk_output_type: console
---

<!-- This is the format for text comments that will be ignored during rendering. Do not put R code in these comments because it will not be ignored. -->

Keywords: `r rmarkdown::metadata$keywords`

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(here)

```

```{=html}
<!-- Highlights: Elsevier requires 3–5 bullet points, each ≤ 85 characters (incl. spaces).
     These are submitted as a separate file at submission; kept here for drafting. -->
```

# Highlights {.unnumbered}

- 

- 

- 

- 

- 

# Introduction

The Middle Palaeolithic was a period of diversity in both hominin populations and lithic technology, and what drove that technological diversity remains contested: do assemblage differences reflect separate cultural traditions, or adaptive responses to different environments? The question has been pursued most intensively in the Mousterian Debate of western Europe, where the variability of Mousterian assemblages has been attributed to cultural tradition \[\], site function \[\], the constraints of locally available raw material \[\], and differing degrees of artefact reduction \[\]. The Quina Mousterian has been central to that debate.

The Quina Mousterian is one of the best-defined Middle Palaeolithic techno-complexes \[\], and what distinguishes it is not a tool type but a techno-economic system, the Quina system, whose defining features form three interlocking dimensions \[\]. The first dimension is blank production. Quina cores receive little preparation and are reduced through alternating, concatenated removals on sub-parallel and secant surfaces, with invasive detachments that optimise the volume of stone converted into blanks \[Bourguignon 1997; Turq 1989, 1992\]. The target product is a thick flake, asymmetric in cross-section, with a sharp cutting edge on one margin and a thick, often cortical, back opposite \[\]. This asymmetry is what allows such a tool to be repeatedly resharpened \[Turq 1992; Hiscock et al. 2009\]. The second dimension is edge maintenance, achieved through Quina retouch. Here the working edge is not merely produced but sustained, and resharpening follows a regular rhythm \[\]. Early episodes detach relatively large resharpening flakes and leave convex scars, giving a plano-convex edge and preserving volume for later work; as retouch proceeds it concentrates on the margin itself, producing smaller concave scars with stepped or hinged terminations and a plano-concave edge that keeps the tool serviceable \[\]. Each generation of retouch therefore restores a working edge without drawing on new raw material. The third dimension is ramification of the reduction sequence \[\]. Resharpening flakes detached during edge maintenance carry cutting edges of their own and can be used directly, as use-wear analysis has shown \[Claud et al. 2012\]; in some cases they were themselves retouched into tools \[\]. Quina flakes, besides being shaped into Quina scrapers, are thick enough to serve as cores for small flakes (i.e. core-on-flake) \[\]. Use-wear on scrapers and resharpening flakes from the same assemblages indicates that the two were put to different tasks, scraping and cutting respectively \[Delpiano et al. 2022\], so that ramification widened the functional range of the toolkit as well as extending its life. Efficient blank production, intensive edge maintenance and ramification together create an economy of stone use in which the thick Quina flake functions as a transportable and highly extendable package of raw material, yielding a large amount of working edge from a small initial investment \[Hiscock et al. 2009\].

Quina technology has been interpreted as an adaptation to high mobility in open environments where resources were patchily distributed and seasonally structured, and this interpretation rests on three lines of evidence. Faunal assemblages from Quina sites are typically dominated by a single gregarious and migratory species, most often reindeer, although equids and bovids predominate in some regions \[Delpech 1996; Delagnes and Rendu 2011; Delpiano et al. 2022; Castel et al. 2017\]. Settlement organisation is correspondingly structured, with residential camps surrounded by task-specific satellite locations given over to hunting and carcass processing \[Delagnes and Rendu 2011\]. At the landscape scale, Quina assemblages occur in open, lightly wooded settings in which prey were distributed across the terrain in a way that was patchy but not random \[Delpech 1996; Hiscock et al. 2009\]. Foragers under these conditions are argued to have relied on seasonal, high-mobility strategies of resource use and tool provisioning \[Hiscock et al. 2009; Delagnes and Rendu 2011; Castel et al. 2017\], and it is here that the technological economy described above becomes advantageous: versatile blanks, tools with long use-lives and cores requiring little preparation together lower the time and energy costs of replacing equipment on the move, and reduce the risk both of being left without an effective cutting edge far from a source of raw material and of carrying a toolkit ill-suited to the tasks that a given season and location happened to present \[Hiscock et al. 2009; Delagnes and Rendu 2011\].

The distinction between the Quina system and Quina retouch alone matters when its distribution is assessed. In the strict sense, with blank production, edge maintenance and ramification operating together, Quina is documented in western and southern Europe within a narrow interval between MIS 4 and early MIS 3, coinciding with pronounced global climatic instability \[Delagnes and Rendu 2011; Delpiano et al. 2022\]. Quina retouch taken on its own is far more widely distributed: in western Asia it has been reported in Acheulean–Yabrudian industries dating from as early as MIS 11 \[\], several hundred thousand years before the European Quina. The occurrence of so distinctive a way of maintaining a working edge in contexts this widely separated in time and space raises the possibility that Quina retouch is a recurrent solution to a recurrent problem rather than the signature of a single tradition. Neither the system nor Quina retouch, however, had been securely documented in the eastern Old World.

Whether East Asia has a Middle Palaeolithic in any meaningful sense has long been contested. On the conventional view, lithic technology in the region remained stable throughout the Palaeolithic, with substantial innovation deferred to the Upper Palaeolithic, so that no phase corresponding to the western Middle Palaeolithic can be recognised \[\]. Recent discoveries have provided new evidence that is reshaping this long-standing understanding. Discoid technology is documented at Lingjing (c. 125–90 ka) \[\], and Levallois technology at Guanyindong (c. 170–80 ka) \[\] and Jinsitai (c. 47–35 ka) \[\]. These finds indicate that the technological scenario of the East Asian Middle Palaeolithic is more complex than the conventional view allows. It is against this background that we identified a complete Quina system at Longtan, dated to c. 60–50 ka \[\], the first Quina site in East Asia to be unambiguously characterised, formally excavated, securely dated and systematically analysed. This discovery not only filled a long-standing gap in the eastern Old World distribution of Quina technology, but also provided new evidence for reassessing the diversity and complexity of the East Asian Middle Palaeolithic. More importantly, Longtan is unlikely to be an isolated case. Since 2010, surface collections bearing Quina features have been recovered from river terraces at a number of localities in the Huangping Basin, where Longtan is situated, and in the neighbouring Binchuan Basin. These finds suggest that Quina technology on the southeastern margin of the Tibetan Plateau may not be confined to a single site, but may instead represent a regional technological phenomenon of some spatial extent.

The southeastern margin of the Tibetan Plateau, where the Huangping and Binchuan basins lie, is the Hengduan Mountains region, an area of pronounced altitudinal gradients, complex landforms and diverse climatic regimes that ranks today among the richest biodiversity hotspots in the world \[\]. Geomorphologically, both basins belong to the dry-hot valley zone of the Jinsha River drainage, a typical fault-controlled valley system \[\]. Under the combined influence of high-relief topography and the monsoon, the region experiences marked seasonal drought and supports a predominantly open vegetation cover \[\]. This setting differs substantially from the cold, open environments of the European Quina sites, and equally from the settings of other Middle Palaeolithic assemblages so far reported in China. It is this regional ecological distinctiveness that makes Longtan and the surrounding Quina localities of particular interest. On the one hand, a technology usually associated with open landscapes, seasonally structured resources and high-mobility subsistence strategies here occurs in a dry-hot valley on the southeastern margin of the Tibetan Plateau, and the relationship between technology and environment in this setting, together with the mechanisms behind it, remains to be examined. On the other hand, all of the localities so far identified are surface collections without absolute dates, and whether they constitute a coherent regional technological entity remains to be tested.

The analysis proceeds in two steps. We ask first whether the surface collections share a consistent technological concept with the Quina technology at Longtan, and can therefore be identified as a single regional technological entity, which we term the Quina Valley. To answer this we systematically analyse the most diagnostic elements of the Quina system, the Quina scrapers and the resharpening flakes removed in maintaining them, and assess their technological consistency with the excavated material from Longtan. Only if the first step can be sustained does the second question arise: whether such an entity can be accounted for by the environmental setting of the region, its available resources and the strategies by which technology was organised. Here we bring together the spatial distribution of the localities, the raw materials used at them and the environmental characteristics of the region. The study clarifies the regional distribution of Quina technology in East Asia and offers a perspective on the relationship between environmental constraint, technological convergence and regional technological organisation in the Middle Palaeolithic.

# Regional Geological and Palaeoenvironmental Setting

The tectonic framework of the study area determines both where its intermontane basins lie and where knappable stone is available within them. The area falls within the Jinsha River drainage in the southern Hengduan Mountains, on the southeastern margin of the Tibetan Plateau, and is controlled by a system of approximately north–south trending faults: continued relative uplift of massifs such as Yulong Snow Mountain and Mount Diancang, alternating with subsidence of the ground between them, has produced a series of north–south trending intermontane basins, among them Lijiang, Heqing and Dali \[Li et al. 2014; An et al. 2011\]. The two basins considered here, Huangping and Binchuan, lie within this same system. This alternation of uplifted ranges and downfaulted basins, combined with the Indian monsoon and the warm, moist airflow from the southwest, gives the deeply incised valley floors a dry-hot valley climate, sheltered from precipitation by the surrounding relief and marked by a sharp division between wet and dry seasons and by high insolation; the modern vegetation there is dominated by drought-tolerant shrub and savanna formations \[\]. The Quina localities discussed here are concentrated in two of these basins, the Huangping Basin in the southwest of Heqing County and the neighbouring Binchuan Basin, each of them cut by tributaries of the Jinsha River: the Caifeng River in the Huangping Basin, and the Sangyuan River together with its tributary the Liandong River in the Binchuan Basin. The two basins differ in bedrock. Permian volcanic rocks are extensively exposed along the western side of the Huangping Basin, with Triassic sedimentary rocks predominating to the east, whereas the Binchuan Basin is largely mantled by unconsolidated Quaternary sediments and exposes its Permian volcanic basement only locally \[\]. The volcanic sequence comprises basalt, tuff and trachyte \[\]. Because the rivers of both basins cut across these outcrops, volcanic cobbles were carried into the gravel bars of the valley floors, and it was these bars, rather than the outcrops themselves, that supplied the stone used at the localities.

The regional monsoon record places the occupation of these basins within an interval of recurrent aridity rather than within a settled climatic phase. The nearest palaeoclimate records covering this period come from the lacustrine sediments of the Heqing Basin, about 60 km north of the Huangping and Binchuan basins and some 650 m higher, which are among the longest and most continuous environmental sequences available for the Asian monsoon domain and are highly sensitive to variation in the Indian Summer Monsoon (ISM) \[Xiao et al. 2010; An et al. 2011\]. Magnetic and geochemical proxies from these sediments divide the interval from 184 to 25 ka into four stages \[Yang et al. 2025\]: relatively arid conditions between c. 184 and 131 ka; a shift to generally humid conditions between 131 and 69 ka; a return to aridity between 69 and 57 ka, corresponding to MIS 4; and a generally humid MIS 3 between 57 and 25 ka, punctuated by repeated arid events. Rapid, millennial-scale weakenings of the monsoon, corresponding closely to North Atlantic cold events, have been resolved in this basin from 52 ka onwards \[Hou et al. 2026\]; these post-date the occupation documented at Longtan, and we cite them to indicate that the monsoon system was liable to interruption of this kind rather than to date any interruption within the period of human presence. These records register monsoon variation at a regional scale rather than local climatic conditions at any one place, and given the distance and altitudinal offset just noted, we treat them here as regional context rather than as a proxy for conditions at the localities themselves. Read at that scale, what matters is not whether a particular stage was on balance wetter or drier, but that the occupation falls where the MIS 4 arid phase gives way to a monsoon recovery that was itself repeatedly interrupted.

Longtan provides the local counterpart to this regional picture, since it preserves both a secure chronology and a record of vegetation change. Optically stimulated luminescence (OSL) dating places its in situ cultural layer at c. 60–50 ka \[\], and Longtan remains the only locality in the study area with an age determination of this kind. The nearby Tianhuadong site, where material of this kind has also been reported, has yielded OSL ages of c. 90–40 ka for its cultural layers \[\], but it has not been systematically analysed, and neither its chronology nor its technological attribution is yet secure enough to carry weight here. The surface collections examined below have no independent age control, and none can be transferred to them by association: what the Longtan dates establish is when the Quina system was present in this landscape, not when any individual surface collection was formed. What Longtan adds beyond a date is a local environmental record. Pollen analysis there documents a transition from coniferous forest to a more open woodland-steppe, with the Quina-bearing layer falling within the phase of vegetation opening \[Ruan et al. 2025\]. Taken together, the two records indicate that people were present in this landscape while it was becoming more open locally, and during an interval when the regional monsoon system was subject to repeated interruption. Whether that setting can account for the technology found within it is a question we return to once the technological results have been presented.

# Materials and methods

## Survey and selection of localities

The surface material analysed here was recovered during four seasons of pedestrian survey in the Huangping and Binchuan basins, carried out in 2010, 2013, 2016 and 2022. Survey followed the main valleys of the two basins, the Caifeng River in the Huangping Basin and the Sangyuan River together with its tributary the Liandong River in the Binchuan Basin, and each flight of river terraces was walked in turn. Survey areas and walking routes were designed to cover the full range of landforms that humans might have used in the past. Routes were recorded continuously using the mobile GPS application 2bulu, localities and artefacts were photographed, and a field form was completed for each locality giving its position, its place within the terrace sequence, its straight-line distance to the present-day river channel, and the number and types of artefacts present.

Because this study is concerned with Quina technology, the localities recorded during survey were then screened, and only those yielding diagnostic Quina scrapers and resharpening flakes were retained. On this criterion, 22 of the 33 localities recorded in the Binchuan Basin (67%) and all five of those recorded in the Huangping Basin contain Quina specimens (\[Table X\]). Two consequences of this design should be stated at the outset. First, the criterion is one of presence rather than of degree: localities were retained on whether Quina pieces occur at them, not on how those pieces were made or how far they had been reduced, so that none of the technological variables analysed below entered the selection. Pieces at the very lightest end of the retouch range are nonetheless less readily recognised as diagnostic, and the sample is therefore unlikely to represent the earliest stages of the reduction continuum. Second, what Section \[\[X\]\] reports are patterns among Quina-bearing localities rather than a map of Quina technology across the two basins; since the visibility of surface localities depends on exposure, vegetation cover and survey intensity, these patterns are read here as trends and not as densities.

## 原料调查与分析方法

To establish what raw material was locally available, gravel bars were sampled at five points along three river channels. At each point an area of 1–2 m² was cleared and every clast longer than 30 mm exposed at the surface was recorded. For each clast we recorded length, width and thickness (mm), roundness on a four-point scale (0–3, from least to most rounded) and shape (sub-oval, tabular or irregular). These records give the relative abundance of each lithology on the gravel bars (n = xx).

To characterise the internal properties of the dominant raw material, thin sections were prepared from five trachyte cobbles and examined petrographically, with mineral composition determined by energy-dispersive spectroscopy. Identifications were carried out by the Guangxi Institute of Geological Survey and the China University of Geosciences (Wuhan). The raw material of each artefact was assigned macroscopically to the same lithological categories, using the surveyed clasts as a reference collection.

Differences in raw material composition, both between the gravel bars and the artefacts and between spatial units, were tested by permutational multivariate analysis of variance (PERMANOVA), stratified at two scales, basin and valley.

Because the artefacts are made overwhelmingly on river cobbles, the straight-line distance from each locality to the nearest present-day channel was used as a proxy for proximity to a source of raw material. These distances are strongly skewed, the groups differ considerably in size, and their variances are unequal, so differences between the two basins were tested with a Mann–Whitney rank sum test and differences among the three valleys with a Kruskal–Wallis rank sum test; medians are reported as the measure of central tendency and the rank tests taken as the primary criterion. Clast size was summarised as the geometric mean of length, width and thickness and compared between lithologies with a Mann–Whitney rank sum test, with the geometric mean size of each lithology and the effect size r reported alongside.

Raw material selection was quantified with Jacobs' electivity index D [@jacobs1974], a modification of Ivlev's index that corrects for the relative abundance of the available resources:

$$D = \frac{r - p}{\,r + p - 2\,r\,p\,}$$

where r is the proportion of a given raw material among the artefacts and p is its proportion among the clasts recorded in the gravel survey. D ranges from −1 (complete avoidance) through 0 (use in proportion to availability) to +1 (complete selection). It was computed for each raw material in the surface Quina scrapers against availability pooled across the gravel survey. Materials present among the artefacts but absent from the survey take D = +1 and are read as evidence of non-local procurement; materials present in the survey but absent from the artefacts take D = −1.

## Techno-typological analysis of Quina retouch products

The techno-typological observations and measurements follow established practice in the study of Quina technology \[Bourguignon 1997; Peresani et al. 2024; Ruan et al. 2025\]. For each scraper we recorded length, width, thickness and mass; blank type; cortex position; platform type and platform depth; interior platform angle (IPA); the number of retouch generations; the numbers of convex scars and concave scars; the geometric index of unifacial reduction (GIUR) \[Kuhn 1990\] and the proportion of the perimeter retouched; the edge angle; and its attribution within Bordes' type list. For each resharpening flake we recorded platform type, depth, IPA and EPA. Statistical tests were chosen according to the type of each variable and the distribution of the data, with parametric or non-parametric procedures used depending on whether the data approximated normality. Relationships between edge angle and GIUR, the proportion of retouched edge and the mean number of retouch generations were examined with Spearman rank correlation, with p-values adjusted by the Benjamini–Hochberg procedure. The difference between the exterior platform angles of the resharpening flakes and the edge angles of the scrapers was tested with Welch's t-test, so that unequal variances would not affect the inference.

To test the technological consistency between the surface Quina scrapers and the excavated material, the Quina scrapers from Longtan were used as the reference, and ordinary scrapers from the same site were included as a technological control group. Six of the recorded variables were selected for multivariate analysis as the most representative: thickness, proportion of the perimeter retouched, GIUR, number of total scars, number of retouch generations and edge angle. All were standardised to z-scores and differences between specimens computed as Euclidean distances. Differences between groups were tested by PERMANOVA, with the proportion of variance explained (R²) reported for each pairwise comparison. The ordinary scrapers serve to calibrate the scale on which those R² values are read: if the same set of variables separates Quina from ordinary scrapers with a high R² while the two Quina groups differ only at a low R², the latter is more likely to reflect a real technological similarity than a lack of statistical power. PERMANOVA compares the positions of group centroids in multivariate space. At the univariate level, mean GIUR, scar count and mean number of retouch generations were compared with Kruskal–Wallis rank sum tests followed by Dunn's post hoc comparisons, and thickness, proportion of the perimeter retouched and edge angle with Welch's ANOVA followed by Welch's t post hoc comparisons; all post hoc comparisons were corrected Bonferroni method. Technological dispersion between groups was tested with PERMDISP. To compare the degree of variation within groups, the coefficients of variation (CV) of length, width, thickness, mass and edge angle were tested for equality across groups using the Krishnamoorthy–Lee modified signed likelihood ratio test (KL-MSLRT), these being ratio-scale variables with a true zero. GIUR, the proportion of the perimeter retouched, scar count and the number of retouch generations were instead compared with Fligner–Killeen tests: the first two are bounded on \[0, 1\], which leaves the CV without a clear interpretation, and the last two are counts whose variance structure does not meet the assumptions required for comparing CVs. To assess how robust the results are, the analyses were repeated with the surface specimens collected from within the area of the Longtan site removed.

To test whether technological variation is systematically structured by the landscape, the analysis was repeated with the locality as the unit, each locality represented by the median value of its Quina scrapers on the same six variables. Four landscape variables were considered: basin (Huangping or Binchuan), geomorphological position (terrace level), distance from the locality to the nearest present-day channel, and assemblage size (number of specimens). The two categorical variables were tested with PERMANOVA and PERMDISP; the two continuous variables with Spearman rank correlation, and with locality-level PERMANOVA on Euclidean distances, assemblage size additionally being examined in bins as a check on robustness. Because the number of units at this level is small (n = 27), most of them yielding only one or two specimens, so that locality medians are not very stable, and because basin and geomorphological position are strongly collinear, these tests have limited statistical power; the results are therefore interpreted mainly from the size and direction of the effect sizes (R², ρ), with the significance tests treated as supporting evidence only.

All statistical analyses were carried out in R 4.5.2 \[R Core Team\]. PERMANOVA and PERMDISP were run with the vegan package \[\], Welch's ANOVA with onewaytests \[\], the KL-MSLRT with cvequality \[\], and Fligner–Killeen tests with the stats package.

## Reproducibility and open materials

Computational reproducibility remains a challenge in archaeological science, yet it underpins both the credibility of quantitative results and their re-use by others [@marwickComputationalReproducibilityArchaeological2017; @marwickOpenScienceArchaeology2017]. All materials needed to reproduce the analyses reported here are therefore available in a version-controlled online repository at \[\[OSF DOI\]\]. These comprise the raw data underlying every reported result together with the complete code used for the analyses and for the figures derived from them. The repository is organised following established practice for research compendia, with a logical folder structure and relative file paths, so that the code runs without manual modification. The data are released under a CC-0 licence, the code under the MIT licence and the figures under CC-BY. Every statistical result reported below, and every figure generated from the data, can therefore be reproduced independently from these materials.

# Results

## Landscape distribution and raw material economy

Quina localities in both basins lie along the valley systems, on river terraces, and at consistently short distances from the channels. In the Huangping Basin they are concentrated on the second and third terraces (T2–T3); in the Binchuan Basin they extend across the second to fourth terraces (T2–T4), with a small number (n = xx) occurring higher, on hilltops. Because the artefacts are made overwhelmingly on river cobbles, the straight-line distance from a locality to the nearest channel serves as a proxy for ease of access to raw material. Across the three valleys, the Sangyuan, Liandong and Caifeng, these distances are short and similarly distributed, with a median of 596 m. Neither comparison detected a difference between spatial units, whether at the scale of the basin (U = \[\[U\]\], p = 0.262) or of the valley (H = \[\[H\]\], df = 2, p = 0.124). Access to river gravels therefore appears to have been broadly comparable across the study area, so that differences in the raw materials actually used at these localities are unlikely to be explained by differential proximity to a source.

Raw material use does not follow raw material abundance. On the gravel bars, sandstone is the most abundant lithology (\[\[%\]\] of \[\[N: 砾石总数\]\] clasts recorded), with trachyte second (\[\[%\]\]); among the artefacts, trachyte accounts for 95.8%. The bars are not uniform in composition, however: raw material composition differs significantly between sampling points (R² = 0.108, pseudo-F = \[\[F\]\], p = 0.001), and stratified analysis attributes this variation to differences between valley systems (R² = \[\[ \]\], p = 0.001) rather than within them. Sandstone-dominated and trachyte-richer bars are therefore distributed unevenly across the study area, in a pattern that follows the contrast in bedrock described in Section 2. If knappers had simply worked whatever lay most abundantly at hand, the raw material composition of the artefacts should vary from valley to valley in the same way.

That expectation is not met. The raw material composition of the Quina scrapers is uniform across the study area, differing neither between basins (PERMANOVA: R² = 0.004, pseudo-F = \[\[ \]\], p = 0.54) nor between valley systems (R² = 0.005, pseudo-F = \[\[ \]\], p = 0.75) (\[\[Fig. X\]\]). The same material was selected wherever people worked, although sandstone is more abundant on the gravel bars and although the composition of those bars varies from valley to valley. Jacobs' electivity index puts the strength of that preference on a common scale: trachyte is strongly selected (D = 0.97) and sandstone strongly avoided (D = −0.98), with \[\[其余岩性的 D \]\] (\[\[Table X\]\]). Because D returns to zero only when a material is used in proportion to its availability, trachyte would have to make up some 96% of the gravel before its use could be read as neutral; no redistribution of availability among the valleys approaches that figure, so neither the direction nor the magnitude of the selection depends on how availability is pooled. The choice of trachyte, in other words, was made against local abundance rather than in step with it.

Three properties of trachyte account for this preference. First, size: trachyte clasts are significantly larger than sandstone clasts (U = 16864, p \< 0.001, r = \[\[效应量\]\]; geometric mean size \[\[ \]\] mm against \[\[ \]\] mm) \[\[N per lithology\]\], giving a larger initial volume from which to prepare a core and sustain prolonged reduction. Second, shape: \[\[%\]\] of trachyte clasts are tabular and \[\[%\]\] irregular (\[\[n\]\] of \[\[N\]\]), against \[\[%\]\] and \[\[%\]\] for sandstone \[\[roundness 结果\]\], and these angular forms offer the acute edges and flat surfaces from which invasive removals can be struck, whereas the smaller and more regular sandstone clasts do not. Third, fracture properties: in thin section the trachyte has a porphyritic texture, with euhedral tabular to prismatic K-feldspar phenocrysts (c. 0.3–5.0 mm, c. 1–10% of the rock) set in a groundmass of K-feldspar microlites, chlorite or altered mafic minerals and magnetite, together with about 50% devitrified glass, forming a hyalopilitic texture \[\[玻晶交织的英文对应词\]\] (\[\[Fig. X\]\]). A groundmass of this kind is likely to promote regular conchoidal fracture and hence sharp, continuous edges, while the crystalline fraction may give the material greater mechanical strength, and so greater durability in use, than a wholly glassy rock.

The advantage of trachyte therefore lies less in its availability than in what it affords: clasts large and angular enough to be worked as cores, in a material whose texture combines predictable fracture with resistance to edge failure. These are the properties on which prolonged resharpening depends, and they are the same properties that the Quina system is understood to require.

## Characterising the Quina retouch products

### Technol-typological features

The 166 Quina scrapers from the surface collections are consistent with the Quina technological concept in each of its three dimensions: in the blanks selected for them, in the way their edges were maintained, and in the ramification of the sequence they belong to.

Blanks. Quina débitage yields thick flakes with markedly asymmetric cross-sections, detached by invasive percussion from cores worked on alternating secant surfaces, and it is the volume held in such a blank that permits the successive stages of edge maintenance that follow \[Bourguignon 1997; Turq 1992\]. The blanks used here carry the attributes that this strategy produces. Most scrapers retain cortex to some degree (\[\[%\]\], n = \[\[ \]\]), and where cortex survives it lies opposite the retouched edge, forming a thick back. Some were made directly on cortical flakes, with cortex covering the whole dorsal face (\[\[n\]\], \[\[%\]\]); among specimens without cortex, a back formed by earlier removals takes its place (\[\[n\]\], \[\[%\]\]). The asymmetric distribution of volume is therefore present whether or not the original surface of the cobble survives. These are not ordinary flakes that happened to be retouched, but blanks carrying a reserve of volume opposite the edge that was to be worked.

Platform observations point the same way. Fifty-two scrapers (31.3%) retain the platform of the original blank, and these are characteristically wide and oblique, with a median platform depth of 13.9 mm \[\[IQR\]\] and a mean IPA of 122° \[\[SD\]\]; both follow from the internal percussion by which thick blanks are removed from a secant flaking surface. Plain platforms predominate (n = 39, 75.0%), with cortical (n = 7, 13.5%) and dihedral platforms (n = 6, 11.5%) less common. Four of the dihedral specimens are *talons à pan*, on which the platform preserves parallel removal scars running approximately orthogonal to the axis of percussion. Few though they are, these carry the most direct information in the assemblage about the reduction strategy behind the blanks: a platform of this kind records a surface that had itself previously served as a flaking surface, which is what the alternating exploitation of two secant planes produces.

刃缘修理。 Quina刮削器最具代表性的技术特征在于其独特的连续修理策略。为了维持厚重刃缘和整体体积结构，Quina修理通常表现为具有阶段性的修理节奏：早期修理以凸型剥离（convex removals）为主，而随着修理推进，逐渐转变为凹型剥离（concave removals），循环往复。这种由凸向凹的修理序列被认为是维持刃缘体积并延长工具使用寿命的重要机制。研究材料中，该修理模式表现十分明显。刮削器平均修理世代为2.76，其中凸型修疤平均数量为4.4个，而凹型修疤平均数量达到18.7个。凸型修疤主要集中于第一修理世代，而凹型修疤则普遍出现于第二和第三修理阶段。部分标本完全由凹型修疤构成，但未观察到完全由凸型修疤组成的案例。这一分布模式表明，研究材料中的刃缘重塑并非随机发生，而是遵循由早期凸型修理向后期凹型修理转变的技术序列。尽管所有标本均符合Quina修理概念，刮削器内部仍存在一定类型学差异（表\[x\]）。这种差异在Bordes类型学分类中有所体现，可能反映了打制者针对不同毛坯条件所采取的技术调整，而非不同技术传统之间的差异。为量化刃缘再加工程度，本文采用GIUR和刃缘修理比例两个指标进行评估。结果显示，GIUR中位值为0.77，表明修理通常深入至毛坯最大厚度区域；刃缘修理比例中位值为0.46，与多数标本保留明显“背”的结构相一致。然而，部分标本表现出更高程度的整体加工，其中包括triple scraper和limace类型。研究材料中的平均刃角为70.0°，该指标更可能反映整体加工强度，而非直接对应工具功能。由于最后一次修理世代的刃角难以独立识别，因此本文测量的是整体刃角。相关分析显示，刃角与GIUR、刃缘修理比例以及修理世代均呈显著单调正相关，进一步支持其主要反映加工程度变化的解释。

再修理石片与分支化。 再修理石片既进一步支持了基纳修理概念,也体现了分支化(ramification)这一基纳体系的另一维度。这类石片通常扁圆、侧廓呈凸起的拱形,背面近端保留有先前的基纳修理痕迹;其台面深度很小(均值 5.3 mm),显示边缘打击的特点。再修理石片的 EPA 平均为 72.8°,与刮削器刃角(70.0°)之间存在一个微小的偏移，这一偏移很可能与地表遗址的不均匀采样有关，而不一定代表实际修理角度的系统差异。再修理石片本身具有锋利刃缘、具备使用潜力。此外,部分刮削器还表现出"克拉克当凹口(Clactonian notch)"式的再修理:这种方式会在刃缘形成较大的凹型修疤,对应的再修理石片也尺寸较大、带有台面,因而具备被进一步加工成工具的潜力。

### 技术一致性分析

为评估地表采集材料是否保持稳定的技术特征，本文以具有明确地层关系和可靠年代控制的龙潭遗址作为参照基准，选取其发掘出土的Quina刮削器作为对照样本，并引入普通刮削器（ordinary scrapers）作为技术差异参照。考虑到高维变量可能引入信息冗余，本文选择6个最具代表性的技术指标进行多变量分析。结果显示，地表采集Quina刮削器与龙潭出土Quina刮削器在整体技术空间中不存在显著差异，而两组Quina刮削器均与普通刮削器表现出显著分离。这一结果具有重要的方法学意义：由于所选变量组合能够有效区分Quina刮削器与普通刮削器（两两比较效应量R² = 0.29–0.36），因此Quina组之间较低的组间差异（R² = 0.004）并非源于变量组合缺乏区分能力，而更可能反映两类Quina材料之间真实的技术相似性。逐变量分析进一步支持这一结果。在单变量尺度上，两个Quina样本组之间均未检测到显著差异，而二者与普通刮削器之间则在全部指标上均表现出显著差异。这表明，地表采集材料与发掘材料之间不仅在综合技术空间中不可区分，在具体技术属性上也表现出高度一致性。上述结果主要反映群组中心位置（centroid）的差异。为进一步评估技术变异范围是否一致，本文进行了PERMDISP分析。结果显示，地表Quina组与龙潭Quina组具有几乎相同的平均质心距离（分别为2.10和2.02，比值约为1.04），且两者之间不存在显著差异（p = 0.53）。相比之下，普通刮削器表现出更低的技术离散度（平均质心距离=1.55），并与两组Quina刮削器均存在显著差异（p均=0.001）。这一结果表明，分析方法能够有效识别真实存在的离散度差异，因此Quina两组之间缺乏显著差异并非由统计灵敏度不足造成。单变量离散度检验进一步验证了这一趋势。KL-MSLRT检验和Fligner–Killeen检验均未发现两个Quina样本组之间存在显著方差差异。为进一步检验结果的稳健性，本文剔除了地表采集样本中来自龙潭遗址附近的材料，并重新进行分析，结果保持一致。这说明观察到的技术一致性并非由单一遗址或局部样本结构驱动，而反映了更广泛的技术稳定性。综合中心位置和离散度两个层面的证据，地表采集Quina刮削器与龙潭发掘Quina刮削器在技术空间中均无法被有效区分，表明二者共享相同的技术逻辑。

遗址-景观层面。进一步分析旨在考察Quina技术变异是否受到景观结构因素的系统性影响。本文选取四个可能影响技术组织的景观变量，包括遗址至河流距离、地貌位置、遗址规模以及盆地归属（黄坪盆地与宾川盆地），并分别采用PERMANOVA评估其对技术变异的解释能力。结果显示，各景观变量对技术差异的解释度均较低。其中，遗址规模（连续变量，R² = 0.006）和盆地归属（R² = 0.011）几乎不贡献技术变异解释，距水距离（R² = 0.031）和地貌位置（R² = 0.073）的解释度相对较高，但仍处于较低水平，且所有变量在置换检验中均未达到统计显著水平。值得注意的是，尽管距水距离对整体技术变异的解释能力有限，逐变量相关分析显示，六个技术指标均与距水距离呈弱正相关关系（Spearman ρ = 0.14–0.37）。这一方向一致性表明，随着遗址距离水源增加，Quina刮削器可能倾向于表现出更高的厚度和更深程度的加工。该趋势与移动过程中携带工具并在远离原料补给区域持续维护、减薄工具的理论预期相一致。然而，由于单个相关系数均未达到显著水平，目前尚无法确认这一趋势是否代表稳定的空间梯度，因此本文仅将其视为未来景观尺度研究中值得检验的行为假设，而非已确定的空间组织规律。总体而言，在当前样本规模和空间分辨率下，尚未发现显著控制Quina技术变异的景观因素。研究区内Quina技术表现出较高程度的跨地点稳定性，其技术组织逻辑并未明显受到遗址规模、盆地归属或局部景观位置的限制。

# Discussion

本研究显示，研究区内存在一套高度一致的基纳技术体系，可归纳为一个区域性技术概念（Quina Valley）。这一结论在原料选择、技术特征与景观分布三个层面均得到支持。在原料经济方面，无论地表采集材料还是龙潭遗址出土标本，均表现出对粗面岩的强烈选择偏好。值得注意的是，当地河滩砾石的原料组成具有显著空间异质性，但这种自然变异并未反映在石制品原料选择中，各遗址均一致偏向粗面岩。在技术层面，研究区基纳刮削器系统性地表现出典型诊断特征，包括厚重且不对称的毛坯结构、由石皮或早期剥片构成的“back”、多世代连续修理，以及凸型与凹型修疤的节奏性交替，同时伴随再修理石片的持续生产。这一组合与欧洲典型Quina技术定义高度一致。在空间尺度上，技术特征与原料选择均未随盆地、地貌位置、距水距离或遗址规模发生系统变化，表明该技术体系具有较强的空间稳定性。综合上述证据，研究区内的基纳材料指向同一技术概念，构成一个可识别的区域性技术实体，即Quina Valley。

这一区域性现象的意义，还在于为龙潭这一长期被讨论的个案提供了新的区域性支撑。作为东亚首个正式发表的基纳遗址，龙潭的技术归属曾引发讨论。Bourguignon et al. (2025)指出，形态相似性并不必然意味着文化亲缘关系，并提出龙潭的剥坯策略可能代表盘状技术的一种变体，同时对其年代与埋藏背景持审慎态度；Ruan et al. (2025)则从技术—经济特征、与盘状技术的系统性差异以及年代与埋藏学证据等方面进行了回应。无论上述争论如何展开，一个孤立遗址在解释能力上始终存在局限。本文的关键贡献在于表明，龙潭并非孤立现象。在鹤庆与宾川两个盆地的多条河谷及多级阶地中，反复出现一套在原料选择、毛坯生产、加工方式与加工程度上均高度一致的基纳组合，这些材料与龙潭共享同一技术概念。尽管这些地表材料缺乏独立的绝对年代约束，但其在区域尺度上的重复出现，从分布意义上为东亚基纳现象提供了独立于单一遗址之外的证据支撑。需要指出的是，基纳风格的石制品并非仅限于本研究区。在更广泛的地理范围内，已有若干遗址报道过类似材料。例如，在中国西南地区有云南富源大河遗址、四川凉山白草坪遗址，在中国北方地区有内蒙古赤峰三龙洞与河北阳原新庙庄等遗址，这些遗址均报道了具有基纳特征的石制品。然而，这些遗址大多尚未开展系统的技术学分析，其组合的技术属性仍有待进一步验证。因此，目前这些材料更适合作为潜在证据线索，而非已确立的技术实体。综合来看，以龙潭为代表的东亚基纳现象，在横断山脉地区已表现为一个可重复识别的区域性技术实体，指向一套可与欧洲基纳体系相比较的技术行为模式。同时，散见于中国其他地区的相关材料，则提示该技术现象在东亚的空间分布可能更为广泛，但其技术一致性仍需未来研究进一步检验。

尽管共享相似的技术行为,目前仍缺乏足够的证据将其与特定人群相联系;因此,现阶段更适宜从环境适应的角度来理解 Quina Valleys。欧洲的基纳遗址已有充分研究,其年代主要集中在 MIS 4 至 MIS 3 早期,对应干冷气候、开阔苔原,以及以驯鹿为主的季节性迁徙动物群。在这样的环境中,觅食者在不同季节、不同地点会遭遇不同的猎物,资源在时空上难以预测;正因如此,欧洲基纳被理解为对这种不确定性的一种应对，这种应对依赖灵活、可反复维护、使用寿命长的工具(Hiscock et al., 2009)。东亚的情况在环境背景上与欧洲存在差异,却面对着同一类适应问题。区域古气候记录显示，该时期整体处于季风回升阶段，但频繁受到千年尺度干旱事件干扰；同时，龙潭孢粉记录表明局地植被正向开阔草地转变。这种区域—局地双重不稳定性，共同构成了一个资源时空分布不确定的环境背景。在这一意义上，欧洲与东亚之间的关键相似性并不在于具体环境类型（冰缘苔原 vs 干热河谷），亦不在于原料类别（燧石 vs 粗面岩），而在于共享的生态结构：开阔环境条件下的高资源波动性与不可预测性。这一共享选择压力为趋同演化提供了机制基础，使得在地理隔离的情况下，两个区域分别演化出功能上相似的技术解决方案成为可能。因此，Quina Valley可被理解为一种在不稳定资源环境中形成的技术—经济适应模式。但同时，目前也无法完全排除技术传播的可能性。尤其考虑到Quina技术在旧大陆西侧与尼安德特人密切相关，其扩散边界仍在不断被重新界定。因此，在当前阶段，对其起源机制应保持开放解释框架。

Quina Valleys的提出，也为理解中国旧石器时代中期技术的多样性与复杂性提供了新的视角。东亚是否存在真正意义上的旧石器时代中期长期存在争议，传统观点强调该地区石器技术在整个旧石器时代的长期稳定性，并据此否认存在可与西欧相对应的中期阶段。近年来，观音洞（约170–80 ka）、灵井（约125–90 ka）以及金斯太（约50 ka）等遗址所报道的勒瓦娄哇、盘状石核等技术，已逐步揭示出这一时期远较传统认识复杂的技术面貌，龙潭遗址基纳体系的确认更为这一复杂格局增添了新的技术类型。然而，此前经过系统技术分析的仅龙潭一处。本研究的意义正在于此：Quina Valleys在区域层面为龙潭的基纳技术提供了独立的复现证据，表明这一技术并非孤立个案，而是在青藏高原东南缘确有存在。与此同时，这一区域性证据也为其他已报道基纳特征、但尚待系统分析的遗址（如云南富源大河遗址）提供了潜在支持。

本研究亦存在若干局限。其一，研究材料主要来源于地表遗址，缺乏可靠的绝对年代控制，目前仅能借助龙潭与天华洞遗址的测年结果提供大致的时间约束。其二，本研究未纳入石核分析，基纳剥坯策略的重建主要依赖刮削器毛坯属性进行间接推断。其三，宾川盆地目前缺乏本地古气候记录，相关环境背景主要参考邻近鹤庆盆地的数据作为区域性替代。其四，Quina Valleys尚缺乏系统的动物考古研究，因此在资源利用与生计结构方面仍缺少关键证据维度。尽管天华洞遗址已有动物化石发现，但相关材料尚未形成可用于系统分析的数据基础。上述限制表明，本研究在年代学约束、技术链完整性及多代理环境重建方面仍存在不确定性，有待未来通过天华洞遗址的进一步发掘，以及在宾川盆地寻找原生地层材料与开展系统动物考古与环境研究加以完善。

# Conclusions

# CRediT author statement {.unnumbered}

# Declaration of competing interest {.unnumbered}

# Funding {.unnumbered}

# Data availability {.unnumbered}

# Acknowledgements {.unnumbered}

\newpage

# References {.unnumbered}

<!-- The reference list is inserted here for Word/HTML output rather than at the very end. -->

::: {#refs}
:::

\newpage

## Colophon {.unnumbered}

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

if ("devtools" %in% installed.packages()) devtools::session_info()
```
