---
title: "The Quina Valley: a regional Middle Palaeolithic technological system on the southeastern margin of the Tibetan Plateau"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Qijun Ruan:
      correspondence: "yes"
      orcid: 0009-0000-2143-5335
      institute:
        - YPICRA
  - Zhenxiu Jia:
      correspondence: "no"
      orcid: 0000-0002-7514-4514
      institute:
        - ALPHA
  - Marco Peresani:
      correspondence: "no"
      orcid: 0000-0001-6562-6336
      institute:
        - FERRARA
        - IGAG
  - Davide Delpiano:
      correspondence: "yes"
      institute:
        - FERRARA
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
# <!-- TODO: update the author list, ORCIDs and affiliations for THIS paper -->
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - YPICRA: Yunnan Provincial Institute of Cultural Relics and Archaeology, Kunming, China
  - FERRARA: Department of Human Studies, Prehistoric and Anthropological Science Unit, University of Ferrara, Ferrara, Italy
  - IGAG: Consiglio Nazionale delle Ricerche-Institute of Environmental Geology and Geoengineering, Laboratory of Palynology and Palaeoecology, Research Group on Vegetation, Climate and Human Stratigraphy, Milan, Italy
title-block-published: "Last updated"
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
  tbl-title: "Table"
  tbl-prefix: "Table"
number-sections: true
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/elsevier-harvard.csl"
keywords:
  - Middle Paleolithic
  - Southeastern Tibetan Plateau
  - Quina technology
  - Quina valleys
abstract: |
  The Middle Paleolithic (~300–40 ka) was a pivotal period in human evolution, characterized by diverse lithic technologies. Conventional views have held that East Asia experienced slow technological development and lacked remarkable Middle Paleolithic innovations. Recent discoveries, however, demonstrate a more complex scenario, with distinctive Middle Paleolithic technologies identified, such as discoidal technology, Levallois technology and Quina technology. Here we present stone artifacts surface-collected from the Caifeng river and Sangyuan river in northwestern Yunnan province, geographically located in southeastern margin of the Tibetan Plateau. These artifacts mainly collected on the second and third river terraces. Diagnostic Quina technological elements have been identified, including plan-secant core reduction pattern, scalariform tool retouching, and branched reduction strategy. Our results demonstrate the existence of “Quina valleys” and expand the spatial range of Quina technological behavior in southwestern China. In addition, our study reinforces the technological diversity and complexity of the Chinese Middle Paleolithic.
editor_options:
  chunk_output_type: console
---

<!-- This is the format for text comments that will be ignored during rendering. Do not put R code in these comments because it will not be ignored. -->

Keywords: `r rmarkdown::metadata$keywords`

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(here)
library(readxl)
library(vegan)
library(rstatix)
library(ggpubr)
library(patchwork)
library(ggrepel)

# A single fixed seed and permutation count govern every permutation, Monte
# Carlo and bootstrap result reported below, so the manuscript is exactly
# reproducible (see the "Materials and methods" note on reproducibility).
set.seed(2226)
PERM <- 9999

# ---- inline-number formatters --------------------------------------------
f  <- function(x, d = 2) formatC(x, format = "f", digits = d)      # fixed dp
fp <- function(p) {                                                # p-values
  if (is.na(p)) "NA"
  else if (p < 0.001) "< 0.001"
  else paste0("= ", formatC(p, format = "f", digits = 3))
}

```

```{=html}
<!-- Highlights: Elsevier requires 3–5 bullet points, each ≤ 85 characters (incl. spaces).
     These are submitted as a separate file at submission; kept here for drafting. -->
```

# Highlights {.unnumbered}

- 

- 

- 

- 

- 

# Introduction

The Middle Palaeolithic was a critical period of diversity in both hominin populations and lithic technology, and what drove that technological diversity remains contested: do assemblage differences reflect separate cultural traditions, or adaptive responses to different environments? The question has been pursued most intensively in the Mousterian Debate of western Europe, where the variability of Mousterian assemblages has been attributed to cultural tradition \[\], site function \[\], the constraints of locally available raw material \[\], and differing degrees of artefact reduction [@dibble1987]. The Quina Mousterian has been central to that debate.

The Quina Mousterian is one of the best-defined Middle Palaeolithic techno-complexes [@delagnes2006; @faivre2017], and what distinguishes it is not a tool type but a techno-economic system, the Quina system, whose defining features form three interlocking dimensions [@delpiano2022; @peresani2023]. The first dimension is blank production. Quina cores receive little preparation and are reduced through alternating, concatenated removals on sub-parallel and secant surfaces, with invasive detachments that optimise the volume of stone converted into blanks [@turq1989; @turq2000]. The target product is a thick flake, asymmetric in cross-section, with a sharp cutting edge on one margin and a thick, often cortical, back opposite \[\]. This asymmetry is what allows such a tool to be repeatedly resharpened [@turq1992; @hiscock2009]. The second dimension is edge maintenance, achieved through Quina retouch. Here the working edge is not merely produced but sustained, and resharpening follows a regular rhythm [@lemoriniScrapersLifeHistory2016]. Early episodes detach relatively large resharpening flakes and leave convex scars, giving a plano-convex edge and preserving volume for later work; as retouch proceeds it concentrates on the margin itself, producing smaller concave scars with stepped or hinged terminations and a plano-concave edge that keeps the tool serviceable \[\]. Each generation of retouch therefore restores a working edge without drawing on new raw material. The third dimension is ramification of the reduction sequence \[\]. Resharpening flakes detached during edge maintenance carry cutting edges of their own and can be used directly, as use-wear analysis has shown [@claud2012]; in some cases they were themselves retouched into tools [@ruanQuinaLithicTechnology2025a]. Quina flakes, besides being shaped into Quina scrapers, are thick enough to serve as cores for small flakes (i.e. core-on-flake) \[\]. Use-wear on scrapers and resharpening flakes from the same assemblages indicates that the two were put to different tasks, scraping and cutting respectively [@delpiano2022], so that ramification widened the functional range of the toolkit as well as extending its life. Efficient blank production, intensive edge maintenance and ramification together create an economy of stone use in which the thick Quina flake functions as a transportable and highly extendable package of raw material, yielding a large amount of working edge from a small initial investment [@hiscock2009].

Quina technology has been interpreted as an adaptation to high mobility in open environments where resources were patchily distributed and seasonally structured, and this interpretation rests on three lines of evidence. Faunal assemblages from Quina sites are typically dominated by a single gregarious and migratory species, most often reindeer, although equids and bovids predominate in some regions [@delpechLenvironnementAnimalMousteriens1996; @delagnesShiftsNeandertalMobility2011; @delpiano2022; @castelNeandertalSubsistenceStrategies2017]. Settlement organisation is correspondingly structured, with residential camps surrounded by task-specific satellite locations given over to hunting and carcass processing [@delagnesShiftsNeandertalMobility2011]. At the landscape scale, Quina assemblages occur in open, lightly wooded settings in which prey were distributed across the terrain in a way that was patchy but not random [@delpechLenvironnementAnimalMousteriens1996; @hiscock2009]. Foragers under these conditions are argued to have relied on seasonal, high-mobility strategies of resource use and tool provisioning [@hiscock2009; @delagnesShiftsNeandertalMobility2011; @castelNeandertalSubsistenceStrategies2017], and it is here that the technological economy described above becomes advantageous: versatile blanks, tools with long use-lives and cores requiring little preparation together lower the time and energy costs of replacing equipment on the move, and reduce the risk both of being left without an effective cutting edge far from a source of raw material and of carrying a toolkit ill-suited to the tasks that a given season and location happened to present [@hiscock2009; @delagnesShiftsNeandertalMobility2011].

The distinction between the Quina system and Quina retouch alone matters when its distribution is assessed. In the strict sense, with blank production, edge maintenance and ramification operating together, Quina is documented in western and southern Europe within a narrow interval between MIS 4 and early MIS 3, coinciding with pronounced global climatic instability [@delagnesShiftsNeandertalMobility2011; @delpiano2022]. Quina retouch taken on its own is far more widely distributed: in western Asia it has been reported in Acheulean–Yabrudian industries dating from as early as MIS 11 \[\], several hundred thousand years before the European Quina. The occurrence of so distinctive a way of maintaining a working edge in contexts this widely separated in time and space raises the possibility that Quina retouch is a recurrent solution to a recurrent problem rather than the signature of a single tradition. Neither the system nor Quina retouch, however, had been securely documented in the eastern Old World.

Whether East Asia has a Middle Palaeolithic in any meaningful sense has long been contested. On the conventional view, lithic technology in the region remained stable throughout the Palaeolithic, with substantial innovation deferred to the Upper Palaeolithic, so that no phase corresponding to the western Middle Palaeolithic can be recognised \[\]. Recent discoveries have provided new evidence that is reshaping this long-standing understanding. Discoid technology is documented at Lingjing (c. 125–90 ka) [@liTechnologicalBehaviorEarly2019; @zhaoEarliestCentripetalFlaking2026], and Levallois technology at Guanyindong (c. 170–80 ka) [@huLateMiddlePleistocene2019] and Jinsitai (c. 47–35 ka) [@liEasternmostMiddlePaleolithic2018]. These finds indicate that the technological scenario of the East Asian Middle Palaeolithic is more complex than the conventional view allows. It is against this background that we identified a complete Quina system at Longtan, dated to c. 60–50 ka [@ruanQuinaLithicTechnology2025a], the first Quina site in East Asia to be unambiguously characterised, formally excavated, securely dated and systematically analysed. This discovery not only filled a long-standing gap in the eastern Old World distribution of Quina technology, but also provided new evidence for reassessing the diversity and complexity of the East Asian Middle Palaeolithic. More importantly, Longtan is unlikely to be an isolated case. Since 2010, surface collections bearing Quina features have been recovered from river terraces at a number of localities in the Huangping Basin, where Longtan is situated, and in the neighbouring Binchuan Basin. These finds suggest that Quina technology on the southeastern margin of the Tibetan Plateau may not be confined to a single site, but may instead represent a regional technological phenomenon of some spatial extent.

The southeastern margin of the Tibetan Plateau, where the Huangping and Binchuan basins lie, is the Hengduan Mountains region, an area of pronounced altitudinal gradients, complex landforms and diverse climatic regimes that ranks today among the richest biodiversity hotspots in the world \[\]. Geomorphologically, both basins belong to the dry-hot valley zone of the Jinsha River drainage, a typical fault-controlled valley system \[\]. Under the combined influence of high-relief topography and the monsoon, the region experiences marked seasonal drought and supports a predominantly open vegetation cover \[\]. This setting differs substantially from the cold, open environments of the European Quina sites, and equally from the settings of other Middle Palaeolithic assemblages so far reported in China. It is this regional ecological distinctiveness that makes Longtan and the surrounding Quina localities of particular interest. On the one hand, a technology usually associated with open landscapes, seasonally structured resources and high-mobility subsistence strategies here occurs in a dry-hot valley on the southeastern margin of the Tibetan Plateau, and the relationship between technology and environment in this setting, together with the mechanisms behind it, remains to be examined. On the other hand, all of the localities so far identified are surface collections without absolute dates, and whether they constitute a coherent regional technological entity remains to be tested.

The analysis proceeds in two steps. We ask first whether the surface collections share a consistent technological concept with the Quina technology at Longtan, and can therefore be identified as a single regional technological entity, which we term the Quina Valley. To answer this we systematically analyse the most diagnostic elements of the Quina system, the Quina scrapers and the resharpening flakes removed in maintaining them, and assess their technological consistency with the excavated material from Longtan. Only if the first step can be sustained does the second question arise: whether such an entity can be accounted for by the environmental setting of the region, its available resources and the strategies by which technology was organised. Here we bring together the spatial distribution of the localities, the raw materials used at them and the environmental characteristics of the region. The study clarifies the regional distribution of Quina technology in East Asia and offers a perspective on the relationship between environmental constraint, technological convergence and regional technological organisation in the Middle Palaeolithic.

# Regional Geological and Palaeoenvironmental Setting

The tectonic framework of the study area determines both where its intermontane basins lie and where knappable stone is available within them. The area falls within the Jinsha River drainage in the southern Hengduan Mountains, on the southeastern margin of the Tibetan Plateau, and is controlled by a system of approximately north–south trending faults: continued relative uplift of massifs such as Yulong Mountains and Diancang Shan, alternating with subsidence of the ground between them, has produced a series of north–south trending intermontane basins, among them Lijiang, Heqing and Dali \[Li et al. 2014; An et al. 2011\]. The two basins considered here, Huangping and Binchuan, lie within this same system. This alternation of uplifted ranges and downfaulted basins, combined with the Indian monsoon and the warm, moist airflow from the southwest, gives the deeply incised valley floors a dry-hot valley climate, sheltered from precipitation by the surrounding relief and marked by a sharp division between wet and dry seasons and by high insolation; the modern vegetation there is dominated by drought-tolerant shrub and savanna formations \[\]. The Quina localities discussed here are concentrated in two of these basins, the Huangping Basin in the southwest of Heqing County and the neighbouring Binchuan Basin, each of them cut by tributaries of the Jinsha River: the Caifeng River in the Huangping Basin, and the Sangyuan River together with its tributary the Liandong River in the Binchuan Basin. The two basins differ in bedrock. Permian volcanic rocks are extensively exposed along the western side of the Huangping Basin, with Triassic sedimentary rocks predominating to the east, whereas the Binchuan Basin is largely mantled by unconsolidated Quaternary sediments and exposes its Permian volcanic basement only locally \[\]. The volcanic sequence comprises basalt, tuff and trachyte \[\]. Because the rivers of both basins cut across these outcrops, volcanic cobbles were carried into the gravel bars of the valley floors, and it was these bars, rather than the outcrops themselves, that supplied the stone used at the localities.

The regional monsoon record places the occupation of these basins within an interval of recurrent aridity rather than within a settled climatic phase. The nearest palaeoclimate records covering this period come from the lacustrine sediments of the Heqing Basin, about 60 km north of the Huangping and Binchuan basins and some 650 m higher, which are among the longest and most continuous environmental sequences available for the Asian monsoon domain and are highly sensitive to variation in the Indian Summer Monsoon (ISM) \[Xiao et al. 2010; An et al. 2011\]. Magnetic and geochemical proxies from these sediments divide the interval from 184 to 25 ka into four stages \[Yang et al. 2025\]: relatively arid conditions between c. 184 and 131 ka; a shift to generally humid conditions between 131 and 69 ka; a return to aridity between 69 and 57 ka, corresponding to MIS 4; and a generally humid MIS 3 between 57 and 25 ka, punctuated by repeated arid events. Rapid, millennial-scale weakenings of the monsoon, corresponding closely to North Atlantic cold events, have been resolved in this basin from 52 ka onwards \[Hou et al. 2026\]; these post-date the occupation documented at Longtan, and we cite them to indicate that the monsoon system was liable to interruption of this kind rather than to date any interruption within the period of human presence. These records register monsoon variation at a regional scale rather than local climatic conditions at any one place, and given the distance and altitudinal offset just noted, we treat them here as regional context rather than as a proxy for conditions at the localities themselves. Read at that scale, what matters is not whether a particular stage was on balance wetter or drier, but that the occupation falls where the MIS 4 arid phase gives way to a monsoon recovery that was itself repeatedly interrupted.

Longtan provides the local counterpart to this regional picture, since it preserves both a secure chronology and a record of vegetation change. Optically stimulated luminescence (OSL) dating places its in situ cultural layer at c. 60–50 ka [@ruanQuinaLithicTechnology2025a], and Longtan remains the only locality in the study area with an age determination of this kind. The nearby Tianhuadong site, where material of this kind has also been reported, has yielded OSL ages of c. 90–40 ka for its cultural layers \[\], but it has not been systematically analysed, and neither its chronology nor its technological attribution is yet secure enough to carry weight here. The surface collections examined below have no independent age control, and none can be transferred to them by association: what the Longtan dates establish is when the Quina system was present in this landscape, not when any individual surface collection was formed. What Longtan adds beyond a date is a local environmental record. Pollen analysis there documents a transition from coniferous forest to a more open woodland-steppe, with the Quina-bearing layer falling within the phase of vegetation opening \[Ruan et al. 2025\]. Taken together, the two records indicate that people were present in this landscape while it was becoming more open locally, and during an interval when the regional monsoon system was subject to repeated interruption. Whether that setting can account for the technology found within it is a question we return to once the technological results have been presented.

# Materials and methods

## Survey and selection of localities

The surface material analysed here was recovered during four seasons of pedestrian survey in the Huangping and Binchuan basins, carried out in 2010, 2013, 2016 and 2022. Survey followed the main valleys of the two basins, the Caifeng River in the Huangping Basin and the Sangyuan River together with its tributary the Liandong River in the Binchuan Basin, and each flight of river terraces was walked in turn. Survey areas and walking routes were designed to cover the full range of landforms that humans might have used in the past. Routes were recorded continuously using the mobile GPS application 2bulu, localities and artefacts were photographed, and a field form was completed for each locality giving its position, its place within the terrace sequence, its straight-line distance to the present-day river channel, and the number and types of artefacts present.

Because this study is concerned with Quina technology, the localities recorded during survey were then screened, and only those yielding diagnostic Quina scrapers and resharpening flakes were retained. On this criterion, 22 of the 33 localities recorded in the Binchuan Basin (67%) and all five of those recorded in the Huangping Basin contain Quina specimens (\[Table X\]). Two consequences of this design should be stated at the outset. First, the criterion is one of presence rather than of degree: localities were retained on whether Quina pieces occur at them, not on how those pieces were made or how far they had been reduced, so that none of the technological variables analysed below entered the selection. Pieces at the very lightest end of the retouch range are nonetheless less readily recognised as diagnostic, and the sample is therefore unlikely to represent the earliest stages of the reduction continuum. Second, what Section \[\[X\]\] reports are patterns among Quina-bearing localities rather than a map of Quina technology across the two basins; since the visibility of surface localities depends on exposure, vegetation cover and survey intensity, these patterns are read here as trends and not as densities.

## Survey and analysis of raw materials

To establish what raw material was locally available, gravel bars were sampled at five points along three river channels. At each point an area of 1–2 m² was cleared and every cobble longer than 30 mm exposed at the surface was recorded. For each cobble we recorded length, width and thickness (mm), roundness on a four-point scale (0–3, from least to most rounded) and shape (sub-oval, tabular or irregular). These records give the relative abundance of each lithology on the gravel bars (n = xx).

To characterise the internal properties of the dominant raw material, thin sections were prepared from five trachyte cobbles and examined petrographically, with mineral composition determined by energy-dispersive spectroscopy. Identifications were carried out by the Guangxi Institute of Geological Survey and the China University of Geosciences (Wuhan). The raw material of each artefact was assigned macroscopically to the same lithological categories, using the surveyed cobbles as a reference collection.

Differences in raw material composition, both between the gravel bars and the artefacts and between spatial units, were tested by permutational multivariate analysis of variance (PERMANOVA), stratified at two scales, basin and valley.

Because the artefacts are made overwhelmingly on river cobbles, the straight-line distance from each locality to the nearest present-day channel was used as a proxy for proximity to a source of raw material. These distances are strongly skewed, the groups differ considerably in size, and their variances are unequal, so differences between the two basins were tested with a Mann–Whitney rank sum test and differences among the three valleys with a Kruskal–Wallis rank sum test; medians are reported as the measure of central tendency and the rank tests taken as the primary criterion. cobble size was summarised as the geometric mean of length, width and thickness and compared between lithologies with a Mann–Whitney rank sum test, with the geometric mean size of each lithology and the effect size r reported alongside.

Raw material selection was quantified with Jacobs' electivity index D [@jacobs1974], a modification of Ivlev's index that corrects for the relative abundance of the available resources:

$$D = \frac{r - p}{\,r + p - 2\,r\,p\,}$$

where r is the proportion of a given raw material among the artefacts and p is its proportion among the cobbles recorded in the gravel survey. D ranges from −1 (complete avoidance) through 0 (use in proportion to availability) to +1 (complete selection). It was computed for each raw material in the surface Quina scrapers against availability pooled across the gravel survey. Materials present among the artefacts but absent from the survey take D = +1 and are read as evidence of non-local procurement; materials present in the survey but absent from the artefacts take D = −1.

## Techno-typological analysis of Quina retouch products

The techno-typological observations and measurements follow established practice in the study of Quina technology \[Bourguignon 1997; Peresani et al. 2024; Ruan et al. 2025\]. For each scraper we recorded length, width, thickness and mass; blank type; cortex position; platform type and platform depth; interior platform angle (IPA); the number of retouch generations; the numbers of convex scars and concave scars; the geometric index of unifacial reduction (GIUR) \[Kuhn 1990\] and the proportion of the perimeter retouched; the edge angle; and its attribution within Bordes' type list. For each resharpening flake we recorded platform type, depth, IPA and EPA. Statistical tests were chosen according to the type of each variable and the distribution of the data, with parametric or non-parametric procedures used depending on whether the data approximated normality. Relationships between edge angle and GIUR, the proportion of retouched edge and the mean number of retouch generations were examined with Spearman rank correlation, with p-values adjusted by the Benjamini–Hochberg procedure. The difference between the exterior platform angles of the resharpening flakes and the edge angles of the scrapers was tested with Welch's t-test, so that unequal variances would not affect the inference.

To test the technological consistency between the surface Quina scrapers and the excavated material, the Quina scrapers from Longtan were used as the reference, and ordinary scrapers (or non-Quina scrapers) from the same site were included as a technological control group. Six of the recorded variables were selected for multivariate analysis as the most representative: thickness, proportion of the perimeter retouched, GIUR, number of total scars, number of retouch generations and edge angle. All were standardised to z-scores and differences between specimens computed as Euclidean distances. Differences between groups were tested by PERMANOVA, with the proportion of variance explained (R²) reported for each pairwise comparison. The ordinary scrapers serve to calibrate the scale on which those R² values are read: if the same set of variables separates Quina from ordinary scrapers with a high R² while the two Quina groups differ only at a low R², the latter is more likely to reflect a real technological similarity than a lack of statistical power. PERMANOVA compares the positions of group centroids in multivariate space. At the univariate level, mean GIUR, scar count and mean number of retouch generations were compared with Kruskal–Wallis rank sum tests followed by Dunn's post hoc comparisons, and thickness, proportion of the perimeter retouched and edge angle with Welch's ANOVA followed by Welch's t post hoc comparisons; all post hoc comparisons were corrected using the Bonferroni method. Technological dispersion between groups was tested with PERMDISP. To compare the degree of variation within groups, the coefficients of variation (CV) of length, width, thickness, mass and edge angle were tested for equality across groups using the Krishnamoorthy–Lee modified signed likelihood ratio test (KL-MSLRT) [@krishnamoorthy2014], these being ratio-scale variables with a true zero. GIUR, the proportion of the perimeter retouched, number of total scars and the number of retouch generations were instead compared with Fligner–Killeen tests: the first two are bounded on \[0, 1\], which leaves the CV without a clear interpretation, and the last two are counts whose variance structure does not meet the assumptions required for comparing CVs. To assess how robust the results are, the analyses were repeated with the surface specimens collected from within the area of the Longtan site removed.

To test whether technological variation is systematically structured by the landscape, the analysis was repeated with the locality as the unit, each locality represented by the median value of its Quina scrapers on the same six variables. Four landscape variables were considered: basin (Huangping or Binchuan), the height of the locality above the present-day channel, distance from the locality to that channel, and assemblage size (number of specimens). Height above the channel was used in preference to a categorical terrace level, and in preference to absolute elevation: it is the continuous quantity of which terrace level is a discretisation, and because it is measured from the local channel rather than from sea level, the difference in base level between the two basins cancels out (Spearman ρ between height and basin = −0.03). Basin, the only categorical variable, was tested with PERMANOVA and PERMDISP; the three continuous variables with Spearman rank correlation, and with locality-level PERMANOVA on Euclidean distances, height and assemblage size additionally being examined in coarse bands as a check on robustness. Because the number of units at this level is small (n = 27), most of them yielding only one or two specimens, so that locality medians are not very stable, and because height above the channel is strongly right-skewed, the three hilltop localities lying far above the terrace sequence, these tests have limited statistical power; the results are therefore interpreted mainly from the size and direction of the effect sizes (R², ρ), with the significance tests treated as supporting evidence only.

All statistical analyses were carried out in R 4.5.2 [@rcoreteam2025]. PERMANOVA and PERMDISP were run with vegan 2.7-2 [@oksanen2025vegan], in both cases with 9999 permutations. The Kruskal–Wallis, Mann–Whitney and Dunn tests, Welch's ANOVA and the Welch's *t* post hoc comparisons, Levene's and Games–Howell tests, and the accompanying effect sizes (ε², rank-biserial *r*, Cohen's *d*) were computed with rstatix 0.7.3 [@kassambara2023rstatix]. The KL-MSLRT was computed with cvequality 0.2.0 [@marwick2019cvequality], using 10^5^ Monte Carlo iterations. Fligner–Killeen tests, Spearman rank correlations and the Benjamini–Hochberg and Bonferroni adjustments were taken from the stats package. Confidence intervals for the coefficients of variation and for the dispersion ratios are bootstrap percentile intervals from 5,000 replicates; the interval for Cohen's *d* is from 1,000 replicates. A fixed random seed (2226) is set at the head of every script and in the manuscript source, so that all permutation, Monte Carlo and bootstrap results reported here are exactly reproducible.

## Reproducibility and open materials

Computational reproducibility remains a challenge in archaeological science, yet it underpins both the credibility of quantitative results and their re-use by others [@marwickComputationalReproducibilityArchaeological2017; @marwickOpenScienceArchaeology2017]. All materials needed to reproduce the analyses reported here are therefore available in a version-controlled online repository at \[\[OSF DOI\]\]. These comprise the raw data underlying every reported result together with the complete code used for the analyses and for the figures derived from them. The repository is organised with a logical folder structure and relative file paths, so that the code runs without manual modification. The data are released under a CC-0 licence, the code under the MIT licence and the figures under CC-BY. Every statistical result reported below, and every figure generated from the data, can therefore be reproduced independently from these materials.

# Results

```{r}
#| label: results-analysis
#| include: false

# ==========================================================================
# All statistical results reported in this section are computed here from the
# raw data. Permutation tests use PERM = 9999 permutations and the fixed seed
# 2226 set in the setup chunk; the seed is re-set before each permutation test
# so results do not depend on chunk-evaluation order. The mirror analysis
# scripts under scripts/ produce the same numbers plus their diagnostic
# figures. Formatters f() and fp() are defined in the setup chunk.
# ==========================================================================

sc_path    <- here::here("data", "Quina_scraper_surface.xlsx")
lt_path    <- here::here("data", "Longtan_lithic_tools.xlsx")
basin_path <- here::here("data", "Raw_mat_basin.xlsx")
site_path  <- here::here("data", "Site_information.xlsx")

drop_sites      <- c("PJDD", "ZKZ")
material_levels <- c("Trachyte", "Sandstone", "Quartz", "Mudstone", "Andesite")
river_levels    <- c("Sangyuan", "Liandong", "Caifeng")
basin_levels    <- c("Binchuan", "Heqing")
variables       <- c("Thickness", "Retouch_length_index", "Ave_GIUR",
                     "N_Scar", "Ave_RG", "Edge_Angle")

harmonise_lithology <- function(x) {
  x <- trimws(as.character(x))
  dplyr::recode(x, "Quartz sandstone" = "Sandstone", "Coarse sandstone" = "Sandstone")
}
strip_basin <- function(x) sub(" basin$", "", trimws(as.character(x)))
gm <- function(x) exp(mean(log(x[is.finite(x) & x > 0])))

q     <- read_excel(sc_path, sheet = "Quina scraper")
rfl   <- read_excel(sc_path, sheet = "Resharpening flake")
sites <- read_excel(site_path); names(sites) <- trimws(names(sites))

# --------------------------------------------------------------------------
# 1. Distance-to-river by basin and valley (site-level; rank tests)
# --------------------------------------------------------------------------
dist_dat <- sites |>
  transmute(Code = trimws(as.character(Code)),
            basin    = factor(strip_basin(basin), levels = basin_levels),
            river_ID = factor(trimws(river_ID),   levels = river_levels),
            d = as.numeric(d_river_m)) |>
  filter(!Code %in% drop_sites, !is.na(d), !is.na(basin), !is.na(river_ID))
dist_median  <- median(dist_dat$d)
mw_basin     <- dist_dat |> wilcox_test(d ~ basin)
dist_basin_U <- mw_basin$statistic; dist_basin_p <- mw_basin$p
kw_river     <- dist_dat |> kruskal_test(d ~ river_ID)
dist_river_H <- kw_river$statistic; dist_river_df <- kw_river$df; dist_river_p <- kw_river$p

# --------------------------------------------------------------------------
# 2. Raw-material composition (available vs used) and Jacobs' D
# --------------------------------------------------------------------------
avail_raw <- harmonise_lithology(read_excel(basin_path, "Sheet1")[["Lithology"]])
comp_avail <- tibble(Material = avail_raw) |>
  filter(!is.na(Material), !Material %in% c("", "NA")) |>
  count(Material) |> mutate(p = 100 * n / sum(n))
n_clast        <- sum(comp_avail$n)
sand_avail_pct <- comp_avail$p[comp_avail$Material == "Sandstone"]
trach_avail_pct<- comp_avail$p[comp_avail$Material == "Trachyte"]

used_raw <- harmonise_lithology(q[["Raw_material"]])
comp_used <- tibble(Material = used_raw) |>
  filter(!is.na(Material), !Material %in% c("", "NA")) |>
  count(Material) |> mutate(r = 100 * n / sum(n))
trach_used_pct <- comp_used$r[comp_used$Material == "Trachyte"]
andesite_n     <- sum(comp_used$n[comp_used$Material == "Andesite"])

jac <- full_join(comp_used |> transmute(Material, r = r / 100),
                 comp_avail |> transmute(Material, p = p / 100), by = "Material") |>
  mutate(r = tidyr::replace_na(r, 0), p = tidyr::replace_na(p, 0),
         D = (r - p) / (r + p - 2 * r * p))
Dget   <- function(m) jac$D[jac$Material == m]
D_trach <- Dget("Trachyte"); D_sand <- Dget("Sandstone")
D_quartz <- Dget("Quartz");  D_mud <- Dget("Mudstone"); D_and <- Dget("Andesite")

# PERMANOVA: used-tool composition ~ basin / valley (artifact-level, Bray-Curtis)
site_key <- sites |>
  transmute(Site_ID = trimws(as.character(Code)),
            basin    = factor(strip_basin(basin), levels = basin_levels),
            river_ID = factor(trimws(river_ID),   levels = river_levels))
scA <- q |>
  transmute(Site_ID = trimws(as.character(Site_ID)),
            Material = harmonise_lithology(Raw_material)) |>
  filter(!is.na(Material), !Material %in% c("", "NA"), !Site_ID %in% drop_sites) |>
  left_join(site_key, by = "Site_ID") |>
  mutate(Material = factor(Material, levels = material_levels))
A_D    <- vegdist(as.matrix(unclass(table(seq_len(nrow(scA)), droplevels(scA$Material)))), "bray")
A_meta <- data.frame(basin = droplevels(scA$basin), river_ID = droplevels(scA$river_ID))
set.seed(2226); usedA_basin <- adonis2(A_D ~ basin,    data = A_meta, permutations = PERM)
set.seed(2226); usedA_river <- adonis2(A_D ~ river_ID, data = A_meta, permutations = PERM)

# PERMANOVA: available clasts ~ sampling point / valley (clast-level heterogeneity)
clasts <- read_excel(basin_path, "Sheet1"); names(clasts) <- trimws(names(clasts))
clasts <- clasts |>
  transmute(Loc = as.character(Loc),
            river_ID = factor(trimws(river_ID), levels = river_levels),
            Material = factor(harmonise_lithology(Lithology), levels = material_levels)) |>
  filter(!is.na(Material))
C_D <- vegdist(as.matrix(unclass(table(seq_len(nrow(clasts)), clasts$Material))), "bray")
set.seed(2226); clastLoc   <- adonis2(C_D ~ Loc,      data = clasts, permutations = PERM)
set.seed(2226); clastRiver <- adonis2(C_D ~ river_ID, data = clasts, permutations = PERM)

# extract each PERMANOVA to plain scalars (the `Pr(>F)` column name cannot be
# reached from inline `r ...` code, so the p-values are pulled out here)
ad3 <- function(a) c(R2 = a$R2[1], F = a$F[1], p = a$`Pr(>F)`[1])
usedA_basin_v <- ad3(usedA_basin); usedA_river_v <- ad3(usedA_river)
clastLoc_v    <- ad3(clastLoc);    clastRiver_v  <- ad3(clastRiver)

# --------------------------------------------------------------------------
# 3. Clast size and shape by raw material
# --------------------------------------------------------------------------
raw <- read_excel(basin_path, "Sheet1"); names(raw) <- trimws(names(raw))
bchr <- trimws(as.character(raw$Breadth)); bchr[bchr == "69..5"] <- "69.5"  # repair typo
cl <- raw |>
  transmute(Lithology = trimws(as.character(Lithology)),
            Material = factor(harmonise_lithology(Lithology), levels = material_levels),
            L  = suppressWarnings(as.numeric(as.character(Length))),
            B  = suppressWarnings(as.numeric(bchr)),
            Th = suppressWarnings(as.numeric(as.character(Thickness))),
            Shape = trimws(as.character(Shape))) |>
  mutate(size = (L * B * Th)^(1 / 3)) |>
  filter(is.finite(size), L > 0, B > 0, Th > 0)
sz  <- cl |> filter(Material %in% c("Trachyte", "Sandstone")) |>
  mutate(G = factor(Material, levels = c("Trachyte", "Sandstone")))
gmT <- gm(sz$size[sz$G == "Trachyte"]); gmS <- gm(sz$size[sz$G == "Sandstone"])
size_wt <- suppressWarnings(wilcox.test(size ~ G, data = sz))
nT <- sum(sz$G == "Trachyte"); nS <- sum(sz$G == "Sandstone")
size_U  <- min(unname(size_wt$statistic), nT * nS - unname(size_wt$statistic))
size_p  <- size_wt$p.value
trs <- cl |> filter(Lithology == "Trachyte")
trach_tabular_pct <- 100 * mean(trs$Shape == "Tablet")
trach_irreg_pct   <- 100 * mean(trs$Shape == "Irregular")

# --------------------------------------------------------------------------
# 4. Quina-scraper techno-typology (descriptive)
# --------------------------------------------------------------------------
n_scraper        <- nrow(q)
cortex_bearing_n <- sum(q$Cortex != "0");  cortex_bearing_pct <- 100 * cortex_bearing_n / n_scraper
nocortex_n       <- sum(q$Cortex == "0");  nocortex_pct       <- 100 * nocortex_n / n_scraper
lateral_n <- sum(q$Cortex_position == "Lateral"); lateral_pct <- 100 * lateral_n / cortex_bearing_n
primary_n <- sum(q$Cortex_position == "Primary"); primary_pct <- 100 * primary_n / n_scraper

plat   <- q |> filter(!is.na(Platform_type))
plat_n <- nrow(plat); plat_pct <- 100 * plat_n / n_scraper
plain_n   <- sum(plat$Platform_type == "Plain");    plain_pct   <- 100 * plain_n / plat_n
natural_n <- sum(plat$Platform_type == "Natural");  natural_pct <- 100 * natural_n / plat_n
dihedral_n<- sum(plat$Platform_type == "Dihedral"); dihedral_pct<- 100 * dihedral_n / plat_n
pd  <- as.numeric(plat$Platform_depth)
pd_median <- median(pd, na.rm = TRUE); pd_q1 <- quantile(pd, .25, na.rm = TRUE); pd_q3 <- quantile(pd, .75, na.rm = TRUE)
ipa <- as.numeric(plat$IPA); ipa_mean <- mean(ipa, na.rm = TRUE); ipa_sd <- sd(ipa, na.rm = TRUE)

cxs <- rowSums(cbind(as.numeric(q$E1_N_CxS), as.numeric(q$E2_N_CxS), as.numeric(q$E3_N_CxS)), na.rm = TRUE)
cvs <- rowSums(cbind(as.numeric(q$E1_N_CvS), as.numeric(q$E2_N_CvS), as.numeric(q$E3_N_CvS)), na.rm = TRUE)
rg  <- as.numeric(q$Ave_RG)
rg_mean  <- mean(rg, na.rm = TRUE);  rg_sd  <- sd(rg, na.rm = TRUE)
cxs_mean <- mean(cxs); cxs_sd <- sd(cxs); cvs_mean <- mean(cvs); cvs_sd <- sd(cvs)

giur <- as.numeric(q$Ave_GIUR); rli <- as.numeric(q$Retouch_length_index); ea <- as.numeric(q$Edge_Angle)
giur_median <- median(giur, na.rm = TRUE); giur_q1 <- quantile(giur, .25, na.rm = TRUE); giur_q3 <- quantile(giur, .75, na.rm = TRUE)
rli_median  <- median(rli,  na.rm = TRUE); rli_q1  <- quantile(rli,  .25, na.rm = TRUE); rli_q3  <- quantile(rli,  .75, na.rm = TRUE)
ea_mean <- mean(ea, na.rm = TRUE); ea_sd <- sd(ea, na.rm = TRUE)

spear <- function(v) {
  ok <- is.finite(ea) & is.finite(v)
  ct <- suppressWarnings(cor.test(ea[ok], v[ok], method = "spearman", exact = FALSE))
  c(rho = unname(ct$estimate), p = ct$p.value)
}
ea_giur <- spear(giur); ea_rli <- spear(rli); ea_rg <- spear(rg)

# --------------------------------------------------------------------------
# 5. Resharpening flakes: EPA vs scraper edge angle (Welch t + Cohen's d)
# --------------------------------------------------------------------------
epa   <- as.numeric(rfl$EPA); pd_rf <- as.numeric(rfl$Platform_depth)
rf_n  <- sum(is.finite(epa))
rf_pd_mean <- mean(pd_rf, na.rm = TRUE); rf_pd_sd <- sd(pd_rf, na.rm = TRUE)
epa_mean   <- mean(epa,  na.rm = TRUE);  epa_sd   <- sd(epa,  na.rm = TRUE)
da <- bind_rows(tibble(Group = "edge", Value = ea), tibble(Group = "epa", Value = epa)) |>
  filter(!is.na(Value)) |> mutate(Group = factor(Group, levels = c("edge", "epa")))
welch  <- rstatix::t_test(da, Value ~ Group, var.equal = FALSE, detailed = TRUE)
welch_t <- welch$statistic; welch_df <- welch$df; welch_p <- welch$p
set.seed(2226)
cohd_res <- rstatix::cohens_d(da, Value ~ Group, var.equal = FALSE, ci = TRUE, nboot = 1000)
cohd <- cohd_res$effsize; cohd_lo <- cohd_res$conf.low; cohd_hi <- cohd_res$conf.high

# --------------------------------------------------------------------------
# 6. Technological consistency: surface vs Longtan (PERMANOVA + PERMDISP)
#    Also assembles the `tc` object consumed by fig-technological-consistency.
# --------------------------------------------------------------------------
grp_levels <- c("SC_Quina", "LT_Quina", "LT_Ordinary")
read_grp <- function(path, sheet, g, site = FALSE) {
  df <- read_excel(path, sheet = sheet)
  df$Group   <- g
  df$Site_ID <- if (site) trimws(as.character(df$Site_ID)) else NA_character_
  df |> mutate(across(all_of(variables), as.numeric)) |> select(Group, Site_ID, all_of(variables))
}
scraper_all <- bind_rows(
  read_grp(sc_path, "Quina scraper",    "SC_Quina",    TRUE),
  read_grp(lt_path, "Quina scraper",    "LT_Quina",    FALSE),
  read_grp(lt_path, "Ordinary scraper", "LT_Ordinary", FALSE)) |>
  mutate(Group = factor(Group, levels = grp_levels))
complete_data <- scraper_all |> filter(if_all(all_of(variables), ~ !is.na(.x)))
n_by_group <- table(complete_data$Group)
ltq_n <- n_by_group[["LT_Quina"]]; lto_n <- n_by_group[["LT_Ordinary"]]
amat  <- complete_data |> select(all_of(variables)) |> scale() |> as.matrix()

pairwise_permanova <- function(data, m) {
  gs <- levels(droplevels(data$Group))
  bind_rows(lapply(combn(gs, 2, simplify = FALSE), function(pair) {
    rows <- data$Group %in% pair; pd <- data[rows, ]; pd$Group <- droplevels(pd$Group)
    set.seed(2226); mm <- adonis2(dist(m[rows, ]) ~ Group, data = pd, permutations = PERM)
    data.frame(Comparison = paste(pair, collapse = " vs "),
               R2 = mm$R2[1], F = mm$F[1], p_value = mm$`Pr(>F)`[1])
  })) |> mutate(p_adjusted = p.adjust(p_value, "BH"))
}
tc_pw <- pairwise_permanova(complete_data, amat)
pwget <- function(comp, col) tc_pw[[col]][tc_pw$Comparison == comp]
tc_scltq_R2 <- pwget("SC_Quina vs LT_Quina", "R2")
tc_scltq_F  <- pwget("SC_Quina vs LT_Quina", "F")
tc_scltq_p  <- pwget("SC_Quina vs LT_Quina", "p_adjusted")
tc_nonq_R2  <- c(pwget("SC_Quina vs LT_Ordinary", "R2"),
                 pwget("LT_Quina vs LT_Ordinary", "R2"))
tc_nonq_p   <- max(pwget("SC_Quina vs LT_Ordinary", "p_adjusted"),
                   pwget("LT_Quina vs LT_Ordinary", "p_adjusted"))

# per-variable tests + post-hoc brackets (panel B of the figure)
variable_long <- complete_data |> select(Group, all_of(variables)) |>
  pivot_longer(all_of(variables), names_to = "Variable", values_to = "Value") |>
  mutate(Variable = factor(Variable, levels = variables))
kw_vars    <- c("Ave_GIUR", "N_Scar", "Ave_RG")
welch_vars <- c("Retouch_length_index", "Thickness", "Edge_Angle")
kw_ph <- variable_long |> filter(Variable %in% kw_vars) |> mutate(Variable = droplevels(Variable)) |>
  group_by(Variable) |> dunn_test(Value ~ Group, p.adjust.method = "bonferroni") |> ungroup()
we_ph <- variable_long |> filter(Variable %in% welch_vars) |> mutate(Variable = droplevels(Variable)) |>
  group_by(Variable) |> pairwise_t_test(Value ~ Group, pool.sd = FALSE, p.adjust.method = "bonferroni") |> ungroup()
posthoc_brackets <- bind_rows(
  kw_ph |> transmute(Variable, group1, group2, p.adj, p.adj.signif),
  we_ph |> transmute(Variable, group1, group2, p.adj, p.adj.signif)) |>
  mutate(Variable = factor(as.character(Variable), levels = variables)) |>
  group_by(Variable) |> mutate(step = row_number()) |> ungroup() |>
  left_join(variable_long |> group_by(Variable) |>
              summarise(ymax = max(Value, na.rm = TRUE),
                        yrange = diff(range(Value, na.rm = TRUE)), .groups = "drop"),
            by = "Variable") |>
  mutate(y.position = ymax + yrange * (0.06 + 0.10 * step))

# PERMDISP (Block A) on the z-scored technical space -> means, pairwise, PCA
run_permdisp <- function(df) {
  mvd <- df |> filter(if_all(all_of(variables), is.finite)) |> mutate(Group = droplevels(Group))
  mat <- scale(as.matrix(mvd[, variables])); d <- dist(mat)
  set.seed(2226); bd <- betadisper(d, mvd$Group)
  set.seed(2226); pt <- permutest(bd, permutations = PERM, pairwise = TRUE)
  pca <- prcomp(mat, center = TRUE, scale. = FALSE); vexp <- pca$sdev^2 / sum(pca$sdev^2) * 100
  list(means = tapply(bd$distances, mvd$Group, mean), overall_p = pt$tab$`Pr(>F)`[1],
       pairwise = pt$pairwise$permuted, n_by_group = table(mvd$Group),
       pca_scores = data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], Group = mvd$Group),
       pca_loadings = data.frame(Variable = rownames(pca$rotation),
                                 PC1 = pca$rotation[, 1], PC2 = pca$rotation[, 2], row.names = NULL),
       var_explained = vexp)
}
permA <- run_permdisp(scraper_all)
pd_scq <- permA$means[["SC_Quina"]]; pd_ltq <- permA$means[["LT_Quina"]]; pd_lto <- permA$means[["LT_Ordinary"]]
permdisp_scltq_p <- permA$pairwise[["SC_Quina-LT_Quina"]]
permdisp_nonq_p  <- max(permA$pairwise[["SC_Quina-LT_Ordinary"]], permA$pairwise[["LT_Quina-LT_Ordinary"]])
# sensitivity: drop surface pieces from within the Longtan/Tianhua site area
permC <- run_permdisp(scraper_all |> filter(!(Group == "SC_Quina" & Site_ID %in% c("LT", "THC"))))
sens_scltq_p <- permC$pairwise[["SC_Quina-LT_Quina"]]
sens_nonq_p  <- max(permC$pairwise[["SC_Quina-LT_Ordinary"]], permC$pairwise[["LT_Quina-LT_Ordinary"]])

tc <- list(pca_scores = permA$pca_scores, pca_loadings = permA$pca_loadings,
           var_explained = permA$var_explained, n_by_group = permA$n_by_group,
           variable_long = variable_long, posthoc_brackets = posthoc_brackets,
           permanova_pairwise = tc_pw,
           permdisp = list(means = permA$means, overall_p = permA$overall_p, pairwise = permA$pairwise))

# --------------------------------------------------------------------------
# 7. Landscape structure of technical variation (locality-level)
# --------------------------------------------------------------------------
site_land <- sites |>
  transmute(Site_ID = trimws(as.character(Code)),
            Basin = factor(strip_basin(basin), levels = basin_levels),
            Distance_to_water = as.numeric(d_river_m),
            Height_above_river = as.numeric(h_river_m))
scL <- q |>
  mutate(Site_ID = trimws(as.character(Site_ID)), across(all_of(variables), as.numeric)) |>
  left_join(site_land, by = "Site_ID") |>
  filter(if_all(all_of(variables), ~ !is.na(.x)))
site_size <- scL |> count(Site_ID, name = "Site_size")
site_df <- scL |> group_by(Site_ID) |>
  summarise(n_art = n(), across(all_of(variables), ~ median(.x, na.rm = TRUE)),
            Basin = first(Basin), Distance_to_water = first(Distance_to_water),
            Height_above_river = first(Height_above_river), .groups = "drop") |>
  left_join(site_size, by = "Site_ID")
n_localities <- nrow(site_df)
dL <- dist(scale(as.matrix(site_df[, variables])))
grad_permanova <- function(col) {
  set.seed(2226); a <- adonis2(dL ~ site_df[[col]], permutations = PERM)
  c(R2 = a$R2[1], F = a$F[1], p = a$`Pr(>F)`[1])
}
ls_size   <- grad_permanova("Site_size")
ls_dist   <- grad_permanova("Distance_to_water")
ls_height <- grad_permanova("Height_above_river")
bdf <- site_df |> filter(!is.na(Basin))
set.seed(2226); aB <- adonis2(dist(scale(as.matrix(bdf[, variables]))) ~ Basin, data = bdf, permutations = PERM)
ls_basin <- c(R2 = aB$R2[1], F = aB$F[1], p = aB$`Pr(>F)`[1])
ls_dist_rho <- vapply(variables, function(v) {
  suppressWarnings(unname(cor.test(site_df[[v]], site_df$Distance_to_water,
                                   method = "spearman", exact = FALSE)$estimate))
}, numeric(1))

```

## Landscape distribution and raw material economy

Quina localities in both basins lie along the valley systems, on river terraces, and at consistently short distances from the channels. In the Huangping Basin they are concentrated on the second and third terraces; in the Binchuan Basin they extend across the second to fourth terraces (T2–T4), with a small number (n = 3) occurring higher, on hilltops. Because the artefacts are made overwhelmingly on river cobbles, the straight-line distance from a locality to the nearest channel serves as a proxy for ease of access to lithic raw material. Across the three valleys, the Sangyuan, Liandong and Caifeng, these distances are short and similarly distributed, with a median of `r f(dist_median, 0)` m. Neither comparison detected a difference between spatial units, whether at the scale of the basin (U = `r f(dist_basin_U, 0)`, p `r fp(dist_basin_p)`) or of the valley (H = `r f(dist_river_H, 2)`, df = `r dist_river_df`, p `r fp(dist_river_p)`). Access to river gravels therefore appears to have been broadly comparable across the study area, so that differences in the raw materials actually used at these localities are unlikely to be explained by differential proximity to a source.

Raw material use does not follow raw material abundance. On the gravel bars, sandstone is the most abundant lithology (`r f(sand_avail_pct, 1)`% of `r n_clast` cobbles), with trachyte second (`r f(trach_avail_pct, 1)`%); among the artefacts, trachyte accounts for `r f(trach_used_pct, 1)`%. The bars are not uniform in composition, however: raw material composition differs significantly between sampling points (R² = `r f(clastLoc_v["R2"], 3)`, pseudo-F = `r f(clastLoc_v["F"], 2)`, p `r fp(clastLoc_v["p"])`), and stratified analysis attributes this variation to differences between valleys (R² = `r f(clastRiver_v["R2"], 3)`, pseudo-F = `r f(clastRiver_v["F"], 2)`, p `r fp(clastRiver_v["p"])`) rather than within them. Sandstone-dominated and trachyte-richer bars are therefore distributed unevenly across the study area, in a pattern that follows the contrast in bedrock described in Section 2 (\[\[Fig. X\]\]). If knappers had simply worked whatever lay most abundantly at hand, the raw material composition of the artefacts should vary from valley to valley in the same way.

That expectation is not met. The raw material composition of the Quina scrapers is uniform across the study area, differing neither between basins (R² = `r f(usedA_basin_v["R2"], 3)`, pseudo-F = `r f(usedA_basin_v["F"], 2)`, p `r fp(usedA_basin_v["p"])`) nor between valleys (R² = `r f(usedA_river_v["R2"], 3)`, pseudo-F = `r f(usedA_river_v["F"], 2)`, p `r fp(usedA_river_v["p"])`) (@fig-raw-material-composition). The same material was selected wherever knappers used, although sandstone is more abundant on the bars and although the composition of those bars varies from valley to valley. Jacobs' electivity index puts the strength of that preference on a common scale: trachyte is strongly selected (D = `r f(D_trach, 2)`) and sandstone strongly avoided (D = `r f(D_sand, 2)`), with quartz and mudstone likewise avoided (D = `r f(D_quartz, 2)` in each case, neither being used at all); andesite returns D = `r f(D_and, 2)`, but only because the `r andesite_n` andesite artefacts have no counterpart among the recorded cobbles, so that its index is undefined in practice rather than informative (\[\[Table X\]\]). Because D returns to zero only when a raw material is used in proportion to its availability, trachyte would have to make up some `r f(trach_used_pct, 0)`% of the gravel before its use could be read as neutral; no redistribution of availability among the valleys approaches that figure, so neither the direction nor the magnitude of the selection depends on how availability is pooled. The selection of trachyte, in other words, was made against local abundance rather than in step with it.

```{r}
#| label: fig-raw-material-composition
#| fig-cap: "Raw-material composition, available versus used, across the three valleys. Each row of stacked bars is one 100% composition: top, the lithology of the river-gravel clasts (available); bottom, the lithology of the surface Quina scrapers (used); with per-cell sample sizes at right. Available composition swings from valley to valley while used composition stays close to 100% trachyte throughout — selection against local abundance."
#| fig-width: 5.9
#| fig-height: 4.25

# port of scripts/figures/raw_material_composition_figure.R
material_colors_fig <- c(Trachyte = "#E07C90", Sandstone = "#6BA8CE",
                         Quartz = "#E6C25C", Mudstone = "#9B87C4", Andesite = "#6FB98E")
layer_levels <- c("River cobbles", "Surface Quina scrapers")

site_key_fig <- sites |>
  transmute(Site_ID = trimws(as.character(Code)),
            river_ID = factor(trimws(river_ID), levels = river_levels))

used_fig <- read_excel(sc_path, sheet = "Quina scraper") |>
  transmute(Site_ID = trimws(as.character(Site_ID)),
            Material = harmonise_lithology(Raw_material)) |>
  filter(!is.na(Material), !Material %in% c("", "NA"), !Site_ID %in% drop_sites) |>
  left_join(site_key_fig, by = "Site_ID") |> filter(!is.na(river_ID)) |>
  transmute(Layer = layer_levels[2], river_ID, Material)

avail_fig <- read_excel(basin_path, sheet = "Sheet1")
names(avail_fig) <- trimws(names(avail_fig))
avail_fig <- avail_fig |>
  transmute(river_ID = factor(trimws(river_ID), levels = river_levels),
            Material = harmonise_lithology(Lithology)) |>
  filter(!is.na(Material), !is.na(river_ID)) |>
  transmute(Layer = layer_levels[1], river_ID, Material)

both_fig <- bind_rows(avail_fig, used_fig) |>
  mutate(Layer = factor(Layer, levels = layer_levels),
         Material = factor(Material, levels = material_levels))

comp_fig <- both_fig |>
  count(Layer, river_ID, Material, name = "n") |>
  complete(Layer, river_ID, Material, fill = list(n = 0)) |>
  group_by(Layer, river_ID) |> mutate(percent = 100 * n / sum(n)) |>
  arrange(desc(as.integer(Material)), .by_group = TRUE) |>
  mutate(pos = cumsum(percent) - percent / 2) |> ungroup()
counts_fig <- both_fig |> count(Layer, river_ID, name = "N")

fig_theme <- theme_minimal(base_size = 9) +
  theme(panel.grid.major = element_line(color = "#E6E8EB", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.5),
        axis.ticks = element_line(color = "#202124", linewidth = 0.3),
        axis.ticks.length = grid::unit(2, "pt"),
        axis.title = element_text(size = 9), axis.text = element_text(color = "#303238", size = 8),
        strip.text = element_text(face = "bold", color = "#202124", size = 9),
        strip.background = element_rect(fill = "#E8E8E8", color = NA),
        legend.title = element_text(size = 8.5), legend.text = element_text(size = 8.5),
        legend.key = element_blank(), legend.key.size = grid::unit(10, "pt"),
        plot.background = element_rect(color = NA, fill = "white"),
        panel.background = element_rect(color = NA, fill = "white"))

valley_order <- rev(river_levels)
p_comp <- ggplot(comp_fig, aes(percent, river_ID, fill = Material)) +
  geom_col(width = 0.72, color = "white", linewidth = 0.3) +
  geom_text(data = subset(comp_fig, percent >= 7),
            aes(x = pos, label = sprintf("%.0f%%", percent)),
            size = 2.5, colour = "white", fontface = "bold") +
  facet_wrap(~ Layer, ncol = 1) +
  scale_fill_manual(values = material_colors_fig, drop = FALSE) +
  scale_y_discrete(limits = valley_order) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.02)),
                     breaks = seq(0, 100, 25), labels = function(x) paste0(x, "%")) +
  labs(x = "Percentage", y = "Valley", fill = "Raw material") +
  fig_theme + theme(panel.grid.major = element_blank(), legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 1))
p_n <- ggplot(counts_fig, aes(N, river_ID)) +
  geom_col(width = 0.62, fill = "#8A8F96") +
  facet_wrap(~ Layer, ncol = 1) + scale_y_discrete(limits = valley_order) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Count (n)", y = NULL) +
  fig_theme + theme(panel.grid.major = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        strip.text = element_blank(), strip.background = element_blank())

p_comp + p_n + plot_layout(widths = c(3.4, 1), guides = "collect") &
  theme(legend.position = "bottom")
```

Three properties of trachyte account for this preference. First, size: trachyte cobbles (GM = `r f(gmT, 1)` mm) are significantly larger than sandstone (GM = `r f(gmS, 1)` mm) (U = `r f(size_U, 0)`, p `r fp(size_p)`), giving a larger initial volume from which to sustain prolonged reduction. Second, shape: `r f(trach_tabular_pct, 1)`% of trachyte cobbles are tabular and `r f(trach_irreg_pct, 1)`% irregular, these angular forms offer the acute natural edges and flat surfaces from which removals can be struck. Third, fracture properties: in thin section the trachyte has a porphyritic texture, with euhedral tabular to prismatic K-feldspar phenocrysts (c. 0.3–5.0 mm, c. 1–10% of the rock) set in a groundmass of K-feldspar microlites, chlorite or altered mafic minerals and magnetite, together with about 50% devitrified glass, forming a intersertal texture (\[\[Fig. X\]\]). A groundmass of this kind is likely to promote regular conchoidal fracture and hence sharp, continuous edges, while the crystalline fraction may give the material greater mechanical strength, and so greater durability in use, than a wholly glassy rock.

The advantage of trachyte therefore lies less in its availability than in what it affords: cobbles large and angular enough to be knapped, in a material whose texture combines predictable fracture with resistance to edge broken. These are the properties on which prolonged resharpening depends, and they are the same properties that the Quina system is understood to require.

## Characterising the Quina retouch products

### Techno-typological features

The `r n_scraper` Quina scrapers from the surface collections are consistent with the Quina technological concept in each of its three dimensions: in the blanks selection, in the retouch strategy, in the ramification.

[Blank selection.]{.underline} Quina flaking yields thick flakes with markedly asymmetric cross-sections, detached by invasive percussion from cores worked on alternating plan-secant surfaces, and it is the volume held in such a blank that permits the successive stages of edge maintenance that follow \[Bourguignon 1997; Turq 1992\]. The blanks used here carry the attributes that this strategy produces. Most scrapers retain cortex to some degree (n = `r cortex_bearing_n`, `r f(cortex_bearing_pct, 1)`%), and where cortex survives it lies most often on a lateral margin (n = `r lateral_n` of `r cortex_bearing_n`, `r f(lateral_pct, 1)`%), opposite the retouched edge, forming a thick back. Some were made directly on cortical flakes, with cortex covering most of the dorsal face (n = `r primary_n`, `r f(primary_pct, 1)`%); among specimens without cortex, a latral thick back usually formed by previous removals (n = `r nocortex_n`, `r f(nocortex_pct, 1)`%). The asymmetric distribution of volume is therefore present whether or not the original surface of the cobble survives. These are not ordinary flakes that happened to be retouched, but blanks carrying a reserve of volume that were selected by knappers.

Platform observations point the same way. In all, `r plat_n` Quina scrapers (`r f(plat_pct, 1)`%) retain the platform of the original blank, and these are characteristically wide and oblique, with a median platform depth of `r f(pd_median, 1)` mm (Q1–Q3 = `r f(pd_q1, 1)`–`r f(pd_q3, 1)` mm) and a mean IPA of `r f(ipa_mean, 0)`° (SD = `r f(ipa_sd, 1)`°); both follow from the invasive percussion by which thick blanks are removed by plan-secant flaking. Plain platforms predominate (n = `r plain_n`, `r f(plain_pct, 1)`%), with cortical (n = `r natural_n`, `r f(natural_pct, 1)`%) and dihedral platforms (n = `r dihedral_n`, `r f(dihedral_pct, 1)`%) less common. Four of the dihedral specimens are *talons lisses à pan*, on which the platform preserves parallel removal scars running approximately orthogonal to the axis of percussion. Few though they are, these carry the most direct information in the assemblage about the reduction strategy behind the blanks: a platform of this kind records a surface that had itself previously served as a flaking surface, which is what the alternating plan-secant exploitation.

[Edge maintenance.]{.underline} The most distinctive feature of a Quina scraper is that the edge was sustained through successive episodes that follow a regular rythum. Retouch usually begins with convex removals and shifts as reduction proceeds towards concave removals, the two alternating across the life of the Quina scraper \[Bourguignon 1997; Peresani et al. 2024\]. That sequence is clearly expressed in this assemblage. Quina scrapers carry a mean of `r f(rg_mean, 2)` retouch generations (SD = `r f(rg_sd, 2)`), with a mean of `r f(cxs_mean, 1)` convex scars (SD = `r f(cxs_sd, 1)`) and `r f(cvs_mean, 1)` concave scars (SD = `r f(cvs_sd, 1)`) per specimen. Convex scars are concentrated in the first generation, concave scars in the second and third. The imbalance between the two, better than four concave removals for every convex one, is the quantitative signature of an edge maintained far beyond its initial production.

Two measures quantify how far that process was carried. The GIUR has a median of `r f(giur_median, 2)` (Q1–Q3 = `r f(giur_q1, 2)`–`r f(giur_q3, 2)`), indicating that retouch commonly extended into the zone of maximum blank thickness; the proportion of the perimeter retouched has a median of `r f(rli_median, 2)` (Q1–Q3 = `r f(rli_q1, 2)`–`r f(rli_q3, 2)`), consistent with the thick back retained on most specimens. A minority were worked more extensively across the whole piece, among them the triple scrapers and *limaces*. Mean edge angle is `r f(ea_mean, 1)`° (SD = `r f(ea_sd, 1)`°). Because the angle produced by the final generation alone cannot be isolated, what we measured is the angle of the edge as a whole, and this is better read as an index of cumulative reduction than as a direct measure of function; edge angle shows significant positive correlation with GIUR (ρ = `r f(ea_giur["rho"], 2)`, p `r fp(ea_giur["p"])`), with the proportion of the perimeter retouched (ρ = `r f(ea_rli["rho"], 2)`, p `r fp(ea_rli["p"])`) and with the number of retouch generations (ρ = `r f(ea_rg["rho"], 2)`, p `r fp(ea_rg["p"])`) (@fig-edge-angle-reduction), which supports that reading.

```{r}
#| label: fig-edge-angle-reduction
#| fig-cap: "Scraper edge angle against the three other reduction measures — proportion of the perimeter retouched, GIUR, and number of retouch generations. Edge angle rises with each, supporting its reading as an index of cumulative reduction. Each panel gives the Spearman rho and its (unadjusted) p; the dashed line is an lm fit shown only as a visual guide to direction. Points on the retouch-generations panel carry a small horizontal jitter for legibility."
#| fig-width: 5.9
#| fig-height: 2.44

# port of scripts/figures/edge_angle_reduction_figure.R
set.seed(2226)
focal_ea    <- "Edge_Angle"
corr_vars   <- c("Retouch_length_index", "Ave_GIUR", "Ave_RG")
corr_labels <- c(Retouch_length_index = "Retouched perimeter",
                 Ave_GIUR = "GIUR", Ave_RG = "Retouch generations")
fmt_p_math  <- function(p) ifelse(p < 0.001, "italic(p) < 0.001",
                                  sprintf("italic(p) == %.3f", p))

quina_ea <- read_excel(sc_path, sheet = "Quina scraper") |>
  mutate(across(all_of(c(focal_ea, corr_vars)), as.numeric))
spearman_one <- function(v) {
  pair <- quina_ea |> select(all_of(c(focal_ea, v))) |> tidyr::drop_na()
  ct <- suppressWarnings(cor.test(pair[[focal_ea]], pair[[v]], method = "spearman", exact = FALSE))
  data.frame(Variable = v, rho = unname(ct$estimate), p_value = ct$p.value)
}
results_ea <- do.call(rbind, lapply(corr_vars, spearman_one))

long_ea <- quina_ea |> select(all_of(c(focal_ea, corr_vars))) |>
  pivot_longer(all_of(corr_vars), names_to = "Variable", values_to = "Value") |>
  filter(!is.na(.data[[focal_ea]]), !is.na(Value)) |>
  mutate(Variable = factor(Variable, levels = corr_vars),
         x_plot = if_else(Variable == "Ave_RG", Value + runif(n(), -0.11, 0.11), Value))
labels_ea <- results_ea |>
  mutate(Variable = factor(Variable, levels = corr_vars),
         label = sprintf("rho == %.2f * ',' ~ %s", rho, fmt_p_math(p_value)))

fig_theme <- theme_minimal(base_size = 9) +
  theme(panel.grid.major = element_line(color = "#E6E8EB", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.5),
        axis.ticks = element_line(color = "#202124", linewidth = 0.3),
        axis.ticks.length = grid::unit(2, "pt"),
        axis.title = element_text(size = 9), axis.text = element_text(color = "#303238", size = 8),
        strip.text = element_text(face = "bold", color = "#202124", size = 8.5),
        strip.background = element_rect(fill = "#E8E8E8", color = NA),
        plot.background = element_rect(color = NA, fill = "white"),
        panel.background = element_rect(color = NA, fill = "white"))

ggplot(long_ea, aes(y = .data[[focal_ea]])) +
  geom_smooth(aes(x = Value), method = "lm", formula = y ~ x, se = TRUE,
              color = "#6BA8CE", fill = "#CBDCEA", linewidth = 0.6, linetype = "dashed") +
  geom_point(aes(x = x_plot), color = "#565693", alpha = 0.5, size = 1.4, shape = 16) +
  geom_text(data = labels_ea, aes(x = -Inf, y = Inf, label = label), parse = TRUE,
            hjust = -0.08, vjust = 1.35, size = 2.5, color = "#202124", inherit.aes = FALSE) +
  facet_wrap(~ Variable, scales = "free_x", ncol = 3, labeller = as_labeller(corr_labels)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.14))) +
  scale_x_continuous(expand = expansion(mult = 0.06)) +
  labs(x = NULL, y = "Edge angle (°)") +
  fig_theme + theme(panel.grid.major = element_blank())
```

Within this common scheme the Quina scrapers vary typologically (\[\[Table X\]\]), and their attributions across Bordes' type list are given in \[\[Table X\]\]. Since every specimen conforms to the Quina retouch, this variation is more plausibly a matter of knappers adjusting to the blanks in front of them than evidence of distinct technological concept [@hiscock2007].

[Ramification sequence.]{.underline} The resharpening flakes (n = `r rf_n`) support the reading of the retouch just described and, at the same time, document the third dimension of the Quina system. They are characteristically flat and oval in plan with an arched profile, and the proximal part of their dorsal face preserves scars of previous Quina retouch; their platforms are shallow (mean PD = `r f(rf_pd_mean, 1)` mm, SD = `r f(rf_pd_sd, 1)` mm), as marginal percussion produces. Their EPAs average `r f(epa_mean, 1)`° (SD = `r f(epa_sd, 1)`°), compared with a mean edge angle of `r f(ea_mean, 1)`° on the Quina scrapers. Although this difference approached significance (Welch's t = `r f(welch_t, 2)`, df = `r f(welch_df, 1)`, p `r fp(welch_p)`), its magnitude was small, with a Cohen's d of `r f(cohd, 2)` (95% CI: `r f(cohd_lo, 2)` to `r f(cohd_hi, 3)`), indicating only a limited separation between Quina scrapers' edge angle and resharpening flakes' EPAs. Since the EPA of a resharpening flake records the angle of the edge from which it was struck, the near identity of the two distributions is what would be expected if these flakes had been detached from Quina scrapers of the kind recovered alongside them; the small residual offset falls within the range that uneven sampling across surface localities would produce, and need not indicate a systematic difference in the angle at which edges were reworked.

Ramification extends beyond the maintenance of edges. Resharpening flakes carry cutting edges of their own and were therefore usable as struck, as use-wear analysis of such flakes at European Quina sites has confirmed [@delpiano2022; @claud2012]. Some Quina scrapers additionally bear "Clactonian notch" removals struck from the edge: broad, invasive scars whose corresponding resharpeninhg flakes are larger, and which are therefore themselves candidates for further modification [@peresani2023]. Material already in hand was thus drawn on repeatedly, first to restore a working edge and then to yield fresh blanks from the tool itself.

### Technological consistency

The surface Quina scrapers cannot be distinguished from the excavated Quina scrapers at Longtan (n = `r ltq_n`), while both are clearly separated from the non-Quina scrapers (n = `r lto_n`) that serve as a control. In the technological space, PERMANOVA finds no difference between the two Quina groups (R² = `r f(tc_scltq_R2, 3)`, pseudo-F = `r f(tc_scltq_F, 2)`, p `r fp(tc_scltq_p)`), whereas each differs from the non-Quina scrapers with an order of magnitude more explained variance (R² = `r f(min(tc_nonq_R2), 2)`–`r f(max(tc_nonq_R2), 2)`, p `r fp(tc_nonq_p)` in both cases) (@fig-technological-consistency).

```{r}
#| label: fig-technological-consistency
#| fig-cap: "Technological consistency of surface Quina scrapers with the excavated Longtan Quina scrapers, against Longtan non-Quina scrapers as a control, on six technical variables. (A) PCA of the z-scored technical space, with convex hulls, group centroids (open circles), variable loadings (arrows), and marginal histograms of the PC1 and PC2 scores by group. (B) Per-variable distributions with pairwise post-hoc brackets (ns / * / ** / *** / ****). The two Quina groups overlap throughout, while both separate from the non-Quina control."
#| fig-width: 5.9
#| fig-height: 8.07

# port of scripts/figures/statistic_figures.R (Fig. TC); reads the `tc` object
# assembled in the results-analysis chunk above.
group_colors <- c(SC_Quina = "#E07C90", LT_Quina = "#E6C25C", LT_Ordinary = "#6BA8CE")
group_labels <- c(SC_Quina = "Surface Quina", LT_Quina = "Longtan Quina",
                  LT_Ordinary = "Longtan non-Quina")
variable_labels <- c(Thickness = "Thickness (mm)", Retouch_length_index = "Retouched perimeter",
                     Ave_GIUR = "GIUR", N_Scar = "Total scars",
                     Ave_RG = "Retouch generations", Edge_Angle = "Edge angle (°)")
variable_labels_bare <- sub(" \\(.*\\)$", "", variable_labels)
names(variable_labels_bare) <- names(variable_labels)

fig_theme <- theme_minimal(base_size = 9) +
  theme(panel.grid.major = element_line(color = "#E6E8EB", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.5),
        axis.ticks = element_line(color = "#202124", linewidth = 0.3),
        axis.ticks.length = grid::unit(2, "pt"),
        axis.title = element_text(size = 9), axis.text = element_text(color = "#303238", size = 8),
        strip.text = element_text(face = "bold", color = "#202124", size = 8.5),
        strip.background = element_rect(fill = "#E8E8E8", color = NA),
        legend.title = element_blank(), legend.text = element_text(size = 8.5),
        legend.key = element_blank(), plot.tag = element_text(face = "bold", size = 11),
        plot.background = element_rect(color = NA, fill = "white"),
        panel.background = element_rect(color = NA, fill = "white"))

scores   <- tc$pca_scores
loadings <- tc$pca_loadings
vexp     <- tc$var_explained
centroids <- scores |> group_by(Group) |> summarise(cx = mean(PC1), cy = mean(PC2), .groups = "drop")
hulls <- scores |> group_by(Group) |> filter(n() >= 3) |>
  group_modify(function(g, k) { h <- chull(g$PC1, g$PC2); bind_rows(g[h, ], g[h[1], ]) }) |> ungroup()
arrow_mult <- 0.85 * min(max(abs(scores$PC1)) / max(abs(loadings$PC1)),
                         max(abs(scores$PC2)) / max(abs(loadings$PC2)))
arrows_df <- loadings |> mutate(x = PC1 * arrow_mult, y = PC2 * arrow_mult,
                                Label = variable_labels_bare[Variable])
EXP_X <- expansion(mult = 0.20); EXP_Y <- expansion(mult = 0.10)

p_a <- ggplot(scores, aes(PC1, PC2, color = Group)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
  geom_polygon(data = hulls, aes(fill = Group, group = Group), alpha = 0.12, color = NA) +
  geom_path(data = hulls, aes(group = Group), linewidth = 0.4, alpha = 0.65) +
  geom_point(size = 1.1, alpha = 0.7, shape = 16) +
  geom_point(data = centroids, aes(cx, cy, fill = Group), shape = 21, color = "#5A5F66",
             size = 2.4, stroke = 0.6, inherit.aes = FALSE) +
  geom_segment(data = arrows_df, aes(x = 0, y = 0, xend = x, yend = y),
               arrow = grid::arrow(length = grid::unit(0.075, "cm"), type = "open", angle = 22),
               linewidth = 0.22, color = "#5A5F66", inherit.aes = FALSE) +
  ggrepel::geom_text_repel(data = arrows_df, aes(x = x, y = y, label = Label),
               size = 2.2, color = "#3A3D42", bg.color = "white", bg.r = 0.12,
               segment.color = "grey55", segment.size = 0.2, min.segment.length = 0.2,
               box.padding = 0.35, point.padding = 0.1, max.overlaps = Inf, seed = 2226,
               inherit.aes = FALSE) +
  scale_color_manual(values = group_colors, labels = group_labels) +
  scale_fill_manual(values = group_colors, labels = group_labels) +
  scale_x_continuous(expand = EXP_X) + scale_y_continuous(expand = EXP_Y) +
  labs(x = sprintf("PC1 (%.1f%%)", vexp[1]), y = sprintf("PC2 (%.1f%%)", vexp[2])) +
  fig_theme +
  theme(panel.grid.major = element_blank(), legend.position = "inside",
        legend.position.inside = c(0.995, 0.995), legend.justification = c(1, 1),
        legend.background = element_rect(fill = alpha("white", 0.85), color = NA),
        legend.margin = margin(2, 4, 2, 2), legend.text = element_text(size = 8),
        legend.key.height = grid::unit(8, "pt")) +
  guides(fill = "none", color = guide_legend(override.aes = list(size = 1.6, alpha = 1), ncol = 1))

marg_theme <- theme_void() +
  theme(plot.background = element_rect(fill = "white", color = NA), plot.margin = margin(0, 0, 0, 0))
p_mx <- ggplot(scores, aes(PC1, after_stat(density), fill = Group, color = Group)) +
  geom_histogram(position = "identity", alpha = 0.35, linewidth = 0.25, bins = 26) +
  scale_fill_manual(values = group_colors) + scale_color_manual(values = group_colors) +
  scale_x_continuous(expand = EXP_X) + scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  marg_theme + guides(fill = "none", color = "none")
p_my <- ggplot(scores, aes(y = PC2, after_stat(density), fill = Group, color = Group)) +
  geom_histogram(position = "identity", alpha = 0.35, linewidth = 0.25, bins = 22) +
  scale_fill_manual(values = group_colors) + scale_color_manual(values = group_colors) +
  scale_y_continuous(expand = EXP_Y) + scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  marg_theme + guides(fill = "none", color = "none")

p_b <- ggplot(tc$variable_long, aes(Group, Value)) +
  geom_jitter(aes(color = Group), width = 0.3, height = 0, size = 0.8, alpha = 0.5, shape = 16) +
  geom_boxplot(color = "black", fill = NA, width = 0.6, linewidth = 0.45, outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 16, size = 1.4, color = "black") +
  ggpubr::stat_pvalue_manual(tc$posthoc_brackets, label = "p.adj.signif", y.position = "y.position",
     tip.length = 0.012, bracket.size = 0.3, label.size = 2.5, color = "#202124") +
  facet_wrap(~ Variable, scales = "free_y", ncol = 3, labeller = as_labeller(variable_labels)) +
  scale_color_manual(values = group_colors, labels = group_labels) +
  scale_x_discrete(labels = group_labels) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) + labs(x = NULL, y = NULL) +
  fig_theme + theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(angle = 20, hjust = 1, size = 7.5)) +
  guides(color = "none")

top_row <- (p_mx + labs(tag = "A") + theme(plot.tag = element_text(face = "bold", size = 11))) +
  plot_spacer() + p_a + p_my +
  plot_layout(design = "AB\nCD", widths = c(5, 1), heights = c(1, 5.5))
top_row / (p_b + labs(tag = "B")) + plot_layout(heights = c(1, 1.5))
```

The control group is what makes this result interpretable. A non-significant difference is not enough by itself evidence of similarity: it may equally reflect variables too blunt to separate anything, or a sample too small to detect what is there. Because the same six variables separate Quina from non-Quina scrapers with relative higher R², they are demonstrably capable of resolving technological differences of the kind at issue here. What this establishes is not that the two groups are identical, which no test can show, but that any difference between them is far smaller than the difference that distinguishes Quina from non-Quina scrapers work.

The univariate results follow the same pattern. None of the six variables differs between the two Quina groups, while every one of them differs between each Quina group and the non-Quina scrapers (@fig-technological-consistency). Agreement therefore holds not only in the multi-dimension technological space but in each of its component measures.

Consistency of this concerns the position of group centroids, and two assemblages could share a centroid while differing in how widely they vary around it. PERMDISP shows that they do not. The mean distance to centroid is almost the same in the two Quina groups (`r f(pd_scq, 2)` and `r f(pd_ltq, 2)`; p `r fp(permdisp_scltq_p)`), whereas the non-Quina scrapers are less dispersed (`r f(pd_lto, 2)`) and differ significantly from both (p `r fp(permdisp_nonq_p)` in each case). Here again the control group establishes that the analysis can detect a real difference in dispersion where one exists, so that the absence of such a difference between the Quina groups is not a matter of sensitivity. The univariate dispersion tests agree: neither the KL-MSLRT nor the Fligner–Killeen tests detect any difference in variability between the two Quina groups (\[\[Table X\]\]).

These results are not driven by the material belongs to Longtan surface collections. Repeating the analyses with the surface specimens collected from within the area of the site removed leaves them unchanged (p `r fp(sens_scltq_p)`, while both remain significantly more dispersed than the non-Quina scrapers at p `r fp(sens_nonq_p)`), so the agreement documented here reflects the surface collections as a whole rather than a single locality or a local peculiarity of sampling.

Taken together, in centroid position and in dispersion alike, the surface Quina scrapers and the excavated Quina scrapers from Longtan cannot be told apart on measurements that readily separate both from other retouched scrapers at the same site. They are most economically understood as the products of a single technological concept.

[Landscape scale.]{.underline} Technological variation is not structured by position in the landscape. Across the `r n_localities` localities that yielded measurable Quina scrapers, no landscape variable accounts for an appreciable share of the variation in the six technological measures: assemblage size (R² = `r f(ls_size["R2"], 3)`, pseudo-F = `r f(ls_size["F"], 3)`, p `r fp(ls_size["p"])`), basin (R² = `r f(ls_basin["R2"], 3)`, pseudo-F = `r f(ls_basin["F"], 3)`, p `r fp(ls_basin["p"])`), distance to the nearest channel (R² = `r f(ls_dist["R2"], 3)`, pseudo-F = `r f(ls_dist["F"], 3)`, p `r fp(ls_dist["p"])`), and height above that channel (R² = `r f(ls_height["R2"], 3)`, pseudo-F = `r f(ls_height["F"], 3)`, p `r fp(ls_height["p"])`). These low R² indicate that landscape variables had little explanatory power for technological variation in Quina scrapers, provide no evidence for a structured spatial influence.

One weak pattern nevertheless runs in a direction worth noting. All six technological measures correlate positively with distance to the nearest channel (Spearman's ρ = `r f(min(ls_dist_rho), 2)`–`r f(max(ls_dist_rho), 2)`; \[\[Fig X\]\]), so that Quina scrapers from localities further from a channel tend to be thicker and more heavily reduced. The direction is the one predicted by work on the European Quina, where prolonged reduction are more pronounced at localities without ready access to abundant raw materials [@turq2000; @hiscock2009]. Two qualifications limit what can be made of this. No individual correlation reaches significance at this sample size, and the six measures are not independent of one another: edge angle covaries with GIUR, with the proportion of retouched edge and with the number of retouch generations (Section 4.2.1; @fig-edge-angle-reduction), so that the concordance of six coefficients does not amount to six independent confirmations. We therefore record the pattern as consistent with an established expectation, and as a hypothesis for landscape-scale work with larger samples, rather than as a demonstrated gradient.

Across the study area, Quina scraper production in essentially the same way, regardless of the basin in which a locality occurs, its elevation above the channel, its distance from water, or the size of the assemblage it has yielded. Such homogeneity under contrasting landscape conditions is not unexpected but is precisely the pattern predicted for a consistently maintained technological system. The absence of landscape structuring should therefore be interpreted not as a lack of detectable variation, but as evidence for the repeated application of a shared technological concept across the region.

# Discussion

The Quina material from these two basins is the product of a single technological concept, which we refer to as the Quina Valley. Three lines of evidence converge on that conclusion. In raw material economy, the surface collections and the excavated assemblage at Longtan alike show a strong preference for trachyte, and that preference holds across the study area even though the gravel bars themselves vary significantly in composition from valley to valley. In technology, the scrapers display the diagnostic attributes of the Quina system: thick, asymmetric blanks with a back formed by cortex or by earlier removals; successive generations of retouch; the rhythmic alternation of convex and concave removals; and the continued production of resharpening flakes, some of which were used or reworked in turn. In their distribution across the landscape, neither technological characteristics nor raw material selection varies systematically with basin, height above the channel, distance to water or assemblage size. What the surface collections record, in other words, is not a scatter of superficially similar pieces but the repeated application of one way of working stone across a defined region.

Longtan is therefore no longer a single case. Its technological attribution has been debated: Bourguignon et al. (2025) observed that morphological resemblance need not imply cultural affinity, suggested that the blank production at Longtan might represent a variant of discoid technology, and expressed reservations about its dating and depositional context; Ruan et al. (2025) responded on techno-economic grounds, on the systematic differences from discoid reduction, and on the chronological and taphonomic evidence. Whatever the outcome of that exchange, a single site is limited in what it can carry. The contribution of the present study is to show that the same way of working stone recurs: across several valleys and several terrace levels in two basins, assemblages appear that agree with Longtan in raw material selection, in blank selection, in the manner of retouch and in the degree to which tools were reduced. These surface collections have no independent age control, and none is claimed for them. What their recurrence establishes is distributional rather than chronological: the Quina system in East Asia is documented at a regional scale, and no longer rests on the evidence of one excavated site.

Quina-style artefacts are not confined to this region. Comparable material has been reported from Dahe in Fuyuan, Yunnan, and Baicaoping in Liangshan, Sichuan, in southwestern China, and from Sanlongdong in Chifeng, Inner Mongolia, and Xinmiaozhuang in Yangyuan, Hebei, in the north \[\]. Most of these assemblages have not been analysed technologically, and whether they represent the Quina system or Quina retouch alone remains open. They are best treated for the moment as leads rather than as established occurrences. The pattern documented here is of a different kind: within the Hengduan Mountains, the Quina system is a recurrent and identifiable regional entity, and the scattered reports elsewhere in China suggest that its distribution in East Asia may prove wider still.

Whether the same technology in two regions implies any connection between the people who made it is a question the evidence cannot settle, and the more productive line is to ask what problem the technology solved. In Europe the answer is reasonably well established. Quina assemblages date largely to MIS 4 and the beginning of MIS 3, in open landscapes with faunas dominated by a single gregarious and migratory species, most often reindeer but equids and bovids in some regions \[Delpech 1996; Delagnes and Rendu 2011; Delpiano et al. 2022\]. The distribution of those animals across the terrain was patchy but not random: prey were partitioned by topography, with reindeer occupying the ground between uplands and plain in a landscape of plateaux dissected by deep valleys, and other species holding lower or more sheltered positions \[Delpech 1996\]. Movement under these conditions was scheduled rather than opportunistic, organised around residential camps and task-specific satellite locations, and the technology has accordingly been read as an answer to two problems at once: the recurrent need to stay equipped while moving away from sources of stone, and the difficulty of anticipating which tasks a given season and location would present \[Hiscock et al. 2009; Delagnes and Rendu 2011\]. The three dimensions of the Quina system answer them in turn. Thick flakes carrying a reserve of volume, struck from cores requiring little preparation, are portable packages of raw material that do not tie a forager to particular outcrops; edge maintenance extends the working life of each piece without recourse to new stone; and ramification allows the same material to yield fresh edges of a different kind as needs change.

The setting on the southeastern margin of the Tibetan Plateau is not the European one, but it posed a problem of the same structure. The regional monsoon record places the occupation at the transition from MIS 4 to MIS 3, when a recovering monsoon was repeatedly interrupted; the pollen record at Longtan shows the local vegetation opening at the same time; and the seasonal aridity characteristic of these dry-hot valleys is likely to have concentrated both water and forage into the valley floors for part of each year. The valleys themselves are deeply incised, with steep altitudinal gradients, so that resources here too were distributed across the terrain in a structured rather than a uniform way. The correspondence with Europe is therefore not one of climate or of raw material, periglacial tundra against dry-hot valley, flint against trachyte, but one of ecological structure: an open landscape in which resources were patchy, seasonally organised and unevenly distributed with respect to topography, requiring frequent and planned movement across it.

A shared problem of this kind provides a mechanism for convergence, and would allow two regions with no contact between them to arrive at technological solutions that are functionally alike. The Quina Valley is on this reading a techno-economic adaptation to a landscape whose resources had to be reached by moving through it. Diffusion cannot be ruled out, however. Quina technology in the western Old World is closely associated with Neanderthals, and the boundaries of their range continue to be redrawn, so the question of origin is better left open at this stage than closed by the argument that is presently most economical.

The Quina Valley bears on a longer-standing question about the East Asian Middle Palaeolithic. The conventional view held that lithic technology in the region was stable across the Palaeolithic and that no phase corresponding to the western European Middle Palaeolithic could be recognised. Levallois and discoid reduction reported at Guanyindong (c. 170–80 ka), Lingjing (c. 125–90 ka) and Jinsitai (c. 47–35 ka) have made that position increasingly difficult to hold \[\], and the Quina system at Longtan added a further technology to the picture. The present study shows that this is not an isolated occurrence but a way of working stone applied repeatedly across a region, and it offers a frame of reference for the assemblages elsewhere in China where Quina-like retouch has been reported but not yet analysed.

Four limitations qualify these results. The surface collections have no independent age control, so the recurrence documented here is a spatial pattern rather than a demonstrated contemporaneity; the shared technological concept, however, holds whether or not the assemblages are of one age. Cores were not analysed and blank production is inferred from the attributes of the blanks themselves, so the first of the three dimensions rests on indirect evidence, while the other two are documented directly and in quantity. No local palaeoclimate record is available for the Binchuan Basin, and the regional setting is inferred from the Heqing sequence, which is treated throughout as context rather than as a local proxy. Most consequentially, no faunal assemblage from the Quina Valley has been analysed, so the mobility and subsistence patterns to which the technology is here related are inferred rather than observed: in the European sequence the argument runs from prey structure through mobility to technology, and here the middle term is untested. Further excavation at Tianhuadong, and the recovery of primary-context assemblages in the Binchuan Basin, would supply the cores, the absolute ages and the faunal record that surface material cannot.

# Conclusions

# CRediT author statement {.unnumbered}

# Declaration of competing interest {.unnumbered}

# Funding {.unnumbered}

# Data availability {.unnumbered}

# Acknowledgements {.unnumbered}

\newpage

# References {.unnumbered}

<!-- The reference list is inserted here for Word/HTML output rather than at the very end. -->

::: {#refs}
:::

\newpage

## Colophon {.unnumbered}

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

if ("devtools" %in% installed.packages()) devtools::session_info()
```
