---
title: "The Middle Paleolithic Quina valleys in southeastern margin of the Tibetan Plateau"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Qijun Ruan:
      correspondence: "yes"
      orcid: 0009-0000-2143-5335
      institute:
        - YPICRA
   - Zhenxiu Jia:
      correspondence: "no"
      orcid: 0000-0002-7514-4514
      institute:
        - ALPHA
   - Marco Peresani:
      correspondence: "no"
      orcid: 0000-0001-6562-6336
      institute:
        - FERRARA
        - IGAG
   - Davide Delpiano:
      correspondence: "yes"
      institute:
        - FERRARA
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
# <!-- TODO: update the author list, ORCIDs and affiliations for THIS paper -->
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - YPICRA: Yunnan Provincial Institute of Cultural Relics and Archaeology, Kunming, China
  - FERRARA: Department of Human Studies, Prehistoric and Anthropological Science Unit, University of Ferrara, Ferrara, Italy
  - IGAG: Consiglio Nazionale delle Ricerche-Institute of Environmental Geology and Geoengineering, Laboratory of Palynology and Palaeoecology, Research Group on Vegetation, Climate and Human Stratigraphy, Milan, Italy
title-block-published: "Last updated"
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
  tbl-title: "Table"
  tbl-prefix: "Table"
number-sections: true
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/elsevier-harvard.csl"
keywords:
  - Middle Paleolithic
  - Southeastern Tibetan Plateau
  - Quina technology
  - Quina valleys
abstract: |
  The Middle Paleolithic (~300–40 ka) was a pivotal period in human evolution, characterized by diverse lithic technologies. Conventional views have held that East Asia experienced slow technological development and lacked remarkable Middle Paleolithic innovations. Recent discoveries, however, demonstrate a more complex scenario, with distinctive Middle Paleolithic technologies identified, such as discoidal technology, Levallois technology and Quina technology. Here we present stone artifacts surface-collected from the Caifeng river and Sangyuan river in northwestern Yunnan province, geographically located in southeastern margin of the Tibetan Plateau. These artifacts mainly collected on the second and third river terraces. Diagnostic Quina technological elements have been identified, including plan-secant core reduction pattern, scalariform tool retouching, and branched reduction strategy. Our results demonstrate the existence of “Quina valleys” and expand the spatial range of Quina technological behavior in southwestern China. In addition, our study reinforces the technological diversity and complexity of the Chinese Middle Paleolithic.
editor_options:
  chunk_output_type: console
---

<!-- This is the format for text comments that will be ignored during rendering. Do not put R code in these comments because it will not be ignored. -->

Keywords: `r rmarkdown::metadata$keywords`

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(here)

```

```{=html}
<!-- Highlights: Elsevier requires 3–5 bullet points, each ≤ 85 characters (incl. spaces).
     These are submitted as a separate file at submission; kept here for drafting. -->
```

# Highlights {.unnumbered}

- 

- 

- 

- 

- 

# Introduction

The Middle Palaeolithic was a period of diversity in both hominin populations and lithic technology, and what drove that technological diversity remains contested: do assemblage differences reflect separate cultural traditions, or adaptive responses to different environments? The question has been pursued most intensively in the Mousterian Debate of western Europe, where the variability of Mousterian assemblages has been attributed to cultural tradition [], site function [], the constraints of locally available raw material [], and differing degrees of artefact reduction []. The Quina Mousterian has been central to that debate.

The Quina Mousterian is one of the best-defined Middle Palaeolithic techno-complexes [], and what distinguishes it is not a tool type but a techno-economic system, the Quina system, whose defining features form three interlocking dimensions []. The first dimension is blank production. Quina cores receive little preparation and are reduced through alternating, concatenated removals on sub-parallel and secant surfaces, with invasive detachments that optimise the volume of stone converted into blanks [Bourguignon 1997; Turq 1989, 1992]. The target product is a thick flake, asymmetric in cross-section, with a sharp cutting edge on one margin and a thick, often cortical, back opposite []. This asymmetry is what allows such a tool to be repeatedly resharpened [Turq 1992; Hiscock et al. 2009]. The second dimension is edge maintenance, achieved through Quina retouch. Here the working edge is not merely produced but sustained, and resharpening follows a regular rhythm []. Early episodes detach relatively large resharpening flakes and leave convex scars, giving a plano-convex edge and preserving volume for later work; as retouch proceeds it concentrates on the margin itself, producing smaller concave scars with stepped or hinged terminations and a plano-concave edge that keeps the tool serviceable []. Each generation of retouch therefore restores a working edge without drawing on new raw material. The third dimension is ramification of the reduction sequence []. Resharpening flakes detached during edge maintenance carry cutting edges of their own and can be used directly, as use-wear analysis has shown [Claud et al. 2012]; in some cases they were themselves retouched into tools []. Quina flakes, besides being shaped into Quina scrapers, are thick enough to serve as cores for small flakes (i.e. core-on-flake) []. Use-wear on scrapers and resharpening flakes from the same assemblages indicates that the two were put to different tasks, scraping and cutting respectively [Delpiano et al. 2022], so that ramification widened the functional range of the toolkit as well as extending its life. Efficient blank production, intensive edge maintenance and ramification together create an economy of stone use in which the thick Quina flake functions as a transportable and highly extendable package of raw material, yielding a large amount of working edge from a small initial investment [Hiscock et al. 2009].

Quina technology has been interpreted as an adaptation to high mobility in open environments where resources were patchily distributed and seasonally structured, and this interpretation rests on three lines of evidence. Faunal assemblages from Quina sites are typically dominated by a single gregarious and migratory species, most often reindeer, although equids and bovids predominate in some regions [Delpech 1996; Delagnes and Rendu 2011; Delpiano et al. 2022; Castel et al. 2017]. Settlement organisation is correspondingly structured, with residential camps surrounded by task-specific satellite locations given over to hunting and carcass processing [Delagnes and Rendu 2011]. At the landscape scale, Quina assemblages occur in open, lightly wooded settings in which prey were distributed across the terrain in a way that was patchy but not random [Delpech 1996; Hiscock et al. 2009]. Foragers under these conditions are argued to have relied on seasonal, high-mobility strategies of resource use and tool provisioning [Hiscock et al. 2009; Delagnes and Rendu 2011; Castel et al. 2017], and it is here that the technological economy described above becomes advantageous: versatile blanks, tools with long use-lives and cores requiring little preparation together lower the time and energy costs of replacing equipment on the move, and reduce the risk both of being left without an effective cutting edge far from a source of raw material and of carrying a toolkit ill-suited to the tasks that a given season and location happened to present [Hiscock et al. 2009; Delagnes and Rendu 2011].

The distinction between the Quina system and Quina retouch alone matters when its distribution is assessed. In the strict sense, with blank production, edge maintenance and ramification operating together, Quina is documented in western and southern Europe within a narrow interval between MIS 4 and early MIS 3, coinciding with pronounced global climatic instability [Delagnes and Rendu 2011; Delpiano et al. 2022]. Quina retouch taken on its own is far more widely distributed: in western Asia it has been reported in Acheulean–Yabrudian industries dating from as early as MIS 11 [], several hundred thousand years before the European Quina. The occurrence of so distinctive a way of maintaining a working edge in contexts this widely separated in time and space raises the possibility that Quina retouch is a recurrent solution to a recurrent problem rather than the signature of a single tradition. Neither the system nor Quina retouch, however, had been securely documented in the eastern Old World.

Whether East Asia has a Middle Palaeolithic in any meaningful sense has long been contested. On the conventional view, lithic technology in the region remained stable throughout the Palaeolithic, with substantial innovation deferred to the Upper Palaeolithic, so that no phase corresponding to the western European Middle Palaeolithic can be recognised []. Recent discoveries have provided new evidence that is reshaping this long-standing understanding. Discoid technology is documented at Lingjing (c. 125–90 ka) [], and Levallois technology at Guanyindong (c. 170–80 ka) [] and Jinsitai (c. 47–35 ka) []. These finds indicate that the technological picture of the East Asian Middle Palaeolithic is far more complex than the conventional view allows. It is against this background that we identified a complete Quina system at Longtan, dated to c. 60–50 ka [], the first Quina site in East Asia to be unambiguously characterised, formally excavated, securely dated and systematically analysed. This discovery not only filled a long-standing gap in the eastern Old World distribution of Quina technology, but also provided new evidence for reassessing the diversity and complexity of the East Asian Middle Palaeolithic. More importantly, Longtan is unlikely to be an isolated case. Since 2010, surface collections bearing Quina features have been recovered from river terraces at a number of localities in the Huangping Basin, where Longtan is situated, and in the neighbouring Binchuan Basin. These finds suggest that Quina technology on the southeastern margin of the Tibetan Plateau may not be confined to a single site, but may instead represent a regional technological phenomenon of some spatial extent.

The southeastern margin of the Tibetan Plateau, where the Huangping and Binchuan basins lie, is the Hengduan Mountains region, an area of pronounced altitudinal gradients, complex landforms and diverse climatic regimes that ranks today among the richest biodiversity hotspots in the world []. Geomorphologically, both basins belong to the dry-hot valley zone of the Jinsha River drainage, a typical fault-controlled valley system []. Under the combined influence of high-relief topography and the monsoon, the region experiences marked seasonal drought and supports a predominantly open vegetation cover []. This setting differs substantially from the cold, open environments of the European Quina sites, and equally from the settings of other Middle Palaeolithic assemblages so far reported in China. It is this regional ecological distinctiveness that makes Longtan and the surrounding Quina localities of particular interest. On the one hand, a technology usually associated with open landscapes, seasonally structured resources and high-mobility subsistence strategies here occurs in a dry-hot valley on the southeastern margin of the Tibetan Plateau, and the relationship between technology and environment in this setting, together with the mechanisms behind it, remains to be examined. On the other hand, all of the localities so far identified are surface collections without absolute dates, and whether they constitute a coherent regional technological entity remains to be tested.

The analysis proceeds in two steps. We ask first whether the surface collections share a consistent technological concept with the Quina technology at Longtan, and can therefore be identified as a single regional technological entity, which we term the Quina Valley. To answer this we systematically analyse the most diagnostic elements of the Quina system, the Quina scrapers and the resharpening flakes removed in maintaining them, and assess their technological consistency with the excavated material from Longtan. Only if the first step can be sustained does the second question arise: whether such an entity can be accounted for by the environmental setting of the region, its available resources and the strategies by which technology was organised. Here we bring together the spatial distribution of the localities, the raw materials used at them and the environmental characteristics of the region. The study clarifies the regional distribution of Quina technology in East Asia and offers a perspective on the relationship between environmental constraint, technological convergence and regional technological organisation in the Middle Palaeolithic.

# Regional Geological and Palaeoenvironmental Setting

The tectonic framework of the study area determines both where its intermontane basins lie and where knappable stone is available within them. The area falls within the Jinsha River drainage in the southern Hengduan Mountains, on the southeastern margin of the Tibetan Plateau, and is controlled by a system of approximately north–south trending faults: continued relative uplift of massifs such as Yulong Snow Mountain and Mount Diancang, alternating with subsidence of the ground between them, has produced a series of north–south trending intermontane basins, among them Lijiang, Heqing, Dali and Binchuan [An et al. 2011; Li and Rioual 2015]. This alternation of uplifted ranges and downfaulted basins, combined with the Indian monsoon and the warm, moist airflow from the southwest, gives the deeply incised valley floors a dry-hot valley climate, sheltered from precipitation by the surrounding relief and marked by a sharp division between wet and dry seasons and by high insolation; the modern vegetation there is dominated by drought-tolerant shrub and savanna formations []. The Quina localities discussed here are concentrated in two of these basins, the Huangping Basin in the southwest of Heqing County and the neighbouring Binchuan Basin, each of them cut by tributaries of the Jinsha River: the Caifeng River in the Huangping Basin, and the Sangyuan River together with its tributary the Liandong River in the Binchuan Basin. The two basins differ in bedrock. Permian volcanic rocks are extensively exposed along the western side of the Huangping Basin, with Triassic sedimentary rocks predominating to the east, whereas the Binchuan Basin is largely mantled by unconsolidated Quaternary sediments and exposes its Permian volcanic basement only locally []. The volcanic sequence comprises basalt, tuff and trachyte []. Because the rivers of both basins cut across these outcrops, volcanic cobbles were carried into the gravel bars of the valley floors, and it was these bars, rather than the outcrops themselves, that supplied the stone used at the localities.

The regional monsoon record places the occupation of these basins within an interval of recurrent aridity rather than within a settled climatic phase. The nearest palaeoclimate record covering this period comes from the lacustrine sediments of the Heqing Basin, about 60 km north of the Huangping and Binchuan basins and some 650 m higher, which is among the longest and most continuous environmental sequences available for the Asian monsoon domain and is highly sensitive to variation in the Indian Summer Monsoon (ISM) [Xiao et al. 2010; An et al. 2011]. Four stages have been recognised within it [Yang et al. 2025]: relatively arid conditions between c. 184 and 131 ka; a shift to generally humid conditions between 131 and 69 ka; a return to aridity between 69 and 57 ka, corresponding to MIS 4; and a generally humid MIS 3 between 57 and 25 ka, punctuated by repeated short-lived arid events. Superimposed on these stages are rapid, millennial-scale fluctuations that correspond closely to North Atlantic cold events [Hou et al. 2026], several of which fall within the period of human occupation documented at Longtan. The record registers monsoon variation at a regional scale rather than local climatic conditions at any one place, and given the distance and altitudinal offset just noted, we treat it here as regional context rather than as a proxy for conditions at the localities themselves. Read at that scale, what matters is not whether a particular stage was on balance wetter or drier, but that the occupation falls at the MIS 4 to MIS 3 transition, when the strengthening of the monsoon did not bring the system to rest: arid episodes continued to recur through the interval in which people were present.

Longtan provides the local counterpart to this regional picture, since it preserves both a secure chronology and a record of vegetation change. Optically stimulated luminescence (OSL) dating places its in situ cultural layer at c. 60–50 ka [], and Longtan remains the only locality in the study area with an age determination of this kind. The nearby Tianhuadong site, where material of this kind has also been reported, has yielded OSL ages of c. 90–40 ka for its cultural layers [], but it has not been systematically analysed, and neither its chronology nor its technological attribution is yet secure enough to carry weight here. The surface collections examined below have no independent age control, and none can be transferred to them by association: what the Longtan dates establish is when the Quina system was present in this landscape, not when any individual surface collection was formed. What Longtan adds beyond a date is a local environmental record. Pollen analysis there documents a transition from coniferous forest to open grassland, with the Quina-bearing layer falling within the phase of vegetation opening [Ruan et al. 2025]. Taken together, the two records indicate that people were present in this landscape while it was becoming more open locally, and while the regional monsoon system continued to be interrupted by arid events. Whether that setting can account for the technology found within it is a question we return to once the technological results have been presented.

# Materials and methods

## 遗址调查与筛选标准

本文的地表材料来自对鹤庆黄坪盆地与宾川盆地的系统踏查，调查分别于2010、2013、2016与2022年四个年度开展。踏查沿两盆地的主要河谷（黄坪的财丰河，宾川的桑园河及其支流炼洞河）展开，系统考察每级河流阶地。调查区域及步行路线的设计旨在覆盖过去人类可能利用的全部地形类型，在复杂且高度异质的景观中兼顾样本代表性与调查工作的可实施性。调查路线通过手机GPS应用2bulu连续记录，拍摄地点和石制品数字照片，并在记录表中填写野外信息，如遗址位置、阶地位置、距河直线距离、石制品数量和类型等。

由于本研究聚焦于基纳技术，我们对踏查记录的地点进行了筛选，仅纳入含有诊断性基纳刮削器和再修理石片的地点。基于这个标准，宾川盆地33处地点中有22处（67%）含基纳材料，黄坪盆地5处地点全部含基纳材料，详细信息如表所示。需要说明，地点的可见性受地表出露条件、植被覆盖与踏查强度影响，因此分布格局仅作趋势性解读，不宜作严格的密度统计。

## 原料调查与分析方法

为评估当地可获得的石料，本研究在3个河道共布设5个砾石采样点，每个采样点面积1-2 m²，收集采样点地表长度大于30 mm的全部砾石进行统计。每件砾石记录以下指标：长度、宽度、厚度（mm），磨圆度（由轻到重分为0–3级）与形状（近椭圆形、板状、不规则）。据此获得各岩性原料在河滩中的相对丰度。为进一步理解原料的内部性质，选取5块粗面岩砾石制作薄片，进行岩相镜检与矿物成分（能谱）分析，由广西壮族自治区地质调查院与中国地质大学（武汉）完成鉴定。

河滩砾石与石制品之间、以及不同空间单元之间的原料组成差异，采用置换多元方差分析（PERMANOVA）检验，并按盆地与河谷两个尺度进行分层分析。

此外，考虑到石制品原料以河滩砾石为主，遗址至河道的最近直线距离被用作接近原料源的代理变量。由于数据存在高度偏态、样本量差异大且方差不齐等特点，因此其在盆地间的差异检验采用 Mann–Whitney检验，在三条河谷间采用 Kruskal–Wallis 检验，以中位数为集中趋势度量，秩检验为主要判据。对于原料的尺寸分析，以每件砾石长、宽、厚的几何平均数为尺寸代理指标，采用 Mann–Whitney U 检验，并报告各岩性的几何平均粒径与效应量 r。

Raw material selection was quantified with Jacobs' electivity index *D* [@jacobs1974], a modification of Ivlev's index that is independent of the relative abundance of the resources:

$$D = \frac{r - p}{\,r + p - 2\,r\,p\,}$$

where $r$ is the proportion of a raw material among the *used* tools and $p$ is its proportion in the *available* basin survey. *D* ranges from $-1$ (complete avoidance) to $+1$ (complete selection), with $0$ indicating use in proportion to availability. *D* was computed for each raw material in the surface Quina-scraper assemblage (SC_Quina) against the basin gravel availability. Materials present in the tools but absent from the basin survey take $D = +1$ and are interpreted as evidence of non-local procurement; materials present in the basin but absent from the tools take $D = -1$.

## Quina修理产品的技术类型分析

基纳刮削器的技术类型学观察与测量遵循基纳技术研究的既有成果 \[Bourguignon 1997；Peresani et al., 2024; Ruan et al., 2025\]。对每件标本记录以下属性：尺寸（长宽厚重）、毛坯类型、石皮位置、台面类型与台面深度、台面内角（IPA）、修理世代数、凸型与凹型修疤数量、GIUR [@kuhn1990] 与刃缘修理比例、刃角，以及 Bordes 类型归属。再修理石片记录台面深度与外台面角（EPA）。统计方法根据变量类型及数据分布特征选择，并依据数据是否近似正态分布采用参数或非参数检验。具体而言，采用 Spearman 秩相关检验（Benjamini校正）分析刃角与 GIUR、刃缘修理比例及平均修理世代之间的关系；再修理石片外台面角（EPA）与刮削器刃角之间的差异采用 Welch t 检验，以避免方差不齐对推断结果的影响。

为检验地表基纳刮削器与龙潭发掘品之间的技术一致性，以龙潭遗址出土基纳刮削器为参照，并引入同一技术背景下的普通刮削器（ordinary scrapers）作为技术对照组。从技术变量中选取 6 个最具代表性的指标（厚度、刃缘修理比例、平均 GIUR、修疤数、平均修理世代和刃角）开展多变量分析；所有变量经 z 标准化后，以欧氏距离计算样本间差异。组间技术差异采用 PERMANOVA 检验，两两比较报告解释变异比例（R²）。普通刮削器作为技术对照组，用于建立解释变异比例的参考尺度：若同一变量能够以较高 R² 区分基纳刮削器与普通刮削器，而不同基纳组之间仅表现出较低 R²，则后者更可能反映真实的技术相似性，而非统计检验功效不足。PERMANOVA 反映的是多变量空间中组质心（centroid）的差异。单变量层面，根据数据结构采用不同统计方法：平均 GIUR、修疤数和平均修理世代采用 Kruskal–Wallis 检验及 Dunn 事后比较；厚度、刃缘修理比例和刃角采用 Welch ANOVA 及 Welch t 事后比较；所有事后比较均采用 Bonferroni 校正以控制假阳性率。组间技术离散度采用 PERMDISP 检验。为比较组内变异程度，长度、宽度、厚度、重量及刃角采用 Krishnamoorthy–Lee Modified Signed Likelihood Ratio Test（KL-MSLRT）检验各组变异系数（CV）是否一致，因为这些指标均属于具有真实零点的比率尺度变量。GIUR、刃缘修理比例、修疤数及修理世代则采用 Fligner–Killeen 检验比较组间离散程度。其中，GIUR 与刃缘修理比例限定于 \[0,1\] 区间，变异系数缺乏可解释性；修疤数与修理世代属于计数变量，其方差结构亦不满足 CV 比较的基本前提。为评估结果的稳健性，剔除地表采集品中来自龙潭遗址范围的样本后重复上述分析。

为检验技术变异是否受到景观结构的系统性影响，以遗址为分析单元，将每个遗址基纳刮削器在上述 6 个技术变量上的中位数作为遗址层面的代表值。进一步考察 4 个景观变量——盆地归属（鹤庆—黄坪盆地与宾川盆地）、地貌位置（阶地层级）、遗址至最近河道距离及遗址规模（标本数量）——对技术变异的影响。分类变量（盆地归属和地貌位置）采用 PERMANOVA 与 PERMDISP 检验；连续变量（距离和规模）采用 Spearman 秩相关（以各遗址标本数量作为权重计算加权秩相关）及基于欧氏距离的遗址级 PERMANOVA 分析，遗址规模同时采用分档分析检验结果稳健性。由于遗址层面的分析单元数量有限（26 个遗址，且多数仅包含 1–2 件标本，遗址中位数稳定性较低），同时盆地归属与地貌位置存在较强共线性，因此相关检验的统计功效受到限制，结果主要依据效应量（R²、ρ）的大小及方向进行解释，显著性检验结果仅作为辅助证据。

上述所有统计分析均在 R 4.5.2 中完成。PERMANOVA 与 PERMDISP 使用 vegan 包，Welch ANOVA 使用 onewaytests 包，KL-MSLRT 使用 cvequality 包，Fligner–Killeen 检验采用 stats 包实现。

## Reproducibility and open materials

Computational reproducibility remains an ongoing challenge in archaeological science, yet it is essential for the credibility and cumulative progress of quantitative research [@marwickComputationalReproducibilityArchaeological2017; @marwickOpenScienceArchaeology2017]. To support reproducibility and re-use, all materials necessary to reproduce the analyses presented here are provided in version-controlled online repository at OSF. These materials include the complete R and Python scripts used for all analyses and visualizations, and the raw datasets underlying all reported results. The repositories are organized following established best practices for research compendia, with a logical folder structure and relative file paths that allow all code to be executed without manual modification. All figures, tables, and statistical results presented in this paper can therefore be independently reproduced from these materials.

# Results

## Quina遗址的景观分布格局与原料经济

Quina遗址在两个盆地内主要沿河谷系统分布。在鹤庆黄坪盆地，遗址集中分布于河流第二、三级阶地；而在宾川盆地，遗址分布范围进一步扩展至第二至第四级阶地，少量遗址位于更高的山体顶部。从遗址与现代河道之间的直线距离来看，桑园河、炼洞河和财丰河三个河谷系统表现出较为一致的空间格局，其中位距离为596 m。采用Mann–Whitney U检验和Kruskal–Wallis检验对不同空间单元内遗址—河流距离进行比较，结果均未发现显著差异（basin: p = 0.262；valley: p = 0.124）。鉴于石制品原料主要来源于河流搬运砾石，遗址与河流之间的距离可作为评估原料获取便利性的空间代理指标。因此，该结果表明，不同河谷系统内遗址对河流资源的空间接近程度总体相似，暗示其原料获取环境可能具有相近的可达性。

然而，原料利用模式并未简单反映自然资源的丰度分布。石制品组合显示，粗面岩（trachyte）是绝对优势原料，占全部石制品的95.8%。相比之下，河滩砾石调查表明，砂岩为自然环境中最丰富的原料类型，粗面岩仅次之，且不同采样点之间的原料组成存在显著空间差异（PERMANOVA: R² = 10.8%, p = 0.001）。进一步的分层PERMANOVA分析显示，该差异主要由不同河谷系统之间的变化所驱动（p = 0.001）。这一结果表明，河流砾石资源本身具有明显的空间异质性，与区域地质背景所反映的原料分布模式相一致。

与自然资源分布形成鲜明对比的是，石制品组合中的原料构成表现出高度的空间稳定性。无论按照盆地划分（R² = 0.004, p = 0.54），还是按照河谷划分（R² = 0.005, p = 0.75），Quina刮削器的原料组成均未表现出显著空间差异。这说明尽管砂岩在河滩环境中具有更高的丰度，且不同河谷之间存在明显的原料组成差异，古人类仍表现出一致性的原料选择行为，即优先利用粗面岩作为主要加工材料。Jacobs选择性指数进一步支持这一判断：粗面岩表现出显著的正向选择（D = 0.97），而砂岩则表现出显著的负向选择（D = −0.98）。

为了探讨这一选择偏好的形成机制，有必要进一步考察不同原料自身的物理属性差异。首先，在尺寸方面，粗面岩砾石显著大于砂岩砾石（U = 16864.00, p \< 0.001），为大型石核制备和持续剥片提供了更大的原料基础。其次，在形态方面，粗面岩砾石主要呈板状（n = 34, 26.4%）和不规则状（n = 39, 30.2%），相较于尺寸较小且形态更规则的砂岩砾石，更适合作为剥片原料使用。进一步的岩石学分析显示，粗面岩具有典型的斑状结构，由斑晶和基质共同组成。斑晶主要为自形板柱状钾长石（粒径约0.3–5.0 mm，占比约1%–10%），基质则由钾长石微晶、绿泥石或蚀变暗色矿物以及磁铁矿组成，并与约50%的脱玻化玻璃质共同形成玻晶交织结构（vitrophyric texture）。该结构可能促进粗面岩形成较为规则的贝壳状断裂，从而产生锋利且连续的刃缘；同时，相较于典型高玻璃质岩石（如黑曜岩），其较高比例的晶质矿物成分可能赋予材料更高的机械强度，提高工具使用过程中的耐久性。

因此，粗面岩的优势并非仅来源于其在河流环境中的可获得性，而更可能与其综合技术性能有关：较大的原料尺寸和适宜的砾石形态为石核加工提供了良好条件，而独特的岩石结构则兼具良好的断裂性能和较高的机械强度。这些属性共同构成了古人类长期优先选择粗面岩进行Quina石器生产的重要物质基础。

## 表征Quina刮削器

### 技术类型特征

本文纳入分析的166件刮削器，在毛坯选择、刃缘修理以及分支化利用三个技术维度上均表现出与Quina技术概念一致的特征。

毛坯生产。 Quina技术通常通过交替剥片（plan-secant）的石核剥片策略，沿石核边缘实施侵入式打击，以获得厚重且具有明显不对称性的石片毛坯。这类厚度较大的非对称毛坯为后续多阶段刃缘修理提供了充足的体积储备，是Quina生产体系区别于其他中旧石器技术的重要特征之一。研究材料中，大多数刮削器均保留不同程度的石皮，其中石皮通常位于加工刃缘的相对侧，形成明显的厚“背”（back）结构。部分标本直接以带石皮石片（cortical flake）作为毛坯，石皮覆盖整个背面；即使在缺乏原始石皮的标本中，部分毛坯仍保留由此前剥片形成的“背”，整体呈现出明显的不对称体积结构。这表明，研究材料中的刮削器毛坯并非简单利用普通石片，而是具有符合Quina技术预期的厚重和不对称特征。此外，部分刮削器仍保留原始石片台面（n = 52, 31.3%），这些台面多表现为宽阔且倾斜的类型，其台面深度中位数达到13.9 mm，平均平台内角（interior platform angle, IPA）为122°。台面类型以素台面（plain platform）为主（n = 39, 75.0%），自然台面（natural platform）（n = 7, 13.5%）和dihedral台面（dihedral platform）（n = 6, 11.5%）相对较少。值得注意的是，二面台面标本中有4件属于talon à pan类型，其台面上保留与打击轴近似正交的平行剥片痕迹。

刃缘修理。 Quina刮削器最具代表性的技术特征在于其独特的连续修理策略。为了维持厚重刃缘和整体体积结构，Quina修理通常表现为具有阶段性的修理节奏：早期修理以凸型剥离（convex removals）为主，而随着修理推进，逐渐转变为凹型剥离（concave removals），循环往复。这种由凸向凹的修理序列被认为是维持刃缘体积并延长工具使用寿命的重要机制。研究材料中，该修理模式表现十分明显。刮削器平均修理世代为2.76，其中凸型修疤平均数量为4.4个，而凹型修疤平均数量达到18.7个。凸型修疤主要集中于第一修理世代，而凹型修疤则普遍出现于第二和第三修理阶段。部分标本完全由凹型修疤构成，但未观察到完全由凸型修疤组成的案例。这一分布模式表明，研究材料中的刃缘重塑并非随机发生，而是遵循由早期凸型修理向后期凹型修理转变的技术序列。尽管所有标本均符合Quina修理概念，刮削器内部仍存在一定类型学差异（表\[x\]）。这种差异在Bordes类型学分类中有所体现，可能反映了打制者针对不同毛坯条件所采取的技术调整，而非不同技术传统之间的差异。为量化刃缘再加工程度，本文采用GIUR和刃缘修理比例两个指标进行评估。结果显示，GIUR中位值为0.77，表明修理通常深入至毛坯最大厚度区域；刃缘修理比例中位值为0.46，与多数标本保留明显“背”的结构相一致。然而，部分标本表现出更高程度的整体加工，其中包括triple scraper和limace类型。研究材料中的平均刃角为70.0°，该指标更可能反映整体加工强度，而非直接对应工具功能。由于最后一次修理世代的刃角难以独立识别，因此本文测量的是整体刃角。相关分析显示，刃角与GIUR、刃缘修理比例以及修理世代均呈显著单调正相关，进一步支持其主要反映加工程度变化的解释。

再修理石片与分支化。 再修理石片既进一步支持了基纳修理概念,也体现了分支化(ramification)这一基纳体系的另一维度。这类石片通常扁圆、侧廓呈凸起的拱形,背面近端保留有先前的基纳修理痕迹;其台面深度很小(均值 5.3 mm),显示边缘打击的特点。再修理石片的 EPA 平均为 72.8°,与刮削器刃角(70.0°)之间存在一个微小的偏移，这一偏移很可能与地表遗址的不均匀采样有关，而不一定代表实际修理角度的系统差异。再修理石片本身具有锋利刃缘、具备使用潜力。此外,部分刮削器还表现出"克拉克当凹口(Clactonian notch)"式的再修理:这种方式会在刃缘形成较大的凹型修疤,对应的再修理石片也尺寸较大、带有台面,因而具备被进一步加工成工具的潜力。

### 技术一致性分析

为评估地表采集材料是否保持稳定的技术特征，本文以具有明确地层关系和可靠年代控制的龙潭遗址作为参照基准，选取其发掘出土的Quina刮削器作为对照样本，并引入普通刮削器（ordinary scrapers）作为技术差异参照。考虑到高维变量可能引入信息冗余，本文选择6个最具代表性的技术指标进行多变量分析。结果显示，地表采集Quina刮削器与龙潭出土Quina刮削器在整体技术空间中不存在显著差异，而两组Quina刮削器均与普通刮削器表现出显著分离。这一结果具有重要的方法学意义：由于所选变量组合能够有效区分Quina刮削器与普通刮削器（两两比较效应量R² = 0.29–0.36），因此Quina组之间较低的组间差异（R² = 0.004）并非源于变量组合缺乏区分能力，而更可能反映两类Quina材料之间真实的技术相似性。逐变量分析进一步支持这一结果。在单变量尺度上，两个Quina样本组之间均未检测到显著差异，而二者与普通刮削器之间则在全部指标上均表现出显著差异。这表明，地表采集材料与发掘材料之间不仅在综合技术空间中不可区分，在具体技术属性上也表现出高度一致性。上述结果主要反映群组中心位置（centroid）的差异。为进一步评估技术变异范围是否一致，本文进行了PERMDISP分析。结果显示，地表Quina组与龙潭Quina组具有几乎相同的平均质心距离（分别为2.10和2.02，比值约为1.04），且两者之间不存在显著差异（p = 0.53）。相比之下，普通刮削器表现出更低的技术离散度（平均质心距离=1.55），并与两组Quina刮削器均存在显著差异（p均=0.001）。这一结果表明，分析方法能够有效识别真实存在的离散度差异，因此Quina两组之间缺乏显著差异并非由统计灵敏度不足造成。单变量离散度检验进一步验证了这一趋势。KL-MSLRT检验和Fligner–Killeen检验均未发现两个Quina样本组之间存在显著方差差异。为进一步检验结果的稳健性，本文剔除了地表采集样本中来自龙潭遗址附近的材料，并重新进行分析，结果保持一致。这说明观察到的技术一致性并非由单一遗址或局部样本结构驱动，而反映了更广泛的技术稳定性。综合中心位置和离散度两个层面的证据，地表采集Quina刮削器与龙潭发掘Quina刮削器在技术空间中均无法被有效区分，表明二者共享相同的技术逻辑。

遗址-景观层面。进一步分析旨在考察Quina技术变异是否受到景观结构因素的系统性影响。本文选取四个可能影响技术组织的景观变量，包括遗址至河流距离、地貌位置、遗址规模以及盆地归属（黄坪盆地与宾川盆地），并分别采用PERMANOVA评估其对技术变异的解释能力。结果显示，各景观变量对技术差异的解释度均较低。其中，遗址规模（连续变量，R² = 0.006）和盆地归属（R² = 0.011）几乎不贡献技术变异解释，距水距离（R² = 0.031）和地貌位置（R² = 0.073）的解释度相对较高，但仍处于较低水平，且所有变量在置换检验中均未达到统计显著水平。值得注意的是，尽管距水距离对整体技术变异的解释能力有限，逐变量相关分析显示，六个技术指标均与距水距离呈弱正相关关系（Spearman ρ = 0.14–0.37）。这一方向一致性表明，随着遗址距离水源增加，Quina刮削器可能倾向于表现出更高的厚度和更深程度的加工。该趋势与移动过程中携带工具并在远离原料补给区域持续维护、减薄工具的理论预期相一致。然而，由于单个相关系数均未达到显著水平，目前尚无法确认这一趋势是否代表稳定的空间梯度，因此本文仅将其视为未来景观尺度研究中值得检验的行为假设，而非已确定的空间组织规律。总体而言，在当前样本规模和空间分辨率下，尚未发现显著控制Quina技术变异的景观因素。研究区内Quina技术表现出较高程度的跨地点稳定性，其技术组织逻辑并未明显受到遗址规模、盆地归属或局部景观位置的限制。

# Discussion

本研究显示，研究区内存在一套高度一致的基纳技术体系，可归纳为一个区域性技术概念（Quina Valley）。这一结论在原料选择、技术特征与景观分布三个层面均得到支持。在原料经济方面，无论地表采集材料还是龙潭遗址出土标本，均表现出对粗面岩的强烈选择偏好。值得注意的是，当地河滩砾石的原料组成具有显著空间异质性，但这种自然变异并未反映在石制品原料选择中，各遗址均一致偏向粗面岩。在技术层面，研究区基纳刮削器系统性地表现出典型诊断特征，包括厚重且不对称的毛坯结构、由石皮或早期剥片构成的“back”、多世代连续修理，以及凸型与凹型修疤的节奏性交替，同时伴随再修理石片的持续生产。这一组合与欧洲典型Quina技术定义高度一致。在空间尺度上，技术特征与原料选择均未随盆地、地貌位置、距水距离或遗址规模发生系统变化，表明该技术体系具有较强的空间稳定性。综合上述证据，研究区内的基纳材料指向同一技术概念，构成一个可识别的区域性技术实体，即Quina Valley。

这一区域性现象的意义，还在于为龙潭这一长期被讨论的个案提供了新的区域性支撑。作为东亚首个正式发表的基纳遗址，龙潭的技术归属曾引发讨论。Bourguignon et al. (2025)指出，形态相似性并不必然意味着文化亲缘关系，并提出龙潭的剥坯策略可能代表盘状技术的一种变体，同时对其年代与埋藏背景持审慎态度；Ruan et al. (2025)则从技术—经济特征、与盘状技术的系统性差异以及年代与埋藏学证据等方面进行了回应。无论上述争论如何展开，一个孤立遗址在解释能力上始终存在局限。本文的关键贡献在于表明，龙潭并非孤立现象。在鹤庆与宾川两个盆地的多条河谷及多级阶地中，反复出现一套在原料选择、毛坯生产、加工方式与加工程度上均高度一致的基纳组合，这些材料与龙潭共享同一技术概念。尽管这些地表材料缺乏独立的绝对年代约束，但其在区域尺度上的重复出现，从分布意义上为东亚基纳现象提供了独立于单一遗址之外的证据支撑。需要指出的是，基纳风格的石制品并非仅限于本研究区。在更广泛的地理范围内，已有若干遗址报道过类似材料。例如，在中国西南地区有云南富源大河遗址、四川凉山白草坪遗址，在中国北方地区有内蒙古赤峰三龙洞与河北阳原新庙庄等遗址，这些遗址均报道了具有基纳特征的石制品。然而，这些遗址大多尚未开展系统的技术学分析，其组合的技术属性仍有待进一步验证。因此，目前这些材料更适合作为潜在证据线索，而非已确立的技术实体。综合来看，以龙潭为代表的东亚基纳现象，在横断山脉地区已表现为一个可重复识别的区域性技术实体，指向一套可与欧洲基纳体系相比较的技术行为模式。同时，散见于中国其他地区的相关材料，则提示该技术现象在东亚的空间分布可能更为广泛，但其技术一致性仍需未来研究进一步检验。

尽管共享相似的技术行为,目前仍缺乏足够的证据将其与特定人群相联系;因此,现阶段更适宜从环境适应的角度来理解 Quina Valleys。欧洲的基纳遗址已有充分研究,其年代主要集中在 MIS 4 至 MIS 3 早期,对应干冷气候、开阔苔原,以及以驯鹿为主的季节性迁徙动物群。在这样的环境中,觅食者在不同季节、不同地点会遭遇不同的猎物,资源在时空上难以预测;正因如此,欧洲基纳被理解为对这种不确定性的一种应对，这种应对依赖灵活、可反复维护、使用寿命长的工具(Hiscock et al., 2009)。东亚的情况在环境背景上与欧洲存在差异,却面对着同一类适应问题。区域古气候记录显示，该时期整体处于季风回升阶段，但频繁受到千年尺度干旱事件干扰；同时，龙潭孢粉记录表明局地植被正向开阔草地转变。这种区域—局地双重不稳定性，共同构成了一个资源时空分布不确定的环境背景。在这一意义上，欧洲与东亚之间的关键相似性并不在于具体环境类型（冰缘苔原 vs 干热河谷），亦不在于原料类别（燧石 vs 粗面岩），而在于共享的生态结构：开阔环境条件下的高资源波动性与不可预测性。这一共享选择压力为趋同演化提供了机制基础，使得在地理隔离的情况下，两个区域分别演化出功能上相似的技术解决方案成为可能。因此，Quina Valley可被理解为一种在不稳定资源环境中形成的技术—经济适应模式。但同时，目前也无法完全排除技术传播的可能性。尤其考虑到Quina技术在旧大陆西侧与尼安德特人密切相关，其扩散边界仍在不断被重新界定。因此，在当前阶段，对其起源机制应保持开放解释框架。

Quina Valleys的提出，也为理解中国旧石器时代中期技术的多样性与复杂性提供了新的视角。东亚是否存在真正意义上的旧石器时代中期长期存在争议，传统观点强调该地区石器技术在整个旧石器时代的长期稳定性，并据此否认存在可与西欧相对应的中期阶段。近年来，观音洞（约170–80 ka）、灵井（约125–90 ka）以及金斯太（约50 ka）等遗址所报道的勒瓦娄哇、盘状石核等技术，已逐步揭示出这一时期远较传统认识复杂的技术面貌，龙潭遗址基纳体系的确认更为这一复杂格局增添了新的技术类型。然而，此前经过系统技术分析的仅龙潭一处。本研究的意义正在于此：Quina Valleys在区域层面为龙潭的基纳技术提供了独立的复现证据，表明这一技术并非孤立个案，而是在青藏高原东南缘确有存在。与此同时，这一区域性证据也为其他已报道基纳特征、但尚待系统分析的遗址（如云南富源大河遗址）提供了潜在支持。

本研究亦存在若干局限。其一，研究材料主要来源于地表遗址，缺乏可靠的绝对年代控制，目前仅能借助龙潭与天华洞遗址的测年结果提供大致的时间约束。其二，本研究未纳入石核分析，基纳剥坯策略的重建主要依赖刮削器毛坯属性进行间接推断。其三，宾川盆地目前缺乏本地古气候记录，相关环境背景主要参考邻近鹤庆盆地的数据作为区域性替代。其四，Quina Valleys尚缺乏系统的动物考古研究，因此在资源利用与生计结构方面仍缺少关键证据维度。尽管天华洞遗址已有动物化石发现，但相关材料尚未形成可用于系统分析的数据基础。上述限制表明，本研究在年代学约束、技术链完整性及多代理环境重建方面仍存在不确定性，有待未来通过天华洞遗址的进一步发掘，以及在宾川盆地寻找原生地层材料与开展系统动物考古与环境研究加以完善。

# Conclusions

# CRediT author statement {.unnumbered}

# Declaration of competing interest {.unnumbered}

# Funding {.unnumbered}

# Data availability {.unnumbered}

# Acknowledgements {.unnumbered}

\newpage

# References {.unnumbered}

<!-- The reference list is inserted here for Word/HTML output rather than at the very end. -->

::: {#refs}
:::

\newpage

## Colophon {.unnumbered}

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

if ("devtools" %in% installed.packages()) devtools::session_info()
```
