---
title: "Supplementary material"
subtitle: "The Quina Landscape: a regional Middle Palaeolithic technological system on the southeastern margin of the Tibetan Plateau"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Qijun Ruan:
      correspondence: "yes"
      orcid: 0009-0000-2143-5335
      institute:
        - YPICRA
  - Davide Delpiano:
      correspondence: "no"
      institute:
        - FERRARA
  - Marco Peresani:
      correspondence: "no"
      orcid: 0000-0001-6562-6336
      institute:
        - FERRARA
        - IGAG
  - Zhenxiu Jia:
      correspondence: "no"
      orcid: 0000-0002-7514-4514
      institute:
        - ALPHA
  - Lijing Yang:
      correspondence: "no"
      institute:
        - BCRMI
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
  - Hao Li:
      correspondence: "yes"
      email: lihao@itpcas.ac.cn
      institute:
        - ALPHA
# <!-- TODO: update the author list, ORCIDs and affiliations for THIS paper -->
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - YPICRA: Yunnan Provincial Institute of Cultural Relics and Archaeology, Kunming, Yunnan, China
  - FERRARA: Department of Human Studies, Prehistoric and Anthropological Science Unit, University of Ferrara, Ferrara, Italy
  - IGAG: Consiglio Nazionale delle Ricerche-Institute of Environmental Geology and Geoengineering, Laboratory of Palynology and Palaeoecology, Research Group on Vegetation, Climate and Human Stratigraphy, Milan, Italy
  - BCRMI: Binchuan Cultural Relics Management Institute, Dali Bai Autonomous Prefecture, Yunnan, China
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
crossref:
  fig-title: "Fig. S"
  fig-prefix: "Fig. S"
  tbl-title: "Table S"
  tbl-prefix: "Table S"
# packages.bib is written by the setup chunk from the installed packages
bibliography:
  - references.bib
  - packages.bib
csl: "../templates/elsevier-harvard.csl"
number-sections: false
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
  - quarto
  # runs after the crossref filter, so that "Table S 1" reads "Table S1"
  - ../templates/supplement-numbering.lua
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

source(here::here("paper", "_analysis.R"))
fpt <- function(p) sub("^= ", "", vapply(p, fp, character(1)))
group_colors <- c(SC_Quina = "#E07C90", LT_Quina = "#E6C25C", LT_Ordinary = "#6BA8CE")
group_labels <- c(SC_Quina = "Surface Quina", LT_Quina = "Longtan Quina",
                  LT_Ordinary = "Longtan non-Quina")
basin_colors   <- c(Binchuan = "#6BA8CE", Heqing = "#E07C90")
grad_low       <- "#6BA8CE"
grad_high      <- "#E07C90"

fig_theme <- theme_minimal(base_size = 9) +
  theme(panel.grid.major = element_line(color = "#E6E8EB", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.5),
        axis.ticks = element_line(color = "#202124", linewidth = 0.3),
        axis.ticks.length = unit(2, "pt"),
        axis.title = element_text(size = 9), axis.text = element_text(color = "#303238", size = 8),
        strip.text = element_text(face = "bold", color = "#202124", size = 8),
        strip.background = element_rect(fill = "#E8E8E8", color = NA),
        legend.title = element_blank(), legend.text = element_text(size = 8),
        legend.key = element_blank(), plot.tag = element_text(face = "bold", size = 11),
        plot.background = element_rect(color = NA, fill = "white"),
        panel.background = element_rect(color = NA, fill = "white"))

si_pkg_tbl <- tribble(
  ~Package,     ~Citation,           ~Purpose,
  "vegan",      "[@R-vegan]",        "Distance matrices, PERMANOVA, PERMDISP and redundancy analysis",
  "rstatix",    "[@R-rstatix]",      "Rank and Welch tests, post hoc comparisons and effect sizes",
  "cvequality", "[@R-cvequality]",   "Krishnamoorthy–Lee test for equality of coefficients of variation",
  "MASS",       "[@MASS2002]",       "Kernel density for the cobble form contours",
  "readxl",     "[@R-readxl]",       "Reading the recording sheets",
  "dplyr",      "[@R-dplyr]",        "Data manipulation",
  "tidyr",      "[@R-tidyr]",        "Reshaping between wide and long form",
  "here",       "[@R-here]",         "Project-relative file paths",
  "ggplot2",    "[@ggplot22016]",    "Graphics",
  "ggpubr",     "[@R-ggpubr]",       "Significance brackets on the distribution panels",
  "ggrepel",    "[@R-ggrepel]",      "Labels placed clear of the points they name",
  "patchwork",  "[@R-patchwork]",    "Composition of the multi-panel figures",
  "scales",     "[@R-scales]",       "Axis and colour-scale transformation",
  "knitr",      "[@R-knitr]",        "Tables and document rendering") |>
  mutate(Version = vapply(Package, function(p) as.character(packageVersion(p)),
                          character(1)), .after = Package)

si_bib <- here("paper", "packages.bib")
knitr::write_bib(c("base", si_pkg_tbl$Package), file = si_bib)
si_lines <- readLines(si_bib)
si_ttl   <- grepl("^  title = ", si_lines)
si_lines[si_ttl] <- sub("^(  title = \\{)([^{:]+)(:)", "\\1{\\2}\\3",
                        si_lines[si_ttl])
si_lines[si_ttl] <- gsub(" ([A-Z])([ ,}])", " {\\1}\\2", si_lines[si_ttl])
writeLines(si_lines, si_bib)
n_fmt <- function(x) format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
```

```{r}
#| label: si-analysis
#| include: false

# quantities needed only in this document, built from the objects _analysis.R leaves

# ---- 1. PERMDISP alongside each raw-material PERMANOVA --------------------
permdisp_one <- function(D, g) {
  g <- droplevels(factor(g))
  set.seed(2226); bd <- suppressWarnings(betadisper(D, g))
  set.seed(2226); pt <- permutest(bd, permutations = PERM)
  c(F = pt$tab$F[1], p = pt$tab$`Pr(>F)`[1])
}
pd_cobbleLoc   <- permdisp_one(C_D, cobbles$Loc)
pd_cobbleRiver <- permdisp_one(C_D, cobbles$river_ID)
pd_usedA_basin <- permdisp_one(A_D, A_meta$basin)
pd_usedA_river <- permdisp_one(A_D, A_meta$river_ID)

# ---- 2. Welch post-hoc statistics ----------------------------------------
we_t <- variable_long |> filter(Variable %in% welch_vars) |>
  mutate(Variable = droplevels(Variable)) |> group_by(Variable) |>
  t_test(Value ~ Group, var.equal = FALSE, p.adjust.method = "bonferroni") |> ungroup()
we_ph_full <- we_ph |> select(Variable, group1, group2, p.adj, p.adj.signif) |>
  left_join(we_t |> select(Variable, group1, group2, statistic, df, p.adj.t = p.adj),
            by = c("Variable", "group1", "group2"))
stopifnot(isTRUE(all.equal(we_ph_full$p.adj, we_ph_full$p.adj.t)))

# ---- 3. Landscape: dispersion between the basins --------------------------
ls_basin_pd <- permdisp_one(dL, site_df$Basin)

# ---- 4. Each measure against each continuous landscape variable -----------
land_cor <- bind_rows(lapply(names(land_vars), function(lv) {
  fam <- bind_rows(lapply(land_variables, function(v) {
    d <- site_df[is.finite(site_df[[v]]) & is.finite(site_df[[lv]]), ]
    ct <- suppressWarnings(cor.test(d[[v]], d[[lv]], method = "spearman", exact = FALSE))
    data.frame(Landscape = lv, variable = v, n = nrow(d),
               rho = unname(ct$estimate), p = ct$p.value)
  }))
  fam$p_adj <- p.adjust(fam$p, "bonferroni")
  fam
}))
stopifnot(isTRUE(all.equal(land_cor$rho[land_cor$Landscape == "Distance_to_water"],
                           ls_dist_cor$rho)),
          isTRUE(all.equal(land_cor$p_adj[land_cor$Landscape == "Distance_to_water"],
                           ls_dist_cor$p_adj)))

# ---- 5. Each measure between the two basins -------------------------------
basin_cmp <- bind_rows(lapply(land_variables, function(v) {
  d <- bdf[is.finite(bdf[[v]]), c("Basin", v)]
  names(d) <- c("Basin", "Value")
  d$Basin <- droplevels(d$Basin)
  wt <- wilcox_test(d, Value ~ Basin)
  data.frame(variable = v, p = wt$p,
             r = suppressWarnings(wilcox_effsize(d, Value ~ Basin))$effsize[1])
}))
basin_cmp$p_adj <- p.adjust(basin_cmp$p, "bonferroni")
basin_cmp <- add_significance(basin_cmp, p.col = "p_adj", output.col = "p.adj.signif")
basin_r_max <- basin_cmp[which.max(basin_cmp$r), ]
basin_r_lab <- variable_labels_land_bare[[basin_r_max$variable]]
if (grepl("^[A-Z][a-z]", basin_r_lab))
  basin_r_lab <- paste0(tolower(substr(basin_r_lab, 1, 1)), substring(basin_r_lab, 2))
basin_iqr_wider <- sum(vapply(land_variables, function(v) {
  q <- tapply(bdf[[v]], droplevels(bdf$Basin), IQR, na.rm = TRUE)
  q[["Binchuan"]] > q[["Huangping"]]
}, logical(1)))
basin_iqr_word <- c("none", "one", "two", "three", "four", "five",
                    "six")[basin_iqr_wider + 1]
```

\newpage

This document accompanies the main text and provides the complete statistical output underlying the reported results, together with the diagnostic and robustness analyses that support their interpretation. It is divided into two parts.

The first part describes the computational environment and parameter settings in detail, ensuring that every reported value can be reproduced exactly. The second part presents the tables and figures, which are arranged in the order in which the results are discussed in the main text and grouped into four sections: the techno-typological characterisation of the retouch products; the technological consistency of the surface collections with the excavated assemblage from Longtan; the organisation of raw material use; and the landscape structure of technological variation. Each section begins with a brief guide to its contents, and each table and figure caption is written to be self-contained, allowing the materials to be understood without reference to the main text.

\newpage

# Computational environment and settings

Every value reported here is computed from the same data, in the same analysis run, and using the same fixed random seed (2226) as its corresponding value in the main text. Both documents are generated from a single analysis file, ensuring that no numerical result can differ between them. The data and code required to reproduce this document are deposited with the research compendium at OSF.

All analyses were conducted in `r R.version.string` [@R-base], with the packages given in @tbl-software.

```{r}
#| label: tbl-software
#| tbl-cap: "The R packages on which the reported results depend, each with the version used in the run that produced this document, the citation the package asks for, and the part of the analysis it carries. The complete session, including the transitive dependencies not listed here, is recorded in the colophon of the main text."

pad <- function(x, n) formatC(x, width = n, flag = "-")

knitr::kable(si_pkg_tbl, row.names = FALSE, align = c("l", "l", "l", "l"),
             col.names = c(pad("Package", 15), pad("Version", 12),
                           pad("Citation", 23), "Purpose"))
```

The random seed (2226) is set once at the beginning of the analysis file and reset immediately before each permutation, Monte Carlo, or bootstrap procedure. This ensures that the results are reproducible and do not depend on the order in which analyses are executed.

Three resampling settings govern the results reported in Part 2. All permutation tests, including PERMANOVA and PERMDISP, use `r n_fmt(PERM)` permutations. Percentile confidence intervals are based on `r n_fmt(B_BOOT)` bootstrap replicates. These intervals are reported in the main text for three quantities: the between-group ratios of the coefficient of variation and Fano factor; Spearman's ρ between each technological measure and distance to the nearest channel; and Cohen's d for the difference between scraper edge angle and the exterior platform angle of resharpening flakes. The Krishnamoorthy–Lee modified signed likelihood ratio test for equality of coefficients of variation is evaluated using `r n_fmt(MSLR_NR)` Monte Carlo iterations. To prevent these simulations from affecting subsequent random-number-dependent results, the random-number generator state is saved and restored around each test. Consequently, the bootstrap confidence intervals are independent of the number or order of Krishnamoorthy–Lee tests performed previously. All families of multiple comparisons are adjusted using the Bonferroni method.

\newpage

# Detailed results

## Techno-typological characterisation of the retouch products

This section describes the retouch products themselves. @tbl-retouch-products gives the metric and reduction attributes of the surface Quina scrapers and of the resharpening flakes that accompany them, and @fig-categorical-attributes gives the categorical attributes of the scrapers — the cover and position of cortex, the type of platform and the direction from which retouch was struck — as the percentage of specimens in each class. @fig-edge-angle-reduction sets edge angle against the three measures of how far reduction was carried, the relationships summarised in the main text. The typological attribution of the same specimens is reported in the main text.

\newpage

```{r}
#| label: tbl-retouch-products
#| tbl-cap: "Metric and reduction attributes of the Quina scrapers and the resharpening flakes from the surface collections. Not every attribute could be measured on every specimen: the platform attributes are recorded only on those that retain the platform of the blank, and GIUR on all but one scraper. The retouch attributes apply to the scrapers alone and the exterior platform angle to the resharpening flakes alone."

attr_row <- function(label, x, d = 1, ds = d) {
  x  <- as.numeric(x); x <- x[is.finite(x)]
  qs <- unname(quantile(x, c(.25, .75)))
  data.frame(Attribute = paste0(IND, label),
             mn  = f(mean(x), d), sd = f(sd(x), d),
             med = f(median(x), ds),
             iqr = paste0(f(qs[1], ds), "–", f(qs[2], ds)))
}
attr_head <- function(label, n_total) {
  data.frame(Attribute = sprintf("**%s** (*n* = %d)", label, n_total),
             mn = "", sd = "", med = "", iqr = "")
}

retouch_products_tbl <- bind_rows(
  attr_head("Quina scrapers", nrow(q)),
  attr_row("Length (mm)",    q$Length,    1),
  attr_row("Width (mm)",     q$Width,     1),
  attr_row("Thickness (mm)", q$Thickness, 1),
  attr_row("Mass (g)",       q$Mass,      1),
  attr_row("Platform depth (mm)", q$Platform_depth, 1),
  attr_row("Platform width (mm)", q$Platform_width, 1),
  attr_row("IPA (°)",             q$IPA,            1),
  attr_row("Retouch generations", q$Ave_RG,    2),
  attr_row(variable_labels[["N_Scar"]], q$N_Scar, 1, ds = 0),
  attr_row("GIUR",                q$Ave_GIUR,  2),
  attr_row("Edge angle (°)",      q$Edge_Angle, 1),
  attr_row("Retouched perimeter", q$Retouch_length_index, 2),
  attr_head("Resharpening flakes", nrow(rfl)),
  attr_row("Length (mm)",    rfl$Length,    1),
  attr_row("Width (mm)",     rfl$Width,     1),
  attr_row("Thickness (mm)", rfl$Thickness, 1),
  attr_row("Mass (g)",       rfl$Mass,      1),
  attr_row("Platform depth (mm)", rfl$Platform_depth, 1),
  attr_row("Platform width (mm)", rfl$Platform_width, 1),
  attr_row("IPA (°)",             rfl$IPA,            1),
  attr_row("EPA (°)",             rfl$EPA,            1))

knitr::kable(retouch_products_tbl, align = c("l", "r", "r", "r", "r"),
             col.names = c("Attribute", "Mean", "SD", "Median", "Q1–Q3"))
```

\newpage

```{r}
#| label: fig-categorical-attributes
#| fig-cap: "Categorical attributes of the surface Quina scrapers: proportion of the dorsal face covered by cortex, position of that cortex, platform type, and the direction from which retouch was struck. Bars run to the percentage of the specimens recorded for that attribute, and the count and percentage are given at the end of each bar."
#| fig-width: 5.9
#| fig-height: 4.9

cat_panel <- function(x, panel, levels_order = NULL, recode_map = NULL) {
  x <- trimws(as.character(x)); x <- x[!is.na(x)]
  if (!is.null(recode_map)) x <- ifelse(x %in% names(recode_map), recode_map[x], x)
  d <- tibble(Level = x) |> count(Level, name = "n")
  d <- if (is.null(levels_order)) arrange(d, desc(n)) else
    d[match(levels_order, d$Level), ] |> filter(!is.na(Level))
  d |> mutate(pct = 100 * n / sum(n), Panel = panel, rank = row_number())
}
cortex_classes <- c("0", "1-25%", "26-50%", "51-75%", "76-99%")
cat_df <- bind_rows(
  cat_panel(q$Cortex, "Cortex proportion", levels_order = cortex_classes),
  cat_panel(q$Cortex_position, "Cortex position", recode_map = c(Non_cortex = "None")),
  cat_panel(q$Platform_type, "Platform type"),
  cat_panel(q$Tot_Direction, "Retouch direction")) |>
  mutate(Panel = factor(Panel, levels = unique(Panel)))
cat_df$Level <- factor(cat_df$Level,
                       levels = cat_df |> arrange(Panel, desc(rank)) |> pull(Level))

cortex_ramp <- setNames(colorRampPalette(c("#CBDCEA", "#2C5F7C"))(length(cortex_classes)),
                        cortex_classes)
panel_flat <- c(`Cortex position` = "#E07C90", `Platform type` = "#E6C25C",
                `Retouch direction` = "#9B87C4")
cat_df <- cat_df |>
  mutate(bar_fill = ifelse(as.character(Panel) == "Cortex proportion",
                           cortex_ramp[as.character(Level)],
                           panel_flat[as.character(Panel)]))

ggplot(cat_df, aes(pct, Level)) +
  geom_col(aes(fill = bar_fill), width = 0.7) +
  geom_text(aes(label = sprintf("n = %d, %.1f%%", n, pct)),
            hjust = -0.12, size = 2.2, color = "#303238") +
  scale_fill_identity() +
  facet_grid(Panel ~ ., scales = "free_y", space = "free_y") +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25),
                     labels = function(z) paste0(z, "%"),
                     expand = expansion(mult = c(0, 0.24))) +
  labs(x = "Percentage", y = NULL) +
  fig_theme +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 7.5),
        strip.text.y = element_text(size = 7.5, angle = -90))
```

\newpage

```{r}
#| label: fig-edge-angle-reduction
#| fig-cap: "Relationship between Quina scraper edge angle and three measures (proportion of perimeter retouched, GIUR, and number of retouch generations). Each panel reports Spearman’s ρ and p-value. The dashed line represents an ordinary least-squares linear fit, shown only as a visual guide to the direction of the relationship."
#| fig-width: 5.9
#| fig-height: 2.44

set.seed(2226)
focal_ea    <- "Edge_Angle"
corr_vars   <- c("Retouch_length_index", "Ave_GIUR", "Ave_RG")
corr_labels <- c(Retouch_length_index = "Retouched perimeter",
                 Ave_GIUR = "GIUR", Ave_RG = "Retouch generations")
fmt_p_math  <- function(p) ifelse(p < 0.001, "italic(p) < 0.001",
                                  sprintf("italic(p) == %.3f", p))

quina_ea <- read_excel(sc_path, sheet = "Quina scraper") |>
  mutate(across(all_of(c(focal_ea, corr_vars)), as.numeric))
spearman_one <- function(v) {
  pair <- quina_ea |> select(all_of(c(focal_ea, v))) |> drop_na()
  ct <- suppressWarnings(cor.test(pair[[focal_ea]], pair[[v]], method = "spearman", exact = FALSE))
  data.frame(Variable = v, rho = unname(ct$estimate), p_value = ct$p.value)
}
results_ea <- do.call(rbind, lapply(corr_vars, spearman_one))
results_ea$p_value <- p.adjust(results_ea$p_value, "bonferroni")   # one family of three

long_ea <- quina_ea |> select(all_of(c(focal_ea, corr_vars))) |>
  pivot_longer(all_of(corr_vars), names_to = "Variable", values_to = "Value") |>
  filter(!is.na(.data[[focal_ea]]), !is.na(Value)) |>
  mutate(Variable = factor(Variable, levels = corr_vars),
         x_plot = if_else(Variable == "Ave_RG", Value + runif(n(), -0.11, 0.11), Value))
labels_ea <- results_ea |>
  mutate(Variable = factor(Variable, levels = corr_vars),
         label = sprintf("rho == %.2f * ',' ~ %s", rho, fmt_p_math(p_value)))

ggplot(long_ea, aes(y = .data[[focal_ea]])) +
  geom_smooth(aes(x = Value), method = "lm", formula = y ~ x, se = TRUE,
              color = "#6BA8CE", fill = "#CBDCEA", linewidth = 0.6, linetype = "dashed") +
  geom_point(aes(x = x_plot), color = "#565693", alpha = 0.5, size = 1.4, shape = 16) +
  geom_text(data = labels_ea, aes(x = -Inf, y = Inf, label = label), parse = TRUE,
            hjust = -0.08, vjust = 1.35, size = 2.5, color = "#202124", inherit.aes = FALSE) +
  facet_wrap(~ Variable, scales = "free_x", ncol = 3, labeller = as_labeller(corr_labels)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.14))) +
  scale_x_continuous(expand = expansion(mult = 0.06)) +
  labs(x = NULL, y = "Edge angle (°)") +
  fig_theme + theme(panel.grid.major = element_blank())
```

\newpage

## Technological consistency

This section compares the surface Quina scrapers with the excavated Quina scrapers from Longtan, against the Longtan non-Quina scrapers as a control. @fig-qq shows the distribution of each of the six technological variables within each group, the check that stands behind the choice of test. @tbl-tc-permanova to @tbl-tc-pca give the full output of the comparison: the pairwise tests of group position and of group spread, the variable-by-variable tests, and the loadings of the ordination. @tbl-tc-sensitivity and @fig-sensitivity-pca then repeat the analysis with the surface specimens collected from within the area of the Longtan site removed.

\newpage

```{r}
#| label: fig-qq
#| fig-cap: "Normal quantile–quantile plots of the six technological variables. In every panel the horizontal axis is the theoretical normal quantile and the vertical axis the sample quantile, and the dashed line passes through the first and third sample quartiles. Inset at the upper left of each panel is a histogram of the same values, plotted in its own right but with its axes suppressed, to give the shape of the distribution alongside its departure from normality."
#| fig-width: 5.9
#| fig-height: 8.4

dist_long <- variable_long |>
  mutate(Variable = factor(variable_labels_bare[as.character(Variable)],
                           levels = unname(variable_labels_bare)),
         Group = factor(group_labels[as.character(Group)],
                        levels = unname(group_labels)))
lab_colors <- setNames(unname(group_colors), unname(group_labels))

INSET_GAP <- unit(1.4, "mm")
inset_theme <- theme_minimal(base_size = 6) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.3),
        axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_rect(fill = "white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        plot.margin = margin(0, 0, 0, 0))

qq_cell <- function(vlab, glab, ylab, carries_legend = FALSE) {
  d <- dist_long[dist_long$Variable == vlab & dist_long$Group == glab, ]
  legend_key <- data.frame(
    x = 0, y = mean(range(d$Value, na.rm = TRUE)),
    Group = factor(levels(dist_long$Group), levels = levels(dist_long$Group)))
  p_hist <- ggplot(d, aes(Value)) +
    geom_histogram(bins = 12, fill = lab_colors[[glab]], alpha = 0.75,
                   color = "#202124", linewidth = 0.1) +
    scale_x_continuous(expand = expansion(mult = 0.03)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
    inset_theme
  p_qq <- ggplot(d, aes(sample = Value, color = Group)) +
    stat_qq(size = 0.7, alpha = 0.6, shape = 16, show.legend = FALSE) +
    stat_qq_line(linewidth = 0.35, color = "#5A5F66", linetype = "dashed",
                 show.legend = FALSE) +
    scale_color_manual(values = unname(group_colors),
                       limits = unname(group_labels), drop = FALSE) +
    scale_x_continuous(expand = expansion(mult = 0.06)) +
    scale_y_continuous(expand = expansion(mult = c(0.06, 0.34))) +
    labs(x = NULL, y = ylab) +
    fig_theme +
    theme(panel.grid.major = element_blank(),
          axis.title.y = element_text(size = 8, face = "bold"),
          axis.text = element_text(size = 6.5)) +
    guides(color = guide_legend(nrow = 1, override.aes = list(size = 1.8, alpha = 1)))
  if (carries_legend) {
    p_qq <- p_qq +
      geom_point(data = legend_key, aes(x = x, y = y, color = Group),
                 inherit.aes = FALSE, alpha = 0, size = 0.7, shape = 16,
                 show.legend = TRUE)
  }
  p_qq + inset_element(p_hist,
                       left   = INSET_GAP,
                       bottom = unit(0.63, "npc"),
                       right  = unit(0.47, "npc"),
                       top    = unit(1, "npc") - INSET_GAP,
                       align_to = "panel", ignore_tag = TRUE)
}

vars_row <- levels(dist_long$Variable)
qq_cells <- unlist(lapply(vars_row, function(v) {
  lapply(seq_along(levels(dist_long$Group)), function(i) {
    qq_cell(v, levels(dist_long$Group)[i], if (i == 1) v else NULL,
            carries_legend = (v == vars_row[1] && i == 1))
  })
}), recursive = FALSE)

wrap_plots(qq_cells, ncol = 3, guides = "collect") &
  theme(legend.position = "bottom", legend.margin = margin(0, 0, 0, 0))
```

\newpage

```{r}
#| label: tbl-tc-permanova
#| tbl-cap: "Pairwise PERMANOVA on the z-scored six-variable technological space, Euclidean distances, 9,999 permutations. Degrees of freedom are given as model, residual. Adjusted *p* is the Bonferroni adjustment across the three comparisons."

pw_label <- function(x) {
  parts <- strsplit(x, " vs ", fixed = TRUE)[[1]]
  paste(group_labels[parts], collapse = " vs ")
}
tc_perm_out <- data.frame(
  Comparison = vapply(tc_pw$Comparison, pw_label, character(1)),
  n = sprintf("%d, %d", tc_pw$n1, tc_pw$n2),
  Df = sprintf("%d, %d", tc_pw$Df, tc_pw$Df_res),
  SS = f(tc_pw$SumOfSqs, 2),
  R2 = f(tc_pw$R2, 3),
  F = f(tc_pw$F, 2),
  p = fpt(tc_pw$p_value),
  padj = fpt(tc_pw$p_adjusted))

knitr::kable(tc_perm_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Comparison", "*n*", "*df*", "SS", "*R*²",
                           "pseudo-*F*", "*p*", "Adjusted *p*"))
```

\newpage

```{r}
#| label: tbl-tc-permdisp
#| tbl-cap: "PERMDISP on the same technological space. The upper panel gives the mean distance of each group to its own centroid, the quantity the test compares; distance is measured in the z-scored space, so its units are standard deviations. The lower panel gives the pairwise tests, each fitted to its pair in turn, with the *p*-value from the permutation of the full three-group model alongside for comparison."

pdd <- permA$distances |> group_by(Group) |>
  summarise(n = n(), m = mean(distance), s = sd(distance), .groups = "drop") |>
  mutate(se = s / sqrt(n))

pd_out <- bind_rows(
  data.frame(Item = "**Distance to centroid, by group**",
             a = "", b = "", c = "", d = "", e = ""),
  data.frame(Item = paste0(IND, group_labels[as.character(pdd$Group)]),
             a = as.character(pdd$n), b = f(pdd$m, 3), c = f(pdd$s, 3),
             d = f(pdd$se, 3), e = ""),
  data.frame(Item = paste0(IND, "*Overall test*"), a = "",
             b = sprintf("%d, %d", permA$overall_df, permA$overall_df_res),
             c = f(permA$overall_F, 2), d = fpt(permA$overall_p), e = ""),
  data.frame(Item = "**Pairwise tests**", a = "", b = "", c = "", d = "", e = ""),
  data.frame(Item = paste0(IND, vapply(permA$pair_tab$Comparison, function(x) {
               paste(group_labels[strsplit(x, "-", fixed = TRUE)[[1]]], collapse = " vs ")
             }, character(1))),
             a = sprintf("%d, %d", permA$pair_tab$n1, permA$pair_tab$n2),
             b = sprintf("%d, %d", permA$pair_tab$Df, permA$pair_tab$Df_res),
             c = f(permA$pair_tab$F, 2),
             d = fpt(permA$pair_tab$p),
             e = fpt(permA$pairwise[permA$pair_tab$Comparison])))

knitr::kable(pd_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("", "*n*", "Mean / *df*", "SD / *F*",
                           "SE / *p*", "*p* (3-group)"))
```

\newpage

```{r}
#| label: tbl-tc-univariate
#| tbl-cap: "Per-variable comparison of the three groups: the omnibus test for each variable, followed by its three post-hoc comparisons. GIUR, total scars and retouch generations were tested with Kruskal–Wallis and Dunn's post-hoc comparisons, the effect size being epsilon-squared; thickness, retouched perimeter and edge angle with Welch's ANOVA and Welch's *t* post-hoc comparisons. All post-hoc *p*-values carry a Bonferroni adjustment within each variable. ns, not significant."

uni_rows <- function(v) {
  is_kw <- v %in% kw_vars
  if (is_kw) {
    o <- kw_omni[kw_omni$Variable == v, ]; e <- kw_eff[kw_eff$Variable == v, ]
    omni <- data.frame(Item = paste0(IND, "*Kruskal–Wallis*"),
                       stat = sprintf("H = %s", f(o$statistic, 2)),
                       df = as.character(o$df), p = fpt(o$p),
                       eff = sprintf("ε² = %s", f(e$effsize, 3)), sig = "")
    ph <- kw_ph[kw_ph$Variable == v, ]
    post <- data.frame(
      Item = paste0(IND, vapply(seq_len(nrow(ph)), function(i)
        paste(group_labels[c(ph$group1[i], ph$group2[i])], collapse = " vs "), character(1))),
      stat = sprintf("Z = %s", f(ph$statistic, 2)),
      df = sprintf("%d, %d", ph$n1, ph$n2),
      p = fpt(ph$p.adj), eff = "", sig = ph$p.adj.signif)
  } else {
    o <- we_omni[we_omni$Variable == v, ]
    omni <- data.frame(Item = paste0(IND, "*Welch's ANOVA*"),
                       stat = sprintf("F = %s", f(o$statistic, 2)),
                       df = sprintf("%s, %s", o$DFn, f(o$DFd, 1)),
                       p = fpt(o$p), eff = "", sig = "")
    ph <- we_ph_full[we_ph_full$Variable == v, ]
    post <- data.frame(
      Item = paste0(IND, vapply(seq_len(nrow(ph)), function(i)
        paste(group_labels[c(ph$group1[i], ph$group2[i])], collapse = " vs "), character(1))),
      stat = sprintf("t = %s", f(ph$statistic, 2)),
      df = f(ph$df, 1), p = fpt(ph$p.adj), eff = "", sig = ph$p.adj.signif)
  }
  bind_rows(data.frame(Item = sprintf("**%s**", variable_labels[[v]]),
                       stat = "", df = "", p = "", eff = "", sig = ""),
            omni, post)
}
uni_out <- bind_rows(lapply(variables, uni_rows))

knitr::kable(uni_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Variable / comparison", "Statistic", "*df* or *n*",
                           "*p*", "Effect size", ""))
```

\newpage

```{r}
#| label: tbl-tc-pca
#| tbl-cap: "Loadings of all six principal components of the z-scored technological space, with the eigenvalue, the percentage of variance and the cumulative percentage for each."

ld <- permA$pca_loadings
pca_out <- bind_rows(
  data.frame(Item = variable_labels[ld$Variable],
             as.data.frame(lapply(ld[, paste0("PC", 1:6)], f, d = 3))),
  data.frame(Item = "**Eigenvalue**",
             as.data.frame(as.list(setNames(f(permA$sdev^2, 2), paste0("PC", 1:6))))),
  data.frame(Item = "**Variance (%)**",
             as.data.frame(as.list(setNames(f(permA$var_explained, 1), paste0("PC", 1:6))))),
  data.frame(Item = "**Cumulative (%)**",
             as.data.frame(as.list(setNames(f(cumsum(permA$var_explained), 1), paste0("PC", 1:6))))))

knitr::kable(pca_out, row.names = FALSE, align = c("l", rep("r", 6)),
             col.names = c("Variable", paste0("PC", 1:6)))
```

\newpage

```{r}
#| label: tbl-tc-sensitivity
#| tbl-cap: "Sensitivity of the technological consistency result to the surface specimens collected from within the area of the Longtan site. The full set is the analysis reported in the main text, with 165 surface specimens; the reduced set excludes those collected within the site area, leaving 150, and is z-scored on its own. PERMANOVA *p*-values carry the Bonferroni adjustment across the three comparisons."

sens_rows <- function(comp_perm, comp_disp) {
  full_p <- tc_pw[tc_pw$Comparison == comp_perm, ]
  red_p  <- tc_pw_excl[tc_pw_excl$Comparison == comp_perm, ]
  full_d <- permA$pair_tab[permA$pair_tab$Comparison == comp_disp, ]
  red_d  <- permC$pair_tab[permC$pair_tab$Comparison == comp_disp, ]
  label  <- paste(group_labels[strsplit(comp_perm, " vs ", fixed = TRUE)[[1]]],
                  collapse = " vs ")
  bind_rows(
    data.frame(Comparison = sprintf("**%s**", label), Analysis = "", fR2 = "", fF = "",
               fp = "", rR2 = "", rF = "", rp = ""),
    data.frame(Comparison = paste0(IND, "PERMANOVA"), Analysis = "centroid position",
               fR2 = f(full_p$R2, 3), fF = f(full_p$F, 2), fp = fpt(full_p$p_adjusted),
               rR2 = f(red_p$R2, 3),  rF = f(red_p$F, 2),  rp = fpt(red_p$p_adjusted)),
    data.frame(Comparison = paste0(IND, "PERMDISP"), Analysis = "dispersion",
               fR2 = "", fF = f(full_d$F, 2), fp = fpt(full_d$p),
               rR2 = "", rF = f(red_d$F, 2),  rp = fpt(red_d$p)))
}
sens_out <- bind_rows(
  sens_rows("SC_Quina vs LT_Quina",    "SC_Quina-LT_Quina"),
  sens_rows("SC_Quina vs LT_Ordinary", "SC_Quina-LT_Ordinary"),
  sens_rows("LT_Quina vs LT_Ordinary", "LT_Quina-LT_Ordinary"))

knitr::kable(sens_out, row.names = FALSE, align = c("l", "l", rep("r", 6)),
             col.names = c("Comparison", "Tests", "*R*²", "*F*", "*p*",
                           "*R*²", "*F*", "*p*"))
```

\newpage

```{r}
#| label: fig-sensitivity-pca
#| fig-cap: "Ordination of the technological space with the surface specimens from within the area of the Longtan site removed, to be compared with panel A of the corresponding figure in the main text. (A) Specimen scores, with convex hulls enclosing each group and filled points marking group centroids. (B) Loadings of the six variables on the same two components, drawn on their own axes so that the specimen scores are not overplotted."
#| fig-width: 5.9
#| fig-height: 3.7

sc_s <- permC$pca_scores
ld_s <- permC$pca_loadings
cent_s <- sc_s |> group_by(Group) |> summarise(cx = mean(PC1), cy = mean(PC2), .groups = "drop")
hull_s <- sc_s |> group_by(Group) |> filter(n() >= 3) |>
  group_modify(function(g, k) { h <- chull(g$PC1, g$PC2); bind_rows(g[h, ], g[h[1], ]) }) |> ungroup()
arr_s <- ld_s |> mutate(Label = variable_labels_bare[Variable])

p_scores <- ggplot(sc_s, aes(PC1, PC2, color = Group)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
  geom_polygon(data = hull_s, aes(fill = Group, group = Group), alpha = 0.18, color = NA) +
  geom_point(size = 1.1, alpha = 0.7, shape = 16) +
  geom_point(data = cent_s, aes(cx, cy, fill = Group), shape = 21, color = "#5A5F66",
             size = 2.4, stroke = 0.6, inherit.aes = FALSE) +
  scale_color_manual(values = group_colors, labels = group_labels) +
  scale_fill_manual(values = group_colors, labels = group_labels) +
  scale_x_continuous(expand = expansion(mult = 0.14)) +
  scale_y_continuous(expand = expansion(mult = 0.10)) +
  labs(x = sprintf("PC1 (%.1f%%)", permC$var_explained[1]),
       y = sprintf("PC2 (%.1f%%)", permC$var_explained[2])) +
  fig_theme +
  theme(panel.grid.major = element_blank(), legend.text = element_text(size = 8)) +
  guides(fill = "none",
         color = guide_legend(nrow = 1, override.aes = list(size = 1.8, alpha = 1)))

circle_ref <- data.frame(theta = seq(0, 2 * pi, length.out = 200)) |>
  mutate(x = cos(theta), y = sin(theta))
lim_s <- 1.12
p_loadings <- ggplot(arr_s, aes(PC1, PC2)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
  geom_path(data = circle_ref, aes(x, y), linewidth = 0.3, color = "grey70",
            linetype = "dashed", inherit.aes = FALSE) +
  geom_segment(aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.075, "cm"), type = "open", angle = 22),
               linewidth = 0.3, color = "#5A5F66") +
  geom_text_repel(aes(label = Label), size = 2.2, color = "#3A3D42",
                  bg.color = "white", bg.r = 0.12, segment.color = "grey55",
                  segment.size = 0.2, min.segment.length = 0.2, box.padding = 0.35,
                  point.padding = 0.1, max.overlaps = Inf, seed = 2226) +
  scale_x_continuous(limits = c(-lim_s, lim_s)) +
  scale_y_continuous(limits = c(-lim_s, lim_s)) +
  labs(x = "PC1 loading", y = "PC2 loading") +
  fig_theme + theme(panel.grid.major = element_blank())

(p_scores + labs(tag = "A")) + (p_loadings + labs(tag = "B")) +
  plot_layout(widths = c(1.7, 1), guides = "collect") &
  theme(legend.position = "bottom", legend.margin = margin(0, 0, 0, 0))
```

\newpage

## Raw material economy

This section reports what was available on the river bars and what was chosen from it. @fig-sample-points shows the sampling points at which the cobbles were recorded, and @fig-thin-section the petrography of the trachyte, the dominant raw material. @tbl-composition gives the raw material composition of the surveyed cobbles and of the surface Quina scrapers, valley by valley, and @tbl-electivity sets the two against one another as an index of selection. @tbl-rawmat-permanova tests whether composition differs between sampling points, valleys and basins, and @tbl-clast-size reports the size of the surveyed cobbles by lithology.

\newpage

![The gravel-bar sampling points. (A) to (E) The five points at which the surveyed cobbles were recorded, each with the cleared quadrat marked out by string and a spirit level laid across it for scale. (F) A general view of one of the sampled channels. Every cobble longer than 30 mm exposed at the surface within the quadrat was recorded.](../figures/sample_point_photo.png){#fig-sample-points width="5.9in"}

\newpage

![Petrography of the trachyte. (A) Hand specimens of five of the sampled trachyte cobbles. (B) Photomicrographs of thin sections cut from them, with the magnification and the scale given in each frame. The rock is porphyritic: euhedral tabular to prismatic K-feldspar phenocrysts, with subordinate quartz, augite and clay minerals, are set in a fine-grained groundmass of K-feldspar microlites and devitrified glass.](../figures/thin_section.png){#fig-thin-section width="5.9in"}

\newpage

```{r}
#| label: tbl-composition
#| tbl-cap: "Raw material composition of the river-bar cobbles and of the surface-collected Quina scrapers, by valley. Cells give the count with the percentage of the row in parentheses."

comp_tbl <- comp_fig |>
  mutate(cell = ifelse(n == 0, "0", sprintf("%d (%s%%)", n, f(percent, 1)))) |>
  select(Layer, river_ID, Material, cell) |>
  pivot_wider(names_from = Material, values_from = cell) |>
  left_join(counts_fig, by = c("Layer", "river_ID"))

comp_out <- bind_rows(lapply(layer_levels, function(lay) {
  d <- comp_tbl[comp_tbl$Layer == lay, ]
  bind_rows(
    data.frame(Valley = sprintf("**%s**", lay),
               Trachyte = "", Sandstone = "", Quartz = "", Mudstone = "",
               Andesite = "", N = ""),
    data.frame(Valley = paste0(IND, as.character(d$river_ID)),
               Trachyte = d$Trachyte, Sandstone = d$Sandstone, Quartz = d$Quartz,
               Mudstone = d$Mudstone, Andesite = d$Andesite, N = as.character(d$N)))
}))

knitr::kable(comp_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r"),
             col.names = c("Valley", "Trachyte", "Sandstone", "Quartz",
                           "Mudstone", "Andesite", "*n*"))
```

\newpage

```{r}
#| label: tbl-electivity
#| tbl-cap: "Jacobs' electivity index for every raw material. *D* runs from −1, complete avoidance, through 0, use in proportion to availability, to +1, complete selection."

elect_tbl <- tibble(Material = material_levels) |>
  left_join(comp_avail |> transmute(Material, an = n, ap = p), by = "Material") |>
  left_join(comp_used  |> transmute(Material, un = n, up = r), by = "Material") |>
  left_join(jac |> transmute(Material, D), by = "Material") |>
  mutate(across(c(an, un), ~ replace_na(.x, 0L)),
         across(c(ap, up), ~ replace_na(.x, 0)),
         Label = ifelse(Material == "Andesite", "Andesite†", Material))

elect_out <- bind_rows(
  data.frame(Material = elect_tbl$Label,
             an = as.character(elect_tbl$an), ap = f(elect_tbl$ap, 1),
             un = as.character(elect_tbl$un), up = f(elect_tbl$up, 1),
             D  = f(elect_tbl$D, 2)),
  data.frame(Material = "**Total**",
             an = as.character(sum(elect_tbl$an)), ap = f(100, 1),
             un = as.character(sum(elect_tbl$un)), up = f(100, 1), D = ""))

knitr::kable(elect_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Raw material", "Cobbles *n*", "Cobbles %",
                           "Scrapers *n*", "Scrapers %", "*D*"))
```

\newpage

```{r}
#| label: tbl-rawmat-permanova
#| tbl-cap: "PERMANOVA and PERMDISP on raw material composition, on Bray–Curtis dissimilarities with 9,999 permutations. Sums of squares, degrees of freedom and residual degrees of freedom are those of the PERMANOVA; the last two columns give the pseudo-*F* and *p* of the test of homogeneity of dispersion on the same dissimilarity matrix. The unit is the individual cobble in the first panel and the individual artefact in the second."

perm_row <- function(model, a, pd) {
  data.frame(Model = paste0(IND, model),
             Df = sprintf("%d, %d", a$Df[1], a$Df[2]),
             SS = f(a$SumOfSqs[1], 2),
             R2 = f(a$R2[1], 3),
             F  = f(a$F[1], 2),
             p  = fpt(a$`Pr(>F)`[1]),
             pdF = f(pd[["F"]], 2),
             pdp = fpt(pd[["p"]]))
}
head_row <- function(label) data.frame(Model = sprintf("**%s**", label),
  Df = "", SS = "", R2 = "", F = "", p = "", pdF = "", pdp = "")

rawmat_out <- bind_rows(
  head_row(sprintf("Available river cobbles (*n* = %d)", n_cobble)),
  perm_row("~ sampling point", cobbleLoc,   pd_cobbleLoc),
  perm_row("~ valley",         cobbleRiver, pd_cobbleRiver),
  head_row(sprintf("Surface Quina scrapers (*n* = %d)", nrow(scA))),
  perm_row("~ basin",  usedA_basin, pd_usedA_basin),
  perm_row("~ valley", usedA_river, pd_usedA_river))

knitr::kable(rawmat_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Model", "*df*", "SS", "*R*²", "pseudo-*F*", "*p*",
                           "PERMDISP *F*", "PERMDISP *p*"))
```

\newpage

```{r}
#| label: tbl-clast-size
#| tbl-cap: "Size of the surveyed river cobbles by lithology. Size is the geometric mean of length, breadth and thickness in millimetres, and GM is the geometric mean of that quantity within each lithology."

size_row <- function(m) {
  x <- cl$size[cl$Material == m]
  if (!length(x)) return(NULL)
  qs <- unname(quantile(x, c(.25, .75)))
  data.frame(Material = m, n = as.character(length(x)),
             GM = f(gm(x), 1), Median = f(median(x), 1),
             IQR = sprintf("%s–%s", f(qs[1], 1), f(qs[2], 1)),
             Range = sprintf("%s–%s", f(min(x), 1), f(max(x), 1)))
}
size_out <- bind_rows(lapply(material_levels, size_row))

knitr::kable(size_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Raw material", "*n*", "GM (mm)", "Median (mm)",
                           "Q1–Q3 (mm)", "Range (mm)"))
```

\newpage

## Landscape structure

This section asks whether the technological variation follows position in the landscape, with the locality rather than the specimen as the unit. @tbl-locality-matrix lists the localities that enter the analysis with their median value on each of the seven measures, and @tbl-landscape-permanova tests each landscape variable against that space. The ordination of the localities, coloured by each landscape variable, is figured in the main text; the two figures here show the same data broken out variable by variable, @fig-landscape-scatter as each measure against each of the continuous variables and @fig-basin-boxplot as a comparison between the two basins.

\newpage

```{r}
#| label: tbl-locality-matrix
#| tbl-cap: "Locality-level analysis frame. Each row is one Quina locality, with the number of Quina scrapers complete on all seven measures and the median of those specimens on each. Distance and height are measured to and above the nearest channel."

loc_out <- site_df |>
  arrange(Basin, desc(n_art), Name) |>
  transmute(Locality = Name, Basin = as.character(Basin),
            n = as.character(n_art),
            d = as.character(round(Distance_to_water)),
            h = as.character(round(Height_above_river)),
            GM = f(GM, 1),
            Th = f(Thickness, 1), RLI = f(Retouch_length_index, 2),
            GIUR = f(Ave_GIUR, 2), NS = f(N_Scar, 0),
            RG = f(Ave_RG, 2), EA = f(Edge_Angle, 1))

knitr::kable(loc_out, row.names = FALSE,
             align = c("l", "l", rep("r", 10)),
             col.names = c("Locality", "Basin", "*n*", "Distance (m)", "Height (m)",
                           unname(variable_labels_land)))
```

\newpage

```{r}
#| label: tbl-landscape-permanova
#| tbl-cap: "Locality-level PERMANOVA for each landscape variable, on Euclidean distances in the z-scored seven-measure space with 9,999 permutations, together with the test of dispersion for the one categorical variable. Each variable is fitted in a model of its own; the four fitted together are given in the main text."

ls_row <- function(label, v, indent = TRUE) {
  data.frame(Term = if (indent) paste0(IND, label) else sprintf("**%s**", label),
             Df = sprintf("%d, %d", as.integer(v["Df"]), as.integer(v["Df_res"])),
             SS = f(v["SumOfSqs"], 2), R2 = f(v["R2"], 3),
             F = f(v["F"], 3), p = fpt(v["p"]))
}
ls_out <- bind_rows(
  data.frame(Term = "**Basin**", Df = "", SS = "", R2 = "", F = "", p = ""),
  ls_row(sprintf("PERMANOVA (Binchuan %d, Huangping %d)",
                 n_basin_site[["Binchuan"]], n_basin_site[["Huangping"]]), ls_basin),
  data.frame(Term = paste0(IND, "PERMDISP"), Df = "", SS = "", R2 = "",
             F = f(ls_basin_pd[["F"]], 3), p = fpt(ls_basin_pd[["p"]])),
  ls_row("Height above channel (m)", ls_height, indent = FALSE),
  ls_row("Distance to channel (m)", ls_dist, indent = FALSE),
  ls_row("Assemblage size (n specimens)", ls_size, indent = FALSE))

knitr::kable(ls_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Landscape variable", "*df*", "SS", "*R*²", "pseudo-*F*", "*p*"))
```

\newpage

```{r}
#| label: fig-landscape-scatter
#| fig-cap: "Each of the seven technological measures against each of the three continuous landscape variables, at the scale of the locality: one row per measure, one column per landscape variable, all at measured values. The dashed line is an ordinary least-squares fit, shown only as a guide to the direction of the relationship. Each panel gives Spearman's ρ with its *p*-value, adjusted by the Bonferroni procedure within each column, every column being one family of seven tests; the distance column is the family reported in the main text."
#| fig-width: 5.9
#| fig-height: 7.8

grid_long <- site_df |>
  select(all_of(c(names(land_vars), land_variables))) |>
  pivot_longer(all_of(land_variables), names_to = "Measure", values_to = "Value") |>
  pivot_longer(all_of(names(land_vars)), names_to = "Landscape", values_to = "X") |>
  mutate(Measure = factor(variable_labels_land[Measure], levels = unname(variable_labels_land)),
         Landscape = factor(land_vars[Landscape], levels = unname(land_vars)))

rho_lab <- land_cor |>
  mutate(Measure = factor(variable_labels_land[variable], levels = unname(variable_labels_land)),
         Landscape = factor(land_vars[Landscape], levels = unname(land_vars)),
         label = sprintf("rho == %.2f * ',' ~ %s", rho,
                         ifelse(p_adj < 0.001, "italic(p) < 0.001",
                                sprintf("italic(p) == %.3f", p_adj))))

ggplot(grid_long, aes(X, Value)) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = "#6BA8CE", fill = "#CBDCEA", linewidth = 0.6, linetype = "dashed") +
  geom_point(color = "#565693", alpha = 0.65, size = 1.2, shape = 16) +
  geom_text(data = rho_lab, aes(x = -Inf, y = Inf, label = label), parse = TRUE,
            hjust = -0.05, vjust = 1.3, size = 2, color = "#202124", inherit.aes = FALSE) +
  facet_grid(Measure ~ Landscape, scales = "free") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.18))) +
  labs(x = NULL, y = NULL) +
  fig_theme + theme(panel.grid.major = element_blank(),
                    strip.text.y = element_text(size = 7, angle = -90))
```

\newpage

```{r}
#| label: fig-basin-boxplot
#| fig-cap: "Each of the seven technological measures between the two basins, at the scale of the locality: one panel per measure, each point one locality entered as the median of its Quina scrapers, the box giving the median and the quartiles with whiskers reaching 1.5 times the interquartile range, and the filled point the mean."
#| fig-width: 5.9
#| fig-height: 5.4

basin_long <- bdf |>
  select(Basin, all_of(land_variables)) |>
  mutate(Basin = droplevels(Basin)) |>
  pivot_longer(all_of(land_variables), names_to = "Measure", values_to = "Value") |>
  mutate(Measure = factor(variable_labels_land[Measure], levels = unname(variable_labels_land)))

basin_brackets <- basin_cmp |>
  transmute(Measure = factor(variable_labels_land[variable], levels = unname(variable_labels_land)),
            group1 = "Binchuan", group2 = "Huangping", label = p.adj.signif) |>
  left_join(basin_long |> group_by(Measure) |>
              summarise(ymax = max(Value, na.rm = TRUE),
                        yrange = diff(range(Value, na.rm = TRUE)), .groups = "drop"),
            by = "Measure") |>
  mutate(y.position = ymax + yrange * 0.10)

ggplot(basin_long, aes(Basin, Value)) +
  geom_jitter(aes(color = Basin), size = 1.7, alpha = 0.6, shape = 16,
              position = position_jitter(width = 0.22, height = 0, seed = 2226)) +
  geom_boxplot(color = "#202124", fill = NA, width = 0.55, linewidth = 0.4,
               outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 16, size = 2, color = "#202124") +
  stat_pvalue_manual(basin_brackets, label = "label", y.position = "y.position",
                     tip.length = 0.012, bracket.size = 0.3, label.size = 2.5,
                     color = "#202124") +
  facet_wrap(~ Measure, scales = "free_y", ncol = 3) +
  scale_color_manual(values = basin_colors, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.16))) +
  labs(x = NULL, y = NULL) +
  fig_theme +
  theme(panel.grid.major = element_blank())
```

\newpage

# References {.unnumbered}

::: {#refs}
:::
