---
title: "Supplementary material 1"
subtitle: "The Quina Landscape: a regional Middle Palaeolithic technological system on the southeastern margin of the Tibetan Plateau"
author:
  - Pei-Yuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Qi-Jun Ruan:
      correspondence: "yes"
      orcid: 0009-0000-2143-5335
      institute:
        - YPICRA
  - Davide Delpiano:
      correspondence: "no"
      institute:
        - FERRARA
  - Marco Peresani:
      correspondence: "no"
      orcid: 0000-0001-6562-6336
      institute:
        - FERRARA
        - IGAG
  - Zhen-Xiu Jia:
      correspondence: "no"
      orcid: 0000-0002-7514-4514
      institute:
        - ALPHA
  - Li-Jing Yang:
      correspondence: "no"
      institute:
        - BCRMI
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
  - Hao Li:
      correspondence: "yes"
      email: lihao@itpcas.ac.cn
      institute:
        - ALPHA
# <!-- TODO: update the author list, ORCIDs and affiliations for THIS paper -->
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - YPICRA: Yunnan Provincial Institute of Cultural Relics and Archaeology, Kunming, Yunnan, China
  - FERRARA: Department of Human Studies, Prehistoric and Anthropological Science Unit, University of Ferrara, Ferrara, Italy
  - IGAG: Consiglio Nazionale delle Ricerche-Institute of Environmental Geology and Geoengineering, Laboratory of Palynology and Palaeoecology, Research Group on Vegetation, Climate and Human Stratigraphy, Milan, Italy
  - BCRMI: Binchuan Cultural Relics Management Institute, Dali Bai Autonomous Prefecture, Yunnan, China
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
crossref:
  fig-title: "Fig. S"
  fig-prefix: "Fig. S"
  tbl-title: "Table S"
  tbl-prefix: "Table S"
# packages.bib is written by the setup chunk from the installed packages
bibliography:
  - references.bib
  - packages.bib
csl: "../templates/elsevier-harvard.csl"
number-sections: false
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
  - quarto
  # runs after the crossref filter, so that "Table S 1" reads "Table S1"
  - ../templates/supplement-numbering.lua
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library(targets)

# The chunks below call ggplot(), mutate() and group_by() directly, so this
# document needs the packages attached in its own session; _analysis.R runs in
# the pipeline's process and hands over objects, not an attached search path.
source(here::here("paper", "_packages.R"))

# knitr evaluates chunks with the working directory set to this document's own
# folder, so the pipeline store has to be named explicitly: an unqualified
# tar_load() looks for paper/_targets/ and fails.
TAR_STORE <- here::here("_targets")

# Every statistical result this document reports is computed by
# paper/_analysis.R, which the pipeline runs as a single target. Loading the
# whole environment leaves the inline references below working unchanged.
tar_load(analysis_env, store = TAR_STORE)
list2env(as.list(analysis_env, all.names = TRUE), envir = globalenv())
fpt <- function(p) sub("^= ", "", vapply(p, fp, character(1)))

# The Bayesian model comes from the pipeline -- the same BARG reference fit the
# manuscript figure draws. Nothing is sampled during a render and nothing can
# be stale. The context keeps its own name because barg_data.R defines
# mod_dat, ROPE_SD and variables, names _analysis.R also uses for its own
# versions of the same things.
tar_load(c(ctx_report, fit_ref, barg_mf), store = TAR_STORE)
barg <- ctx_report
brms_ref   <- fit_ref
brms_coef  <- barg$coef_table(brms_ref)
brms_icc   <- barg$icc_table(brms_ref)
brms_n     <- barg$n_spec
brms_n_loc <- barg$n_loc
brms_cri_excl_zero <- sum(brms_coef$CrI_lo > 0 | brms_coef$CrI_hi < 0)
# convergence is quoted from the manifest the pipeline builds beside the fits
brms_manifest <- barg_mf
brms_diag_ref <- with(brms_manifest, diagnostics[diagnostics$fit == "ref", ])
brms_div       <- brms_diag_ref$divergent
brms_treedepth <- brms_diag_ref$max_treedepth
brms_max_rhat  <- brms_diag_ref$max_rhat
brms_min_ess   <- min(brms_diag_ref$min_ess_bulk, brms_diag_ref$min_ess_tail)
group_colors <- c(SC_Quina = "#E07C90", LT_Quina = "#E6C25C", LT_Ordinary = "#6BA8CE")
group_labels <- c(SC_Quina = "Surface Quina", LT_Quina = "Longtan Quina",
                  LT_Ordinary = "Longtan non-Quina")
basin_colors   <- c(Binchuan = "#6BA8CE", Huangping = "#E07C90")
grad_low       <- "#6BA8CE"
grad_high      <- "#E07C90"

fig_theme <- theme_minimal(base_size = 9) +
  theme(panel.grid.major = element_line(color = "#E6E8EB", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.5),
        axis.ticks = element_line(color = "#202124", linewidth = 0.3),
        axis.ticks.length = unit(2, "pt"),
        axis.title = element_text(size = 9), axis.text = element_text(color = "#303238", size = 8),
        strip.text = element_text(face = "bold", color = "#202124", size = 8),
        strip.background = element_rect(fill = "#E8E8E8", color = NA),
        legend.title = element_blank(), legend.text = element_text(size = 8),
        legend.key = element_blank(), plot.tag = element_text(face = "bold", size = 11),
        plot.background = element_rect(color = NA, fill = "white"),
        panel.background = element_rect(color = NA, fill = "white"))

si_pkg_tbl <- tribble(
  ~Package,     ~Citation,           ~Purpose,
  "vegan",      "[@R-vegan]",        "Distance matrices, PERMANOVA and PERMDISP",
  "WdStar",     "[@R-WdStar]",       "Welch-type distance-based MANOVA, the *W*~d~\\* statistic",
  "TOSTER",     "[@R-TOSTER]",       "Two one-sided tests, the specimen-level equivalence cross-check",
  "rstatix",    "[@R-rstatix]",      "Rank and Welch tests, post hoc comparisons and effect sizes",
  "cvequality", "[@R-cvequality]",   "Krishnamoorthy–Lee test for equality of coefficients of variation",
  "brms",       "[@R-brms]",         "Bayesian multivariate multilevel model at the specimen level",
  "posterior",  "[@R-posterior]",    "Posterior summaries and convergence diagnostics for that model",
  "MASS",       "[@MASS2002]",       "Kernel density for the cobble form contours",
  "readxl",     "[@R-readxl]",       "Reading the recording sheets",
  "dplyr",      "[@R-dplyr]",        "Data manipulation",
  "tidyr",      "[@R-tidyr]",        "Reshaping between wide and long form",
  "here",       "[@R-here]",         "Project-relative file paths",
  "ggplot2",    "[@ggplot22016]",    "Graphics",
  "ggpubr",     "[@R-ggpubr]",       "Significance brackets on the distribution panels",
  "ggrepel",    "[@R-ggrepel]",      "Labels placed clear of the points they name",
  "patchwork",  "[@R-patchwork]",    "Composition of the multi-panel figures",
  "scales",     "[@R-scales]",       "Axis and colour-scale transformation",
  "knitr",      "[@R-knitr]",        "Tables and document rendering") |>
  mutate(Version = vapply(Package, function(p) as.character(packageVersion(p)),
                          character(1)), .after = Package)

si_bib <- here("paper", "packages.bib")
knitr::write_bib(c("base", si_pkg_tbl$Package), file = si_bib)
si_lines <- readLines(si_bib)
si_ttl   <- grepl("^  title = ", si_lines)
si_lines[si_ttl] <- sub("^(  title = \\{)([^{:]+)(:)", "\\1{\\2}\\3",
                        si_lines[si_ttl])
si_lines[si_ttl] <- gsub(" ([A-Z])([ ,}])", " {\\1}\\2", si_lines[si_ttl])
writeLines(si_lines, si_bib)
n_fmt <- function(x) format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
```

```{r}
#| label: si-analysis
#| include: false

# quantities needed only in this document, from the objects _analysis.R leaves

# ---- 1. PERMDISP alongside each raw-material PERMANOVA ----
permdisp_one <- function(D, g) {
  g <- droplevels(factor(g))
  set.seed(2226); bd <- suppressWarnings(betadisper(D, g))
  set.seed(2226); pt <- permutest(bd, permutations = PERM)
  c(F = pt$tab$F[1], p = pt$tab$`Pr(>F)`[1])
}
pd_cobbleLoc   <- permdisp_one(C_D, cobbles$Loc)
pd_cobbleRiver <- permdisp_one(C_D, cobbles$river_ID)
pd_usedA_basin <- permdisp_one(A_D, A_meta$basin)
pd_usedA_river <- permdisp_one(A_D, A_meta$river_ID)

# ---- 2. Welch post-hoc statistics ----
we_t <- variable_long |> filter(Variable %in% welch_vars) |>
  mutate(Variable = droplevels(Variable)) |> group_by(Variable) |>
  t_test(Value ~ Group, var.equal = FALSE, p.adjust.method = "bonferroni") |> ungroup()
we_ph_full <- we_ph |> select(Variable, group1, group2, p.adj, p.adj.signif) |>
  left_join(we_t |> select(Variable, group1, group2, statistic, df, p.adj.t = p.adj),
            by = c("Variable", "group1", "group2"))
stopifnot(isTRUE(all.equal(we_ph_full$p.adj, we_ph_full$p.adj.t)))

# ---- 3. The field record of the surveyed cobbles ----
# Tabulated as the survey record; nothing is tested on them, the form analysis
# taking form from the three axes instead.
round_labels <- c("0 (angular)", "1", "2", "3 (rounded)")
cob <- read_excel(basin_path, sheet = "Sheet1")
names(cob) <- trimws(names(cob))
cob <- cob |>
  transmute(Point = factor(paste0("Loc", Loc)),
            Material = factor(harmonise_lithology(Lithology), levels = material_levels),
            Round_num = as.integer(Roundness),
            Roundness = factor(Round_num, levels = 0:3, labels = round_labels,
                               ordered = TRUE),
            Shape = factor(trimws(as.character(Shape)))) |>
  filter(!is.na(Material), !is.na(Roundness), !is.na(Shape)) |>
  droplevels()
n_cob_round <- nrow(cob)

# `cl` from _analysis.R already drops cobbles without a complete set of axes
axis_labels <- c(L = "Length", B = "Width", Th = "Thickness")
n_cob_axes  <- nrow(cl)

# count with row percentage; a label gives the block a bold heading row
ct_cells <- function(tb, label = NULL) {
  pct  <- 100 * prop.table(tb, 1)
  body <- data.frame(Row = paste0(if (is.null(label)) "" else IND, rownames(tb)),
                     check.names = FALSE)
  for (j in colnames(tb))
    body[[j]] <- sprintf("%d (%s%%)", tb[, j], f(pct[, j], 1))
  body$n <- as.character(rowSums(tb))
  if (is.null(label)) return(body)
  head <- body[1, ]; head[] <- ""; head$Row <- sprintf("**%s**", label)
  bind_rows(head, body)
}
```

\newpage

# Introduction

This document accompanies the main text and provides the complete statistical output underlying the reported results, together with the diagnostic and robustness analyses that support their interpretation. It is divided into three parts.

The first part describes the computational environment and parameter settings in detail, ensuring that every reported value can be reproduced exactly. The second part presents the tables and figures, which are arranged in the order in which the results are discussed in the main text and grouped into four sections: the techno-typological characterisation of the retouch products; the technological consistency of the surface collections with the excavated assemblage from Longtan; the organisation of raw material use; and the landscape structure of technological variation. Each section begins with a brief guide to its contents, and each table and figure caption is written to be self-contained, allowing the materials to be understood without reference to the main text. The third part carries a single sensitivity analysis that cuts across those sections, repeating every PERMANOVA reported here and in the main text with a statistic that does not assume the compared groups to be equally dispersed.

\newpage

# Part 1: Computational environment and settings

Every value reported here is computed from the same data as presented in the main manuscript, in the same analysis run, and using the same fixed random seed (2226) as its corresponding value in the main text. The data and code required to reproduce this document are deposited with the research compendium at OSF.

All analyses were conducted in `r R.version.string` [@R-base], with the packages given in @tbl-software.

```{r}
#| label: tbl-software
#| tbl-cap: "The R packages on which the reported results depend, each with the version used in the run that produced this document, the citation the package asks for, and the part of the analysis it carries. The complete session, including the transitive dependencies not listed here, is recorded in the colophon of the main text."

pad <- function(x, n) formatC(x, width = n, flag = "-")

knitr::kable(si_pkg_tbl, row.names = FALSE, align = c("l", "l", "l", "l"),
             col.names = c(pad("Package", 15), pad("Version", 12),
                           pad("Citation", 23), "Purpose"))
```

The random seed (2226) is set once at the beginning of the analysis file and reset immediately before each permutation, Monte Carlo, or bootstrap procedure. This ensures that the results are reproducible and do not depend on the order in which analyses are executed.

Three resampling settings govern the results reported in Part 2. All permutation tests, including PERMANOVA and PERMDISP, use `r n_fmt(PERM)` permutations. Percentile confidence intervals are based on `r n_fmt(B_BOOT)` bootstrap replicates. These intervals are reported in the main text for two quantities: the between-group ratios of the coefficient of variation and Fano factor, and Cohen's d for the difference between scraper edge angle and the exterior platform angle of resharpening flakes. The Krishnamoorthy–Lee modified signed likelihood ratio test for equality of coefficients of variation is evaluated using `r n_fmt(MSLR_NR)` Monte Carlo iterations. To prevent these simulations from affecting subsequent random-number-dependent results, the random-number generator state is saved and restored around each test. Consequently, the bootstrap confidence intervals are independent of the number or order of Krishnamoorthy–Lee tests performed previously. All families of multiple comparisons are adjusted using the Bonferroni method.

\newpage

# Part 2: Detailed results

## Techno-typological characterisation of the Quina retouch products

This section describes the characteristics of the Quina retouch products. @tbl-retouch-products reports the quantified attributes of the surface-collected Quina scrapers and resharpening flakes. @fig-categorical-attributes shows the categorical attributes of the scrapers, including cortex coverage and position, platform type, and the direction of retouch, expressed as the percentage of specimens in each category. @fig-edge-angle-reduction plots edge angle against three measures of reduction intensity; the corresponding relationships are discussed in the main text.

\newpage

```{r}
#| label: tbl-retouch-products
#| tbl-cap: "Metric and reduction attributes of the Quina scrapers and the resharpening flakes from the surface collections. Not every attribute could be measured on every specimen: the platform attributes are recorded only on those that retain the platform of the blank, and GIUR on all but one scraper. The retouch attributes apply to the scrapers alone and the exterior platform angle to the resharpening flakes alone."

attr_row <- function(label, x, d = 1, ds = d) {
  x  <- as.numeric(x); x <- x[is.finite(x)]
  qs <- unname(quantile(x, c(.25, .75)))
  data.frame(Attribute = paste0(IND, label),
             mn  = f(mean(x), d), sd = f(sd(x), d),
             med = f(median(x), ds),
             iqr = paste0(f(qs[1], ds), "–", f(qs[2], ds)))
}
attr_head <- function(label, n_total) {
  data.frame(Attribute = sprintf("**%s** (*n* = %d)", label, n_total),
             mn = "", sd = "", med = "", iqr = "")
}

retouch_products_tbl <- bind_rows(
  attr_head("Quina scrapers", nrow(q)),
  attr_row("Length (mm)",    q$Length,    1),
  attr_row("Width (mm)",     q$Width,     1),
  attr_row("Thickness (mm)", q$Thickness, 1),
  attr_row("Mass (g)",       q$Mass,      1),
  attr_row("Platform depth (mm)", q$Platform_depth, 1),
  attr_row("Platform width (mm)", q$Platform_width, 1),
  attr_row("IPA (°)",             q$IPA,            1),
  attr_row("Retouch generations", q$Ave_RG,    2),
  attr_row(variable_labels[["N_Scar"]], q$N_Scar, 1, ds = 0),
  attr_row("GIUR",                q$Ave_GIUR,  2),
  attr_row("Edge angle (°)",      q$Edge_Angle, 1),
  attr_row("Retouched perimeter", q$Retouch_length_index, 2),
  attr_head("Resharpening flakes", nrow(rfl)),
  attr_row("Length (mm)",    rfl$Length,    1),
  attr_row("Width (mm)",     rfl$Width,     1),
  attr_row("Thickness (mm)", rfl$Thickness, 1),
  attr_row("Mass (g)",       rfl$Mass,      1),
  attr_row("Platform depth (mm)", rfl$Platform_depth, 1),
  attr_row("Platform width (mm)", rfl$Platform_width, 1),
  attr_row("IPA (°)",             rfl$IPA,            1),
  attr_row("EPA (°)",             rfl$EPA,            1))

knitr::kable(retouch_products_tbl, align = c("l", "r", "r", "r", "r"),
             col.names = c("Attribute", "Mean", "SD", "Median", "Q1–Q3"))
```

\newpage

```{r}
#| label: fig-categorical-attributes
#| fig-cap: "Categorical attributes of the surface Quina scrapers: proportion of the dorsal face covered by cortex, position of that cortex, platform type, and the direction from which retouch was struck. Bars run to the percentage of the specimens recorded for that attribute, and the count and percentage are given at the end of each bar."
#| fig-width: 5.9
#| fig-height: 4.9

cat_panel <- function(x, panel, levels_order = NULL, recode_map = NULL) {
  x <- trimws(as.character(x)); x <- x[!is.na(x)]
  if (!is.null(recode_map)) x <- ifelse(x %in% names(recode_map), recode_map[x], x)
  d <- tibble(Level = x) |> count(Level, name = "n")
  d <- if (is.null(levels_order)) arrange(d, desc(n)) else
    d[match(levels_order, d$Level), ] |> filter(!is.na(Level))
  d |> mutate(pct = 100 * n / sum(n), Panel = panel, rank = row_number())
}
cortex_classes <- c("0", "1-25%", "26-50%", "51-75%", "76-99%")
cat_df <- bind_rows(
  cat_panel(q$Cortex, "Cortex proportion", levels_order = cortex_classes),
  cat_panel(q$Cortex_position, "Cortex position", recode_map = c(Non_cortex = "None")),
  cat_panel(q$Platform_type, "Platform type"),
  cat_panel(q$Tot_Direction, "Retouch direction")) |>
  mutate(Panel = factor(Panel, levels = unique(Panel)))
cat_df$Level <- factor(cat_df$Level,
                       levels = cat_df |> arrange(Panel, desc(rank)) |> pull(Level))

cortex_ramp <- setNames(colorRampPalette(c("#CBDCEA", "#2C5F7C"))(length(cortex_classes)),
                        cortex_classes)
panel_flat <- c(`Cortex position` = "#E07C90", `Platform type` = "#E6C25C",
                `Retouch direction` = "#9B87C4")
cat_df <- cat_df |>
  mutate(bar_fill = ifelse(as.character(Panel) == "Cortex proportion",
                           cortex_ramp[as.character(Level)],
                           panel_flat[as.character(Panel)]))

ggplot(cat_df, aes(pct, Level)) +
  geom_col(aes(fill = bar_fill), width = 0.7) +
  geom_text(aes(label = sprintf("n = %d, %.1f%%", n, pct)),
            hjust = -0.12, size = 2.2, color = "#303238") +
  scale_fill_identity() +
  facet_grid(Panel ~ ., scales = "free_y", space = "free_y") +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25),
                     labels = function(z) paste0(z, "%"),
                     expand = expansion(mult = c(0, 0.24))) +
  labs(x = "Percentage", y = NULL) +
  fig_theme +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 7.5),
        strip.text.y = element_text(size = 7.5, angle = -90))
```

\newpage

```{r}
#| label: fig-edge-angle-reduction
#| fig-cap: "Relationship between Quina scraper edge angle and three measures. Each panel reports Spearman’s ρ and p-value. The dashed line represents an ordinary least-squares linear fit, shown only as a visual guide to the direction of the relationship."
#| fig-width: 5.9
#| fig-height: 2.44

set.seed(2226)
focal_ea    <- "Edge_Angle"
corr_vars   <- c("Retouch_length_index", "Ave_GIUR", "Ave_RG")
corr_labels <- c(Retouch_length_index = "Retouched perimeter",
                 Ave_GIUR = "GIUR", Ave_RG = "Retouch generations")
fmt_p_math  <- function(p) ifelse(p < 0.001, "italic(p) < 0.001",
                                  sprintf("italic(p) == %.3f", p))

quina_ea <- read_excel(sc_path, sheet = "Quina scraper") |>
  mutate(across(all_of(c(focal_ea, corr_vars)), as.numeric))
spearman_one <- function(v) {
  pair <- quina_ea |> select(all_of(c(focal_ea, v))) |> drop_na()
  ct <- suppressWarnings(cor.test(pair[[focal_ea]], pair[[v]], method = "spearman", exact = FALSE))
  data.frame(Variable = v, rho = unname(ct$estimate), p_value = ct$p.value)
}
results_ea <- do.call(rbind, lapply(corr_vars, spearman_one))
results_ea$p_value <- p.adjust(results_ea$p_value, "bonferroni")   # one family of three

long_ea <- quina_ea |> select(all_of(c(focal_ea, corr_vars))) |>
  pivot_longer(all_of(corr_vars), names_to = "Variable", values_to = "Value") |>
  filter(!is.na(.data[[focal_ea]]), !is.na(Value)) |>
  mutate(Variable = factor(Variable, levels = corr_vars),
         x_plot = if_else(Variable == "Ave_RG", Value + runif(n(), -0.11, 0.11), Value))
labels_ea <- results_ea |>
  mutate(Variable = factor(Variable, levels = corr_vars),
         label = sprintf("rho == %.2f * ',' ~ %s", rho, fmt_p_math(p_value)))

ggplot(long_ea, aes(y = .data[[focal_ea]])) +
  geom_smooth(aes(x = Value), method = "lm", formula = y ~ x, se = TRUE,
              color = "#6BA8CE", fill = "#CBDCEA", linewidth = 0.6, linetype = "dashed") +
  geom_point(aes(x = x_plot), color = "#565693", alpha = 0.5, size = 1.4, shape = 16) +
  geom_text(data = labels_ea, aes(x = -Inf, y = Inf, label = label), parse = TRUE,
            hjust = -0.08, vjust = 1.35, size = 2.5, color = "#202124", inherit.aes = FALSE) +
  facet_wrap(~ Variable, scales = "free_x", ncol = 3, labeller = as_labeller(corr_labels)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.14))) +
  scale_x_continuous(expand = expansion(mult = 0.06)) +
  labs(x = NULL, y = "Edge angle (°)") +
  fig_theme + theme(panel.grid.major = element_blank())
```

\newpage

## Technological consistency

This section compares the surface-collected Quina scrapers with the excavated Quina scrapers from Longtan, using the Longtan non-Quina scrapers as a control group. @fig-qq shows the distributions of the six technological variables for each group and provides the basis for assessing the assumptions underlying the statistical comparisons. @tbl-tc-permanova to @tbl-tc-pca provide the detailed results, including pairwise tests of differences in group position and dispersion, tests for individual variables, and the loadings from the ordination. @tbl-tc-sensitivity and @fig-jackknife repeat the analysis after sequentially excluding each locality in turn. @fig-pvalue-curve shows the full equivalence *p*-value curves, and @tbl-toster reports the specimen-level equivalence tests alongside the locality-level analysis.

\newpage

```{r}
#| label: fig-qq
#| fig-cap: "Normal quantile–quantile plots of the six technological variables. In every panel the horizontal axis is the theoretical normal quantile and the vertical axis the sample quantile, and the dashed line passes through the first and third sample quartiles. Inset at the upper left of each panel is a histogram of the same values, plotted in its own right but with its axes suppressed, to give the shape of the distribution alongside its departure from normality."
#| fig-width: 5.9
#| fig-height: 8.4

dist_long <- variable_long |>
  mutate(Variable = factor(variable_labels_bare[as.character(Variable)],
                           levels = unname(variable_labels_bare)),
         Group = factor(group_labels[as.character(Group)],
                        levels = unname(group_labels)))
lab_colors <- setNames(unname(group_colors), unname(group_labels))

INSET_GAP <- unit(1.4, "mm")
inset_theme <- theme_minimal(base_size = 6) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.3),
        axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_rect(fill = "white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        plot.margin = margin(0, 0, 0, 0))

qq_cell <- function(vlab, glab, ylab, carries_legend = FALSE) {
  d <- dist_long[dist_long$Variable == vlab & dist_long$Group == glab, ]
  legend_key <- data.frame(
    x = 0, y = mean(range(d$Value, na.rm = TRUE)),
    Group = factor(levels(dist_long$Group), levels = levels(dist_long$Group)))
  p_hist <- ggplot(d, aes(Value)) +
    geom_histogram(bins = 12, fill = lab_colors[[glab]], alpha = 0.75,
                   color = "#202124", linewidth = 0.1) +
    scale_x_continuous(expand = expansion(mult = 0.03)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
    inset_theme
  p_qq <- ggplot(d, aes(sample = Value, color = Group)) +
    stat_qq(size = 0.7, alpha = 0.6, shape = 16, show.legend = FALSE) +
    stat_qq_line(linewidth = 0.35, color = "#5A5F66", linetype = "dashed",
                 show.legend = FALSE) +
    scale_color_manual(values = unname(group_colors),
                       limits = unname(group_labels), drop = FALSE) +
    scale_x_continuous(expand = expansion(mult = 0.06)) +
    scale_y_continuous(expand = expansion(mult = c(0.06, 0.34))) +
    labs(x = NULL, y = ylab) +
    fig_theme +
    theme(panel.grid.major = element_blank(),
          axis.title.y = element_text(size = 8, face = "bold"),
          axis.text = element_text(size = 6.5)) +
    guides(color = guide_legend(nrow = 1, override.aes = list(size = 1.8, alpha = 1)))
  if (carries_legend) {
    p_qq <- p_qq +
      geom_point(data = legend_key, aes(x = x, y = y, color = Group),
                 inherit.aes = FALSE, alpha = 0, size = 0.7, shape = 16,
                 show.legend = TRUE)
  }
  p_qq + inset_element(p_hist,
                       left   = INSET_GAP,
                       bottom = unit(0.63, "npc"),
                       right  = unit(0.47, "npc"),
                       top    = unit(1, "npc") - INSET_GAP,
                       align_to = "panel", ignore_tag = TRUE)
}

vars_row <- levels(dist_long$Variable)
qq_cells <- unlist(lapply(vars_row, function(v) {
  lapply(seq_along(levels(dist_long$Group)), function(i) {
    qq_cell(v, levels(dist_long$Group)[i], if (i == 1) v else NULL,
            carries_legend = (v == vars_row[1] && i == 1))
  })
}), recursive = FALSE)

wrap_plots(qq_cells, ncol = 3, guides = "collect") &
  theme(legend.position = "bottom", legend.margin = margin(0, 0, 0, 0))
```

\newpage

```{r}
#| label: tbl-tc-permanova
#| tbl-cap: "Pairwise PERMANOVA on the z-scored six-variable technological space, Euclidean distances, 9,999 permutations. Degrees of freedom are given as model, residual. Adjusted *p* is the Bonferroni adjustment across the three comparisons."

pw_label <- function(x) {
  parts <- strsplit(x, " vs ", fixed = TRUE)[[1]]
  paste(group_labels[parts], collapse = " vs ")
}
tc_perm_out <- data.frame(
  Comparison = vapply(tc_pw$Comparison, pw_label, character(1)),
  n = sprintf("%d, %d", tc_pw$n1, tc_pw$n2),
  Df = sprintf("%d, %d", tc_pw$Df, tc_pw$Df_res),
  SS = f(tc_pw$SumOfSqs, 2),
  R2 = f(tc_pw$R2, 3),
  F = f(tc_pw$F, 2),
  p = fpt(tc_pw$p_value),
  padj = fpt(tc_pw$p_adjusted))

knitr::kable(tc_perm_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Comparison", "*n*", "*df*", "SS", "*R*²",
                           "pseudo-*F*", "*p*", "Adjusted *p*"))
```

\newpage

```{r}
#| label: tbl-tc-permdisp
#| tbl-cap: "PERMDISP on the same technological space. The upper panel gives the mean distance of each group to its own centroid, the quantity the test compares; distance is measured in the z-scored space, so its units are standard deviations. The lower panel gives the pairwise tests, each fitted to its pair in turn, with the *p*-value from the permutation of the full three-group model alongside for comparison."

pdd <- permA$distances |> group_by(Group) |>
  summarise(n = n(), m = mean(distance), s = sd(distance), .groups = "drop") |>
  mutate(se = s / sqrt(n))

pd_out <- bind_rows(
  data.frame(Item = "**Distance to centroid, by group**",
             a = "", b = "", c = "", d = "", e = ""),
  data.frame(Item = paste0(IND, group_labels[as.character(pdd$Group)]),
             a = as.character(pdd$n), b = f(pdd$m, 3), c = f(pdd$s, 3),
             d = f(pdd$se, 3), e = ""),
  data.frame(Item = paste0(IND, "*Overall test*"), a = "",
             b = sprintf("%d, %d", permA$overall_df, permA$overall_df_res),
             c = f(permA$overall_F, 2), d = fpt(permA$overall_p), e = ""),
  data.frame(Item = "**Pairwise tests**", a = "", b = "", c = "", d = "", e = ""),
  data.frame(Item = paste0(IND, vapply(permA$pair_tab$Comparison, function(x) {
               paste(group_labels[strsplit(x, "-", fixed = TRUE)[[1]]], collapse = " vs ")
             }, character(1))),
             a = sprintf("%d, %d", permA$pair_tab$n1, permA$pair_tab$n2),
             b = sprintf("%d, %d", permA$pair_tab$Df, permA$pair_tab$Df_res),
             c = f(permA$pair_tab$F, 2),
             d = fpt(permA$pair_tab$p),
             e = fpt(permA$pairwise[permA$pair_tab$Comparison])))

knitr::kable(pd_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("", "*n*", "Mean / *df*", "SD / *F*",
                           "SE / *p*", "*p* (3-group)"))
```

\newpage

```{r}
#| label: tbl-tc-univariate
#| tbl-cap: "Per-variable comparison of the three groups: the omnibus test for each variable, followed by its three post-hoc comparisons. GIUR, total scars and retouch generations were tested with Kruskal–Wallis and Dunn's post-hoc comparisons, the effect size being epsilon-squared; thickness, retouched perimeter and edge angle with Welch's ANOVA and Welch's *t* post-hoc comparisons. All post-hoc *p*-values carry a Bonferroni adjustment within each variable. ns, not significant."

uni_rows <- function(v) {
  is_kw <- v %in% kw_vars
  if (is_kw) {
    o <- kw_omni[kw_omni$Variable == v, ]; e <- kw_eff[kw_eff$Variable == v, ]
    omni <- data.frame(Item = paste0(IND, "*Kruskal–Wallis*"),
                       stat = sprintf("H = %s", f(o$statistic, 2)),
                       df = as.character(o$df), p = fpt(o$p),
                       eff = sprintf("ε² = %s", f(e$effsize, 3)), sig = "")
    ph <- kw_ph[kw_ph$Variable == v, ]
    post <- data.frame(
      Item = paste0(IND, vapply(seq_len(nrow(ph)), function(i)
        paste(group_labels[c(ph$group1[i], ph$group2[i])], collapse = " vs "), character(1))),
      stat = sprintf("Z = %s", f(ph$statistic, 2)),
      df = sprintf("%d, %d", ph$n1, ph$n2),
      p = fpt(ph$p.adj), eff = "", sig = ph$p.adj.signif)
  } else {
    o <- we_omni[we_omni$Variable == v, ]
    omni <- data.frame(Item = paste0(IND, "*Welch's ANOVA*"),
                       stat = sprintf("F = %s", f(o$statistic, 2)),
                       df = sprintf("%s, %s", o$DFn, f(o$DFd, 1)),
                       p = fpt(o$p), eff = "", sig = "")
    ph <- we_ph_full[we_ph_full$Variable == v, ]
    post <- data.frame(
      Item = paste0(IND, vapply(seq_len(nrow(ph)), function(i)
        paste(group_labels[c(ph$group1[i], ph$group2[i])], collapse = " vs "), character(1))),
      stat = sprintf("t = %s", f(ph$statistic, 2)),
      df = f(ph$df, 1), p = fpt(ph$p.adj), eff = "", sig = ph$p.adj.signif)
  }
  bind_rows(data.frame(Item = sprintf("**%s**", variable_labels[[v]]),
                       stat = "", df = "", p = "", eff = "", sig = ""),
            omni, post)
}
uni_out <- bind_rows(lapply(variables, uni_rows))

knitr::kable(uni_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Variable / comparison", "Statistic", "*df* or *n*",
                           "*p*", "Effect size", ""))
```

\newpage

```{r}
#| label: tbl-tc-pca
#| tbl-cap: "Loadings of all six principal components of the z-scored technological space, with the eigenvalue, the percentage of variance and the cumulative percentage for each."

ld <- permA$pca_loadings
pca_out <- bind_rows(
  data.frame(Item = variable_labels[ld$Variable],
             as.data.frame(lapply(ld[, paste0("PC", 1:6)], f, d = 3))),
  data.frame(Item = "**Eigenvalue**",
             as.data.frame(as.list(setNames(f(permA$sdev^2, 2), paste0("PC", 1:6))))),
  data.frame(Item = "**Variance (%)**",
             as.data.frame(as.list(setNames(f(permA$var_explained, 1), paste0("PC", 1:6))))),
  data.frame(Item = "**Cumulative (%)**",
             as.data.frame(as.list(setNames(f(cumsum(permA$var_explained), 1), paste0("PC", 1:6))))))

knitr::kable(pca_out, row.names = FALSE, align = c("l", rep("r", 6)),
             col.names = c("Variable", paste0("PC", 1:6)))
```

\newpage

```{r}
#| label: tbl-tc-sensitivity
#| tbl-cap: "Sensitivity of the technological consistency result to the surface specimens collected from within the area of the Longtan site. The full set is the analysis reported in the main text, with 165 surface specimens; the reduced set excludes those collected within the site area, leaving 150, and is z-scored on its own. PERMANOVA *p*-values carry the Bonferroni adjustment across the three comparisons."

sens_rows <- function(comp_perm, comp_disp) {
  full_p <- tc_pw[tc_pw$Comparison == comp_perm, ]
  red_p  <- tc_pw_excl[tc_pw_excl$Comparison == comp_perm, ]
  full_d <- permA$pair_tab[permA$pair_tab$Comparison == comp_disp, ]
  red_d  <- permC$pair_tab[permC$pair_tab$Comparison == comp_disp, ]
  label  <- paste(group_labels[strsplit(comp_perm, " vs ", fixed = TRUE)[[1]]],
                  collapse = " vs ")
  bind_rows(
    data.frame(Comparison = sprintf("**%s**", label), Analysis = "", fR2 = "", fF = "",
               fp = "", rR2 = "", rF = "", rp = ""),
    data.frame(Comparison = paste0(IND, "PERMANOVA"), Analysis = "centroid position",
               fR2 = f(full_p$R2, 3), fF = f(full_p$F, 2), fp = fpt(full_p$p_adjusted),
               rR2 = f(red_p$R2, 3),  rF = f(red_p$F, 2),  rp = fpt(red_p$p_adjusted)),
    data.frame(Comparison = paste0(IND, "PERMDISP"), Analysis = "dispersion",
               fR2 = "", fF = f(full_d$F, 2), fp = fpt(full_d$p),
               rR2 = "", rF = f(red_d$F, 2),  rp = fpt(red_d$p)))
}
sens_out <- bind_rows(
  sens_rows("SC_Quina vs LT_Quina",    "SC_Quina-LT_Quina"),
  sens_rows("SC_Quina vs LT_Ordinary", "SC_Quina-LT_Ordinary"),
  sens_rows("LT_Quina vs LT_Ordinary", "LT_Quina-LT_Ordinary"))

knitr::kable(sens_out, row.names = FALSE, align = c("l", "l", rep("r", 6)),
             col.names = c("Comparison", "Tests", "*R*²", "*F*", "*p*",
                           "*R*²", "*F*", "*p*"))
```

\newpage

```{r}
#| label: fig-jackknife
#| fig-cap: "Leave-one-locality-out jackknife of technological consistency. (A) R² from pairwise PERMANOVAs after removing each locality, with the full-sample R² shown by dashed lines. Marker fill indicates the unadjusted p value. (B) Ordination after the eight iterations removing the most specimens, showing group centroids and 95% normal ellipses."
#| fig-width: 7.5
#| fig-height: 6.2
#| out-width: "5.9in"

# Drawn from `jack` and `permA`, both computed in _analysis.R. Nothing is re-run.
INK <- "#202124"; GREY <- "#5A5F66"
CMP <- c("Surface Quina\nvs Longtan Quina", "Surface Quina\nvs Longtan non-Quina")
SIG_COLS <- c("p ≤ 0.05" = "#E07C90", "p > 0.05" = "#6BA8CE")

# ---- panel A: one row per iteration ----
jk <- jack_i
jk_ord  <- jk$Dropped[order(-jk$n_dropped, jk$Dropped)]   # most specimens first
jk$lab  <- factor(jk$Dropped, levels = rev(jk_ord))       # so the largest sits on top
jk_long <- bind_rows(
  data.frame(lab = jk$lab, Comparison = CMP[1], R2 = jk$R2_q, p = jk$p_q),
  data.frame(lab = jk$lab, Comparison = CMP[2], R2 = jk$R2_o, p = jk$p_o)) |>
  mutate(Comparison = factor(Comparison, levels = CMP),
         psig = factor(ifelse(p <= 0.05, names(SIG_COLS)[1], names(SIG_COLS)[2]),
                       levels = names(SIG_COLS)))
jk_vline <- data.frame(Comparison = factor(CMP, levels = CMP),
                       R2full = c(jack$R2_q[1], jack$R2_o[1]))

# x starts at zero in both columns so the near-zero R2 reads as near zero
JK_XMAX <- 0.35
# breaks per column: a common break set collided at 0.010 against 0.0
jk_breaks <- function(lims) {
  if (max(lims, na.rm = TRUE) < 0.05) c(0, 0.005) else c(0, 0.15, 0.30)
}
# the columns are pinned to different x ranges; blank corners give each row its own
jk_blank <- data.frame(
  Comparison = factor(rep(CMP, each = 2), levels = CMP),
  R2 = c(0, 0.01, 0, JK_XMAX),
  lab = factor(levels(jk_long$lab)[1], levels = levels(jk_long$lab)))

# top row: the y scale is free and unlabelled, only the width of the axis is read
jk_dens <- ggplot(jk_long, aes(x = R2)) +
  geom_density(fill = "#E1E3E6", colour = "#9AA0A6", linewidth = 0.3) +
  geom_vline(data = jk_vline, aes(xintercept = R2full), inherit.aes = FALSE,
             linetype = "dashed", linewidth = 0.35, colour = GREY) +
  geom_blank(data = jk_blank, aes(x = R2), inherit.aes = FALSE) +
  scale_x_continuous(breaks = jk_breaks, expand = expansion(mult = c(0, 0.04))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  facet_wrap(~ Comparison, scales = "free") +
  labs(x = NULL, y = NULL, tag = "A") +
  fig_theme +
  theme(strip.text = element_text(face = "bold", colour = INK, size = 6.8,
                                  lineheight = 1.05),
        axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.spacing.x = unit(5, "pt"),
        plot.margin = margin(4, 2, 0, 4))

jk_forest <- ggplot(jk_long, aes(x = R2, y = lab)) +
  geom_vline(data = jk_vline, aes(xintercept = R2full), inherit.aes = FALSE,
             linetype = "dashed", linewidth = 0.35, colour = GREY) +
  geom_point(aes(shape = Comparison, fill = psig), size = 1.9,
             colour = "grey25", stroke = 0.3) +
  scale_shape_manual(values = setNames(c(21, 22), CMP), name = NULL) +
  scale_fill_manual(values = SIG_COLS, name = NULL, drop = FALSE) +
  geom_blank(data = jk_blank, aes(x = R2, y = lab), inherit.aes = FALSE) +
  # limits are pinned so no partial layer can reorder the localities
  scale_y_discrete(limits = levels(jk_long$lab),
                   expand = expansion(add = c(0.7, 0.7))) +
  scale_x_continuous(breaks = jk_breaks, expand = expansion(mult = c(0, 0.04))) +
  facet_wrap(~ Comparison, scales = "free_x") +
  guides(fill = guide_legend(override.aes = list(shape = 21, size = 2.4)),
         shape = "none") +
  labs(x = expression(R^2 ~ "after dropping that locality"), y = NULL) +
  fig_theme +
  theme(strip.text = element_blank(), strip.background = element_blank(),
        axis.text.x = element_text(size = 6.5), axis.text.y = element_text(size = 7),
        panel.spacing.x = unit(5, "pt"),
        plot.margin = margin(2, 2, 4, 4))

# ---- panel B: where the surface centroid moves ----
# The ordination is the one _analysis.R fits on the full pooled data; every
# iteration is projected into that fixed basis. Refitting per iteration would let
# eigenvector signs flip and show motion that is an artefact of the fit.
GRP_COLS <- setNames(unname(group_colors), unname(group_labels))
pcs <- permA$pca_scores |>
  transmute(PC1, PC2, Site_ID,
            Group = factor(unname(group_labels[as.character(Group)]),
                           levels = unname(group_labels)))
vexp_jk <- permA$var_explained
SC_LAB  <- unname(group_labels[["SC_Quina"]])

# the eight localities whose iterations remove the most specimens
SENS_N   <- 8
sens_loc <- jk_ord[seq_len(SENS_N)]
# each facet holds the data as it stands AFTER that locality is dropped
sens_dat <- do.call(rbind, lapply(sens_loc, function(l) {
  keep <- !(pcs$Group == SC_LAB & pcs$Site_ID %in% l)
  data.frame(pcs[keep, c("PC1", "PC2", "Group")],
             Facet = factor(l, levels = sens_loc))
}))
sens_cen <- aggregate(cbind(PC1, PC2) ~ Facet + Group, sens_dat, mean)

pc_xr <- range(pcs$PC1); pc_yr <- range(pcs$PC2)
jk_B <- ggplot(sens_dat, aes(PC1, PC2)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.25, colour = GREY) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.25, colour = GREY) +
  geom_point(aes(colour = Group), size = 0.3, alpha = 0.3) +
  stat_ellipse(aes(colour = Group), type = "norm", level = 0.95,
               linewidth = 0.3, show.legend = FALSE) +
  geom_point(data = sens_cen, aes(PC1, PC2, fill = Group), inherit.aes = FALSE,
             shape = 21, size = 1.7, stroke = 0.35, colour = INK,
             show.legend = FALSE) +
  scale_colour_manual(values = GRP_COLS, name = NULL) +
  scale_fill_manual(values = GRP_COLS, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 1.7, alpha = 1))) +
  facet_wrap(~ Facet, ncol = 2) +
  coord_fixed(xlim = c(pc_xr[1] - 0.5, pc_xr[2] + 0.5),
              ylim = c(pc_yr[1] - 0.6, pc_yr[2] + 0.6), expand = FALSE) +
  labs(x = paste0("PC1 (", f(vexp_jk[1], 1), "%)"),
       y = paste0("PC2 (", f(vexp_jk[2], 1), "%)"), tag = "B") +
  fig_theme +
  theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold", colour = INK, size = 6.6),
        axis.text = element_text(size = 6.2),
        panel.spacing = unit(2.5, "pt"),
        plot.margin = margin(4, 4, 4, 2))

# One flat grid, not nested: B spans from the top of the density strip to the foot of the forest
jk_dens + jk_forest + jk_B +
  plot_layout(design = "AC\nBC", widths = c(1, 1.6), heights = c(1, 7),
              guides = "collect") &
  theme(legend.position = "bottom", legend.box = "horizontal",
        legend.justification = "center",
        legend.text = element_text(size = 6.6),
        legend.key.size = unit(0.30, "cm"),
        legend.box.spacing = unit(2, "pt"),
        legend.margin = margin(0, 4, 0, 4))
```

\newpage

```{r}
#| label: fig-pvalue-curve
#| fig-cap: "Equivalence p-value curves for the difference between surface and Longtan Quina groups. (A) Multivariate centroid distance in six-variable z-scored space. (B) Six individual measures on a common ρ scale. (C) The same measures rescaled to their own units. The dashed line marks α = 0.05, the open circle the equivalence threshold, and the solid line Δ = 1, corresponding to the established Quina–non-Quina contrast."
#| fig-width: 5.9
#| fig-height: 7.8

# curves read off `eq_pcurve` from _analysis.R; rho is non-negative, so all are one-sided
INK <- "#202124"; GREY <- "#5A5F66"; TXT <- "#303238"
PAL <- c("#E07C90", "#6BA8CE", "#E6C25C", "#9B87C4", "#6FB98E", "#E69F00")
MULTI_LAB <- "Multivariate (centroid)"

curve_theme <- fig_theme + theme(panel.grid.major.x = element_blank(),
                                 plot.margin = margin(6, 10, 6, 6))

# single variables ordered by Delta at p = 0.05, so colour runs left to right
pc_d05 <- setNames(eq_pc_d05, eq_pc_lev)
uni     <- eq_pc_lev[seq_len(length(variables))]
uni     <- uni[order(pc_d05[uni])]
pc_ord  <- c(uni, MULTI_LAB)
pc_cols <- setNames(c(PAL, INK), pc_ord)
lw_all  <- setNames(c(rep(0.5, length(uni)), 1.1), pc_ord)

pcurve <- eq_pcurve
pcurve$Measure <- factor(as.character(pcurve$Measure), levels = pc_ord)
pc_cross <- data.frame(Measure = factor(pc_ord, levels = pc_ord),
                       Delta = unname(pc_d05[pc_ord]), p = 0.05)
ANN_A <- "alpha = 0.05"

pc_base <- function(dfc, dfx) {
  ggplot(dfc, aes(x = Delta, y = p, colour = Measure)) +
    geom_hline(yintercept = 0.05, linetype = "dashed",
               linewidth = 0.45, colour = GREY) +
    geom_vline(xintercept = 1, linewidth = 0.5, colour = GREY) +
    geom_line(aes(linewidth = Measure)) +
    geom_point(data = dfx, shape = 21, size = 1.9, stroke = 0.6,
               fill = "white", show.legend = FALSE) +
    annotate("text", x = 1.19, y = 0.05, label = ANN_A, hjust = 1, vjust = -0.6,
             size = 2.4, colour = GREY) +
    scale_linewidth_manual(values = lw_all, guide = "none") +
    scale_x_continuous(limits = c(0, 1.2), breaks = seq(0, 1.2, 0.2),
                       expand = expansion(mult = c(0.01, 0.02))) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                       expand = expansion(mult = c(0.01, 0.02))) +
    labs(x = expression("Equivalence bound " * Delta ~ "(units of " * rho * ")"),
         y = "Equivalence test p-value") +
    curve_theme
}

# ---- panel A: the multivariate centroid alone ----
pc_A <- pc_base(pcurve[pcurve$Measure == MULTI_LAB, ],
                pc_cross[pc_cross$Measure == MULTI_LAB, ]) +
  scale_colour_manual(values = pc_cols, guide = "none") +
  labs(x = NULL, tag = "A")

# ---- panel B: the six single variables ----
leg_lab <- setNames(sub(" \\(.*\\)$", "", uni), uni)
pc_B <- pc_base(pcurve[pcurve$Measure %in% uni, ],
                pc_cross[pc_cross$Measure %in% uni, ]) +
  scale_colour_manual(values = pc_cols, breaks = uni, labels = leg_lab, name = NULL) +
  guides(colour = guide_legend(override.aes = list(linewidth = 0.9))) +
  labs(x = NULL, tag = "B") +
  theme(legend.position = "inside",
        legend.position.inside = c(0.99, 0.99),
        legend.justification.inside = c(1, 1),
        legend.background = element_rect(fill = "white", colour = NA),
        legend.text = element_text(size = 6.6, colour = TXT),
        legend.key.width = unit(0.45, "cm"),
        legend.key.height = unit(0.34, "cm"),
        legend.margin = margin(2, 4, 2, 2))

# ---- panel C: the same six curves on their own original-unit axes ----
# rho and the original unit differ by a constant factor, so these are the panel B
# curves with the x axis rescaled; nothing is recomputed.
den_by_lab <- setNames(unname(eq_den0), unname(variable_labels[names(eq_den0)]))
ccurve <- do.call(rbind, lapply(uni, function(m) {
  s <- pcurve[pcurve$Measure == m, ]
  data.frame(Measure = m, Delta_raw = s$Delta * den_by_lab[[m]], p = s$p)
}))
ccurve$Measure <- factor(ccurve$Measure, levels = uni)
unit_suffix <- c(Thickness = " mm", Retouch_length_index = "", Ave_GIUR = "",
                 N_Scar = " scars", Ave_RG = " gen", Edge_Angle = "°")
unit_digits <- c(Thickness = 1, Retouch_length_index = 3, Ave_GIUR = 3,
                 N_Scar = 1, Ave_RG = 2, Edge_Angle = 1)
var_of_lab  <- setNames(names(variable_labels), unname(variable_labels))
c_cross <- data.frame(Measure = factor(uni, levels = uni),
                      Delta_raw = unname(pc_d05[uni] * den_by_lab[uni]), p = 0.05)
c_cross$lab <- vapply(seq_along(uni), function(i) {
  v <- var_of_lab[[uni[i]]]
  paste0(f(c_cross$Delta_raw[i], unit_digits[[v]]), unit_suffix[[v]])
}, character(1))
c_vline <- data.frame(Measure = factor(uni, levels = uni),
                      x = unname(den_by_lab[uni]))
c_vtext <- data.frame(Measure = factor(uni[1], levels = uni),
                      x = unname(den_by_lab[uni[1]]), lab = "Quina vs non-Quina")
pc_C <- ggplot(ccurve, aes(x = Delta_raw, y = p, colour = Measure)) +
  geom_hline(yintercept = 0.05, linetype = "dashed",
             linewidth = 0.45, colour = GREY) +
  geom_vline(data = c_vline, aes(xintercept = x), inherit.aes = FALSE,
             linewidth = 0.5, colour = GREY) +
  geom_line(linewidth = 0.6) +
  geom_point(data = c_cross, shape = 21, size = 1.8, stroke = 0.6, fill = "white") +
  geom_text(data = c_cross, aes(x = Delta_raw, y = p, label = lab),
            inherit.aes = FALSE, hjust = -0.12, vjust = -0.75,
            size = 2.2, colour = TXT) +
  geom_text(data = c_vtext, aes(x = x, y = 0.96, label = lab), inherit.aes = FALSE,
            angle = 90, hjust = 1, vjust = -0.45, size = 2.0, colour = TXT) +
  scale_colour_manual(values = pc_cols, guide = "none") +
  scale_x_continuous(n.breaks = 3, expand = expansion(mult = c(0.01, 0.04))) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25),
                     expand = expansion(mult = c(0.01, 0.02))) +
  facet_wrap(~ Measure, ncol = 3, scales = "free_x") +
  labs(x = NULL, y = "Equivalence test p-value", tag = "C") +
  curve_theme +
  theme(strip.text = element_text(face = "bold", colour = INK, size = 6.6),
        axis.text.x = element_text(size = 6.4),
        panel.spacing.x = unit(5, "pt"))

pc_A / pc_B / pc_C + plot_layout(heights = c(1, 2, 2.2))
```

\newpage

```{r}
#| label: si-toster
#| include: false

# The same bound in both: Delta is the rho bound of @fig-pvalue-curve carried back
# into each measure's own units. What differs is the unit of resampling, and so
# the width of the interval the test inverts.
tost_bound <- setNames(unname(eq_pc_d05[seq_along(variables)]) *
                         unname(eq_den0[variables]), variables)
tost_ci <- function(o) unname(unlist(o$effsize["Raw", c("lower.ci", "upper.ci")]))
tost_p  <- function(o) max(o$TOST[c("TOST Lower", "TOST Upper"), "p.value"])

set.seed(2226)
tost_tbl <- bind_rows(lapply(variables, function(v) {
  bd <- tost_bound[[v]]
  x  <- eq_Msc[, v]; y <- eq_Mlq[, v]
  w  <- TOSTER::t_TOST(x = x, y = y, eqb = bd, var.equal = FALSE)      # Welch
  bt <- TOSTER::boot_t_TOST(x = x, y = y, eqb = bd, R = 1999)          # studentised
  data.frame(Variable = v, bound = bd, diff = unname(eq_diff0[[v]]),
             cl_lo = eq_diff_ci90[1, v], cl_hi = eq_diff_ci90[2, v],
             w_lo = tost_ci(w)[1],  w_hi = tost_ci(w)[2],  w_p = tost_p(w),
             b_lo = tost_ci(bt)[1], b_hi = tost_ci(bt)[2], b_p = tost_p(bt))
}))
tost_tbl <- tost_tbl |>
  mutate(cl_w = cl_hi - cl_lo, w_w = w_hi - w_lo, b_w = b_hi - b_lo,
         ratio_w = w_w / cl_w, ratio_b = b_w / cl_w)

tost_n_narrower <- sum(tost_tbl$ratio_w <= 1)
tost_ratio_med  <- median(tost_tbl$ratio_w)
tost_ratio_min  <- min(tost_tbl$ratio_w); tost_ratio_max <- max(tost_tbl$ratio_w)
# how many of the six each form of the test declares equivalent at this bound
tost_n_w <- sum(tost_tbl$w_p <= 0.05); tost_n_b <- sum(tost_tbl$b_p <= 0.05)
tost_margin <- max(c(tost_tbl$w_p, tost_tbl$b_p)) - 0.05
icc_of <- function(r) 100 * brms_icc$Median[brms_icc$Response == r]
```

\newpage

```{r}
#| label: tbl-toster
#| tbl-cap: "Specimen-level equivalence tests of surface and Longtan Quina groups alongside the locality-level analysis."

tost_dig <- c(Thickness = 2, Retouch_length_index = 3, Ave_GIUR = 3,
              N_Scar = 2, Ave_RG = 2, Edge_Angle = 2)
tost_ci_txt <- function(lo, hi, d) sprintf("%s to %s", f(lo, d), f(hi, d))
tost_out <- data.frame(
  Measure = unname(variable_labels[tost_tbl$Variable]),
  bound   = vapply(seq_len(nrow(tost_tbl)),
                   function(i) f(tost_tbl$bound[i], tost_dig[[tost_tbl$Variable[i]]]),
                   character(1)),
  diff    = vapply(seq_len(nrow(tost_tbl)),
                   function(i) f(tost_tbl$diff[i], tost_dig[[tost_tbl$Variable[i]]]),
                   character(1)),
  cl      = vapply(seq_len(nrow(tost_tbl)),
                   function(i) tost_ci_txt(tost_tbl$cl_lo[i], tost_tbl$cl_hi[i],
                                           tost_dig[[tost_tbl$Variable[i]]]), character(1)),
  we      = vapply(seq_len(nrow(tost_tbl)),
                   function(i) tost_ci_txt(tost_tbl$w_lo[i], tost_tbl$w_hi[i],
                                           tost_dig[[tost_tbl$Variable[i]]]), character(1)),
  wp      = fpt(tost_tbl$w_p),
  bo      = vapply(seq_len(nrow(tost_tbl)),
                   function(i) tost_ci_txt(tost_tbl$b_lo[i], tost_tbl$b_hi[i],
                                           tost_dig[[tost_tbl$Variable[i]]]), character(1)),
  bp      = fpt(tost_tbl$b_p),
  ratio   = f(tost_tbl$ratio_w, 2))

knitr::kable(tost_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Measure", "Δ", "Difference", "90% CI, by locality",
                           "90% CI, Welch", "*p*", "90% CI, bootstrap", "*p*",
                           "Width ratio"))
```
\newpage

## Raw material economy

This section documents the raw materials available on the river gravel bars and their representation in the surface assemblages. Two figures give the field setting: @fig-sample-points shows the locations at which cobbles were sampled, and @fig-thin-section illustrates the petrographic characteristics of trachyte.

The analytical results follow. @tbl-composition reports the raw material composition of the surveyed cobbles and surface Quina scrapers for each valley, and @tbl-electivity quantifies their relationship as an index of selection. @fig-electivity-sensitivity assesses the sensitivity of this index to the sampling scheme by recalculating it after omitting each sampling point and each valley in turn. @tbl-rawmat-permanova tests for differences in raw material composition among sampling points, valleys, and basins, and @tbl-clast-size reports the sizes of the surveyed cobbles by lithology.

The last three tables are the field record of the same cobbles. @tbl-cobble-dimensions, @tbl-cobble-roundness, and @tbl-cobble-shape give the three measured axes, roundness scored using the Powers scale, and the shape category assigned in the field.

\newpage

![The gravel-bar sampling points. (A) to (E) The five points at which the surveyed cobbles were recorded, each with the cleared quadrat marked out by string and a spirit level laid across it for scale. (F) A general view of one of the sampled channels. Every cobble longer than 30 mm exposed at the surface within the quadrat was recorded.](../figures/sample_point_photo.png){#fig-sample-points width="5.9in"}

\newpage

![Petrography of the trachyte. (A) Hand specimens of five of the sampled trachyte cobbles. (B) Photomicrographs of thin sections cut from them, with the magnification and the scale given in each frame. The rock is porphyritic: euhedral tabular to prismatic K-feldspar phenocrysts, with subordinate quartz, augite and clay minerals, are set in a fine-grained groundmass of K-feldspar microlites and devitrified glass.](../figures/thin_section.png){#fig-thin-section width="5.9in"}

\newpage

```{r}
#| label: tbl-composition
#| tbl-cap: "Raw material composition of the river-bar cobbles and of the surface-collected Quina scrapers, by valley. Cells give the count with the percentage of the row in parentheses."

comp_tbl <- comp_fig |>
  mutate(cell = ifelse(n == 0, "0", sprintf("%d (%s%%)", n, f(percent, 1)))) |>
  select(Layer, river_ID, Material, cell) |>
  pivot_wider(names_from = Material, values_from = cell) |>
  left_join(counts_fig, by = c("Layer", "river_ID"))

comp_out <- bind_rows(lapply(layer_levels, function(lay) {
  d <- comp_tbl[comp_tbl$Layer == lay, ]
  bind_rows(
    data.frame(Valley = sprintf("**%s**", lay),
               Trachyte = "", Sandstone = "", Quartz = "", Mudstone = "",
               Andesite = "", N = ""),
    data.frame(Valley = paste0(IND, as.character(d$river_ID)),
               Trachyte = d$Trachyte, Sandstone = d$Sandstone, Quartz = d$Quartz,
               Mudstone = d$Mudstone, Andesite = d$Andesite, N = as.character(d$N)))
}))

knitr::kable(comp_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r"),
             col.names = c("Valley", "Trachyte", "Sandstone", "Quartz",
                           "Mudstone", "Andesite", "*n*"))
```

\newpage

```{r}
#| label: tbl-electivity
#| tbl-cap: "Jacobs' electivity index for every raw material. *D* runs from −1, complete avoidance, through 0, use in proportion to availability, to +1, complete selection."

elect_tbl <- tibble(Material = material_levels) |>
  left_join(comp_avail |> transmute(Material, an = n, ap = p), by = "Material") |>
  left_join(comp_used  |> transmute(Material, un = n, up = r), by = "Material") |>
  left_join(jac |> transmute(Material, D), by = "Material") |>
  mutate(across(c(an, un), ~ replace_na(.x, 0L)),
         across(c(ap, up), ~ replace_na(.x, 0)),
         Label = ifelse(Material == "Andesite", "Andesite†", Material))

elect_out <- bind_rows(
  data.frame(Material = elect_tbl$Label,
             an = as.character(elect_tbl$an), ap = f(elect_tbl$ap, 1),
             un = as.character(elect_tbl$un), up = f(elect_tbl$up, 1),
             D  = f(elect_tbl$D, 2)),
  data.frame(Material = "**Total**",
             an = as.character(sum(elect_tbl$an)), ap = f(100, 1),
             un = as.character(sum(elect_tbl$un)), up = f(100, 1), D = ""))

knitr::kable(elect_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Raw material", "Cobbles *n*", "Cobbles %",
                           "Scrapers *n*", "Scrapers %", "*D*"))
```

\newpage

```{r}
#| label: fig-electivity-sensitivity
#| fig-cap: "Sensitivity of Jacobs' electivity index to gravel survey structure. Jacobs' D is shown across the observed range of gravel compositions, with artefact composition (r) fixed at its region-wide value. Results are grouped by sampling point and valley; shading indicates the observed range of gravel composition, and the dashed line marks the neutral point."
#| fig-width: 5.9
#| fig-height: 6.2

# r, the artefact proportion, is held fixed at the region-wide `comp_used`; only p,
# the proportion among the surveyed cobbles, moves, so every marker lies on one
# curve. Sampling point and valley are two groupings of the same cobbles.
INK <- "#202124"; GREY <- "#5A5F66"
material_colors_fig <- c(Trachyte = "#E07C90", Sandstone = "#6BA8CE",
                         Quartz = "#E6C25C", Mudstone = "#9B87C4",
                         Andesite = "#6FB98E")
pctf <- function(x) paste0(f(100 * x, 1), "%")

jacobs_D  <- function(r, p) (r - p) / (r + p - 2 * r * p)
prop_of   <- function(df, m) mean(df$Material == m)
materials <- c("Trachyte", "Sandstone")
r_obs     <- setNames(comp_used$r[match(materials, comp_used$Material)] / 100, materials)
p_pooled  <- setNames(comp_avail$p[match(materials, comp_avail$Material)] / 100, materials)

# leave-one-out needs at least three groups: with two it repeats the single-group marker
SETS <- c("Pooled", "Leave-one-out", "Single unit")
build_sets <- function(group, unit) {
  ids <- levels(factor(group))
  n   <- length(ids)
  stopifnot(n >= 3)
  pts <- bind_rows(lapply(materials, function(m) bind_rows(
    tibble(Material = m, Set = SETS[1], Unit = NA_character_,
           p = unname(p_pooled[[m]])),
    tibble(Material = m, Set = SETS[2], Unit = ids,
           p = vapply(ids, function(g) prop_of(cobbles[group != g, ], m),
                      numeric(1), USE.NAMES = FALSE)),
    tibble(Material = m, Set = SETS[3], Unit = ids,
           p = vapply(ids, function(g) prop_of(cobbles[group == g, ], m),
                      numeric(1), USE.NAMES = FALSE))))) |>
    mutate(D = jacobs_D(unname(r_obs[Material]), p),
           Set = factor(Set, levels = SETS))
  list(pts = pts, n = n, unit = unit)
}

curve_panel <- function(m, S) {
  col <- unname(material_colors_fig[m])
  r   <- unname(r_obs[[m]])
  crv <- tibble(p = seq(0, 1, length.out = 2001)) |> mutate(D = jacobs_D(r, p))
  dat <- S$pts[S$pts$Material == m, ]
  D_p <- dat$D[dat$Set == SETS[1]]
  rng <- range(dat$p)

  set_shape <- setNames(c(21, 22, 24), SETS)
  set_size  <- setNames(c(2.9, 1.8, 1.8), SETS)
  set_fill  <- setNames(c(col, "white", "white"), SETS)

  # annotations go to the side of the line that the data leave empty
  hi    <- r > 0.5
  lab_h <- if (hi) 1.22 else -0.22
  rug_s <- if (hi) "b" else "t"

  ggplot(crv, aes(p, D)) +
    # the interval the gravel survey actually covers
    annotate("rect", xmin = rng[1], xmax = rng[2], ymin = -Inf, ymax = Inf,
             fill = col, alpha = 0.12) +
    geom_hline(yintercept = 0, colour = INK, linewidth = 0.4) +
    geom_vline(xintercept = r, colour = GREY, linetype = "dashed",
               linewidth = 0.45) +
    annotate("text", x = r, y = 0.40, hjust = lab_h, vjust = 0.5,
             size = 2.3, colour = GREY, lineheight = 1.05,
             label = paste0("neutral point\np = r = ", pctf(r))) +
    geom_line(colour = col, linewidth = 0.85) +
    # rug: marker positions stay readable where the curve markers crowd
    geom_rug(data = dat[dat$Set != SETS[1], ], aes(x = p), inherit.aes = FALSE,
             sides = rug_s, length = unit(3.5, "pt"), colour = GREY,
             linewidth = 0.35) +
    geom_point(data = dat, aes(p, D, shape = Set, fill = Set, size = Set),
               colour = INK, stroke = 0.45) +
    annotate("text", x = unname(p_pooled[[m]]), y = D_p, hjust = 0.5,
             vjust = if (hi) 2.4 else -1.6, size = 2.6,
             colour = INK, fontface = "bold",
             label = paste0("D = ", f(D_p, 2))) +
    scale_shape_manual(values = set_shape, drop = FALSE) +
    scale_fill_manual(values = set_fill, drop = FALSE) +
    scale_size_manual(values = set_size, drop = FALSE, guide = "none") +
    scale_x_continuous(breaks = seq(0, 1, 0.25),
                       labels = function(x) paste0(100 * x, "%"),
                       expand = expansion(mult = c(0.02, 0.02))) +
    scale_y_continuous(breaks = seq(-1, 1, 0.5),
                       expand = expansion(mult = c(0.02, 0.02))) +
    # clip rather than drop: D is exactly +-1 at the ends of the sweep
    coord_cartesian(xlim = c(0, 1), ylim = c(-1, 1)) +
    # the fill scale differs between panels, so the legend is carried by shape
    # alone, with fixed key fills so they still read solid-vs-open
    guides(fill = "none",
           shape = guide_legend(override.aes = list(
             fill = c(INK, "white", "white"), size = unname(set_size)))) +
    labs(title = sprintf("%s · %d %ss", m, S$n, S$unit),
         x = paste0("p, ", tolower(m), " among the surveyed cobbles"),
         y = "Jacobs' electivity index D") +
    fig_theme +
    theme(plot.title = element_text(face = "bold", size = 8.5, colour = INK),
          axis.title = element_text(size = 8),
          plot.margin = margin(5, 8, 3, 5))
}

S_point  <- build_sets(cobbles$Loc,      "sampling point")
S_valley <- build_sets(cobbles$river_ID, "valley")

# upper row: grouped by sampling point. lower row: grouped by valley.
(curve_panel("Trachyte", S_point)  | curve_panel("Sandstone", S_point)) /
  (curve_panel("Trachyte", S_valley) | curve_panel("Sandstone", S_valley)) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom", legend.box.spacing = unit(3, "pt"))
```

\newpage

```{r}
#| label: tbl-rawmat-permanova
#| tbl-cap: "PERMANOVA and PERMDISP on raw material composition, on Bray–Curtis dissimilarities with 9,999 permutations. Sums of squares, degrees of freedom and residual degrees of freedom are those of the PERMANOVA; the last two columns give the pseudo-*F* and *p* of the test of homogeneity of dispersion on the same dissimilarity matrix. The unit is the individual cobble in the first panel and the individual artefact in the second."

perm_row <- function(model, a, pd) {
  data.frame(Model = paste0(IND, model),
             Df = sprintf("%d, %d", a$Df[1], a$Df[2]),
             SS = f(a$SumOfSqs[1], 2),
             R2 = f(a$R2[1], 3),
             F  = f(a$F[1], 2),
             p  = fpt(a$`Pr(>F)`[1]),
             pdF = f(pd[["F"]], 2),
             pdp = fpt(pd[["p"]]))
}
head_row <- function(label) data.frame(Model = sprintf("**%s**", label),
  Df = "", SS = "", R2 = "", F = "", p = "", pdF = "", pdp = "")

rawmat_out <- bind_rows(
  head_row(sprintf("Available river cobbles (*n* = %d)", n_cobble)),
  perm_row("~ sampling point", cobbleLoc,   pd_cobbleLoc),
  perm_row("~ valley",         cobbleRiver, pd_cobbleRiver),
  head_row(sprintf("Surface Quina scrapers (*n* = %d)", nrow(scA))),
  perm_row("~ basin",  usedA_basin, pd_usedA_basin),
  perm_row("~ valley", usedA_river, pd_usedA_river))

knitr::kable(rawmat_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Model", "*df*", "SS", "*R*²", "pseudo-*F*", "*p*",
                           "PERMDISP *F*", "PERMDISP *p*"))
```

\newpage

```{r}
#| label: tbl-clast-size
#| tbl-cap: "Size of the surveyed river cobbles by lithology. Size is the geometric mean of length, breadth and thickness in millimetres, and GM is the geometric mean of that quantity within each lithology."

size_row <- function(m) {
  x <- cl$size[cl$Material == m]
  if (!length(x)) return(NULL)
  qs <- unname(quantile(x, c(.25, .75)))
  data.frame(Material = m, n = as.character(length(x)),
             GM = f(gm(x), 1), Median = f(median(x), 1),
             IQR = sprintf("%s–%s", f(qs[1], 1), f(qs[2], 1)),
             Range = sprintf("%s–%s", f(min(x), 1), f(max(x), 1)))
}
size_out <- bind_rows(lapply(material_levels, size_row))

knitr::kable(size_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Raw material", "*n*", "GM (mm)", "Median (mm)",
                           "Q1–Q3 (mm)", "Range (mm)"))
```

\newpage

```{r}
#| label: tbl-cobble-dimensions
#| tbl-cap: "The three axes of the surveyed river cobbles by lithology, in millimetres, as recorded in the field. The panels are the same cobbles measured on each axis in turn; length, width and thickness are the axes as recorded, not the ordered a ≥ b ≥ c of the form analysis, which derives that ordering from these three. The lower end of the length column reflects the survey threshold, cobbles longer than 30 mm having been recorded, rather than the size distribution of the deposit; one cobble was recorded a little below that threshold and is retained here."

dim_row <- function(v, m) {
  x <- cl[[v]][cl$Material == m]
  x <- x[is.finite(x)]
  if (!length(x)) return(NULL)
  qs <- unname(quantile(x, c(.25, .75)))
  data.frame(Group = paste0(IND, m), n = as.character(length(x)),
             Median = f(median(x), 1),
             IQR = sprintf("%s–%s", f(qs[1], 1), f(qs[2], 1)),
             Range = sprintf("%s–%s", f(min(x), 1), f(max(x), 1)))
}
dim_out <- bind_rows(lapply(names(axis_labels), function(v) bind_rows(
  data.frame(Group = sprintf("**%s**", axis_labels[[v]]),
             n = "", Median = "", IQR = "", Range = ""),
  bind_rows(lapply(material_levels, function(m) dim_row(v, m))))))

knitr::kable(dim_out, row.names = FALSE, align = c("l", "r", "r", "r", "r"),
             col.names = c("Axis / raw material", "*n*", "Median (mm)",
                           "Q1–Q3 (mm)", "Range (mm)"))
```

\newpage

```{r}
#| label: tbl-cobble-roundness
#| tbl-cap: "Roundness of the surveyed river cobbles, recorded in the field on the Powers scale as the integer codes 0 to 3 with 0 angular and 3 rounded. The two panels are the same cobbles partitioned two ways, by lithology and by sampling point. Cells give the count with the percentage of the row in parentheses."

rnd_out <- bind_rows(
  ct_cells(table(cob$Material, cob$Roundness), "By lithology"),
  ct_cells(table(cob$Point, cob$Roundness), "By sampling point"))

knitr::kable(rnd_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Group", round_labels, "*n*"))
```

\newpage

```{r}
#| label: tbl-cobble-shape
#| tbl-cap: "Field-assigned shape of the surveyed river cobbles by lithology. The term is a nominal field record with three categories, and is distinct from the quantitative form analysis reported in the main text, which takes form from the three axes rather than from the field term. Cells give the count with the percentage of the row in parentheses."

shape_out <- ct_cells(table(cob$Material, cob$Shape))

knitr::kable(shape_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r"),
             col.names = c("Raw material", levels(cob$Shape), "*n*"))
```

\newpage

## Landscape structure

This section examines whether Quina scraper technological variation is associated with landscape position. @tbl-brms-icc reports the proportion of variance attributable to differences among localities for each of the seven measures, based on a single multivariate multilevel model fitted to the 166 Quina scrapers from 26 localities. Individual specimens were used rather than locality medians to retain within-locality variation and avoid giving localities with different sample sizes equal weight.

The intraclass correlation was calculated for each posterior draw as the locality-level variance divided by the total variance, with both components evaluated on the response-specific link scale. The resulting ICCs are reported in @tbl-brms-icc. Model specification, priors and convergence diagnostics are reported in Supplementary material 2.

\newpage

```{r}
#| label: tbl-brms-icc
#| tbl-cap: "Intraclass correlations for the attributes from the Bayesian multivariate multilevel model. ICCs represent the proportion of variance attributable to differences among localities, calculated for each posterior draw on the response-specific link scale."

icc_out <- data.frame(
  Response = unname(barg$resp_lab[brms_icc$Response]),
  fam = unname(barg$fam_lab[brms_icc$Family]),
  sd  = f(brms_icc$sd_median, 3),
  icc = f(brms_icc$Median, 3),
  ci  = sprintf("%s to %s", f(brms_icc$CrI_lo, 3), f(brms_icc$CrI_hi, 3)))

knitr::kable(icc_out, row.names = FALSE, align = c("l", "l", "r", "r", "r"),
             col.names = c("Measure", "Likelihood", "SD (locality)", "ICC",
                           "95% CrI"))
```

\newpage

## The localities not analysed

The survey covered more localities than were included in the analysis. @tbl-excluded-localities lists the eleven that were excluded; none yielded Quina retouch products. They are included here to preserve the complete survey record, but no analyses reported in this paper use data from these localities.

\newpage

```{r}
#| label: tbl-excluded-localities
#| tbl-cap: "The eleven surveyed localities that are not analysed in this paper. No Quina retouch products, neither Quina scrapers nor resharpening flakes, were recorded at any of them, so none meets the criterion for inclusion set out in the main text and none enters any analysis reported in this paper; they are listed for completeness of the survey record. All eleven lie in the Binchuan Basin. Landform is the position within the terrace sequence, as in Table 1 of the main text, with the one locality standing well above that sequence entered as a hilltop. Coordinates are degrees, minutes and seconds, elevation is metres above sea level, and height is metres above the local channel. The sheet these rows come from carries no artefact counts and no distance to the channel, neither having been recorded for a locality that yielded nothing to analyse."

dms <- function(x, hemi) {
  tot <- round(x * 3600, 1)
  d   <- floor(tot / 3600)
  m   <- floor((tot - d * 3600) / 60)
  s   <- tot - d * 3600 - m * 60
  sprintf("%d°%02d′%04.1f″%s", d, m, s, hemi)
}

# sheet 2 of the site table: localities that yielded no Quina retouch products
excl_sites <- read_excel(site_path, sheet = 2)
names(excl_sites) <- trimws(names(excl_sites))

excl_out <- excl_sites |>
  transmute(Locality = gsub("_", " ", trimws(as.character(name))),
            Valley   = factor(trimws(as.character(river_ID)), levels = river_levels),
            Landform = trimws(as.character(geomorph)),
            lat  = dms(as.numeric(lat), "N"),
            lon  = dms(as.numeric(lon), "E"),
            elev = as.character(round(as.numeric(elev_m))),
            h    = as.character(round(as.numeric(h_river_m)))) |>
  arrange(Valley) |>
  mutate(Valley = as.character(Valley))

knitr::kable(excl_out, row.names = FALSE,
             align = c("l", "l", "l", "r", "r", "r", "r"),
             col.names = c("Locality", "Valley", "Landform", "Latitude", "Longitude",
                           "Elevation (m)", "Height (m)"))
```

\newpage

# Part 3: Sensitivity of the PERMANOVA results to unequal dispersion

```{r}
#| label: si-wds
#| include: false

# ---- Wd*, the Welch-type correction to PERMANOVA ----
# `wds_of()`, `wds_pair()` and the three technological comparisons come from
# _analysis.R; added here are the raw material designs, the cobble form, and the
# same three comparisons on the reduced set. Last chunk in the document to draw
# random numbers, so nothing in Parts 1-2 depends on it.
wds_cobbleLoc   <- wds_of(C_D, cobbles$Loc)
wds_cobbleRiver <- wds_of(C_D, cobbles$river_ID)
wds_usedA_basin <- wds_of(A_D, A_meta$basin)
wds_usedA_river <- wds_of(A_D, A_meta$river_ID)
wds_form        <- wds_of(D_form, sz$G)
wds_tc_excl <- lapply(tc_pairs, function(pr) wds_pair(amat_excl, complete_excl$Group, pr))

# the Bonferroni family of the reduced set, the counterpart of wds_tc_padj
wds_tc_excl_padj <- p.adjust(vapply(wds_tc_excl, function(x) x[["p"]], numeric(1)),
                             "bonferroni")

# the counts quoted in the text
tc_cmp     <- vapply(tc_pairs, paste, character(1), collapse = " vs ")
wds_perm_p <- c(cobbleLoc$`Pr(>F)`[1], cobbleRiver$`Pr(>F)`[1],
                usedA_basin$`Pr(>F)`[1], usedA_river$`Pr(>F)`[1], form_p,
                tc_pw$p_value[match(tc_cmp, tc_pw$Comparison)],
                tc_pw_excl$p_value[match(tc_cmp, tc_pw_excl$Comparison)])
wds_wd_p <- c(wds_cobbleLoc[["p"]], wds_cobbleRiver[["p"]], wds_usedA_basin[["p"]],
              wds_usedA_river[["p"]], wds_form[["p"]],
              vapply(wds_tc,      function(x) x[["p"]], numeric(1)),
              vapply(wds_tc_excl, function(x) x[["p"]], numeric(1)))
n_wds_total   <- length(wds_wd_p)
n_wds_tested  <- sum(!is.na(wds_wd_p))
n_wds_agree   <- sum((wds_perm_p <= 0.05) == (wds_wd_p <= 0.05), na.rm = TRUE)
wds_basin_undef_pct <- 100 * wds_usedA_basin[["undef"]]
n_sangyuan_scraper  <- sum(A_meta$river_ID == "Sangyuan", na.rm = TRUE)
```

PERMANOVA assesses differences in group position but assumes comparable within-group dispersion. Because PERMDISP indicates unequal dispersion in several designs in Part 2, we repeated the relevant analyses using the Welch-type W~d~* statistic, which accounts for group-specific dispersion [@alekseyenko2016; @hamidi2019], implemented in WdStar [@R-WdStar].

@tbl-wds reports W~d~* alongside every PERMANOVA reported here and in the main text, using the same data and `r n_fmt(PERM)` permutations. Of the `r n_wds_total` models, `r n_wds_tested` were suitable for W~d~*, and all `r n_wds_agree` gave the same conclusions as PERMANOVA. Bonferroni adjustment also leaves all pairwise conclusions unchanged.

W~d~* is defined only for grouping factors, not continuous predictors. It is also undefined when within-group dispersion is zero, as in the Sangyuan valley comparison and in `r f(wds_basin_undef_pct, 1)`% of basin-model permutations. Neither case shows significant dispersion differences by PERMDISP (p `r fp(pd_usedA_basin[["p"]])` and `r fp(pd_usedA_river[["p"]])`), so this does not affect the reported interpretation.

\newpage

```{r}
#| label: tbl-wds
#| tbl-cap: "PERMANOVA results with PERMDISP and the Welch-type W~d~* statistic, using the same dissimilarities, grouping factors, and 9,999 permutations. The PERMANOVA and PERMDISP columns reproduce the results reported in Part 2; p-values are unadjusted, and Bonferroni correction does not change any conclusion. n gives the group sizes in model order. An em dash indicates that W~d~* could not be computed because within-group dispersion was zero."

wnum <- function(x, d = 2) if (is.finite(x)) f(x, d) else "—"
wpv  <- function(p) if (is.na(p)) "—" else fpt(p)
wrow <- function(label, n_txt, pF, pp, dF, dp, w) {
  data.frame(Model = paste0(IND, label), n = n_txt,
             F = wnum(pF), p = wpv(pp), pdF = wnum(dF), pdp = wpv(dp),
             wF = wnum(w[["stat"]]), wp = wpv(w[["p"]]))
}
whead <- function(label) data.frame(Model = sprintf("**%s**", label), n = "",
  F = "", p = "", pdF = "", pdp = "", wF = "", wp = "")
gsz <- function(g) paste(as.integer(table(droplevels(factor(g)))), collapse = ", ")
tc_wrow <- function(pw, pt, wl, i) {
  pr <- tc_pairs[[i]]
  a  <- pw[pw$Comparison == paste(pr, collapse = " vs "), ]
  d  <- pt[pt$Comparison == paste(pr, collapse = "-"), ]
  wrow(paste(group_labels[pr], collapse = " vs "), sprintf("%d, %d", a$n1, a$n2),
       a$F, a$p_value, d$F, d$p, wl[[i]])
}

wds_out <- bind_rows(
  whead("Raw material composition, Bray–Curtis"),
  wrow("Available river cobbles ~ sampling point", gsz(cobbles$Loc),
       cobbleLoc$F[1], cobbleLoc$`Pr(>F)`[1],
       pd_cobbleLoc[["F"]], pd_cobbleLoc[["p"]], wds_cobbleLoc),
  wrow("Available river cobbles ~ valley", gsz(cobbles$river_ID),
       cobbleRiver$F[1], cobbleRiver$`Pr(>F)`[1],
       pd_cobbleRiver[["F"]], pd_cobbleRiver[["p"]], wds_cobbleRiver),
  wrow("Surface Quina scrapers ~ basin", gsz(A_meta$basin),
       usedA_basin$F[1], usedA_basin$`Pr(>F)`[1],
       pd_usedA_basin[["F"]], pd_usedA_basin[["p"]], wds_usedA_basin),
  wrow("Surface Quina scrapers ~ valley", gsz(A_meta$river_ID),
       usedA_river$F[1], usedA_river$`Pr(>F)`[1],
       pd_usedA_river[["F"]], pd_usedA_river[["p"]], wds_usedA_river),
  whead("Cobble form, Aitchison"),
  wrow("Trachyte vs sandstone", gsz(sz$G),
       form_F, form_p, form_disp_F, form_disp_p, wds_form),
  whead("Technological consistency, Euclidean in the z-scored six-measure space"),
  tc_wrow(tc_pw, permA$pair_tab, wds_tc, 1),
  tc_wrow(tc_pw, permA$pair_tab, wds_tc, 2),
  tc_wrow(tc_pw, permA$pair_tab, wds_tc, 3),
  whead("The same, on the reduced set excluding the Longtan site area"),
  tc_wrow(tc_pw_excl, permC$pair_tab, wds_tc_excl, 1),
  tc_wrow(tc_pw_excl, permC$pair_tab, wds_tc_excl, 2),
  tc_wrow(tc_pw_excl, permC$pair_tab, wds_tc_excl, 3))

knitr::kable(wds_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Model or comparison", "*n*", "pseudo-*F*", "*p*",
                           "PERMDISP *F*", "PERMDISP *p*", "*W*~d~\\*", "*p*"))
```

\newpage

# References {.unnumbered}

::: {#refs}
:::
