---
title: "Supplementary material"
subtitle: "The Quina Landscape: a regional Middle Palaeolithic technological system on the southeastern margin of the Tibetan Plateau"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Qijun Ruan:
      correspondence: "yes"
      orcid: 0009-0000-2143-5335
      institute:
        - YPICRA
  - Davide Delpiano:
      correspondence: "no"
      institute:
        - FERRARA
  - Marco Peresani:
      correspondence: "no"
      orcid: 0000-0001-6562-6336
      institute:
        - FERRARA
        - IGAG
  - Zhenxiu Jia:
      correspondence: "no"
      orcid: 0000-0002-7514-4514
      institute:
        - ALPHA
  - Lijing Yang:
      correspondence: "no"
      institute:
        - BCRMI
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
  - Hao Li:
      correspondence: "yes"
      email: lihao@itpcas.ac.cn
      institute:
        - ALPHA
# <!-- TODO: update the author list, ORCIDs and affiliations for THIS paper -->
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - YPICRA: Yunnan Provincial Institute of Cultural Relics and Archaeology, Kunming, Yunnan, China
  - FERRARA: Department of Human Studies, Prehistoric and Anthropological Science Unit, University of Ferrara, Ferrara, Italy
  - IGAG: Consiglio Nazionale delle Ricerche-Institute of Environmental Geology and Geoengineering, Laboratory of Palynology and Palaeoecology, Research Group on Vegetation, Climate and Human Stratigraphy, Milan, Italy
  - BCRMI: Binchuan Cultural Relics Management Institute, Dali Bai Autonomous Prefecture, Yunnan, China
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
crossref:
  fig-title: "Fig. S"
  fig-prefix: "Fig. S"
  tbl-title: "Table S"
  tbl-prefix: "Table S"
# packages.bib is written by the setup chunk from the installed packages
bibliography:
  - references.bib
  - packages.bib
csl: "../templates/elsevier-harvard.csl"
number-sections: false
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
  - quarto
  # runs after the crossref filter, so that "Table S 1" reads "Table S1"
  - ../templates/supplement-numbering.lua
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

source(here::here("paper", "_analysis.R"))
fpt <- function(p) sub("^= ", "", vapply(p, fp, character(1)))
group_colors <- c(SC_Quina = "#E07C90", LT_Quina = "#E6C25C", LT_Ordinary = "#6BA8CE")
group_labels <- c(SC_Quina = "Surface Quina", LT_Quina = "Longtan Quina",
                  LT_Ordinary = "Longtan non-Quina")
basin_colors   <- c(Binchuan = "#6BA8CE", Huangping = "#E07C90")
grad_low       <- "#6BA8CE"
grad_high      <- "#E07C90"

fig_theme <- theme_minimal(base_size = 9) +
  theme(panel.grid.major = element_line(color = "#E6E8EB", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.5),
        axis.ticks = element_line(color = "#202124", linewidth = 0.3),
        axis.ticks.length = unit(2, "pt"),
        axis.title = element_text(size = 9), axis.text = element_text(color = "#303238", size = 8),
        strip.text = element_text(face = "bold", color = "#202124", size = 8),
        strip.background = element_rect(fill = "#E8E8E8", color = NA),
        legend.title = element_blank(), legend.text = element_text(size = 8),
        legend.key = element_blank(), plot.tag = element_text(face = "bold", size = 11),
        plot.background = element_rect(color = NA, fill = "white"),
        panel.background = element_rect(color = NA, fill = "white"))

si_pkg_tbl <- tribble(
  ~Package,     ~Citation,           ~Purpose,
  "vegan",      "[@R-vegan]",        "Distance matrices, PERMANOVA and PERMDISP",
  "WdStar",     "[@R-WdStar]",       "Welch-type distance-based MANOVA, the *W*~d~\\* statistic",
  "TOSTER",     "[@R-TOSTER]",       "Two one-sided tests, the specimen-level equivalence cross-check",
  "rstatix",    "[@R-rstatix]",      "Rank and Welch tests, post hoc comparisons and effect sizes",
  "cvequality", "[@R-cvequality]",   "Krishnamoorthy–Lee test for equality of coefficients of variation",
  "brms",       "[@R-brms]",         "Bayesian multivariate multilevel model at the specimen level",
  "posterior",  "[@R-posterior]",    "Posterior summaries and convergence diagnostics for that model",
  "MASS",       "[@MASS2002]",       "Kernel density for the cobble form contours",
  "readxl",     "[@R-readxl]",       "Reading the recording sheets",
  "dplyr",      "[@R-dplyr]",        "Data manipulation",
  "tidyr",      "[@R-tidyr]",        "Reshaping between wide and long form",
  "here",       "[@R-here]",         "Project-relative file paths",
  "ggplot2",    "[@ggplot22016]",    "Graphics",
  "ggpubr",     "[@R-ggpubr]",       "Significance brackets on the distribution panels",
  "ggrepel",    "[@R-ggrepel]",      "Labels placed clear of the points they name",
  "patchwork",  "[@R-patchwork]",    "Composition of the multi-panel figures",
  "scales",     "[@R-scales]",       "Axis and colour-scale transformation",
  "knitr",      "[@R-knitr]",        "Tables and document rendering") |>
  mutate(Version = vapply(Package, function(p) as.character(packageVersion(p)),
                          character(1)), .after = Package)

si_bib <- here("paper", "packages.bib")
knitr::write_bib(c("base", si_pkg_tbl$Package), file = si_bib)
si_lines <- readLines(si_bib)
si_ttl   <- grepl("^  title = ", si_lines)
si_lines[si_ttl] <- sub("^(  title = \\{)([^{:]+)(:)", "\\1{\\2}\\3",
                        si_lines[si_ttl])
si_lines[si_ttl] <- gsub(" ([A-Z])([ ,}])", " {\\1}\\2", si_lines[si_ttl])
writeLines(si_lines, si_bib)
n_fmt <- function(x) format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
```

```{r}
#| label: si-analysis
#| include: false

# quantities needed only in this document, built from the objects _analysis.R leaves

# ---- 1. PERMDISP alongside each raw-material PERMANOVA --------------------
permdisp_one <- function(D, g) {
  g <- droplevels(factor(g))
  set.seed(2226); bd <- suppressWarnings(betadisper(D, g))
  set.seed(2226); pt <- permutest(bd, permutations = PERM)
  c(F = pt$tab$F[1], p = pt$tab$`Pr(>F)`[1])
}
pd_cobbleLoc   <- permdisp_one(C_D, cobbles$Loc)
pd_cobbleRiver <- permdisp_one(C_D, cobbles$river_ID)
pd_usedA_basin <- permdisp_one(A_D, A_meta$basin)
pd_usedA_river <- permdisp_one(A_D, A_meta$river_ID)

# ---- 2. Welch post-hoc statistics ----------------------------------------
we_t <- variable_long |> filter(Variable %in% welch_vars) |>
  mutate(Variable = droplevels(Variable)) |> group_by(Variable) |>
  t_test(Value ~ Group, var.equal = FALSE, p.adjust.method = "bonferroni") |> ungroup()
we_ph_full <- we_ph |> select(Variable, group1, group2, p.adj, p.adj.signif) |>
  left_join(we_t |> select(Variable, group1, group2, statistic, df, p.adj.t = p.adj),
            by = c("Variable", "group1", "group2"))
stopifnot(isTRUE(all.equal(we_ph_full$p.adj, we_ph_full$p.adj.t)))

# ---- 3. Cobble roundness and field-assigned shape -------------------------
# both were recorded on every surveyed cobble in the field; the quantitative
# form analysis takes form from the axes instead, so neither enters it
round_labels <- c("0 (angular)", "1", "2", "3 (rounded)")
cob <- read_excel(basin_path, sheet = "Sheet1")
names(cob) <- trimws(names(cob))
cob <- cob |>
  transmute(Point = factor(paste0("Loc", Loc)),
            Material = factor(harmonise_lithology(Lithology), levels = material_levels),
            Round_num = as.integer(Roundness),
            Roundness = factor(Round_num, levels = 0:3, labels = round_labels,
                               ordered = TRUE),
            Shape = factor(trimws(as.character(Shape)))) |>
  filter(!is.na(Material), !is.na(Roundness), !is.na(Shape)) |>
  droplevels()
n_cob_round <- nrow(cob)
n_cob_quartz <- sum(cob$Material == "Quartz")
n_cob_mud    <- sum(cob$Material == "Mudstone")

# the tests are on the two lithologies the manuscript compares; quartz and
# mudstone are carried in the tables as counts only
cob_ts <- cob |> filter(Material %in% c("Trachyte", "Sandstone")) |> droplevels()
rnd_nT <- sum(cob_ts$Material == "Trachyte")
rnd_nS <- sum(cob_ts$Material == "Sandstone")

# roundness is ordinal: Mann-Whitney U, with the rank-biserial correlation
# 2U/(n1 n2) - 1, positive when the trachyte cobbles rank the more rounded
rnd_mw <- suppressWarnings(wilcox.test(Round_num ~ Material, data = cob_ts))
rnd_U  <- unname(rnd_mw$statistic)
rnd_rb <- 2 * rnd_U / (rnd_nT * rnd_nS) - 1

# shape is nominal: Fisher exact with a Monte Carlo p, and Cramer's V. Its RNG
# use is insulated, as the Krishnamoorthy-Lee test's is, so that nothing drawn
# later in the document depends on how many replicates it took
shp_tab <- table(cob_ts$Material, cob_ts$Shape)
shp_fisher <- local({
  old <- if (exists(".Random.seed", envir = .GlobalEnv))
    get(".Random.seed", envir = .GlobalEnv) else NULL
  on.exit(if (!is.null(old)) assign(".Random.seed", old, envir = .GlobalEnv))
  set.seed(2226)
  fisher.test(shp_tab, simulate.p.value = TRUE, B = FISHER_B)
})
shp_V <- sqrt(unname(suppressWarnings(chisq.test(shp_tab))$statistic) /
                (sum(shp_tab) * (min(dim(shp_tab)) - 1)))

# count with row percentage: the cell of both cobble-descriptor tables. With a
# label the block is given a bold heading row and its rows are indented under it
ct_cells <- function(tb, label = NULL) {
  pct  <- 100 * prop.table(tb, 1)
  body <- data.frame(Row = paste0(if (is.null(label)) "" else IND, rownames(tb)),
                     check.names = FALSE)
  for (j in colnames(tb))
    body[[j]] <- sprintf("%d (%s%%)", tb[, j], f(pct[, j], 1))
  body$n <- as.character(rowSums(tb))
  if (is.null(label)) return(body)
  head <- body[1, ]; head[] <- ""; head$Row <- sprintf("**%s**", label)
  bind_rows(head, body)
}
```

\newpage

# Introduction

This document accompanies the main text and provides the complete statistical output underlying the reported results, together with the diagnostic and robustness analyses that support their interpretation. It is divided into three parts.

The first part describes the computational environment and parameter settings in detail, ensuring that every reported value can be reproduced exactly. The second part presents the tables and figures, which are arranged in the order in which the results are discussed in the main text and grouped into four sections: the techno-typological characterisation of the retouch products; the technological consistency of the surface collections with the excavated assemblage from Longtan; the organisation of raw material use; and the landscape structure of technological variation. Each section begins with a brief guide to its contents, and each table and figure caption is written to be self-contained, allowing the materials to be understood without reference to the main text. The third part carries a single sensitivity analysis that cuts across those sections, repeating every PERMANOVA reported here and in the main text with a statistic that does not assume the compared groups to be equally dispersed.

\newpage

# Part 1: Computational environment and settings

Every value reported here is computed from the same data as presented in the main manuscript, in the same analysis run, and using the same fixed random seed (2226) as its corresponding value in the main text. The data and code required to reproduce this document are deposited with the research compendium at OSF.

All analyses were conducted in `r R.version.string` [@R-base], with the packages given in @tbl-software.

```{r}
#| label: tbl-software
#| tbl-cap: "The R packages on which the reported results depend, each with the version used in the run that produced this document, the citation the package asks for, and the part of the analysis it carries. The complete session, including the transitive dependencies not listed here, is recorded in the colophon of the main text."

pad <- function(x, n) formatC(x, width = n, flag = "-")

knitr::kable(si_pkg_tbl, row.names = FALSE, align = c("l", "l", "l", "l"),
             col.names = c(pad("Package", 15), pad("Version", 12),
                           pad("Citation", 23), "Purpose"))
```

The random seed (2226) is set once at the beginning of the analysis file and reset immediately before each permutation, Monte Carlo, or bootstrap procedure. This ensures that the results are reproducible and do not depend on the order in which analyses are executed.

Three resampling settings govern the results reported in Part 2. All permutation tests, including PERMANOVA and PERMDISP, use `r n_fmt(PERM)` permutations. Percentile confidence intervals are based on `r n_fmt(B_BOOT)` bootstrap replicates. These intervals are reported in the main text for two quantities: the between-group ratios of the coefficient of variation and Fano factor, and Cohen's d for the difference between scraper edge angle and the exterior platform angle of resharpening flakes. The Krishnamoorthy–Lee modified signed likelihood ratio test for equality of coefficients of variation is evaluated using `r n_fmt(MSLR_NR)` Monte Carlo iterations. To prevent these simulations from affecting subsequent random-number-dependent results, the random-number generator state is saved and restored around each test. Consequently, the bootstrap confidence intervals are independent of the number or order of Krishnamoorthy–Lee tests performed previously. All families of multiple comparisons are adjusted using the Bonferroni method.

\newpage

# Part 2: Detailed results

## Techno-typological characterisation of the retouch products

This section describes the retouch products themselves. @tbl-retouch-products gives the metric and reduction attributes of the surface Quina scrapers and of the resharpening flakes that accompany them, and @fig-categorical-attributes gives the categorical attributes of the scrapers — the cover and position of cortex, the type of platform and the direction from which retouch was struck — as the percentage of specimens in each class. @fig-edge-angle-reduction sets edge angle against the three measures of how far reduction was carried, the relationships summarised in the main text. The typological attribution of the same specimens is reported in the main text.

\newpage

```{r}
#| label: tbl-retouch-products
#| tbl-cap: "Metric and reduction attributes of the Quina scrapers and the resharpening flakes from the surface collections. Not every attribute could be measured on every specimen: the platform attributes are recorded only on those that retain the platform of the blank, and GIUR on all but one scraper. The retouch attributes apply to the scrapers alone and the exterior platform angle to the resharpening flakes alone."

attr_row <- function(label, x, d = 1, ds = d) {
  x  <- as.numeric(x); x <- x[is.finite(x)]
  qs <- unname(quantile(x, c(.25, .75)))
  data.frame(Attribute = paste0(IND, label),
             mn  = f(mean(x), d), sd = f(sd(x), d),
             med = f(median(x), ds),
             iqr = paste0(f(qs[1], ds), "–", f(qs[2], ds)))
}
attr_head <- function(label, n_total) {
  data.frame(Attribute = sprintf("**%s** (*n* = %d)", label, n_total),
             mn = "", sd = "", med = "", iqr = "")
}

retouch_products_tbl <- bind_rows(
  attr_head("Quina scrapers", nrow(q)),
  attr_row("Length (mm)",    q$Length,    1),
  attr_row("Width (mm)",     q$Width,     1),
  attr_row("Thickness (mm)", q$Thickness, 1),
  attr_row("Mass (g)",       q$Mass,      1),
  attr_row("Platform depth (mm)", q$Platform_depth, 1),
  attr_row("Platform width (mm)", q$Platform_width, 1),
  attr_row("IPA (°)",             q$IPA,            1),
  attr_row("Retouch generations", q$Ave_RG,    2),
  attr_row(variable_labels[["N_Scar"]], q$N_Scar, 1, ds = 0),
  attr_row("GIUR",                q$Ave_GIUR,  2),
  attr_row("Edge angle (°)",      q$Edge_Angle, 1),
  attr_row("Retouched perimeter", q$Retouch_length_index, 2),
  attr_head("Resharpening flakes", nrow(rfl)),
  attr_row("Length (mm)",    rfl$Length,    1),
  attr_row("Width (mm)",     rfl$Width,     1),
  attr_row("Thickness (mm)", rfl$Thickness, 1),
  attr_row("Mass (g)",       rfl$Mass,      1),
  attr_row("Platform depth (mm)", rfl$Platform_depth, 1),
  attr_row("Platform width (mm)", rfl$Platform_width, 1),
  attr_row("IPA (°)",             rfl$IPA,            1),
  attr_row("EPA (°)",             rfl$EPA,            1))

knitr::kable(retouch_products_tbl, align = c("l", "r", "r", "r", "r"),
             col.names = c("Attribute", "Mean", "SD", "Median", "Q1–Q3"))
```

\newpage

```{r}
#| label: fig-categorical-attributes
#| fig-cap: "Categorical attributes of the surface Quina scrapers: proportion of the dorsal face covered by cortex, position of that cortex, platform type, and the direction from which retouch was struck. Bars run to the percentage of the specimens recorded for that attribute, and the count and percentage are given at the end of each bar."
#| fig-width: 5.9
#| fig-height: 4.9

cat_panel <- function(x, panel, levels_order = NULL, recode_map = NULL) {
  x <- trimws(as.character(x)); x <- x[!is.na(x)]
  if (!is.null(recode_map)) x <- ifelse(x %in% names(recode_map), recode_map[x], x)
  d <- tibble(Level = x) |> count(Level, name = "n")
  d <- if (is.null(levels_order)) arrange(d, desc(n)) else
    d[match(levels_order, d$Level), ] |> filter(!is.na(Level))
  d |> mutate(pct = 100 * n / sum(n), Panel = panel, rank = row_number())
}
cortex_classes <- c("0", "1-25%", "26-50%", "51-75%", "76-99%")
cat_df <- bind_rows(
  cat_panel(q$Cortex, "Cortex proportion", levels_order = cortex_classes),
  cat_panel(q$Cortex_position, "Cortex position", recode_map = c(Non_cortex = "None")),
  cat_panel(q$Platform_type, "Platform type"),
  cat_panel(q$Tot_Direction, "Retouch direction")) |>
  mutate(Panel = factor(Panel, levels = unique(Panel)))
cat_df$Level <- factor(cat_df$Level,
                       levels = cat_df |> arrange(Panel, desc(rank)) |> pull(Level))

cortex_ramp <- setNames(colorRampPalette(c("#CBDCEA", "#2C5F7C"))(length(cortex_classes)),
                        cortex_classes)
panel_flat <- c(`Cortex position` = "#E07C90", `Platform type` = "#E6C25C",
                `Retouch direction` = "#9B87C4")
cat_df <- cat_df |>
  mutate(bar_fill = ifelse(as.character(Panel) == "Cortex proportion",
                           cortex_ramp[as.character(Level)],
                           panel_flat[as.character(Panel)]))

ggplot(cat_df, aes(pct, Level)) +
  geom_col(aes(fill = bar_fill), width = 0.7) +
  geom_text(aes(label = sprintf("n = %d, %.1f%%", n, pct)),
            hjust = -0.12, size = 2.2, color = "#303238") +
  scale_fill_identity() +
  facet_grid(Panel ~ ., scales = "free_y", space = "free_y") +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25),
                     labels = function(z) paste0(z, "%"),
                     expand = expansion(mult = c(0, 0.24))) +
  labs(x = "Percentage", y = NULL) +
  fig_theme +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 7.5),
        strip.text.y = element_text(size = 7.5, angle = -90))
```

\newpage

```{r}
#| label: fig-edge-angle-reduction
#| fig-cap: "Relationship between Quina scraper edge angle and three measures (proportion of perimeter retouched, GIUR, and number of retouch generations). Each panel reports Spearman’s ρ and p-value. The dashed line represents an ordinary least-squares linear fit, shown only as a visual guide to the direction of the relationship."
#| fig-width: 5.9
#| fig-height: 2.44

set.seed(2226)
focal_ea    <- "Edge_Angle"
corr_vars   <- c("Retouch_length_index", "Ave_GIUR", "Ave_RG")
corr_labels <- c(Retouch_length_index = "Retouched perimeter",
                 Ave_GIUR = "GIUR", Ave_RG = "Retouch generations")
fmt_p_math  <- function(p) ifelse(p < 0.001, "italic(p) < 0.001",
                                  sprintf("italic(p) == %.3f", p))

quina_ea <- read_excel(sc_path, sheet = "Quina scraper") |>
  mutate(across(all_of(c(focal_ea, corr_vars)), as.numeric))
spearman_one <- function(v) {
  pair <- quina_ea |> select(all_of(c(focal_ea, v))) |> drop_na()
  ct <- suppressWarnings(cor.test(pair[[focal_ea]], pair[[v]], method = "spearman", exact = FALSE))
  data.frame(Variable = v, rho = unname(ct$estimate), p_value = ct$p.value)
}
results_ea <- do.call(rbind, lapply(corr_vars, spearman_one))
results_ea$p_value <- p.adjust(results_ea$p_value, "bonferroni")   # one family of three

long_ea <- quina_ea |> select(all_of(c(focal_ea, corr_vars))) |>
  pivot_longer(all_of(corr_vars), names_to = "Variable", values_to = "Value") |>
  filter(!is.na(.data[[focal_ea]]), !is.na(Value)) |>
  mutate(Variable = factor(Variable, levels = corr_vars),
         x_plot = if_else(Variable == "Ave_RG", Value + runif(n(), -0.11, 0.11), Value))
labels_ea <- results_ea |>
  mutate(Variable = factor(Variable, levels = corr_vars),
         label = sprintf("rho == %.2f * ',' ~ %s", rho, fmt_p_math(p_value)))

ggplot(long_ea, aes(y = .data[[focal_ea]])) +
  geom_smooth(aes(x = Value), method = "lm", formula = y ~ x, se = TRUE,
              color = "#6BA8CE", fill = "#CBDCEA", linewidth = 0.6, linetype = "dashed") +
  geom_point(aes(x = x_plot), color = "#565693", alpha = 0.5, size = 1.4, shape = 16) +
  geom_text(data = labels_ea, aes(x = -Inf, y = Inf, label = label), parse = TRUE,
            hjust = -0.08, vjust = 1.35, size = 2.5, color = "#202124", inherit.aes = FALSE) +
  facet_wrap(~ Variable, scales = "free_x", ncol = 3, labeller = as_labeller(corr_labels)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.14))) +
  scale_x_continuous(expand = expansion(mult = 0.06)) +
  labs(x = NULL, y = "Edge angle (°)") +
  fig_theme + theme(panel.grid.major = element_blank())
```

\newpage

## Technological consistency

This section compares the surface Quina scrapers with the excavated Quina scrapers from Longtan, against the Longtan non-Quina scrapers as a control. @fig-qq shows the distribution of each of the six technological variables within each group, the check that stands behind the choice of test. @tbl-tc-permanova to @tbl-tc-pca give the full output of the comparison: the pairwise tests of group position and of group spread, the variable-by-variable tests, the reduction-stage profile drawn from retouch generations, and the loadings of the ordination. @tbl-tc-sensitivity and @fig-jackknife then repeat the analysis with the surface specimens collected from within the area of the Longtan site removed. @fig-pvalue-curve gives the full equivalence *p*-value curves.

Retouch generations is fractional and bounded rather than a count, being the mean number of superimposed retouch layers over a specimen's edges, so the two Quina groups are compared not only in the dispersion of that value but also in the profile it gives across reduction stages (@tbl-tc-rg-profile). The two profiles do not differ (Fisher's exact test with a Monte Carlo *p* over `r n_fmt(FISHER_B)` replicates, *p* `r fp(rg_fisher$p.value)`, Cramér's *V* = `r f(rg_V, 2)`; *n* = `r rg_n[["SC_Quina"]]` surface and `r rg_n[["LT_Quina"]]` Longtan Quina scrapers with a recorded value).

\newpage

```{r}
#| label: fig-qq
#| fig-cap: "Normal quantile–quantile plots of the six technological variables. In every panel the horizontal axis is the theoretical normal quantile and the vertical axis the sample quantile, and the dashed line passes through the first and third sample quartiles. Inset at the upper left of each panel is a histogram of the same values, plotted in its own right but with its axes suppressed, to give the shape of the distribution alongside its departure from normality."
#| fig-width: 5.9
#| fig-height: 8.4

dist_long <- variable_long |>
  mutate(Variable = factor(variable_labels_bare[as.character(Variable)],
                           levels = unname(variable_labels_bare)),
         Group = factor(group_labels[as.character(Group)],
                        levels = unname(group_labels)))
lab_colors <- setNames(unname(group_colors), unname(group_labels))

INSET_GAP <- unit(1.4, "mm")
inset_theme <- theme_minimal(base_size = 6) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.3),
        axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_rect(fill = "white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        plot.margin = margin(0, 0, 0, 0))

qq_cell <- function(vlab, glab, ylab, carries_legend = FALSE) {
  d <- dist_long[dist_long$Variable == vlab & dist_long$Group == glab, ]
  legend_key <- data.frame(
    x = 0, y = mean(range(d$Value, na.rm = TRUE)),
    Group = factor(levels(dist_long$Group), levels = levels(dist_long$Group)))
  p_hist <- ggplot(d, aes(Value)) +
    geom_histogram(bins = 12, fill = lab_colors[[glab]], alpha = 0.75,
                   color = "#202124", linewidth = 0.1) +
    scale_x_continuous(expand = expansion(mult = 0.03)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
    inset_theme
  p_qq <- ggplot(d, aes(sample = Value, color = Group)) +
    stat_qq(size = 0.7, alpha = 0.6, shape = 16, show.legend = FALSE) +
    stat_qq_line(linewidth = 0.35, color = "#5A5F66", linetype = "dashed",
                 show.legend = FALSE) +
    scale_color_manual(values = unname(group_colors),
                       limits = unname(group_labels), drop = FALSE) +
    scale_x_continuous(expand = expansion(mult = 0.06)) +
    scale_y_continuous(expand = expansion(mult = c(0.06, 0.34))) +
    labs(x = NULL, y = ylab) +
    fig_theme +
    theme(panel.grid.major = element_blank(),
          axis.title.y = element_text(size = 8, face = "bold"),
          axis.text = element_text(size = 6.5)) +
    guides(color = guide_legend(nrow = 1, override.aes = list(size = 1.8, alpha = 1)))
  if (carries_legend) {
    p_qq <- p_qq +
      geom_point(data = legend_key, aes(x = x, y = y, color = Group),
                 inherit.aes = FALSE, alpha = 0, size = 0.7, shape = 16,
                 show.legend = TRUE)
  }
  p_qq + inset_element(p_hist,
                       left   = INSET_GAP,
                       bottom = unit(0.63, "npc"),
                       right  = unit(0.47, "npc"),
                       top    = unit(1, "npc") - INSET_GAP,
                       align_to = "panel", ignore_tag = TRUE)
}

vars_row <- levels(dist_long$Variable)
qq_cells <- unlist(lapply(vars_row, function(v) {
  lapply(seq_along(levels(dist_long$Group)), function(i) {
    qq_cell(v, levels(dist_long$Group)[i], if (i == 1) v else NULL,
            carries_legend = (v == vars_row[1] && i == 1))
  })
}), recursive = FALSE)

wrap_plots(qq_cells, ncol = 3, guides = "collect") &
  theme(legend.position = "bottom", legend.margin = margin(0, 0, 0, 0))
```

\newpage

```{r}
#| label: tbl-tc-permanova
#| tbl-cap: "Pairwise PERMANOVA on the z-scored six-variable technological space, Euclidean distances, 9,999 permutations. Degrees of freedom are given as model, residual. Adjusted *p* is the Bonferroni adjustment across the three comparisons."

pw_label <- function(x) {
  parts <- strsplit(x, " vs ", fixed = TRUE)[[1]]
  paste(group_labels[parts], collapse = " vs ")
}
tc_perm_out <- data.frame(
  Comparison = vapply(tc_pw$Comparison, pw_label, character(1)),
  n = sprintf("%d, %d", tc_pw$n1, tc_pw$n2),
  Df = sprintf("%d, %d", tc_pw$Df, tc_pw$Df_res),
  SS = f(tc_pw$SumOfSqs, 2),
  R2 = f(tc_pw$R2, 3),
  F = f(tc_pw$F, 2),
  p = fpt(tc_pw$p_value),
  padj = fpt(tc_pw$p_adjusted))

knitr::kable(tc_perm_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Comparison", "*n*", "*df*", "SS", "*R*²",
                           "pseudo-*F*", "*p*", "Adjusted *p*"))
```

\newpage

```{r}
#| label: tbl-tc-permdisp
#| tbl-cap: "PERMDISP on the same technological space. The upper panel gives the mean distance of each group to its own centroid, the quantity the test compares; distance is measured in the z-scored space, so its units are standard deviations. The lower panel gives the pairwise tests, each fitted to its pair in turn, with the *p*-value from the permutation of the full three-group model alongside for comparison."

pdd <- permA$distances |> group_by(Group) |>
  summarise(n = n(), m = mean(distance), s = sd(distance), .groups = "drop") |>
  mutate(se = s / sqrt(n))

pd_out <- bind_rows(
  data.frame(Item = "**Distance to centroid, by group**",
             a = "", b = "", c = "", d = "", e = ""),
  data.frame(Item = paste0(IND, group_labels[as.character(pdd$Group)]),
             a = as.character(pdd$n), b = f(pdd$m, 3), c = f(pdd$s, 3),
             d = f(pdd$se, 3), e = ""),
  data.frame(Item = paste0(IND, "*Overall test*"), a = "",
             b = sprintf("%d, %d", permA$overall_df, permA$overall_df_res),
             c = f(permA$overall_F, 2), d = fpt(permA$overall_p), e = ""),
  data.frame(Item = "**Pairwise tests**", a = "", b = "", c = "", d = "", e = ""),
  data.frame(Item = paste0(IND, vapply(permA$pair_tab$Comparison, function(x) {
               paste(group_labels[strsplit(x, "-", fixed = TRUE)[[1]]], collapse = " vs ")
             }, character(1))),
             a = sprintf("%d, %d", permA$pair_tab$n1, permA$pair_tab$n2),
             b = sprintf("%d, %d", permA$pair_tab$Df, permA$pair_tab$Df_res),
             c = f(permA$pair_tab$F, 2),
             d = fpt(permA$pair_tab$p),
             e = fpt(permA$pairwise[permA$pair_tab$Comparison])))

knitr::kable(pd_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("", "*n*", "Mean / *df*", "SD / *F*",
                           "SE / *p*", "*p* (3-group)"))
```

\newpage

```{r}
#| label: tbl-tc-univariate
#| tbl-cap: "Per-variable comparison of the three groups: the omnibus test for each variable, followed by its three post-hoc comparisons. GIUR, total scars and retouch generations were tested with Kruskal–Wallis and Dunn's post-hoc comparisons, the effect size being epsilon-squared; thickness, retouched perimeter and edge angle with Welch's ANOVA and Welch's *t* post-hoc comparisons. All post-hoc *p*-values carry a Bonferroni adjustment within each variable. ns, not significant."

uni_rows <- function(v) {
  is_kw <- v %in% kw_vars
  if (is_kw) {
    o <- kw_omni[kw_omni$Variable == v, ]; e <- kw_eff[kw_eff$Variable == v, ]
    omni <- data.frame(Item = paste0(IND, "*Kruskal–Wallis*"),
                       stat = sprintf("H = %s", f(o$statistic, 2)),
                       df = as.character(o$df), p = fpt(o$p),
                       eff = sprintf("ε² = %s", f(e$effsize, 3)), sig = "")
    ph <- kw_ph[kw_ph$Variable == v, ]
    post <- data.frame(
      Item = paste0(IND, vapply(seq_len(nrow(ph)), function(i)
        paste(group_labels[c(ph$group1[i], ph$group2[i])], collapse = " vs "), character(1))),
      stat = sprintf("Z = %s", f(ph$statistic, 2)),
      df = sprintf("%d, %d", ph$n1, ph$n2),
      p = fpt(ph$p.adj), eff = "", sig = ph$p.adj.signif)
  } else {
    o <- we_omni[we_omni$Variable == v, ]
    omni <- data.frame(Item = paste0(IND, "*Welch's ANOVA*"),
                       stat = sprintf("F = %s", f(o$statistic, 2)),
                       df = sprintf("%s, %s", o$DFn, f(o$DFd, 1)),
                       p = fpt(o$p), eff = "", sig = "")
    ph <- we_ph_full[we_ph_full$Variable == v, ]
    post <- data.frame(
      Item = paste0(IND, vapply(seq_len(nrow(ph)), function(i)
        paste(group_labels[c(ph$group1[i], ph$group2[i])], collapse = " vs "), character(1))),
      stat = sprintf("t = %s", f(ph$statistic, 2)),
      df = f(ph$df, 1), p = fpt(ph$p.adj), eff = "", sig = ph$p.adj.signif)
  }
  bind_rows(data.frame(Item = sprintf("**%s**", variable_labels[[v]]),
                       stat = "", df = "", p = "", eff = "", sig = ""),
            omni, post)
}
uni_out <- bind_rows(lapply(variables, uni_rows))

knitr::kable(uni_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Variable / comparison", "Statistic", "*df* or *n*",
                           "*p*", "Effect size", ""))
```

\newpage

```{r}
#| label: tbl-tc-rg-profile
#| tbl-cap: "Reduction-stage profile of the two Quina groups, from retouch generations. Retouch generations is the mean number of superimposed retouch layers over a specimen's edges, so it is fractional and bounded rather than a count, and the groups are compared as a profile across reduction stages rather than by a dispersion statistic. Cut points fall between the integers, so that an integer value sits inside a bin and never on a boundary; starting from the finest such scheme, any bin holding fewer than five specimens in either group is merged into its neighbour. Cells give the count with the percentage of the row in parentheses."

rg_out <- ct_cells(rg_tab)
rg_out[[1]] <- unname(group_labels[rg_out[[1]]])

knitr::kable(rg_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r"),
             col.names = c("Group", rg_bin_labels, "*n*"))
```

\newpage

```{r}
#| label: tbl-tc-pca
#| tbl-cap: "Loadings of all six principal components of the z-scored technological space, with the eigenvalue, the percentage of variance and the cumulative percentage for each."

ld <- permA$pca_loadings
pca_out <- bind_rows(
  data.frame(Item = variable_labels[ld$Variable],
             as.data.frame(lapply(ld[, paste0("PC", 1:6)], f, d = 3))),
  data.frame(Item = "**Eigenvalue**",
             as.data.frame(as.list(setNames(f(permA$sdev^2, 2), paste0("PC", 1:6))))),
  data.frame(Item = "**Variance (%)**",
             as.data.frame(as.list(setNames(f(permA$var_explained, 1), paste0("PC", 1:6))))),
  data.frame(Item = "**Cumulative (%)**",
             as.data.frame(as.list(setNames(f(cumsum(permA$var_explained), 1), paste0("PC", 1:6))))))

knitr::kable(pca_out, row.names = FALSE, align = c("l", rep("r", 6)),
             col.names = c("Variable", paste0("PC", 1:6)))
```

\newpage

```{r}
#| label: tbl-tc-sensitivity
#| tbl-cap: "Sensitivity of the technological consistency result to the surface specimens collected from within the area of the Longtan site. The full set is the analysis reported in the main text, with 165 surface specimens; the reduced set excludes those collected within the site area, leaving 150, and is z-scored on its own. PERMANOVA *p*-values carry the Bonferroni adjustment across the three comparisons."

sens_rows <- function(comp_perm, comp_disp) {
  full_p <- tc_pw[tc_pw$Comparison == comp_perm, ]
  red_p  <- tc_pw_excl[tc_pw_excl$Comparison == comp_perm, ]
  full_d <- permA$pair_tab[permA$pair_tab$Comparison == comp_disp, ]
  red_d  <- permC$pair_tab[permC$pair_tab$Comparison == comp_disp, ]
  label  <- paste(group_labels[strsplit(comp_perm, " vs ", fixed = TRUE)[[1]]],
                  collapse = " vs ")
  bind_rows(
    data.frame(Comparison = sprintf("**%s**", label), Analysis = "", fR2 = "", fF = "",
               fp = "", rR2 = "", rF = "", rp = ""),
    data.frame(Comparison = paste0(IND, "PERMANOVA"), Analysis = "centroid position",
               fR2 = f(full_p$R2, 3), fF = f(full_p$F, 2), fp = fpt(full_p$p_adjusted),
               rR2 = f(red_p$R2, 3),  rF = f(red_p$F, 2),  rp = fpt(red_p$p_adjusted)),
    data.frame(Comparison = paste0(IND, "PERMDISP"), Analysis = "dispersion",
               fR2 = "", fF = f(full_d$F, 2), fp = fpt(full_d$p),
               rR2 = "", rF = f(red_d$F, 2),  rp = fpt(red_d$p)))
}
sens_out <- bind_rows(
  sens_rows("SC_Quina vs LT_Quina",    "SC_Quina-LT_Quina"),
  sens_rows("SC_Quina vs LT_Ordinary", "SC_Quina-LT_Ordinary"),
  sens_rows("LT_Quina vs LT_Ordinary", "LT_Quina-LT_Ordinary"))

knitr::kable(sens_out, row.names = FALSE, align = c("l", "l", rep("r", 6)),
             col.names = c("Comparison", "Tests", "*R*²", "*F*", "*p*",
                           "*R*²", "*F*", "*p*"))
```

\newpage

```{r}
#| label: fig-jackknife
#| fig-cap: "Leave-one-locality-out jackknife of the technological consistency result. Each iteration removes every surface specimen from one locality and re-runs the two pairwise PERMANOVAs that involve surface material, in the z-scored space of the full pooled data; the standardisation and the ordination are both fitted once, on that full set, and held fixed, so that nothing in the picture moves with the specimens being dropped. (A) *R*² after dropping each locality, one column per comparison and one row per locality, ordered by the number of specimens the iteration removes. The dashed rule is the *R*² of the full sample and the fill of each marker gives the unadjusted *p* of that iteration. Both panels begin at zero, so that the near-zero *R*² of the Quina-to-Quina comparison reads as near zero rather than being stretched across the panel. Above each column is a kernel density of the same iterations on the same axis, filled grey; its vertical scale is free and unlabelled, since nothing is read off the height of the peak, only the width of the axis it occupies, which is the point: every iteration falls within a narrow band. (B) The ordination as it stands after each of the eight iterations that remove the most specimens: the specimens that remain, a 95% normal ellipse for each group, and each group's centroid as a filled marker."
#| fig-width: 7.5
#| fig-height: 6.2
#| out-width: "5.9in"

# Both panels are drawn from objects _analysis.R has already computed: `jack`,
# the leave-one-locality-out PERMANOVA table, and `permA`, the ordination fitted
# once on the full pooled z-scored data. Nothing is re-run here.
INK <- "#202124"; GREY <- "#5A5F66"
CMP <- c("Surface Quina\nvs Longtan Quina", "Surface Quina\nvs Longtan non-Quina")
SIG_COLS <- c("p ≤ 0.05" = "#E07C90", "p > 0.05" = "#6BA8CE")

# ---- panel A: one row per iteration ---------------------------------------
jk <- jack_i
jk_ord  <- jk$Dropped[order(-jk$n_dropped, jk$Dropped)]   # most specimens first
jk$lab  <- factor(jk$Dropped, levels = rev(jk_ord))       # so the largest sits on top
jk_long <- bind_rows(
  data.frame(lab = jk$lab, Comparison = CMP[1], R2 = jk$R2_q, p = jk$p_q),
  data.frame(lab = jk$lab, Comparison = CMP[2], R2 = jk$R2_o, p = jk$p_o)) |>
  mutate(Comparison = factor(Comparison, levels = CMP),
         psig = factor(ifelse(p <= 0.05, names(SIG_COLS)[1], names(SIG_COLS)[2]),
                       levels = names(SIG_COLS)))
jk_vline <- data.frame(Comparison = factor(CMP, levels = CMP),
                       R2full = c(jack$R2_q[1], jack$R2_o[1]))

# The x axis starts at zero in both columns, so that the near-zero R2 of the
# Quina-to-Quina comparison reads as near zero rather than being stretched
# across the panel by an automatic scale.
JK_XMAX <- 0.35
# Breaks are chosen per column: with a common break set the left column ended
# on 0.010 right where the right column began on 0.0 and the two labels
# collided. Panel A is narrow, so each column carries only what it can hold.
jk_breaks <- function(lims) {
  if (max(lims, na.rm = TRUE) < 0.05) c(0, 0.005) else c(0, 0.15, 0.30)
}
# The two columns are pinned to different x ranges. The density strip and the
# forest below it are given the same blank corners, so the two rows of panel A
# stand on one x axis per column.
jk_blank <- data.frame(
  Comparison = factor(rep(CMP, each = 2), levels = CMP),
  R2 = c(0, 0.01, 0, JK_XMAX),
  lab = factor(levels(jk_long$lab)[1], levels = levels(jk_long$lab)))

# top row: how tightly the iterations land. The y scale is free and unlabelled;
# nothing is read off the height of the peak, only the width of the axis it
# occupies.
jk_dens <- ggplot(jk_long, aes(x = R2)) +
  geom_density(fill = "#E1E3E6", colour = "#9AA0A6", linewidth = 0.3) +
  geom_vline(data = jk_vline, aes(xintercept = R2full), inherit.aes = FALSE,
             linetype = "dashed", linewidth = 0.35, colour = GREY) +
  geom_blank(data = jk_blank, aes(x = R2), inherit.aes = FALSE) +
  scale_x_continuous(breaks = jk_breaks, expand = expansion(mult = c(0, 0.04))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  facet_wrap(~ Comparison, scales = "free") +
  labs(x = NULL, y = NULL, tag = "A") +
  fig_theme +
  theme(strip.text = element_text(face = "bold", colour = INK, size = 6.8,
                                  lineheight = 1.05),
        axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.spacing.x = unit(5, "pt"),
        plot.margin = margin(4, 2, 0, 4))

# bottom row: the iterations themselves. The column headings sit on the density
# strip above, so this row carries no strip of its own.
jk_forest <- ggplot(jk_long, aes(x = R2, y = lab)) +
  geom_vline(data = jk_vline, aes(xintercept = R2full), inherit.aes = FALSE,
             linetype = "dashed", linewidth = 0.35, colour = GREY) +
  geom_point(aes(shape = Comparison, fill = psig), size = 1.9,
             colour = "grey25", stroke = 0.3) +
  scale_shape_manual(values = setNames(c(21, 22), CMP), name = NULL) +
  scale_fill_manual(values = SIG_COLS, name = NULL, drop = FALSE) +
  geom_blank(data = jk_blank, aes(x = R2, y = lab), inherit.aes = FALSE) +
  # limits are pinned so no partial layer can reorder the localities
  scale_y_discrete(limits = levels(jk_long$lab),
                   expand = expansion(add = c(0.7, 0.7))) +
  scale_x_continuous(breaks = jk_breaks, expand = expansion(mult = c(0, 0.04))) +
  facet_wrap(~ Comparison, scales = "free_x") +
  guides(fill = guide_legend(override.aes = list(shape = 21, size = 2.4)),
         shape = "none") +
  labs(x = expression(R^2 ~ "after dropping that locality"), y = NULL) +
  fig_theme +
  theme(strip.text = element_blank(), strip.background = element_blank(),
        axis.text.x = element_text(size = 6.5), axis.text.y = element_text(size = 7),
        panel.spacing.x = unit(5, "pt"),
        plot.margin = margin(2, 2, 4, 4))

# ---- panel B: where the surface centroid moves ----------------------------
# The ordination is the one fitted in _analysis.R on the full pooled data, and
# every iteration is projected into that fixed basis. Refitting per iteration
# would let the eigenvector signs and order flip between runs and would show
# motion that is an artefact of the fit rather than of the data.
GRP_COLS <- setNames(unname(group_colors), unname(group_labels))
pcs <- permA$pca_scores |>
  transmute(PC1, PC2, Site_ID,
            Group = factor(unname(group_labels[as.character(Group)]),
                           levels = unname(group_labels)))
vexp_jk <- permA$var_explained
SC_LAB  <- unname(group_labels[["SC_Quina"]])

# the eight localities whose iterations remove the most specimens
SENS_N   <- 8
sens_loc <- jk_ord[seq_len(SENS_N)]
# each facet holds the data as it stands AFTER that locality is dropped
sens_dat <- do.call(rbind, lapply(sens_loc, function(l) {
  keep <- !(pcs$Group == SC_LAB & pcs$Site_ID %in% l)
  data.frame(pcs[keep, c("PC1", "PC2", "Group")],
             Facet = factor(l, levels = sens_loc))
}))
sens_cen <- aggregate(cbind(PC1, PC2) ~ Facet + Group, sens_dat, mean)

pc_xr <- range(pcs$PC1); pc_yr <- range(pcs$PC2)
jk_B <- ggplot(sens_dat, aes(PC1, PC2)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.25, colour = GREY) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.25, colour = GREY) +
  geom_point(aes(colour = Group), size = 0.3, alpha = 0.3) +
  stat_ellipse(aes(colour = Group), type = "norm", level = 0.95,
               linewidth = 0.3, show.legend = FALSE) +
  geom_point(data = sens_cen, aes(PC1, PC2, fill = Group), inherit.aes = FALSE,
             shape = 21, size = 1.7, stroke = 0.35, colour = INK,
             show.legend = FALSE) +
  scale_colour_manual(values = GRP_COLS, name = NULL) +
  scale_fill_manual(values = GRP_COLS, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 1.7, alpha = 1))) +
  facet_wrap(~ Facet, ncol = 2) +
  coord_fixed(xlim = c(pc_xr[1] - 0.5, pc_xr[2] + 0.5),
              ylim = c(pc_yr[1] - 0.6, pc_yr[2] + 0.6), expand = FALSE) +
  labs(x = paste0("PC1 (", f(vexp_jk[1], 1), "%)"),
       y = paste0("PC2 (", f(vexp_jk[2], 1), "%)"), tag = "B") +
  fig_theme +
  theme(panel.grid = element_blank(),
        strip.text = element_text(face = "bold", colour = INK, size = 6.6),
        axis.text = element_text(size = 6.2),
        panel.spacing = unit(2.5, "pt"),
        plot.margin = margin(4, 4, 4, 2))

# One flat layout rather than a nested one. The density strip and the forest are
# rows 1 and 2 of the left column and panel B spans both rows, so all three are
# sized against one grid: B's panel area runs from the top of the density strip
# to the foot of the forest.
jk_dens + jk_forest + jk_B +
  plot_layout(design = "AC\nBC", widths = c(1, 1.6), heights = c(1, 7),
              guides = "collect") &
  theme(legend.position = "bottom", legend.box = "horizontal",
        legend.justification = "center",
        legend.text = element_text(size = 6.6),
        legend.key.size = unit(0.30, "cm"),
        legend.box.spacing = unit(2, "pt"),
        legend.margin = margin(0, 4, 0, 4))
```

\newpage

```{r}
#| label: fig-pvalue-curve
#| fig-cap: "Equivalence *p*-value curves for the ratio ρ = |mean(SC Quina) − mean(LT Quina)| / |mean(LT Quina) − mean(LT non-Quina)|. For each measure, *p*(Δ) is the proportion of cluster-bootstrap replicates (*R* = 20,000) with ρ ≥ Δ; equivalence holds at level α when *p* ≤ α. (A) The multivariate centroid (Euclidean distance in z-scored six-variable space). (B) The six single variables on the same ρ axis. (C) The same six curves, one per facet, with the *x* axis rescaled to the variable's own units. In all panels the dashed horizontal rule is α = 0.05, the hollow circle marks the crossing point, and the solid vertical rule is Δ = 1, the bound at which the two Quina groups would differ as much as Quina differs from non-Quina."
#| fig-width: 5.9
#| fig-height: 7.8

# The curves are read straight off `eq_pcurve`, the cluster-bootstrap p-value
# function computed in _analysis.R; no test is re-run per bound. rho is
# non-negative, so every curve here is one-sided.
INK <- "#202124"; GREY <- "#5A5F66"; TXT <- "#303238"
PAL <- c("#E07C90", "#6BA8CE", "#E6C25C", "#9B87C4", "#6FB98E", "#E69F00")
MULTI_LAB <- "Multivariate (centroid)"

# The vertical major grid is dropped: these curves are read off the y axis, so
# the horizontal rules help and the vertical ones only add clutter.
curve_theme <- fig_theme + theme(panel.grid.major.x = element_blank(),
                                 plot.margin = margin(6, 10, 6, 6))

# single variables ordered by Delta at p = 0.05, so colour runs left to right
pc_d05 <- setNames(eq_pc_d05, eq_pc_lev)
uni     <- eq_pc_lev[seq_len(length(variables))]
uni     <- uni[order(pc_d05[uni])]
pc_ord  <- c(uni, MULTI_LAB)
pc_cols <- setNames(c(PAL, INK), pc_ord)
lw_all  <- setNames(c(rep(0.5, length(uni)), 1.1), pc_ord)

pcurve <- eq_pcurve
pcurve$Measure <- factor(as.character(pcurve$Measure), levels = pc_ord)
pc_cross <- data.frame(Measure = factor(pc_ord, levels = pc_ord),
                       Delta = unname(pc_d05[pc_ord]), p = 0.05)
ANN_A <- "alpha = 0.05"

# The Delta = 1 line is left unlabelled in panels A and B and carries its text
# only once, on the first facet of panel C.
pc_base <- function(dfc, dfx) {
  ggplot(dfc, aes(x = Delta, y = p, colour = Measure)) +
    geom_hline(yintercept = 0.05, linetype = "dashed",
               linewidth = 0.45, colour = GREY) +
    geom_vline(xintercept = 1, linewidth = 0.5, colour = GREY) +
    geom_line(aes(linewidth = Measure)) +
    geom_point(data = dfx, shape = 21, size = 1.9, stroke = 0.6,
               fill = "white", show.legend = FALSE) +
    annotate("text", x = 1.19, y = 0.05, label = ANN_A, hjust = 1, vjust = -0.6,
             size = 2.4, colour = GREY) +
    scale_linewidth_manual(values = lw_all, guide = "none") +
    scale_x_continuous(limits = c(0, 1.2), breaks = seq(0, 1.2, 0.2),
                       expand = expansion(mult = c(0.01, 0.02))) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                       expand = expansion(mult = c(0.01, 0.02))) +
    labs(x = expression("Equivalence bound " * Delta ~ "(units of " * rho * ")"),
         y = "Equivalence test p-value") +
    curve_theme
}

# ---- panel A: the multivariate centroid alone ------------------------------
pc_A <- pc_base(pcurve[pcurve$Measure == MULTI_LAB, ],
                pc_cross[pc_cross$Measure == MULTI_LAB, ]) +
  scale_colour_manual(values = pc_cols, guide = "none") +
  labs(x = NULL, tag = "A")

# ---- panel B: the six single variables -------------------------------------
# The legend carries the variable-to-colour mapping only and sits inside the
# panel, in the empty upper right.
leg_lab <- setNames(sub(" \\(.*\\)$", "", uni), uni)
pc_B <- pc_base(pcurve[pcurve$Measure %in% uni, ],
                pc_cross[pc_cross$Measure %in% uni, ]) +
  scale_colour_manual(values = pc_cols, breaks = uni, labels = leg_lab, name = NULL) +
  guides(colour = guide_legend(override.aes = list(linewidth = 0.9))) +
  labs(x = NULL, tag = "B") +
  theme(legend.position = "inside",
        legend.position.inside = c(0.99, 0.99),
        legend.justification.inside = c(1, 1),
        legend.background = element_rect(fill = "white", colour = NA),
        legend.text = element_text(size = 6.6, colour = TXT),
        legend.key.width = unit(0.45, "cm"),
        legend.key.height = unit(0.34, "cm"),
        legend.margin = margin(2, 4, 2, 2))

# ---- panel C: the same six curves on their own original-unit axes ----------
# rho is a difference divided by that variable's denominator, so rho and the
# original unit differ by a constant factor; the curves here are the panel B
# curves with the x axis rescaled, nothing is recomputed.
den_by_lab <- setNames(unname(eq_den0), unname(variable_labels[names(eq_den0)]))
ccurve <- do.call(rbind, lapply(uni, function(m) {
  s <- pcurve[pcurve$Measure == m, ]
  data.frame(Measure = m, Delta_raw = s$Delta * den_by_lab[[m]], p = s$p)
}))
ccurve$Measure <- factor(ccurve$Measure, levels = uni)
unit_suffix <- c(Thickness = " mm", Retouch_length_index = "", Ave_GIUR = "",
                 N_Scar = " scars", Ave_RG = " gen", Edge_Angle = "°")
unit_digits <- c(Thickness = 1, Retouch_length_index = 3, Ave_GIUR = 3,
                 N_Scar = 1, Ave_RG = 2, Edge_Angle = 1)
var_of_lab  <- setNames(names(variable_labels), unname(variable_labels))
c_cross <- data.frame(Measure = factor(uni, levels = uni),
                      Delta_raw = unname(pc_d05[uni] * den_by_lab[uni]), p = 0.05)
c_cross$lab <- vapply(seq_along(uni), function(i) {
  v <- var_of_lab[[uni[i]]]
  paste0(f(c_cross$Delta_raw[i], unit_digits[[v]]), unit_suffix[[v]])
}, character(1))
c_vline <- data.frame(Measure = factor(uni, levels = uni),
                      x = unname(den_by_lab[uni]))
c_vtext <- data.frame(Measure = factor(uni[1], levels = uni),
                      x = unname(den_by_lab[uni[1]]), lab = "Quina vs non-Quina")
pc_C <- ggplot(ccurve, aes(x = Delta_raw, y = p, colour = Measure)) +
  geom_hline(yintercept = 0.05, linetype = "dashed",
             linewidth = 0.45, colour = GREY) +
  geom_vline(data = c_vline, aes(xintercept = x), inherit.aes = FALSE,
             linewidth = 0.5, colour = GREY) +
  geom_line(linewidth = 0.6) +
  geom_point(data = c_cross, shape = 21, size = 1.8, stroke = 0.6, fill = "white") +
  geom_text(data = c_cross, aes(x = Delta_raw, y = p, label = lab),
            inherit.aes = FALSE, hjust = -0.12, vjust = -0.75,
            size = 2.2, colour = TXT) +
  geom_text(data = c_vtext, aes(x = x, y = 0.96, label = lab), inherit.aes = FALSE,
            angle = 90, hjust = 1, vjust = -0.45, size = 2.0, colour = TXT) +
  scale_colour_manual(values = pc_cols, guide = "none") +
  scale_x_continuous(n.breaks = 3, expand = expansion(mult = c(0.01, 0.04))) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25),
                     expand = expansion(mult = c(0.01, 0.02))) +
  facet_wrap(~ Measure, ncol = 3, scales = "free_x") +
  labs(x = NULL, y = "Equivalence test p-value", tag = "C") +
  curve_theme +
  # the six facets are narrow, so the strip and tick text come down a step
  theme(strip.text = element_text(face = "bold", colour = INK, size = 6.6),
        axis.text.x = element_text(size = 6.4),
        panel.spacing.x = unit(5, "pt"))

# Panel B's legend sits inside its own panel, so no guide is collected at the
# assembly level and all three tiers keep the full plot width.
pc_A / pc_B / pc_C + plot_layout(heights = c(1, 2, 2.2))
```

\newpage

```{r}
#| label: si-toster
#| include: false

# The equivalence analysis of this section against the conventional one. The
# bound is the same in both: Delta is the rho bound of @fig-pvalue-curve
# carried back into each measure's own units, rho being a difference divided by
# that measure's denominator. What differs is the unit of resampling, and so
# the width of the interval the test inverts.
tost_bound <- setNames(unname(eq_pc_d05[seq_along(variables)]) *
                         unname(eq_den0[variables]), variables)
tost_ci <- function(o) unname(unlist(o$effsize["Raw", c("lower.ci", "upper.ci")]))
tost_p  <- function(o) max(o$TOST[c("TOST Lower", "TOST Upper"), "p.value"])

set.seed(2226)
tost_tbl <- bind_rows(lapply(variables, function(v) {
  bd <- tost_bound[[v]]
  x  <- eq_Msc[, v]; y <- eq_Mlq[, v]
  w  <- TOSTER::t_TOST(x = x, y = y, eqb = bd, var.equal = FALSE)      # Welch
  bt <- TOSTER::boot_t_TOST(x = x, y = y, eqb = bd, R = 1999)          # studentized
  data.frame(Variable = v, bound = bd, diff = unname(eq_diff0[[v]]),
             cl_lo = eq_diff_ci90[1, v], cl_hi = eq_diff_ci90[2, v],
             w_lo = tost_ci(w)[1],  w_hi = tost_ci(w)[2],  w_p = tost_p(w),
             b_lo = tost_ci(bt)[1], b_hi = tost_ci(bt)[2], b_p = tost_p(bt))
}))
tost_tbl <- tost_tbl |>
  mutate(cl_w = cl_hi - cl_lo, w_w = w_hi - w_lo, b_w = b_hi - b_lo,
         ratio_w = w_w / cl_w, ratio_b = b_w / cl_w)

tost_n_narrower <- sum(tost_tbl$ratio_w <= 1)
tost_ratio_med  <- median(tost_tbl$ratio_w)
tost_ratio_min  <- min(tost_tbl$ratio_w); tost_ratio_max <- max(tost_tbl$ratio_w)
# how many of the six each form of the test declares equivalent at this bound,
# and by how much in p the ones that fall outside miss it
tost_n_w <- sum(tost_tbl$w_p <= 0.05); tost_n_b <- sum(tost_tbl$b_p <= 0.05)
tost_margin <- max(c(tost_tbl$w_p, tost_tbl$b_p)) - 0.05
icc_of <- function(r) 100 * brms_icc$ICC[brms_icc$Response == r]
```

The equivalence test reported above resamples the surface material by locality: whole localities are drawn with replacement, each with all of its specimens, because the `r n_by_group[["SC_Quina"]]` surface scrapers come from `r eq_n_loc` localities and are not that many independent observations. The multilevel model of Part 2 puts the locality share of the variance at `r f(icc_of("Thickness"), 0)`% for thickness and `r f(icc_of("NScar"), 0)`% for the scar count (@tbl-brms-icc), so the assumption of independent specimens is not one this material supports. The cost of that choice is a wider interval, and @tbl-toster is there to say how much wider.

The comparison holds the equivalence bound fixed and changes the treatment of uncertainty. For each measure Δ is the ρ bound of @fig-pvalue-curve carried back into that measure's own units, and the test is repeated at that same Δ with TOSTER [@R-TOSTER; @lakens2017; @lakens2018], in its parametric Welch form and in its studentized bootstrap form, both of which take the specimen as the unit. The specimen-level interval is the narrower on all `r tost_n_narrower` measures, by a median factor of `r f(tost_ratio_med, 2)` (`r f(tost_ratio_min, 2)` to `r f(tost_ratio_max, 2)`). At this Δ the Welch form declares equivalence on `r tost_n_w` of the six and the bootstrap form on `r tost_n_b`, and the measures that fall outside miss by less than `r f(tost_margin, 3)` in *p*. That is what a bound of this kind should be expected to produce: Δ is by construction the smallest bound at which the analysis reported here reaches *p* = 0.05, so it sits exactly on the edge of the claim and a change of method lands on one side of the edge or the other. **The analysis reported in this paper is the locality one**; the TOSTER columns record what treating the specimens as independent would have bought.

\newpage

```{r}
#| label: tbl-toster
#| tbl-cap: "The equivalence tests of this section at the specimen level, beside the locality-level analysis that is reported. Δ is the equivalence bound, the ρ bound of @fig-pvalue-curve carried back into the measure's own units; the difference is the surface mean minus the Longtan Quina mean, in those same units. Each interval is the 90% interval that a TOST at α = 0.05 inverts: equivalence holds when the interval lies inside ±Δ. The first is the percentile interval of the cluster bootstrap of this paper, over 20,000 replicates in which localities are drawn as whole clusters; the second and third are TOSTER's, from Welch's *t* and from a studentized bootstrap over 1,999 replicates, both treating the specimens as independent. The last column is the width of TOSTER's Welch interval as a fraction of the cluster interval's, and *p* is the TOST *p*-value, the larger of the two one-sided values. The equivalence statement of @fig-pvalue-curve is made on ρ, the difference divided by the reference contrast, with that contrast re-estimated in every replicate; the locality interval given here is on the difference alone, and is set beside TOSTER's to compare their widths rather than to restate that statement."

tost_dig <- c(Thickness = 2, Retouch_length_index = 3, Ave_GIUR = 3,
              N_Scar = 2, Ave_RG = 2, Edge_Angle = 2)
tost_ci_txt <- function(lo, hi, d) sprintf("%s to %s", f(lo, d), f(hi, d))
tost_out <- data.frame(
  Measure = unname(variable_labels[tost_tbl$Variable]),
  bound   = vapply(seq_len(nrow(tost_tbl)),
                   function(i) f(tost_tbl$bound[i], tost_dig[[tost_tbl$Variable[i]]]),
                   character(1)),
  diff    = vapply(seq_len(nrow(tost_tbl)),
                   function(i) f(tost_tbl$diff[i], tost_dig[[tost_tbl$Variable[i]]]),
                   character(1)),
  cl      = vapply(seq_len(nrow(tost_tbl)),
                   function(i) tost_ci_txt(tost_tbl$cl_lo[i], tost_tbl$cl_hi[i],
                                           tost_dig[[tost_tbl$Variable[i]]]), character(1)),
  we      = vapply(seq_len(nrow(tost_tbl)),
                   function(i) tost_ci_txt(tost_tbl$w_lo[i], tost_tbl$w_hi[i],
                                           tost_dig[[tost_tbl$Variable[i]]]), character(1)),
  wp      = fpt(tost_tbl$w_p),
  bo      = vapply(seq_len(nrow(tost_tbl)),
                   function(i) tost_ci_txt(tost_tbl$b_lo[i], tost_tbl$b_hi[i],
                                           tost_dig[[tost_tbl$Variable[i]]]), character(1)),
  bp      = fpt(tost_tbl$b_p),
  ratio   = f(tost_tbl$ratio_w, 2))

knitr::kable(tost_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Measure", "Δ", "Difference", "90% CI, by locality",
                           "90% CI, Welch", "*p*", "90% CI, bootstrap", "*p*",
                           "Width ratio"))
```
\newpage

## Raw material economy

This section reports what was available on the river bars and what was chosen from it. @fig-sample-points shows the sampling points at which the cobbles were recorded, and @fig-thin-section the petrography of the trachyte, the dominant raw material. @tbl-composition gives the raw material composition of the surveyed cobbles and of the surface Quina scrapers, valley by valley, and @tbl-electivity sets the two against one another as an index of selection. @fig-electivity-sensitivity then asks whether that index depends on how the gravel was sampled, by recomputing it with each sampling point, and each valley, left out in turn. @tbl-rawmat-permanova tests whether composition differs between sampling points, valleys and basins, and @tbl-clast-size reports the size of the surveyed cobbles by lithology. @tbl-cobble-roundness and @tbl-cobble-shape give the two descriptors recorded in the field on the same cobbles, roundness on the Powers scale and the field-assigned shape term.

Roundness was recorded as the integer codes 0 to 3 of the Powers scale, 0 angular and 3 rounded, and the field-assigned shape as one of three nominal terms; both are complete for all `r n_cob_round` surveyed cobbles. Of the two lithologies the manuscript compares, the trachyte cobbles are the more rounded (Mann–Whitney *U* = `r f(rnd_U, 0)`, *p* `r fp(rnd_mw$p.value)`, rank-biserial correlation = `r f(rnd_rb, 2)`, *n* = `r rnd_nT` trachyte and `r rnd_nS` sandstone), while the shape terms are distributed alike between them (Fisher's exact test with a Monte Carlo *p* over `r n_fmt(FISHER_B)` replicates, *p* `r fp(shp_fisher$p.value)`, Cramér's *V* = `r f(shp_V, 2)`). Quartz (*n* = `r n_cob_quartz`) and mudstone (*n* = `r n_cob_mud`) are too few to test and are given in the tables as counts only.

\newpage

![The gravel-bar sampling points. (A) to (E) The five points at which the surveyed cobbles were recorded, each with the cleared quadrat marked out by string and a spirit level laid across it for scale. (F) A general view of one of the sampled channels. Every cobble longer than 30 mm exposed at the surface within the quadrat was recorded.](../figures/sample_point_photo.png){#fig-sample-points width="5.9in"}

\newpage

![Petrography of the trachyte. (A) Hand specimens of five of the sampled trachyte cobbles. (B) Photomicrographs of thin sections cut from them, with the magnification and the scale given in each frame. The rock is porphyritic: euhedral tabular to prismatic K-feldspar phenocrysts, with subordinate quartz, augite and clay minerals, are set in a fine-grained groundmass of K-feldspar microlites and devitrified glass.](../figures/thin_section.png){#fig-thin-section width="5.9in"}

\newpage

```{r}
#| label: tbl-composition
#| tbl-cap: "Raw material composition of the river-bar cobbles and of the surface-collected Quina scrapers, by valley. Cells give the count with the percentage of the row in parentheses."

comp_tbl <- comp_fig |>
  mutate(cell = ifelse(n == 0, "0", sprintf("%d (%s%%)", n, f(percent, 1)))) |>
  select(Layer, river_ID, Material, cell) |>
  pivot_wider(names_from = Material, values_from = cell) |>
  left_join(counts_fig, by = c("Layer", "river_ID"))

comp_out <- bind_rows(lapply(layer_levels, function(lay) {
  d <- comp_tbl[comp_tbl$Layer == lay, ]
  bind_rows(
    data.frame(Valley = sprintf("**%s**", lay),
               Trachyte = "", Sandstone = "", Quartz = "", Mudstone = "",
               Andesite = "", N = ""),
    data.frame(Valley = paste0(IND, as.character(d$river_ID)),
               Trachyte = d$Trachyte, Sandstone = d$Sandstone, Quartz = d$Quartz,
               Mudstone = d$Mudstone, Andesite = d$Andesite, N = as.character(d$N)))
}))

knitr::kable(comp_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r"),
             col.names = c("Valley", "Trachyte", "Sandstone", "Quartz",
                           "Mudstone", "Andesite", "*n*"))
```

\newpage

```{r}
#| label: tbl-electivity
#| tbl-cap: "Jacobs' electivity index for every raw material. *D* runs from −1, complete avoidance, through 0, use in proportion to availability, to +1, complete selection."

elect_tbl <- tibble(Material = material_levels) |>
  left_join(comp_avail |> transmute(Material, an = n, ap = p), by = "Material") |>
  left_join(comp_used  |> transmute(Material, un = n, up = r), by = "Material") |>
  left_join(jac |> transmute(Material, D), by = "Material") |>
  mutate(across(c(an, un), ~ replace_na(.x, 0L)),
         across(c(ap, up), ~ replace_na(.x, 0)),
         Label = ifelse(Material == "Andesite", "Andesite†", Material))

elect_out <- bind_rows(
  data.frame(Material = elect_tbl$Label,
             an = as.character(elect_tbl$an), ap = f(elect_tbl$ap, 1),
             un = as.character(elect_tbl$un), up = f(elect_tbl$up, 1),
             D  = f(elect_tbl$D, 2)),
  data.frame(Material = "**Total**",
             an = as.character(sum(elect_tbl$an)), ap = f(100, 1),
             un = as.character(sum(elect_tbl$un)), up = f(100, 1), D = ""))

knitr::kable(elect_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Raw material", "Cobbles *n*", "Cobbles %",
                           "Scrapers *n*", "Scrapers %", "*D*"))
```

\newpage

```{r}
#| label: fig-electivity-sensitivity
#| fig-cap: "Sensitivity of Jacobs' electivity index to the gravel survey. *r*, the proportion of a lithology among the artefacts, is a property of the artefacts and is held fixed at its region-wide value throughout (trachyte 95.8%, sandstone 1.8% of the 166 surface Quina scrapers); only *p*, its proportion among the surveyed cobbles, moves. Every marker therefore lies on the same curve within a panel, and the curve gives *D* for any composition the gravel might have had. The upper row groups the 469 cobbles by sampling point, the lower row by valley; the two rows are otherwise identical. Within each panel the filled circle is the pooled survey, the squares are the five, or three, leave-one-out estimates, each pooling the remaining units, and the triangles are the units taken one at a time; the ticks on the panel edge repeat the marker positions on the *p* axis. The shaded band spans the interval the survey covers and the dashed line marks the neutral point *p* = *r*, at which *D* = 0 and the lithology would be used exactly in proportion to its availability."
#| fig-width: 5.9
#| fig-height: 6.2

# r, the proportion of a lithology among the artefacts, is a property of the
# artefacts and is held fixed at the region-wide value of `comp_used`; only p,
# the proportion among the surveyed cobbles, moves. Every marker therefore lies
# on one and the same curve. The sampling point and the valley are two
# groupings of the same cobbles, so one builder serves both rows.
INK <- "#202124"; GREY <- "#5A5F66"
material_colors_fig <- c(Trachyte = "#E07C90", Sandstone = "#6BA8CE",
                         Quartz = "#E6C25C", Mudstone = "#9B87C4",
                         Andesite = "#6FB98E")
pctf <- function(x) paste0(f(100 * x, 1), "%")

jacobs_D  <- function(r, p) (r - p) / (r + p - 2 * r * p)
prop_of   <- function(df, m) mean(df$Material == m)
materials <- c("Trachyte", "Sandstone")
r_obs     <- setNames(comp_used$r[match(materials, comp_used$Material)] / 100, materials)
p_pooled  <- setNames(comp_avail$p[match(materials, comp_avail$Material)] / 100, materials)

# Leave-one-out needs at least three groups to say anything: with two, dropping
# one leaves the other, so the leave-one-out marker and the single-group marker
# of the other group are the same number.
SETS <- c("Pooled", "Leave-one-out", "Single unit")
build_sets <- function(group, unit) {
  ids <- levels(factor(group))
  n   <- length(ids)
  stopifnot(n >= 3)
  pts <- bind_rows(lapply(materials, function(m) bind_rows(
    tibble(Material = m, Set = SETS[1], Unit = NA_character_,
           p = unname(p_pooled[[m]])),
    tibble(Material = m, Set = SETS[2], Unit = ids,
           p = vapply(ids, function(g) prop_of(cobbles[group != g, ], m),
                      numeric(1), USE.NAMES = FALSE)),
    tibble(Material = m, Set = SETS[3], Unit = ids,
           p = vapply(ids, function(g) prop_of(cobbles[group == g, ], m),
                      numeric(1), USE.NAMES = FALSE))))) |>
    mutate(D = jacobs_D(unname(r_obs[Material]), p),
           Set = factor(Set, levels = SETS))
  list(pts = pts, n = n, unit = unit)
}

curve_panel <- function(m, S) {
  col <- unname(material_colors_fig[m])
  r   <- unname(r_obs[[m]])
  crv <- tibble(p = seq(0, 1, length.out = 2001)) |> mutate(D = jacobs_D(r, p))
  dat <- S$pts[S$pts$Material == m, ]
  D_p <- dat$D[dat$Set == SETS[1]]
  rng <- range(dat$p)

  set_shape <- setNames(c(21, 22, 24), SETS)
  set_size  <- setNames(c(2.9, 1.8, 1.8), SETS)
  set_fill  <- setNames(c(col, "white", "white"), SETS)

  # r sits close to one end of the axis, and the markers sit at the top of the
  # panel for a selected material and at the bottom for an avoided one. Both
  # annotations therefore go to the side of the line, and on the side of the
  # curve, that the data leave empty.
  hi    <- r > 0.5
  lab_h <- if (hi) 1.22 else -0.22
  rug_s <- if (hi) "b" else "t"

  ggplot(crv, aes(p, D)) +
    # the interval the gravel survey actually covers
    annotate("rect", xmin = rng[1], xmax = rng[2], ymin = -Inf, ymax = Inf,
             fill = col, alpha = 0.12) +
    geom_hline(yintercept = 0, colour = INK, linewidth = 0.4) +
    geom_vline(xintercept = r, colour = GREY, linetype = "dashed",
               linewidth = 0.45) +
    annotate("text", x = r, y = 0.40, hjust = lab_h, vjust = 0.5,
             size = 2.3, colour = GREY, lineheight = 1.05,
             label = paste0("neutral point\np = r = ", pctf(r))) +
    geom_line(colour = col, linewidth = 0.85) +
    # rug: the marker positions stay readable off the axis even where the
    # curve markers crowd
    geom_rug(data = dat[dat$Set != SETS[1], ], aes(x = p), inherit.aes = FALSE,
             sides = rug_s, length = unit(3.5, "pt"), colour = GREY,
             linewidth = 0.35) +
    geom_point(data = dat, aes(p, D, shape = Set, fill = Set, size = Set),
               colour = INK, stroke = 0.45) +
    annotate("text", x = unname(p_pooled[[m]]), y = D_p, hjust = 0.5,
             vjust = if (hi) 2.4 else -1.6, size = 2.6,
             colour = INK, fontface = "bold",
             label = paste0("D = ", f(D_p, 2))) +
    scale_shape_manual(values = set_shape, drop = FALSE) +
    scale_fill_manual(values = set_fill, drop = FALSE) +
    scale_size_manual(values = set_size, drop = FALSE, guide = "none") +
    scale_x_continuous(breaks = seq(0, 1, 0.25),
                       labels = function(x) paste0(100 * x, "%"),
                       expand = expansion(mult = c(0.02, 0.02))) +
    scale_y_continuous(breaks = seq(-1, 1, 0.5),
                       expand = expansion(mult = c(0.02, 0.02))) +
    # clip rather than drop: D is exactly +-1 at the ends of the sweep, and a
    # scale limit would discard those rows
    coord_cartesian(xlim = c(0, 1), ylim = c(-1, 1)) +
    # The pooled marker takes the lithology colour, so the fill scale differs
    # between panels and patchwork cannot collect a fill guide. The legend is
    # therefore carried by shape alone, which is identical in all four panels,
    # with the keys given a fixed fill so that they still read solid-vs-open.
    guides(fill = "none",
           shape = guide_legend(override.aes = list(
             fill = c(INK, "white", "white"), size = unname(set_size)))) +
    labs(title = sprintf("%s · %d %ss", m, S$n, S$unit),
         x = paste0("p, ", tolower(m), " among the surveyed cobbles"),
         y = "Jacobs' electivity index D") +
    fig_theme +
    theme(plot.title = element_text(face = "bold", size = 8.5, colour = INK),
          axis.title = element_text(size = 8),
          plot.margin = margin(5, 8, 3, 5))
}

S_point  <- build_sets(cobbles$Loc,      "sampling point")
S_valley <- build_sets(cobbles$river_ID, "valley")

# upper row: grouped by sampling point. lower row: grouped by valley.
(curve_panel("Trachyte", S_point)  | curve_panel("Sandstone", S_point)) /
  (curve_panel("Trachyte", S_valley) | curve_panel("Sandstone", S_valley)) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom", legend.box.spacing = unit(3, "pt"))
```

\newpage

```{r}
#| label: tbl-rawmat-permanova
#| tbl-cap: "PERMANOVA and PERMDISP on raw material composition, on Bray–Curtis dissimilarities with 9,999 permutations. Sums of squares, degrees of freedom and residual degrees of freedom are those of the PERMANOVA; the last two columns give the pseudo-*F* and *p* of the test of homogeneity of dispersion on the same dissimilarity matrix. The unit is the individual cobble in the first panel and the individual artefact in the second."

perm_row <- function(model, a, pd) {
  data.frame(Model = paste0(IND, model),
             Df = sprintf("%d, %d", a$Df[1], a$Df[2]),
             SS = f(a$SumOfSqs[1], 2),
             R2 = f(a$R2[1], 3),
             F  = f(a$F[1], 2),
             p  = fpt(a$`Pr(>F)`[1]),
             pdF = f(pd[["F"]], 2),
             pdp = fpt(pd[["p"]]))
}
head_row <- function(label) data.frame(Model = sprintf("**%s**", label),
  Df = "", SS = "", R2 = "", F = "", p = "", pdF = "", pdp = "")

rawmat_out <- bind_rows(
  head_row(sprintf("Available river cobbles (*n* = %d)", n_cobble)),
  perm_row("~ sampling point", cobbleLoc,   pd_cobbleLoc),
  perm_row("~ valley",         cobbleRiver, pd_cobbleRiver),
  head_row(sprintf("Surface Quina scrapers (*n* = %d)", nrow(scA))),
  perm_row("~ basin",  usedA_basin, pd_usedA_basin),
  perm_row("~ valley", usedA_river, pd_usedA_river))

knitr::kable(rawmat_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Model", "*df*", "SS", "*R*²", "pseudo-*F*", "*p*",
                           "PERMDISP *F*", "PERMDISP *p*"))
```

\newpage

```{r}
#| label: tbl-clast-size
#| tbl-cap: "Size of the surveyed river cobbles by lithology. Size is the geometric mean of length, breadth and thickness in millimetres, and GM is the geometric mean of that quantity within each lithology."

size_row <- function(m) {
  x <- cl$size[cl$Material == m]
  if (!length(x)) return(NULL)
  qs <- unname(quantile(x, c(.25, .75)))
  data.frame(Material = m, n = as.character(length(x)),
             GM = f(gm(x), 1), Median = f(median(x), 1),
             IQR = sprintf("%s–%s", f(qs[1], 1), f(qs[2], 1)),
             Range = sprintf("%s–%s", f(min(x), 1), f(max(x), 1)))
}
size_out <- bind_rows(lapply(material_levels, size_row))

knitr::kable(size_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Raw material", "*n*", "GM (mm)", "Median (mm)",
                           "Q1–Q3 (mm)", "Range (mm)"))
```

\newpage

```{r}
#| label: tbl-cobble-roundness
#| tbl-cap: "Roundness of the surveyed river cobbles, recorded in the field on the Powers scale as the integer codes 0 to 3 with 0 angular and 3 rounded. The two panels are the same cobbles partitioned two ways, by lithology and by sampling point. Cells give the count with the percentage of the row in parentheses."

rnd_out <- bind_rows(
  ct_cells(table(cob$Material, cob$Roundness), "By lithology"),
  ct_cells(table(cob$Point, cob$Roundness), "By sampling point"))

knitr::kable(rnd_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Group", round_labels, "*n*"))
```

\newpage

```{r}
#| label: tbl-cobble-shape
#| tbl-cap: "Field-assigned shape of the surveyed river cobbles by lithology. The term is a nominal field record with three categories, and is distinct from the quantitative form analysis reported in the main text, which takes form from the three axes rather than from the field term. Cells give the count with the percentage of the row in parentheses."

shape_out <- ct_cells(table(cob$Material, cob$Shape))

knitr::kable(shape_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r"),
             col.names = c("Raw material", levels(cob$Shape), "*n*"))
```

\newpage

## Landscape structure

This section asks whether the technological variation follows position in the landscape. @tbl-locality-matrix lists the localities that enter the analysis, with the number of Quina scrapers each yielded and the median of those scrapers on each of the seven measures. @tbl-brms-coef and @tbl-brms-icc then give the model reported in the main text, which puts the question to the specimens themselves rather than to those locality medians.

The locality is a convenient summary but a poor unit of analysis: pooling a locality to its median discards the spread within it and weights a locality of two specimens as heavily as one of twenty. The three landscape predictors were therefore fitted to the `r n_brms` individual scrapers from `r n_brms_loc` localities in a single multivariate multilevel model (@tbl-brms-coef). Of its `r nrow(brms_coef)` landscape coefficients, `r brms_cri_excl_zero` have a 95% credible interval clear of zero. The locality itself carries a modest share of the variance in every measure, with intraclass correlations from `r f(min(brms_icc$ICC), 2)` to `r f(max(brms_icc$ICC), 2)` (@tbl-brms-icc). The sampler converged: `r brms_div` divergent transitions, `r brms_treedepth` iterations at maximum treedepth, a maximum R-hat of `r f(brms_max_rhat, 3)` and a minimum effective sample size of `r n_fmt(round(brms_min_ess))` across `r n_fmt(brms_ndraws)` post-warmup draws, from `r BRMS_CHAINS` chains of `r n_fmt(BRMS_ITER)` iterations.

The two tables summarise the fit; the model is reported at length in the Supplementary Bayesian Report, written under the Bayesian Analysis Reporting Guidelines [@kruschke2021] and deposited with the repository as `paper/barg/barg_report.html`. That report carries the prior specification in full, the prior and posterior predictive checks, the complete convergence diagnostics, the decisions taken against a region of practical equivalence, and a sensitivity analysis over six prior specifications. The figure in the main text is drawn from it. Its reference fit differs from the one tabulated here only in the scale of the slope prior, which is set to the link-scale standard deviation of each response rather than to one; the two agree to within a percentage point on every intraclass correlation.

\newpage

```{r}
#| label: tbl-locality-matrix
#| tbl-cap: "Locality-level analysis frame. Each row is one Quina locality, with the number of Quina scrapers complete on all seven measures and the median of those specimens on each. Distance and height are measured to and above the nearest channel."

loc_out <- site_df |>
  arrange(Basin, desc(n_art), Name) |>
  transmute(Locality = Name, Basin = as.character(Basin),
            n = as.character(n_art),
            d = as.character(round(Distance_to_water)),
            h = as.character(round(Height_above_river)),
            GM = f(GM, 1),
            Th = f(Thickness, 1), RLI = f(Retouch_length_index, 2),
            GIUR = f(Ave_GIUR, 2), NS = f(N_Scar, 0),
            RG = f(Ave_RG, 2), EA = f(Edge_Angle, 1))

knitr::kable(loc_out, row.names = FALSE,
             align = c("l", "l", rep("r", 10)),
             col.names = c("Locality", "Basin", "*n*", "Distance (m)", "Height (m)",
                           unname(variable_labels_land)))
```

\newpage

```{r}
#| label: tbl-brms-coef
#| tbl-cap: "Landscape terms in the specimen-level multivariate multilevel model. The unit is the individual Quina scraper rather than the locality median, and the seven measures are fitted together in one model, each with the likelihood its scale calls for, with locality carried as a random intercept correlated across responses. Basin is contrasted against Binchuan and the two continuous predictors are z-standardised, so each coefficient is the change on the link scale of its own response per standard deviation of the predictor. The ROPE is a region of practical equivalence of half-width one tenth of the standard deviation of that response on the same scale, and the last column gives the percentage of the posterior that falls inside it."

brms_out <- bind_rows(lapply(brms_resps, function(r) {
  d <- brms_coef[brms_coef$Response == r, ]
  bind_rows(
    data.frame(Term = sprintf("**%s (%s)**", brms_resp_labels[[r]],
                              brms_fam_labels[[d$Family[1]]]),
               med = "", ci = "", rope = "", inr = ""),
    data.frame(Term = paste0(IND, d$Predictor),
               med  = f(d$Median, 3),
               ci   = sprintf("%s to %s", f(d$CrI_lower, 3), f(d$CrI_upper, 3)),
               rope = f(d$ROPE_halfwidth, 3),
               inr  = f(d$pct_in_ROPE, 1)))
}))

knitr::kable(brms_out, row.names = FALSE, align = c("l", "r", "r", "r", "r"),
             col.names = c("Response / landscape term", "Median", "95% CrI",
                           "ROPE ±", "% in ROPE"))
```

\newpage

```{r}
#| label: tbl-brms-icc
#| tbl-cap: "Share of the variance in each measure that lies between localities rather than within them, from the same model. The intraclass correlation is computed for every posterior draw as the locality variance over the sum of the locality and residual variances, both on the link scale of the response. Residual variance is σ² for the gaussian and lognormal responses, log(1 + 1/μ + 1/shape) for the negative binomial, and the sum of the two trigamma terms for the beta responses, each evaluated at the model intercept."

icc_out <- data.frame(
  Response = unname(brms_resp_labels[brms_icc$Response]),
  fam = unname(brms_fam_labels[brms_icc$Family]),
  sd  = f(brms_icc$sd_Locality, 3),
  icc = f(brms_icc$ICC, 3),
  ci  = sprintf("%s to %s", f(brms_icc$ICC_lower, 3), f(brms_icc$ICC_upper, 3)))

knitr::kable(icc_out, row.names = FALSE, align = c("l", "l", "r", "r", "r"),
             col.names = c("Measure", "Likelihood", "SD (locality)", "ICC",
                           "95% CrI"))
```

\newpage

## The localities not analysed

The survey recorded more localities than the analysis uses. @tbl-locality-matrix lists those that enter it; @tbl-excluded-localities lists the eleven that do not, none of which yielded Quina retouch products. They are given here so that the survey record can be read in full; nothing reported in this paper is computed from them.

\newpage

```{r}
#| label: tbl-excluded-localities
#| tbl-cap: "The eleven surveyed localities that are not analysed in this paper. No Quina retouch products, neither Quina scrapers nor resharpening flakes, were recorded at any of them, so none meets the criterion for inclusion set out in the main text and none enters any analysis reported here or in @tbl-locality-matrix; they are listed for completeness of the survey record. All eleven lie in the Binchuan Basin. Landform is the position within the terrace sequence, as in Table 1 of the main text, with the one locality standing well above that sequence entered as a hilltop. Coordinates are degrees, minutes and seconds, elevation is metres above sea level, and height is metres above the local channel. The sheet these rows come from carries no artefact counts and no distance to the channel, neither having been recorded for a locality that yielded nothing to analyse."

dms <- function(x, hemi) {
  tot <- round(x * 3600, 1)
  d   <- floor(tot / 3600)
  m   <- floor((tot - d * 3600) / 60)
  s   <- tot - d * 3600 - m * 60
  sprintf("%d°%02d′%04.1f″%s", d, m, s, hemi)
}

# sheet 2 of the site table: the localities the same survey recorded but which
# yielded no Quina retouch products
excl_sites <- read_excel(site_path, sheet = 2)
names(excl_sites) <- trimws(names(excl_sites))

excl_out <- excl_sites |>
  transmute(Locality = gsub("_", " ", trimws(as.character(name))),
            Valley   = factor(trimws(as.character(river_ID)), levels = river_levels),
            Landform = trimws(as.character(geomorph)),
            lat  = dms(as.numeric(lat), "N"),
            lon  = dms(as.numeric(lon), "E"),
            elev = as.character(round(as.numeric(elev_m))),
            h    = as.character(round(as.numeric(h_river_m)))) |>
  arrange(Valley) |>
  mutate(Valley = as.character(Valley))

knitr::kable(excl_out, row.names = FALSE,
             align = c("l", "l", "l", "r", "r", "r", "r"),
             col.names = c("Locality", "Valley", "Landform", "Latitude", "Longitude",
                           "Elevation (m)", "Height (m)"))
```

\newpage

# Part 3: Sensitivity of the PERMANOVA results to unequal dispersion

```{r}
#| label: si-wds
#| include: false

# ---- Wd*, the Welch-type correction to PERMANOVA --------------------------
# Alekseyenko's statistic, from the authors' own package [@alekseyenko2016;
# @hamidi2019; @R-WdStar]. `wds_of()` and `wds_pair()`, the three pairwise
# comparisons of the technological space (`wds_tc`) and their Bonferroni
# adjustment come from _analysis.R, which computes them because the manuscript
# reports them; what is added here is the raw material designs, the cobble
# form, and the same three comparisons on the reduced set.
# This chunk is the last in the document to draw random numbers, so nothing
# reported in Part 1 or Part 2 depends on it.
wds_cobbleLoc   <- wds_of(C_D, cobbles$Loc)
wds_cobbleRiver <- wds_of(C_D, cobbles$river_ID)
wds_usedA_basin <- wds_of(A_D, A_meta$basin)
wds_usedA_river <- wds_of(A_D, A_meta$river_ID)
wds_form        <- wds_of(D_form, sz$G)
wds_tc_excl <- lapply(tc_pairs, function(pr) wds_pair(amat_excl, complete_excl$Group, pr))

# the Bonferroni family of the reduced set, the counterpart of wds_tc_padj
wds_tc_excl_padj <- p.adjust(vapply(wds_tc_excl, function(x) x[["p"]], numeric(1)),
                             "bonferroni")

# the counts quoted in the text
tc_cmp     <- vapply(tc_pairs, paste, character(1), collapse = " vs ")
wds_perm_p <- c(cobbleLoc$`Pr(>F)`[1], cobbleRiver$`Pr(>F)`[1],
                usedA_basin$`Pr(>F)`[1], usedA_river$`Pr(>F)`[1], form_p,
                tc_pw$p_value[match(tc_cmp, tc_pw$Comparison)],
                tc_pw_excl$p_value[match(tc_cmp, tc_pw_excl$Comparison)])
wds_wd_p <- c(wds_cobbleLoc[["p"]], wds_cobbleRiver[["p"]], wds_usedA_basin[["p"]],
              wds_usedA_river[["p"]], wds_form[["p"]],
              vapply(wds_tc,      function(x) x[["p"]], numeric(1)),
              vapply(wds_tc_excl, function(x) x[["p"]], numeric(1)))
n_wds_total   <- length(wds_wd_p)
n_wds_tested  <- sum(!is.na(wds_wd_p))
n_wds_agree   <- sum((wds_perm_p <= 0.05) == (wds_wd_p <= 0.05), na.rm = TRUE)
wds_basin_undef_pct <- 100 * wds_usedA_basin[["undef"]]
n_sangyuan_scraper  <- sum(A_meta$river_ID == "Sangyuan", na.rm = TRUE)
```

PERMANOVA reads a difference between groups as a difference in position, and it does so on the assumption that the groups are equally dispersed about their own centroids. PERMDISP shows that assumption to fail in several of the designs reported in Part 2, and where it fails the reading is no longer clean: a rejection may follow from unequal spread rather than from position, and a failure to reject may hide a difference in position that the unequal spread has masked. The *W*~d~\* statistic is a Welch-type correction to PERMANOVA which weights each group by its own dispersion instead of pooling them, so it does not carry that assumption [@alekseyenko2016; @hamidi2019]. It is computed here with the authors' own implementation, the `WdStar` package [@R-WdStar]. Setting it beside each PERMANOVA separates the two readings.

@tbl-wds repeats every PERMANOVA reported in this supplement and in the main text with *W*~d~\* alongside, on the same dissimilarities, the same grouping and the same `r n_fmt(PERM)` permutations. Of the `r n_wds_total` models, `r n_wds_tested` can be tested this way, and all `r n_wds_agree` of them reach the conclusion their PERMANOVA reaches.

Every PERMANOVA that rejects does so with PERMDISP rejecting as well, and each of them survives the correction at *p* < 0.001 with a *W*~d~\* statistic of the same order as the pseudo-*F* it replaces; those differences are differences in position and not in spread alone. Within each of the two families of three pairwise comparisons the Bonferroni adjustment leaves every conclusion unchanged.

Two limits should be stated. *W*~d~\* takes a grouping factor, so a continuous predictor has no *W*~d~\* counterpart and none is given for one. And the statistic divides by the within-group dispersion, which the raw material designs can drive to zero: the `r n_sangyuan_scraper` surface Quina scrapers of the Sangyuan valley are all trachyte, so that group carries no internal dissimilarity and the statistic is undefined for the valley model, while `r f(wds_basin_undef_pct, 1)` per cent of the permutations of the basin model fall into the same degeneracy and leave its *p*-value undefined. Neither of those two designs has a dispersion problem to begin with (PERMDISP *p* `r fp(pd_usedA_basin[["p"]])` and `r fp(pd_usedA_river[["p"]])`), so nothing in their interpretation turns on the correction.

\newpage

```{r}
#| label: tbl-wds
#| tbl-cap: "Every PERMANOVA reported in this supplement and in the main text, with PERMDISP and the Welch-type *W*~d~\\* statistic on the same dissimilarities, the same grouping and 9,999 permutations. PERMANOVA and PERMDISP columns repeat the values given in the tables of Part 2; *p*-values are unadjusted, and the Bonferroni adjustment within the two families of three pairwise comparisons changes no conclusion in either column. The *n* column gives the size of each group in the order the model fits them. An em dash marks a quantity that cannot be computed: *W*~d~\\* divides by the within-group dispersion, which is zero for a group whose members share a single raw material."

wnum <- function(x, d = 2) if (is.finite(x)) f(x, d) else "—"
wpv  <- function(p) if (is.na(p)) "—" else fpt(p)
wrow <- function(label, n_txt, pF, pp, dF, dp, w) {
  data.frame(Model = paste0(IND, label), n = n_txt,
             F = wnum(pF), p = wpv(pp), pdF = wnum(dF), pdp = wpv(dp),
             wF = wnum(w[["stat"]]), wp = wpv(w[["p"]]))
}
whead <- function(label) data.frame(Model = sprintf("**%s**", label), n = "",
  F = "", p = "", pdF = "", pdp = "", wF = "", wp = "")
gsz <- function(g) paste(as.integer(table(droplevels(factor(g)))), collapse = ", ")
tc_wrow <- function(pw, pt, wl, i) {
  pr <- tc_pairs[[i]]
  a  <- pw[pw$Comparison == paste(pr, collapse = " vs "), ]
  d  <- pt[pt$Comparison == paste(pr, collapse = "-"), ]
  wrow(paste(group_labels[pr], collapse = " vs "), sprintf("%d, %d", a$n1, a$n2),
       a$F, a$p_value, d$F, d$p, wl[[i]])
}

wds_out <- bind_rows(
  whead("Raw material composition, Bray–Curtis"),
  wrow("Available river cobbles ~ sampling point", gsz(cobbles$Loc),
       cobbleLoc$F[1], cobbleLoc$`Pr(>F)`[1],
       pd_cobbleLoc[["F"]], pd_cobbleLoc[["p"]], wds_cobbleLoc),
  wrow("Available river cobbles ~ valley", gsz(cobbles$river_ID),
       cobbleRiver$F[1], cobbleRiver$`Pr(>F)`[1],
       pd_cobbleRiver[["F"]], pd_cobbleRiver[["p"]], wds_cobbleRiver),
  wrow("Surface Quina scrapers ~ basin", gsz(A_meta$basin),
       usedA_basin$F[1], usedA_basin$`Pr(>F)`[1],
       pd_usedA_basin[["F"]], pd_usedA_basin[["p"]], wds_usedA_basin),
  wrow("Surface Quina scrapers ~ valley", gsz(A_meta$river_ID),
       usedA_river$F[1], usedA_river$`Pr(>F)`[1],
       pd_usedA_river[["F"]], pd_usedA_river[["p"]], wds_usedA_river),
  whead("Cobble form, Aitchison"),
  wrow("Trachyte vs sandstone", gsz(sz$G),
       form_F, form_p, form_disp_F, form_disp_p, wds_form),
  whead("Technological consistency, Euclidean in the z-scored six-measure space"),
  tc_wrow(tc_pw, permA$pair_tab, wds_tc, 1),
  tc_wrow(tc_pw, permA$pair_tab, wds_tc, 2),
  tc_wrow(tc_pw, permA$pair_tab, wds_tc, 3),
  whead("The same, on the reduced set excluding the Longtan site area"),
  tc_wrow(tc_pw_excl, permC$pair_tab, wds_tc_excl, 1),
  tc_wrow(tc_pw_excl, permC$pair_tab, wds_tc_excl, 2),
  tc_wrow(tc_pw_excl, permC$pair_tab, wds_tc_excl, 3))

knitr::kable(wds_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Model or comparison", "*n*", "pseudo-*F*", "*p*",
                           "PERMDISP *F*", "PERMDISP *p*", "*W*~d~\\*", "*p*"))
```

\newpage

# References {.unnumbered}

::: {#refs}
:::
