---
title: "Supplementary material"
subtitle: "The Quina Valley: a regional Middle Palaeolithic technological system on the southeastern margin of the Tibetan Plateau"
author:
  - Peiyuan Xiao:
      correspondence: "no"
      orcid: 0009-0000-9733-5875
      institute:
        - ALPHA
        - UCAS
        - UW
  - Qijun Ruan:
      correspondence: "yes"
      orcid: 0009-0000-2143-5335
      institute:
        - YPICRA
  - Zhenxiu Jia:
      correspondence: "no"
      orcid: 0000-0002-7514-4514
      institute:
        - ALPHA
  - Marco Peresani:
      correspondence: "no"
      orcid: 0000-0001-6562-6336
      institute:
        - FERRARA
        - IGAG
  - Davide Delpiano:
      correspondence: "yes"
      institute:
        - FERRARA
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - UW
# <!-- the author list mirrors paper/manuscript.qmd; update both together -->
institute:
  - ALPHA: Alpine Paleoecology and Human Adaptation Group (ALPHA Group), State Key Laboratory of Tibetan Plateau Earth System, Environment and Resources, Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing, China
  - UCAS: University of Chinese Academy of Sciences, Beijing, China
  - UW: Department of Anthropology, University of Washington, Seattle, WA, USA
  - YPICRA: Yunnan Provincial Institute of Cultural Relics and Archaeology, Kunming, China
  - FERRARA: Department of Human Studies, Prehistoric and Anthropological Science Unit, University of Ferrara, Ferrara, Italy
  - IGAG: Consiglio Nazionale delle Ricerche-Institute of Environmental Geology and Geoengineering, Laboratory of Palynology and Palaeoecology, Research Group on Vegetation, Climate and Human Stratigraphy, Milan, Italy
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
crossref:
  fig-title: "Fig. S"
  fig-prefix: "Fig. S"
  tbl-title: "Table S"
  tbl-prefix: "Table S"
number-sections: false
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
  - quarto
  # runs after the crossref filter, so that "Table S 1" reads "Table S1"
  - ../templates/supplement-numbering.lua
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

# Every result in this document is computed by paper/_analysis.R, the same file
# the manuscript sources, so the two cannot disagree. The few quantities needed
# only here -- the dispersion tests that accompany the raw-material PERMANOVAs,
# the omnibus statistics behind the per-variable post-hoc comparisons, and the
# two further correlation families of the landscape figure -- are computed in
# the next chunk from the objects it leaves behind.
#
# Apart from the note that opens it and the short guide that opens each of its
# four sections, this document carries no running text: it is the tables and
# figures with their captions alone, each caption written to stand on its own.
source(here::here("paper", "_analysis.R"))

# p-value formatter for table cells: fp() with its leading "= " removed
fpt <- function(p) sub("^= ", "", vapply(p, fp, character(1)))

group_colors <- c(SC_Quina = "#E07C90", LT_Quina = "#E6C25C", LT_Ordinary = "#6BA8CE")
group_labels <- c(SC_Quina = "Surface Quina", LT_Quina = "Longtan Quina",
                  LT_Ordinary = "Longtan non-Quina")
# variable_labels and variable_labels_bare come from _analysis.R, the same set
# the manuscript uses, so that no variable is named two ways across the two
# documents
# the blue and the pink of the shared palette, which are also the two ends of
# the gradient used for the continuous landscape variables
basin_colors   <- c(Binchuan = "#6BA8CE", Heqing = "#E07C90")
grad_low       <- "#6BA8CE"
grad_high      <- "#E07C90"

fig_theme <- theme_minimal(base_size = 9) +
  theme(panel.grid.major = element_line(color = "#E6E8EB", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.5),
        axis.ticks = element_line(color = "#202124", linewidth = 0.3),
        axis.ticks.length = unit(2, "pt"),
        axis.title = element_text(size = 9), axis.text = element_text(color = "#303238", size = 8),
        strip.text = element_text(face = "bold", color = "#202124", size = 8),
        strip.background = element_rect(fill = "#E8E8E8", color = NA),
        legend.title = element_blank(), legend.text = element_text(size = 8),
        legend.key = element_blank(), plot.tag = element_text(face = "bold", size = 11),
        plot.background = element_rect(color = NA, fill = "white"),
        panel.background = element_rect(color = NA, fill = "white"))
```

```{r}
#| label: si-analysis
#| include: false

# ==========================================================================
# Quantities needed only in this document. Everything here is built from the
# objects created in _analysis.R, on the same seed and the same PERM.
# ==========================================================================

# ---- 1. PERMDISP alongside each raw-material PERMANOVA --------------------
# A significant PERMANOVA with a small R2 can reflect unequal dispersion
# rather than a shift of centroid, so each composition model is repeated as a
# test of dispersion on the same dissimilarity.
permdisp_one <- function(D, g) {
  g <- droplevels(factor(g))
  set.seed(2226); bd <- suppressWarnings(betadisper(D, g))
  set.seed(2226); pt <- permutest(bd, permutations = PERM)
  c(F = pt$tab$F[1], p = pt$tab$`Pr(>F)`[1])
}
pd_cobbleLoc   <- permdisp_one(C_D, cobbles$Loc)
pd_cobbleRiver <- permdisp_one(C_D, cobbles$river_ID)
pd_usedA_basin <- permdisp_one(A_D, A_meta$basin)
pd_usedA_river <- permdisp_one(A_D, A_meta$river_ID)

# ---- 2. Welch post-hoc statistics ----------------------------------------
# pairwise_t_test() returns adjusted p-values but no test statistic, so the
# same comparisons are re-run with t_test() to recover t and its df. The two
# procedures are the same Welch test with the same Bonferroni adjustment; the
# check below fails the render if the adjusted p-values ever diverge.
we_t <- variable_long |> filter(Variable %in% welch_vars) |>
  mutate(Variable = droplevels(Variable)) |> group_by(Variable) |>
  t_test(Value ~ Group, var.equal = FALSE, p.adjust.method = "bonferroni") |> ungroup()
we_ph_full <- we_ph |> select(Variable, group1, group2, p.adj, p.adj.signif) |>
  left_join(we_t |> select(Variable, group1, group2, statistic, df, p.adj.t = p.adj),
            by = c("Variable", "group1", "group2"))
stopifnot(isTRUE(all.equal(we_ph_full$p.adj, we_ph_full$p.adj.t)))

# ---- 3. Landscape: dispersion between the basins --------------------------
# Basin is the one categorical landscape variable and so the only one for which
# a test of dispersion is defined. The other three enter every analysis at
# their measured values; they are nowhere discretised.
ls_basin_pd <- permdisp_one(dL, site_df$Basin)
n_basin_site <- table(droplevels(site_df$Basin))

# ---- 4. Locality-level ordination ----------------------------------------
site_pca <- prcomp(scale(as.matrix(site_df[, variables])), center = TRUE, scale. = FALSE)
site_vexp <- site_pca$sdev^2 / sum(site_pca$sdev^2) * 100
site_scores <- bind_cols(site_df, as.data.frame(site_pca$x[, 1:2]))

# ---- 5. Each measure against each continuous landscape variable -----------
# The distance family is the one the paper reports; the other two are given
# here so that every panel of the scatter figure carries its coefficient. Each
# is one family of six tests and carries its own Bonferroni adjustment, exactly
# as the distance family does in _analysis.R. The check below fails the render
# if the distance rows ever depart from the values reported in the paper.
land_vars <- c(Height_above_river = "Height above channel (m)",
               Site_size          = "Assemblage size (n specimens)",
               Distance_to_water  = "Distance to channel (m)")
land_cor <- bind_rows(lapply(names(land_vars), function(lv) {
  fam <- bind_rows(lapply(variables, function(v) {
    d <- site_df[is.finite(site_df[[v]]) & is.finite(site_df[[lv]]), ]
    ct <- suppressWarnings(cor.test(d[[v]], d[[lv]], method = "spearman", exact = FALSE))
    data.frame(Landscape = lv, variable = v, n = nrow(d),
               rho = unname(ct$estimate), p = ct$p.value)
  }))
  fam$p_adj <- p.adjust(fam$p, "bonferroni")
  fam
}))
stopifnot(isTRUE(all.equal(land_cor$rho[land_cor$Landscape == "Distance_to_water"],
                           ls_dist_cor$rho)),
          isTRUE(all.equal(land_cor$p_adj[land_cor$Landscape == "Distance_to_water"],
                           ls_dist_cor$p_adj)))

# ---- 6. Each landscape variable as a surface over the ordination ----------
# A thin-plate spline fitted over the ordination plane, as ordisurf fits one,
# and drawn as the background of the locality-level figure. select = TRUE lets
# the smooth be shrunk out of the model altogether, so a flat surface is a
# conclusion the fit reached rather than one imposed on it. The grid is clipped
# to the convex hull of the localities: colour would be an extrapolation
# anywhere else.
hull_i  <- chull(site_scores$PC1, site_scores$PC2)
hull_xy <- as.matrix(site_scores[c(hull_i, hull_i[1]), c("PC1", "PC2")])
# the grid is kept whole and the cells outside the hull set to NA, rather than
# dropped: a complete grid is what geom_raster needs, and it draws the missing
# cells as nothing at all
surf_one <- function(v, n = 250) {
  y <- site_scores[[v]]
  g <- mgcv::gam(y ~ s(PC1, PC2, k = 10), data = site_scores,
                 method = "REML", select = TRUE)
  s <- summary(g)
  rx <- range(site_scores$PC1); ry <- range(site_scores$PC2)
  grd <- expand.grid(
    PC1 = seq(rx[1] - 0.08 * diff(rx), rx[2] + 0.08 * diff(rx), length.out = n),
    PC2 = seq(ry[1] - 0.08 * diff(ry), ry[2] + 0.08 * diff(ry), length.out = n))
  grd$fit <- as.numeric(predict(g, newdata = grd))
  grd$fit[!as.logical(mgcv::in.out(hull_xy, as.matrix(grd[, c("PC1", "PC2")])))] <- NA_real_
  # the fit of each surface is reported in the caption rather than in the panel
  list(grid = grd, dev = 100 * s$dev.expl, p = s$s.table[1, "p-value"])
}
surfaces <- setNames(lapply(names(land_vars), surf_one), names(land_vars))
surf_dev_max <- max(vapply(surfaces, function(z) z$dev, numeric(1)))
surf_p_min   <- min(vapply(surfaces, function(z) z$p,   numeric(1)))

# ---- 7. Each measure between the two basins -------------------------------
# Basin is the categorical landscape variable, so it cannot enter the scatter
# figure with the other three; these are the univariate comparisons behind the
# boxplot that carries it. bdf is the frame the basin PERMANOVA is fitted on in
# _analysis.R, so figure and test cannot disagree. Each measure is compared with
# a Mann-Whitney rank sum test carrying the rank-biserial r; the six are one
# family and take a Bonferroni adjustment, as each correlation family above
# does. The figure shows the outcome as the bracket the main-text boxplots use,
# so the numbers themselves are reported in its caption.
basin_cmp <- bind_rows(lapply(variables, function(v) {
  d <- bdf[is.finite(bdf[[v]]), c("Basin", v)]
  names(d) <- c("Basin", "Value")
  d$Basin <- droplevels(d$Basin)
  wt <- wilcox_test(d, Value ~ Basin)
  data.frame(variable = v, p = wt$p,
             r = suppressWarnings(wilcox_effsize(d, Value ~ Basin))$effsize[1])
}))
basin_cmp$p_adj <- p.adjust(basin_cmp$p, "bonferroni")
basin_cmp <- add_significance(basin_cmp, p.col = "p_adj", output.col = "p.adj.signif")
basin_r_max <- basin_cmp[which.max(basin_cmp$r), ]
# the label of that measure, set to run inside a sentence: an ordinary name is
# lowercased, an abbreviation such as GIUR left as it stands
basin_r_lab <- variable_labels_bare[[basin_r_max$variable]]
if (grepl("^[A-Z][a-z]", basin_r_lab))
  basin_r_lab <- paste0(tolower(substr(basin_r_lab, 1, 1)), substring(basin_r_lab, 2))
# the boxes are wider in Binchuan on most measures, which is what the
# significant dispersion test records; counted here rather than asserted
basin_iqr_wider <- sum(vapply(variables, function(v) {
  q <- tapply(bdf[[v]], droplevels(bdf$Basin), IQR, na.rm = TRUE)
  q[["Binchuan"]] > q[["Heqing"]]
}, logical(1)))
# spelled out, the count being small enough to read as a word in the caption
basin_iqr_word <- c("none", "one", "two", "three", "four", "five",
                    "six")[basin_iqr_wider + 1]
```

This document reports the full statistical output behind the results summarised in the main text, together with the diagnostic and robustness analyses that support them. Every number here is computed from the same data, in the same run and on the same fixed seed (2226) as the corresponding number in the paper: both documents source a single analysis file, so a value can never differ between them. Permutation tests use 9,999 permutations throughout, and every family of multiple comparisons carries a Bonferroni adjustment. The data and the code needed to regenerate this document are deposited with the research compendium.

The tables and figures follow the order of the Results and are grouped into four sections: raw material economy first, then the techno-typological characterisation of the retouch products, then the technological consistency of the surface collections with the excavated material at Longtan, and last the landscape structure of the technological variation. Each section opens with a short guide to what it contains. Each caption is written to stand on its own, so that no table or figure depends on the text of the paper to be read.

\newpage

# Raw material economy

This section reports what was available on the river bars and what was chosen from it. @tbl-composition gives the raw material composition of the surveyed cobbles and of the surface Quina scrapers, valley by valley, and @tbl-electivity sets the two against one another as an index of selection. @tbl-rawmat-permanova tests whether composition differs between sampling points, valleys and basins, and @tbl-clast-size reports the size of the surveyed cobbles by lithology.

\newpage

```{r}
#| label: tbl-composition
#| tbl-cap: "Raw material composition of the river-bar cobbles and of the surface-collected Quina scrapers, by valley. Cells give the count with the percentage of the row in parentheses."

comp_tbl <- comp_fig |>
  mutate(cell = ifelse(n == 0, "0", sprintf("%d (%s%%)", n, f(percent, 1)))) |>
  select(Layer, river_ID, Material, cell) |>
  pivot_wider(names_from = Material, values_from = cell) |>
  left_join(counts_fig, by = c("Layer", "river_ID"))

comp_out <- bind_rows(lapply(layer_levels, function(lay) {
  d <- comp_tbl[comp_tbl$Layer == lay, ]
  bind_rows(
    data.frame(Valley = sprintf("**%s**", lay),
               Trachyte = "", Sandstone = "", Quartz = "", Mudstone = "",
               Andesite = "", N = ""),
    data.frame(Valley = paste0(IND, as.character(d$river_ID)),
               Trachyte = d$Trachyte, Sandstone = d$Sandstone, Quartz = d$Quartz,
               Mudstone = d$Mudstone, Andesite = d$Andesite, N = as.character(d$N)))
}))

knitr::kable(comp_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r"),
             col.names = c("Valley", "Trachyte", "Sandstone", "Quartz",
                           "Mudstone", "Andesite", "*n*"))
```

\newpage

```{r}
#| label: tbl-electivity
#| tbl-cap: "Jacobs' electivity index for every raw material. *D* runs from −1, complete avoidance, through 0, use in proportion to availability, to +1, complete selection."

elect_tbl <- tibble(Material = material_levels) |>
  left_join(comp_avail |> transmute(Material, an = n, ap = p), by = "Material") |>
  left_join(comp_used  |> transmute(Material, un = n, up = r), by = "Material") |>
  left_join(jac |> transmute(Material, D), by = "Material") |>
  mutate(across(c(an, un), ~ replace_na(.x, 0L)),
         across(c(ap, up), ~ replace_na(.x, 0)),
         Label = ifelse(Material == "Andesite", "Andesite†", Material))

elect_out <- bind_rows(
  data.frame(Material = elect_tbl$Label,
             an = as.character(elect_tbl$an), ap = f(elect_tbl$ap, 1),
             un = as.character(elect_tbl$un), up = f(elect_tbl$up, 1),
             D  = f(elect_tbl$D, 2)),
  data.frame(Material = "**Total**",
             an = as.character(sum(elect_tbl$an)), ap = f(100, 1),
             un = as.character(sum(elect_tbl$un)), up = f(100, 1), D = ""))

knitr::kable(elect_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Raw material", "Cobbles *n*", "Cobbles %",
                           "Scrapers *n*", "Scrapers %", "*D*"))
```

\newpage

```{r}
#| label: tbl-rawmat-permanova
#| tbl-cap: "PERMANOVA and PERMDISP on raw material composition, on Bray–Curtis dissimilarities with 9,999 permutations. Sums of squares, degrees of freedom and residual degrees of freedom are those of the PERMANOVA; the last two columns give the pseudo-*F* and *p* of the test of homogeneity of dispersion on the same dissimilarity matrix. The unit is the individual cobble in the first panel and the individual artefact in the second."

perm_row <- function(model, a, pd) {
  data.frame(Model = paste0(IND, model),
             Df = sprintf("%d, %d", a$Df[1], a$Df[2]),
             SS = f(a$SumOfSqs[1], 2),
             R2 = f(a$R2[1], 3),
             F  = f(a$F[1], 2),
             p  = fpt(a$`Pr(>F)`[1]),
             pdF = f(pd[["F"]], 2),
             pdp = fpt(pd[["p"]]))
}
head_row <- function(label) data.frame(Model = sprintf("**%s**", label),
  Df = "", SS = "", R2 = "", F = "", p = "", pdF = "", pdp = "")

rawmat_out <- bind_rows(
  head_row(sprintf("Available river cobbles (*n* = %d)", n_cobble)),
  perm_row("~ sampling point", cobbleLoc,   pd_cobbleLoc),
  perm_row("~ valley",         cobbleRiver, pd_cobbleRiver),
  head_row(sprintf("Surface Quina scrapers (*n* = %d)", nrow(scA))),
  perm_row("~ basin",  usedA_basin, pd_usedA_basin),
  perm_row("~ valley", usedA_river, pd_usedA_river))

knitr::kable(rawmat_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Model", "*df*", "SS", "*R*²", "pseudo-*F*", "*p*",
                           "PERMDISP *F*", "PERMDISP *p*"))
```

\newpage

```{r}
#| label: tbl-clast-size
#| tbl-cap: "Size of the surveyed river cobbles by lithology. Size is the geometric mean of length, breadth and thickness in millimetres, and GM is the geometric mean of that quantity within each lithology."

size_row <- function(m) {
  x <- cl$size[cl$Material == m]
  if (!length(x)) return(NULL)
  qs <- unname(quantile(x, c(.25, .75)))
  data.frame(Material = m, n = as.character(length(x)),
             GM = f(gm(x), 1), Median = f(median(x), 1),
             IQR = sprintf("%s–%s", f(qs[1], 1), f(qs[2], 1)),
             Range = sprintf("%s–%s", f(min(x), 1), f(max(x), 1)))
}
size_out <- bind_rows(lapply(material_levels, size_row))

knitr::kable(size_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Raw material", "*n*", "GM (mm)", "Median (mm)",
                           "Q1–Q3 (mm)", "Range (mm)"))
```

\newpage

# Techno-typological characterisation of the retouch products

This section describes the retouch products themselves. @tbl-retouch-products gives the metric and reduction attributes of the surface Quina scrapers and of the resharpening flakes that accompany them, and @fig-categorical-attributes gives the categorical attributes of the scrapers — the cover and position of cortex, the type of platform and the direction from which retouch was struck — as the percentage of specimens in each class. The typological attribution of the same specimens is reported in the main text.

\newpage

```{r}
#| label: tbl-retouch-products
#| tbl-cap: "Metric and reduction attributes of the Quina scrapers and the resharpening flakes from the surface collections. Not every attribute could be measured on every specimen: the platform attributes are recorded only on those that retain the platform of the blank, and GIUR on all but one scraper. The retouch attributes apply to the scrapers alone and the exterior platform angle to the resharpening flakes alone."

# one row per attribute; d = decimals for mean/SD, ds for the order statistics
attr_row <- function(label, x, d = 1, ds = d) {
  x  <- as.numeric(x); x <- x[is.finite(x)]
  qs <- unname(quantile(x, c(.25, .75)))
  data.frame(Attribute = paste0(IND, label),
             mn  = f(mean(x), d), sd = f(sd(x), d),
             med = f(median(x), ds),
             iqr = paste0(f(qs[1], ds), "–", f(qs[2], ds)))
}
# panel heading, keeping the two classes of retouch product apart
attr_head <- function(label, n_total) {
  data.frame(Attribute = sprintf("**%s** (*n* = %d)", label, n_total),
             mn = "", sd = "", med = "", iqr = "")
}

retouch_products_tbl <- bind_rows(
  attr_head("Quina scrapers", nrow(q)),
  attr_row("Length (mm)",    q$Length,    1),
  attr_row("Width (mm)",     q$Width,     1),
  attr_row("Thickness (mm)", q$Thickness, 1),
  attr_row("Mass (g)",       q$Mass,      1),
  attr_row("Platform depth (mm)", q$Platform_depth, 1),
  attr_row("Platform width (mm)", q$Platform_width, 1),
  attr_row("IPA (°)",             q$IPA,            1),
  attr_row("Retouch generations", q$Ave_RG,    2),
  attr_row(variable_labels[["N_Scar"]], q$N_Scar, 1, ds = 0),
  attr_row("GIUR",                q$Ave_GIUR,  2),
  attr_row("Edge angle (°)",      q$Edge_Angle, 1),
  attr_row("Retouched perimeter", q$Retouch_length_index, 2),
  attr_head("Resharpening flakes", nrow(rfl)),
  attr_row("Length (mm)",    rfl$Length,    1),
  attr_row("Width (mm)",     rfl$Width,     1),
  attr_row("Thickness (mm)", rfl$Thickness, 1),
  attr_row("Mass (g)",       rfl$Mass,      1),
  attr_row("Platform depth (mm)", rfl$Platform_depth, 1),
  attr_row("Platform width (mm)", rfl$Platform_width, 1),
  attr_row("IPA (°)",             rfl$IPA,            1),
  attr_row("EPA (°)",             rfl$EPA,            1))

knitr::kable(retouch_products_tbl, align = c("l", "r", "r", "r", "r"),
             col.names = c("Attribute", "Mean", "SD", "Median", "Q1–Q3"))
```

\newpage

```{r}
#| label: fig-categorical-attributes
#| fig-cap: "Categorical attributes of the surface Quina scrapers: proportion of the dorsal face covered by cortex, position of that cortex, platform type, and the direction from which retouch was struck. Bars run to the percentage of the specimens recorded for that attribute, and the count and percentage are given at the end of each bar."
#| fig-width: 5.9
#| fig-height: 4.9

# one row per category, ordered within its panel: by frequency where the
# categories are nominal, in their own order where they are not
cat_panel <- function(x, panel, levels_order = NULL, recode_map = NULL) {
  x <- trimws(as.character(x)); x <- x[!is.na(x)]
  if (!is.null(recode_map)) x <- ifelse(x %in% names(recode_map), recode_map[x], x)
  d <- tibble(Level = x) |> count(Level, name = "n")
  d <- if (is.null(levels_order)) arrange(d, desc(n)) else
    d[match(levels_order, d$Level), ] |> filter(!is.na(Level))
  d |> mutate(pct = 100 * n / sum(n), Panel = panel, rank = row_number())
}
cortex_classes <- c("0", "1-25%", "26-50%", "51-75%", "76-99%")
cat_df <- bind_rows(
  cat_panel(q$Cortex, "Cortex proportion", levels_order = cortex_classes),
  cat_panel(q$Cortex_position, "Cortex position", recode_map = c(Non_cortex = "None")),
  cat_panel(q$Platform_type, "Platform type"),
  cat_panel(q$Tot_Direction, "Retouch direction")) |>
  mutate(Panel = factor(Panel, levels = unique(Panel)))
# a discrete axis places its first level at the foot of the panel, so the
# display order is reversed here to put the leading category at the top
cat_df$Level <- factor(cat_df$Level,
                       levels = cat_df |> arrange(Panel, desc(rank)) |> pull(Level))

# colour carries the ordering where there is one: the cortex classes take a
# light-to-dark ramp, and each nominal panel a single colour of the shared
# palette, which distinguishes the panels without implying an order within them
cortex_ramp <- setNames(colorRampPalette(c("#CBDCEA", "#2C5F7C"))(length(cortex_classes)),
                        cortex_classes)
panel_flat <- c(`Cortex position` = "#E07C90", `Platform type` = "#E6C25C",
                `Retouch direction` = "#9B87C4")
cat_df <- cat_df |>
  mutate(bar_fill = ifelse(as.character(Panel) == "Cortex proportion",
                           cortex_ramp[as.character(Level)],
                           panel_flat[as.character(Panel)]))

ggplot(cat_df, aes(pct, Level)) +
  geom_col(aes(fill = bar_fill), width = 0.7) +
  geom_text(aes(label = sprintf("n = %d, %.1f%%", n, pct)),
            hjust = -0.12, size = 2.2, color = "#303238") +
  scale_fill_identity() +
  facet_grid(Panel ~ ., scales = "free_y", space = "free_y") +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25),
                     labels = function(z) paste0(z, "%"),
                     expand = expansion(mult = c(0, 0.24))) +
  labs(x = "Percentage", y = NULL) +
  fig_theme +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 7.5),
        strip.text.y = element_text(size = 7.5, angle = -90))
```

\newpage

# Technological consistency

This section compares the surface Quina scrapers with the excavated Quina scrapers from Longtan, against the Longtan non-Quina scrapers as a control. @fig-qq shows the distribution of each of the six technological variables within each group, the check that stands behind the choice of test. @tbl-tc-permanova to @tbl-tc-pca give the full output of the comparison: the pairwise tests of group position and of group spread, the variable-by-variable tests, and the loadings of the ordination. @tbl-tc-sensitivity and @fig-sensitivity-pca then repeat the analysis with the surface specimens collected from within the area of the Longtan site removed.

\newpage

```{r}
#| label: fig-qq
#| fig-cap: "Normal quantile–quantile plots of the six technological variables. In every panel the horizontal axis is the theoretical normal quantile and the vertical axis the sample quantile, and the dashed line passes through the first and third sample quartiles. Inset at the upper left of each panel is a histogram of the same values, plotted in its own right but with its axes suppressed, to give the shape of the distribution alongside its departure from normality."
#| fig-width: 5.9
#| fig-height: 8.4

dist_long <- variable_long |>
  mutate(Variable = factor(variable_labels_bare[as.character(Variable)],
                           levels = unname(variable_labels_bare)),
         Group = factor(group_labels[as.character(Group)],
                        levels = unname(group_labels)))
lab_colors <- setNames(unname(group_colors), unname(group_labels))

# Each cell of the figure is built as its own plot rather than as one facet of
# a whole, so that a histogram drawn as a plot in its own right can be inset
# into it. inset_element places the inset as a fraction of the panel, so every
# one of them lands in the same corner at the same size whatever the scales of
# the panel it sits in.
INSET_GAP <- unit(1.4, "mm")     # the same clearance at the left and the top
inset_theme <- theme_minimal(base_size = 6) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "#202124", fill = NA, linewidth = 0.3),
        axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_rect(fill = "white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        plot.margin = margin(0, 0, 0, 0))

# Each cell holds one group only, so a legend built from it has a key for that
# group and two blank ones, and three such legends would be collected side by
# side. The legend is therefore carried by a single cell, which is given three
# fully transparent points, one per group, placed inside its own range so that
# they move nothing. override.aes makes the keys solid.
qq_cell <- function(vlab, glab, ylab, carries_legend = FALSE) {
  d <- dist_long[dist_long$Variable == vlab & dist_long$Group == glab, ]
  legend_key <- data.frame(
    x = 0, y = mean(range(d$Value, na.rm = TRUE)),
    Group = factor(levels(dist_long$Group), levels = levels(dist_long$Group)))
  p_hist <- ggplot(d, aes(Value)) +
    geom_histogram(bins = 12, fill = lab_colors[[glab]], alpha = 0.75,
                   color = "#202124", linewidth = 0.1) +
    scale_x_continuous(expand = expansion(mult = 0.03)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
    inset_theme
  p_qq <- ggplot(d, aes(sample = Value, color = Group)) +
    stat_qq(size = 0.7, alpha = 0.6, shape = 16, show.legend = FALSE) +
    stat_qq_line(linewidth = 0.35, color = "#5A5F66", linetype = "dashed",
                 show.legend = FALSE) +
    scale_color_manual(values = unname(group_colors),
                       limits = unname(group_labels), drop = FALSE) +
    scale_x_continuous(expand = expansion(mult = 0.06)) +
    # headroom under the inset, which is opaque and would otherwise cover points
    scale_y_continuous(expand = expansion(mult = c(0.06, 0.34))) +
    labs(x = NULL, y = ylab) +
    fig_theme +
    theme(panel.grid.major = element_blank(),
          axis.title.y = element_text(size = 8, face = "bold"),
          axis.text = element_text(size = 6.5)) +
    guides(color = guide_legend(nrow = 1, override.aes = list(size = 1.8, alpha = 1)))
  if (carries_legend) {
    p_qq <- p_qq +
      geom_point(data = legend_key, aes(x = x, y = y, color = Group),
                 inherit.aes = FALSE, alpha = 0, size = 0.7, shape = 16,
                 show.legend = TRUE)
  }
  # the gap is given in millimetres rather than as a fraction of the panel, so
  # that the inset stands the same distance from the left edge as from the top;
  # the panels are wider than they are tall, and equal fractions would not be
  # equal distances
  p_qq + inset_element(p_hist,
                       left   = INSET_GAP,
                       bottom = unit(0.63, "npc"),
                       right  = unit(0.47, "npc"),
                       top    = unit(1, "npc") - INSET_GAP,
                       align_to = "panel", ignore_tag = TRUE)
}

# the group is carried by the legend along the foot of the figure rather than
# by a strip over every column, the group names being too long to sit there;
# the variable names the row from the left
vars_row <- levels(dist_long$Variable)
qq_cells <- unlist(lapply(vars_row, function(v) {
  lapply(seq_along(levels(dist_long$Group)), function(i) {
    qq_cell(v, levels(dist_long$Group)[i], if (i == 1) v else NULL,
            carries_legend = (v == vars_row[1] && i == 1))
  })
}), recursive = FALSE)

wrap_plots(qq_cells, ncol = 3, guides = "collect") &
  theme(legend.position = "bottom", legend.margin = margin(0, 0, 0, 0))
```

\newpage

```{r}
#| label: tbl-tc-permanova
#| tbl-cap: "Pairwise PERMANOVA on the z-scored six-variable technological space, Euclidean distances, 9,999 permutations. Degrees of freedom are given as model, residual. Adjusted *p* is the Bonferroni adjustment across the three comparisons."

pw_label <- function(x) {
  parts <- strsplit(x, " vs ", fixed = TRUE)[[1]]
  paste(group_labels[parts], collapse = " vs ")
}
tc_perm_out <- data.frame(
  Comparison = vapply(tc_pw$Comparison, pw_label, character(1)),
  n = sprintf("%d, %d", tc_pw$n1, tc_pw$n2),
  Df = sprintf("%d, %d", tc_pw$Df, tc_pw$Df_res),
  SS = f(tc_pw$SumOfSqs, 2),
  R2 = f(tc_pw$R2, 3),
  F = f(tc_pw$F, 2),
  p = fpt(tc_pw$p_value),
  padj = fpt(tc_pw$p_adjusted))

knitr::kable(tc_perm_out, row.names = FALSE,
             align = c("l", "r", "r", "r", "r", "r", "r", "r"),
             col.names = c("Comparison", "*n*", "*df*", "SS", "*R*²",
                           "pseudo-*F*", "*p*", "Adjusted *p*"))
```

\newpage

```{r}
#| label: tbl-tc-permdisp
#| tbl-cap: "PERMDISP on the same technological space. The upper panel gives the mean distance of each group to its own centroid, the quantity the test compares; distance is measured in the z-scored space, so its units are standard deviations. The lower panel gives the pairwise tests, each fitted to its pair in turn, with the *p*-value from the permutation of the full three-group model alongside for comparison."

pdd <- permA$distances |> group_by(Group) |>
  summarise(n = n(), m = mean(distance), s = sd(distance), .groups = "drop") |>
  mutate(se = s / sqrt(n))

pd_out <- bind_rows(
  data.frame(Item = "**Distance to centroid, by group**",
             a = "", b = "", c = "", d = "", e = ""),
  data.frame(Item = paste0(IND, group_labels[as.character(pdd$Group)]),
             a = as.character(pdd$n), b = f(pdd$m, 3), c = f(pdd$s, 3),
             d = f(pdd$se, 3), e = ""),
  data.frame(Item = paste0(IND, "*Overall test*"), a = "",
             b = sprintf("%d, %d", permA$overall_df, permA$overall_df_res),
             c = f(permA$overall_F, 2), d = fpt(permA$overall_p), e = ""),
  data.frame(Item = "**Pairwise tests**", a = "", b = "", c = "", d = "", e = ""),
  data.frame(Item = paste0(IND, vapply(permA$pair_tab$Comparison, function(x) {
               paste(group_labels[strsplit(x, "-", fixed = TRUE)[[1]]], collapse = " vs ")
             }, character(1))),
             a = sprintf("%d, %d", permA$pair_tab$n1, permA$pair_tab$n2),
             b = sprintf("%d, %d", permA$pair_tab$Df, permA$pair_tab$Df_res),
             c = f(permA$pair_tab$F, 2),
             d = fpt(permA$pair_tab$p),
             e = fpt(permA$pairwise[permA$pair_tab$Comparison])))

knitr::kable(pd_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("", "*n*", "Mean / *df*", "SD / *F*",
                           "SE / *p*", "*p* (3-group)"))
```

\newpage

```{r}
#| label: tbl-tc-univariate
#| tbl-cap: "Per-variable comparison of the three groups: the omnibus test for each variable, followed by its three post-hoc comparisons. GIUR, total scars and retouch generations were tested with Kruskal–Wallis and Dunn's post-hoc comparisons, the effect size being epsilon-squared; thickness, retouched perimeter and edge angle with Welch's ANOVA and Welch's *t* post-hoc comparisons. All post-hoc *p*-values carry a Bonferroni adjustment within each variable. ns, not significant."

uni_rows <- function(v) {
  is_kw <- v %in% kw_vars
  if (is_kw) {
    o <- kw_omni[kw_omni$Variable == v, ]; e <- kw_eff[kw_eff$Variable == v, ]
    omni <- data.frame(Item = paste0(IND, "*Kruskal–Wallis*"),
                       stat = sprintf("H = %s", f(o$statistic, 2)),
                       df = as.character(o$df), p = fpt(o$p),
                       eff = sprintf("ε² = %s", f(e$effsize, 3)), sig = "")
    ph <- kw_ph[kw_ph$Variable == v, ]
    post <- data.frame(
      Item = paste0(IND, vapply(seq_len(nrow(ph)), function(i)
        paste(group_labels[c(ph$group1[i], ph$group2[i])], collapse = " vs "), character(1))),
      stat = sprintf("Z = %s", f(ph$statistic, 2)),
      df = sprintf("%d, %d", ph$n1, ph$n2),
      p = fpt(ph$p.adj), eff = "", sig = ph$p.adj.signif)
  } else {
    o <- we_omni[we_omni$Variable == v, ]
    omni <- data.frame(Item = paste0(IND, "*Welch's ANOVA*"),
                       stat = sprintf("F = %s", f(o$statistic, 2)),
                       df = sprintf("%s, %s", o$DFn, f(o$DFd, 1)),
                       p = fpt(o$p), eff = "", sig = "")
    ph <- we_ph_full[we_ph_full$Variable == v, ]
    post <- data.frame(
      Item = paste0(IND, vapply(seq_len(nrow(ph)), function(i)
        paste(group_labels[c(ph$group1[i], ph$group2[i])], collapse = " vs "), character(1))),
      stat = sprintf("t = %s", f(ph$statistic, 2)),
      df = f(ph$df, 1), p = fpt(ph$p.adj), eff = "", sig = ph$p.adj.signif)
  }
  bind_rows(data.frame(Item = sprintf("**%s**", variable_labels[[v]]),
                       stat = "", df = "", p = "", eff = "", sig = ""),
            omni, post)
}
uni_out <- bind_rows(lapply(variables, uni_rows))

knitr::kable(uni_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Variable / comparison", "Statistic", "*df* or *n*",
                           "*p*", "Effect size", ""))
```

\newpage

```{r}
#| label: tbl-tc-pca
#| tbl-cap: "Loadings of all six principal components of the z-scored technological space, with the eigenvalue, the percentage of variance and the cumulative percentage for each."

ld <- permA$pca_loadings
pca_out <- bind_rows(
  data.frame(Item = variable_labels[ld$Variable],
             as.data.frame(lapply(ld[, paste0("PC", 1:6)], f, d = 3))),
  data.frame(Item = "**Eigenvalue**",
             as.data.frame(as.list(setNames(f(permA$sdev^2, 2), paste0("PC", 1:6))))),
  data.frame(Item = "**Variance (%)**",
             as.data.frame(as.list(setNames(f(permA$var_explained, 1), paste0("PC", 1:6))))),
  data.frame(Item = "**Cumulative (%)**",
             as.data.frame(as.list(setNames(f(cumsum(permA$var_explained), 1), paste0("PC", 1:6))))))

knitr::kable(pca_out, row.names = FALSE, align = c("l", rep("r", 6)),
             col.names = c("Variable", paste0("PC", 1:6)))
```

\newpage

```{r}
#| label: tbl-tc-sensitivity
#| tbl-cap: "Sensitivity of the technological consistency result to the surface specimens collected from within the area of the Longtan site. The full set is the analysis reported in the main text, with 165 surface specimens; the reduced set excludes those collected within the site area, leaving 150, and is z-scored on its own. PERMANOVA *p*-values carry the Bonferroni adjustment across the three comparisons."

sens_rows <- function(comp_perm, comp_disp) {
  full_p <- tc_pw[tc_pw$Comparison == comp_perm, ]
  red_p  <- tc_pw_excl[tc_pw_excl$Comparison == comp_perm, ]
  full_d <- permA$pair_tab[permA$pair_tab$Comparison == comp_disp, ]
  red_d  <- permC$pair_tab[permC$pair_tab$Comparison == comp_disp, ]
  label  <- paste(group_labels[strsplit(comp_perm, " vs ", fixed = TRUE)[[1]]],
                  collapse = " vs ")
  bind_rows(
    data.frame(Comparison = sprintf("**%s**", label), Analysis = "", fR2 = "", fF = "",
               fp = "", rR2 = "", rF = "", rp = ""),
    data.frame(Comparison = paste0(IND, "PERMANOVA"), Analysis = "centroid position",
               fR2 = f(full_p$R2, 3), fF = f(full_p$F, 2), fp = fpt(full_p$p_adjusted),
               rR2 = f(red_p$R2, 3),  rF = f(red_p$F, 2),  rp = fpt(red_p$p_adjusted)),
    data.frame(Comparison = paste0(IND, "PERMDISP"), Analysis = "dispersion",
               fR2 = "", fF = f(full_d$F, 2), fp = fpt(full_d$p),
               rR2 = "", rF = f(red_d$F, 2),  rp = fpt(red_d$p)))
}
sens_out <- bind_rows(
  sens_rows("SC_Quina vs LT_Quina",    "SC_Quina-LT_Quina"),
  sens_rows("SC_Quina vs LT_Ordinary", "SC_Quina-LT_Ordinary"),
  sens_rows("LT_Quina vs LT_Ordinary", "LT_Quina-LT_Ordinary"))

knitr::kable(sens_out, row.names = FALSE, align = c("l", "l", rep("r", 6)),
             col.names = c("Comparison", "Tests", "*R*²", "*F*", "*p*",
                           "*R*²", "*F*", "*p*"))
```

\newpage

```{r}
#| label: fig-sensitivity-pca
#| fig-cap: "Ordination of the technological space with the surface specimens from within the area of the Longtan site removed, to be compared with panel A of the corresponding figure in the main text. (A) Specimen scores, with convex hulls enclosing each group and filled points marking group centroids. (B) Loadings of the six variables on the same two components, drawn on their own axes so that the specimen scores are not overplotted."
#| fig-width: 5.9
#| fig-height: 3.7

sc_s <- permC$pca_scores
ld_s <- permC$pca_loadings
cent_s <- sc_s |> group_by(Group) |> summarise(cx = mean(PC1), cy = mean(PC2), .groups = "drop")
hull_s <- sc_s |> group_by(Group) |> filter(n() >= 3) |>
  group_modify(function(g, k) { h <- chull(g$PC1, g$PC2); bind_rows(g[h, ], g[h[1], ]) }) |> ungroup()
arr_s <- ld_s |> mutate(Label = variable_labels_bare[Variable])

p_scores <- ggplot(sc_s, aes(PC1, PC2, color = Group)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
  geom_polygon(data = hull_s, aes(fill = Group, group = Group), alpha = 0.12, color = NA) +
  geom_path(data = hull_s, aes(group = Group), linewidth = 0.4, alpha = 0.65) +
  geom_point(size = 1.1, alpha = 0.7, shape = 16) +
  geom_point(data = cent_s, aes(cx, cy, fill = Group), shape = 21, color = "#5A5F66",
             size = 2.4, stroke = 0.6, inherit.aes = FALSE) +
  scale_color_manual(values = group_colors, labels = group_labels) +
  scale_fill_manual(values = group_colors, labels = group_labels) +
  scale_x_continuous(expand = expansion(mult = 0.14)) +
  scale_y_continuous(expand = expansion(mult = 0.10)) +
  labs(x = sprintf("PC1 (%.1f%%)", permC$var_explained[1]),
       y = sprintf("PC2 (%.1f%%)", permC$var_explained[2])) +
  fig_theme +
  theme(panel.grid.major = element_blank(), legend.text = element_text(size = 8)) +
  guides(fill = "none",
         color = guide_legend(nrow = 1, override.aes = list(size = 1.8, alpha = 1)))

# the loadings on their own axes, so that the arrows need no arbitrary scaling
# factor to be readable over the point cloud. Each column of the rotation has
# unit norm across all six components, so the unit circle is the reference an
# arrow would reach if these two components accounted for the variable in full.
circle_ref <- data.frame(theta = seq(0, 2 * pi, length.out = 200)) |>
  mutate(x = cos(theta), y = sin(theta))
lim_s <- 1.12
p_loadings <- ggplot(arr_s, aes(PC1, PC2)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
  geom_path(data = circle_ref, aes(x, y), linewidth = 0.3, color = "grey70",
            linetype = "dashed", inherit.aes = FALSE) +
  geom_segment(aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.075, "cm"), type = "open", angle = 22),
               linewidth = 0.3, color = "#5A5F66") +
  geom_text_repel(aes(label = Label), size = 2.2, color = "#3A3D42",
                  bg.color = "white", bg.r = 0.12, segment.color = "grey55",
                  segment.size = 0.2, min.segment.length = 0.2, box.padding = 0.35,
                  point.padding = 0.1, max.overlaps = Inf, seed = 2226) +
  scale_x_continuous(limits = c(-lim_s, lim_s)) +
  scale_y_continuous(limits = c(-lim_s, lim_s)) +
  labs(x = "PC1 loading", y = "PC2 loading") +
  fig_theme + theme(panel.grid.major = element_blank())

(p_scores + labs(tag = "A")) + (p_loadings + labs(tag = "B")) +
  plot_layout(widths = c(1.7, 1), guides = "collect") &
  theme(legend.position = "bottom", legend.margin = margin(0, 0, 0, 0))
```

\newpage

# Landscape structure

This section asks whether the technological variation follows position in the landscape, with the locality rather than the specimen as the unit. @tbl-locality-matrix lists the localities that enter the analysis with their median value on each of the six measures, and @tbl-landscape-permanova tests each landscape variable against that space. The three figures show the same data in turn: @fig-site-pca as an ordination of the localities coloured by each landscape variable, @fig-landscape-scatter as each measure against each of the continuous variables, and @fig-basin-boxplot as a comparison between the two basins.

\newpage

```{r}
#| label: tbl-locality-matrix
#| tbl-cap: "Locality-level analysis frame. Each row is one Quina locality, with the number of Quina scrapers complete on all six variables and the median of those specimens on each. Distance and height are measured to and above the nearest channel."

loc_out <- site_df |>
  arrange(Basin, desc(n_art), Name) |>
  transmute(Locality = Name, Basin = as.character(Basin),
            n = as.character(n_art),
            d = as.character(round(Distance_to_water)),
            h = as.character(round(Height_above_river)),
            Th = f(Thickness, 1), RLI = f(Retouch_length_index, 2),
            GIUR = f(Ave_GIUR, 2), NS = f(N_Scar, 0),
            RG = f(Ave_RG, 2), EA = f(Edge_Angle, 1))

knitr::kable(loc_out, row.names = FALSE,
             align = c("l", "l", rep("r", 9)),
             col.names = c("Locality", "Basin", "*n*", "Distance (m)", "Height (m)",
                           unname(variable_labels)))
```

\newpage

```{r}
#| label: tbl-landscape-permanova
#| tbl-cap: "Locality-level PERMANOVA for each landscape variable, on Euclidean distances in the z-scored six-variable space with 9,999 permutations, together with the test of dispersion for the one categorical variable."

ls_row <- function(label, v, indent = TRUE) {
  data.frame(Term = if (indent) paste0(IND, label) else sprintf("**%s**", label),
             Df = sprintf("%d, %d", as.integer(v["Df"]), as.integer(v["Df_res"])),
             SS = f(v["SumOfSqs"], 2), R2 = f(v["R2"], 3),
             F = f(v["F"], 3), p = fpt(v["p"]))
}
ls_out <- bind_rows(
  data.frame(Term = "**Basin**", Df = "", SS = "", R2 = "", F = "", p = ""),
  ls_row(sprintf("PERMANOVA (Binchuan %d, Heqing %d)",
                 n_basin_site[["Binchuan"]], n_basin_site[["Heqing"]]), ls_basin),
  data.frame(Term = paste0(IND, "PERMDISP"), Df = "", SS = "", R2 = "",
             F = f(ls_basin_pd[["F"]], 3), p = fpt(ls_basin_pd[["p"]])),
  ls_row("Height above channel (m)", ls_height, indent = FALSE),
  ls_row("Distance to channel (m)", ls_dist, indent = FALSE),
  ls_row("Assemblage size (n specimens)", ls_size, indent = FALSE))

knitr::kable(ls_out, row.names = FALSE, align = c("l", "r", "r", "r", "r", "r"),
             col.names = c("Landscape variable", "*df*", "SS", "*R*²", "pseudo-*F*", "*p*"))
```

\newpage

```{r}
#| label: fig-site-pca
#| fig-cap: "Locality-level ordination of the six technological measures, each locality entered as the median of its Quina scrapers, coloured by (A) basin, (B) height above the channel, (C) assemblage size and (D) distance to the nearest channel. The four panels are the same ordination and differ only in the landscape variable shown, which is named in the legend beneath each. The three continuous variables are carried by the colour gradient at their measured values, the form in which they enter every test reported here; the gradient runs between the two colours that distinguish the basins in (A), the colour scale is square-root, all three being right-skewed, and the legend ticks are in the original units. Behind the localities of (B) to (D), on that same scale, is a thin-plate spline of the variable fitted over the ordination plane and clipped to the area the localities cover. A landscape gradient would show as colour running in a direction across a panel. Basin, the one categorical variable, has hulls enclosing each group and filled points marking group centroids."
#| fig-width: 5.9
#| fig-height: 5.8

# Basin is the only categorical landscape variable; the other three enter the
# tests at their measured values, so they are shown here as gradients rather
# than in classes of our own choosing. The landscape variable is named in the
# legend under each panel rather than in a title over it.
site_panel <- function(grp, cols, legend_title) {
  d <- site_scores |> mutate(G = .data[[grp]]) |> filter(!is.na(G))
  cen <- d |> group_by(G) |> summarise(cx = mean(PC1), cy = mean(PC2), .groups = "drop")
  hul <- d |> group_by(G) |> filter(n() >= 3) |>
    group_modify(function(g, k) { h <- chull(g$PC1, g$PC2); bind_rows(g[h, ], g[h[1], ]) }) |> ungroup()
  ggplot(d, aes(PC1, PC2, color = G)) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
    geom_polygon(data = hul, aes(fill = G, group = G), alpha = 0.12, color = NA) +
    geom_point(size = 1.5, alpha = 0.85, shape = 16) +
    geom_point(data = cen, aes(cx, cy, fill = G), shape = 21, color = "#5A5F66",
               size = 2.2, stroke = 0.5, inherit.aes = FALSE) +
    scale_color_manual(values = cols, name = legend_title) +
    scale_fill_manual(values = cols, name = legend_title) +
    labs(x = sprintf("PC1 (%.1f%%)", site_vexp[1]),
         y = sprintf("PC2 (%.1f%%)", site_vexp[2])) +
    fig_theme +
    theme(panel.grid.major = element_blank(), legend.position = "bottom",
          legend.title.position = "top",
          legend.title = element_text(size = 7.5, hjust = 0.5),
          legend.key.size = unit(7, "pt"), legend.text = element_text(size = 7),
          legend.margin = margin(0, 0, 0, 0)) +
    guides(fill = "none", color = guide_legend(nrow = 1, override.aes = list(size = 1.6)))
}
# the gradient runs between the blue and the pink of the shared palette, the
# same two colours that distinguish the basins in panel A, and carries both the
# fitted surface and the localities drawn over it. A square-root scale:
# all three continuous variables are right-skewed (18 of the localities yielded
# one or two specimens, and three sit far above the terrace sequence), so a
# linear ramp would leave most points one colour.
site_grad_panel <- function(var, legend_title, breaks) {
  s <- surfaces[[var]]
  # both scales carry the same limits, so that the flat surface is drawn at the
  # colour its value deserves and not stretched across the whole ramp; the
  # background is lightened, which keeps a point of average value visible
  # against it
  lims <- range(site_scores[[var]], na.rm = TRUE)
  ggplot(site_scores, aes(PC1, PC2, color = .data[[var]])) +
    geom_raster(data = s$grid, aes(PC1, PC2, fill = fit), alpha = 0.5,
                inherit.aes = FALSE) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3, color = "grey55") +
    geom_point(size = 1.8, shape = 16) +
    scale_fill_gradient(low = grad_low, high = grad_high, transform = "sqrt",
                        limits = lims, oob = scales::squish,
                        na.value = "transparent", guide = "none") +
    scale_color_gradient(low = grad_low, high = grad_high, name = legend_title,
                         transform = "sqrt", breaks = breaks,
                         limits = lims, oob = scales::squish) +
    labs(x = sprintf("PC1 (%.1f%%)", site_vexp[1]),
         y = sprintf("PC2 (%.1f%%)", site_vexp[2])) +
    fig_theme +
    theme(panel.grid.major = element_blank(), legend.position = "bottom",
          legend.title.position = "top",
          legend.title = element_text(size = 7.5, hjust = 0.5),
          legend.text = element_text(size = 7),
          legend.key.width = unit(34, "pt"), legend.key.height = unit(5, "pt"),
          legend.margin = margin(0, 0, 0, 0)) +
    guides(color = guide_colourbar())
}

(site_panel("Basin", basin_colors, "Basin") + labs(tag = "A", x = NULL)) +
  (site_grad_panel("Height_above_river", "Height above channel (m)",
                   c(30, 60, 100, 150)) + labs(tag = "B", x = NULL, y = NULL)) +
  (site_grad_panel("Site_size", "Assemblage size (n specimens)",
                   c(1, 3, 10, 30)) + labs(tag = "C")) +
  (site_grad_panel("Distance_to_water", "Distance to channel (m)",
                   c(250, 500, 1000, 2000)) + labs(tag = "D", y = NULL)) +
  plot_layout(ncol = 2)
```

\newpage

```{r}
#| label: fig-landscape-scatter
#| fig-cap: "Each of the six technological measures against each of the three continuous landscape variables, at the scale of the locality: one row per measure, one column per landscape variable, all at measured values. The dashed line is an ordinary least-squares fit, shown only as a guide to the direction of the relationship. Each panel gives Spearman's ρ with its *p*-value, adjusted by the Bonferroni procedure within each column, every column being one family of six tests; the distance column is the family reported in the main text."
#| fig-width: 5.9
#| fig-height: 7.8

# the three landscape variables enter as measured, so the display is a scatter
# throughout; nothing here is binned
grid_long <- site_df |>
  select(all_of(c(names(land_vars), variables))) |>
  pivot_longer(all_of(variables), names_to = "Measure", values_to = "Value") |>
  pivot_longer(all_of(names(land_vars)), names_to = "Landscape", values_to = "X") |>
  mutate(Measure = factor(variable_labels[Measure], levels = unname(variable_labels)),
         Landscape = factor(land_vars[Landscape], levels = unname(land_vars)))

rho_lab <- land_cor |>
  mutate(Measure = factor(variable_labels[variable], levels = unname(variable_labels)),
         Landscape = factor(land_vars[Landscape], levels = unname(land_vars)),
         label = sprintf("rho == %.2f * ',' ~ %s", rho,
                         ifelse(p_adj < 0.001, "italic(p) < 0.001",
                                sprintf("italic(p) == %.3f", p_adj))))

ggplot(grid_long, aes(X, Value)) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = "#6BA8CE", fill = "#CBDCEA", linewidth = 0.6, linetype = "dashed") +
  geom_point(color = "#565693", alpha = 0.65, size = 1.2, shape = 16) +
  geom_text(data = rho_lab, aes(x = -Inf, y = Inf, label = label), parse = TRUE,
            hjust = -0.05, vjust = 1.3, size = 2, color = "#202124", inherit.aes = FALSE) +
  facet_grid(Measure ~ Landscape, scales = "free") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.18))) +
  labs(x = NULL, y = NULL) +
  fig_theme + theme(panel.grid.major = element_blank(),
                    strip.text.y = element_text(size = 7, angle = -90))
```

\newpage

```{r}
#| label: fig-basin-boxplot
#| fig-cap: "Each of the six technological measures between the two basins, at the scale of the locality: one panel per measure, each point one locality entered as the median of its Quina scrapers, the box giving the median and the quartiles with whiskers reaching 1.5 times the interquartile range, and the filled point the mean."
#| fig-width: 5.9
#| fig-height: 5.4

# one row per locality, as everywhere in this section: the points are the
# medians the tests are run on
basin_long <- bdf |>
  select(Basin, all_of(variables)) |>
  mutate(Basin = droplevels(Basin)) |>
  pivot_longer(all_of(variables), names_to = "Measure", values_to = "Value") |>
  mutate(Measure = factor(variable_labels[Measure], levels = unname(variable_labels)))

# the bracket sits above the highest point of its panel, placed as the brackets
# of the main-text boxplots are and carrying the same significance labels
basin_brackets <- basin_cmp |>
  transmute(Measure = factor(variable_labels[variable], levels = unname(variable_labels)),
            group1 = "Binchuan", group2 = "Heqing", label = p.adj.signif) |>
  left_join(basin_long |> group_by(Measure) |>
              summarise(ymax = max(Value, na.rm = TRUE),
                        yrange = diff(range(Value, na.rm = TRUE)), .groups = "drop"),
            by = "Measure") |>
  mutate(y.position = ymax + yrange * 0.10)

ggplot(basin_long, aes(Basin, Value)) +
  geom_jitter(aes(color = Basin), size = 1.7, alpha = 0.6, shape = 16,
              position = position_jitter(width = 0.22, height = 0, seed = 2226)) +
  geom_boxplot(color = "#202124", fill = NA, width = 0.55, linewidth = 0.4,
               outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 16, size = 2, color = "#202124") +
  stat_pvalue_manual(basin_brackets, label = "label", y.position = "y.position",
                     tip.length = 0.012, bracket.size = 0.3, label.size = 2.5,
                     color = "#202124") +
  facet_wrap(~ Measure, scales = "free_y", ncol = 3) +
  scale_color_manual(values = basin_colors, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.16))) +
  labs(x = NULL, y = NULL) +
  fig_theme +
  theme(panel.grid.major = element_blank())
```
